### 2021-04-14 draft/draft.tex
supplementary material created and some methods added

\synctex=1
\documentclass[a4paper,10pt]{article}

%% TO-DO: Achim and Schill in acknowl
%% TO-DO: link to repo in paper and suppl mat
%% TO-DO: my new grant
%% TO-DO: give URLs to, e.g, Crooks reference: mybibwithurl?

%% For tracking changes
%% https://tex.stackexchange.com/questions/65453/track-changes-in-latex

% packages

% %% txfonts breaks amsmath.
% %% http://www.texfaq.org/FAQ-alreadydef

\usepackage{gitinfo2}

\usepackage{savesym}
\usepackage{amsmath}
\savesymbol{iint}
\usepackage{txfonts}
\restoresymbol{TXF}{iint}

\usepackage[utf8]{inputenc} % allows usage of spanish special characters
\usepackage[spanish,english]{babel} % english dictionary for proper hyphenation
%\usepackage{txfonts} % nice fonts
\usepackage{authblk} % author affiliations
\usepackage{graphicx} % images
\usepackage{float} % image positioning
\usepackage{subcaption} % collages of multiple images
%%\usepackage[hidelinks]{hyperref} % hyperlinks
\usepackage{hyperref}
% \usepackage{cite} % contraction of references, use
%                   % \usepackage[superscript]{cite} for superscript
%                   % citations
%% cite does not follow the usual pre and post citation
\usepackage[square,numbers,sort&compress]{natbib}
\usepackage{xcolor} % text color
\usepackage[margin=3cm]{geometry} % margins
\usepackage{lineno} % line numbers
\usepackage{csquotes} % paragraph quotes
\usepackage{multirow} % joining rows in tables
\usepackage{array}
\usepackage[normalem]{ulem}
\usepackage{todonotes}
%%\usepackage{amsmath}: something breaks. What?
\usepackage{pdflscape} %% Landscape
\usepackage{ulem} % strike-through text

%% cross references with suppl mat
\usepackage{nameref}
\usepackage{zref-xr,zref-user}
\zxrsetup{toltxlabel}
\zexternaldocument*[S-]{supp}
\hypersetup{
  colorlinks = true,
  citecolor=  black,
  linkcolor = {blue},
  filecolor = cyan %% controls color of external ref, if used
}

% customization
\setcounter{Maxaffil}{0} % affiliations
\renewcommand\Affilfont{\itshape\small} % style of affiliations text
\captionsetup[figure]{font=small, labelfont={bf}, labelformat={default}, labelsep=period, name={Fig.}} % custom figure captions
\makeatletter\renewcommand{\@biblabel}[1]{#1.}\makeatother % change [number] for number. in reference list
\newcommand{\idea}[1]{\textcolor{red}{#1}}
\newcommand{\mr}{\textcolor{red}{\textbf{[missing ref(s)]}}}
\newcommand{\change}[2]{\textcolor{red}{\sout{#1}}\textcolor{blue}{#2}}
\linenumbers % add line numbers to whole doc

% title, authors, affiliations
\title{Conditional prediction of consecutive tumor evolution using cancer
  progression models: What genotype comes next?}
%\title{Conditional prediction of short-term tumor evolution using cancer progression models}
%\title{Prediction of consecutive tumor evolution using cancer progression models}
%\title{Conditional prediction of upcoming tumor evolution using cancer progression models}
%\title{Partial prediction of tumor evolutionary trajectories using cancer progression models}
%\title{Using cancer progression models to predict upcoming tumor evolution}
\author[1,2,3]{Juan Diaz-Colunga}
\author[1,2,$\dagger$]{Ramon Diaz-Uriarte}
\affil[1]{Dpt. of Biochemistry, School of Medicine, Universidad Autónoma de Madrid, Madrid, Spain}
\affil[2]{Instituto de Investigaciones Biomédicas `Alberto Sols'
  (UAM-CSIC), Madrid, Spain}
\affil[3]{Department of Ecology \& Evolutionary Biology and Microbial
  Sciences Institute, Yale University, New Haven, CT, USA}
%\affil[3]{Dpt. of Ecology and Evolutionary Biology, Yale University, New Haven, Connecticut, USA}
%\affil[4]{Microbial Sciences Institute, Yale University, New Haven, Connecticut, USA}
\affil[$\dagger$]{To whom correspondence should be addressed: \normalfont ramon.diaz@iib.uam.es}
%\date{}

\begin{document}

% \date{\gitAuthorDate\ {\footnotesize (Release\gitRels: Rev:
%      \gitAbbrevHash)}}

\maketitle

\begin{abstract}

  Accurate prediction of tumor progression is key for adaptive
  therapy and precision medicine. Cancer progression models (CPMs) can be
  used to infer dependencies in mutation accumulation from cross-sectional
  data and provide predictions of tumor progression paths. But their
  performance when predicting the complete evolutionary paths is limited
  by violations of assumptions and the size of available data
  sets. Instead of predicting full tumor progression paths, we can focus
  on short-term predictions, more relevant for diagnostic and therapeutic
  purposes. Here we examine if five distinct CPMs can be used to answer
  the question ``Given that a genotype with $\mathrm{n}$ mutations has
  been observed, what genotype with $\mathrm{n}+1$ mutations is next in
  the path of tumor progression'' or, shortly, ``What genotype comes
  next''. Using simulated data we find that under specific combinations of
  genotype and fitness landscape characteristics CPMs can provide
  predictions of short-term evolution that closely match the true
  probabilities, and that some genotype characteristics (fitness and
  probability of being a local fitness maximum) can be much more relevant
  than global features. Thus, CPMs can provide short-term predictions even
  when global, long-term predictions are not possible because fitness
  landscape- and evolutionary model-specific assumptions are violated.
  When good performance is possible, we observe significant variation in
  the quality of predictions of different methods. Genotype-specific and
  global fitness landscape characteristics are required to determine which
  method provides best results in each case.  Application of these methods
  to 25 cancer data sets shows that their use is hampered by lack of the
  information needed to make principled decisions about method choice and
  what predictions to trust. Fruitful use of these methods for short-term
  predictions requires adapting method's use to local genotype
  characteristics and obtaining reliable indicators of performance; it
  will also be necessary to clarify the interpretation of the method's results
  when key assumptions do not hold.

\end{abstract}

%% bioniformatics:  A maximum of 150 words is recommended. https://academic.oup.com/bioinformatics/pages/instructions_for_authors
%% PLoS Comp Biol: I think around 300

%% Length of ms: For bioinformatics: Original papers: up to 7 pages, this
%% is approx. 5000 words, not including figures,
%%
%% Do as pdftotext draft.pdf - | wc -w
%% Maybe rm refs and figures before word counting
%% Place \end{document} before refs; uncomment it.

%% As of now, this is twice as long as it should be for Bioinformatics:
%% 9400 words withouth figures.

\section{Introduction}\label{sec:intro}

Accurately predicting tumor evolution would allow for better prognosis and
treatment. Unfortunately, the stochastic components in key factors that
govern tumor progression (mutation, genetic drift or clonal selection
\cite{greaves_evolutionary_2015,lipinski_cancer_2016,williams2018}, as
well as non-genetic variability \cite{Brock2009} and the tumor
microenvironment \cite{Albini2007}) fundamentally limit
predictability. This is also common to other problems
\cite{lassig_predicting_2017, losos2018} such as the evolution of
antibiotic resistance \cite{toprak2012} or virulence \cite{Day2004}, where
even small increases in predictive power could be of critical importance
for the design of treatment strategies \cite{palmer2013}.

% what are cancer progression models?
Cross-sectional high-throughput data are widely available for many cancer
types, and Cancer Progression Models (CPMs) \cite{Beerenwinkel2014,
  beerenwinkel_computational_2016}, originally developed to identify
restrictions in the order of accumulation of mutations in tumors
using cross-sectional data, could thus be key tools for predicting cancer
progression. CPMs (implicitly) encode all the possible mutational
trajectories that are compatible with the restrictions and, for some
methods, provide probabilities of different paths (figure
\ref{fig:fig1}A-B). However their performance to infer the complete
evolutionary trajectories, from the root genotype with no driver mutations
to a final, fixated genotype, is limited
\cite{diaz-uriarte2019a,hosseini2019a}; this has been attributed mainly to
violations of the underlying assumptions of the models (fitness landscapes
with a single fitness maximum in the genotype with all loci mutated,
strong selection week mutation regime \cite{gillespie1983}) and the
difficulty to acquire
large, unbiased data sets to power them \cite{diaz-uriarte2019a}.

Predicting the complete evolutionary path, however, might not be the most
relevant objective. Detection of a specific tumor composition at a given
point in time automatically discards all mutational pathways that do not
visit the observed state, reducing the amount of possible evolutionary
trajectories that can be followed from then on and thus potentially
facilitating the task of predicting further tumor progression. In this
work we address the question of how accurately CPMs can predict short-term
cancer evolution after a genotype is detected: what is the state that
comes consecutively after or ``what genotype comes next''. Even if the
predictability of the complete path is low because of violations of method
assumptions, genotype-specific features could lead to local improvements
in short-term predictability \cite{bank_predictability_2016,
  visser2018,ferretti2018}. Moreover, in contrast to the more general,
non-local descriptors, these genotype-specific factors can provide more
detailed insight regarding the requirements for the methods to be
reliable. In fact, conditional predictions on the observed genotype are
arguably much more relevant than complete evolutionary path predictions
for both adaptive therapy \cite[e.g.,][]{melnikov2020,hansen2020a,
  stankova2019a} and for SMART (sequential, multiple assignment,
randomized trial) designs in dynamic treatment regimes for personalized
medicine \cite{tsiatis2020,chakraborty2013}.

[Fig.~\ref{fig:fig1} about here]

% Rather than analyzing full evolutionary
% trajectories, as has been done in other work
% \cite{diaz-uriarte2019a,hosseini2019a}, we focus here on the question
% ``what genotype comes next'', the partial paths from an observed tumor
% state to the state that comes consecutively after.

% This might, on the one hand, lead to improvements in predictive power,
% even when the predictability of the complete path is low because of
% violations of method assumptions: genotype-specific features could lead to
% local improvements in short-term predictability
% \cite{bank_predictability_2016, visser2018,ferretti2018}. On the other
% hand, conditional predictions on the observed genotype are much more
% relevant clinically and for treatment purposes in adaptive therapy
% \cite[e.g.,][]{melnikov2020,hansen2020a, stankova2019a} and for SMART
% (sequential, multiple assignment, randomized trial) designs in dynamic
% treatment regimes for personalized medicine
% \cite{tsiatis2020,chakraborty2013}.

% , accurately predicting the next genotype could be among
% the tools for the physician to become, in a game theory context, the
% ``rational leader'' that anticipates and steers to limit cancer's
% resistance strategies \cite{stankova2019a}; moreover, observed genotypes
% and predicted next genotypes could be of use in SMART (sequential,
% multiple assignment, randomized trial) designs in dynamic treatment
% regimes for personalized medicine \cite{tsiatis2020,chakraborty2013}.

% \cite[prenote][posnote]{hansen2020a}

% accurately predicting the
% next genotype would play a key role in adaptive therapy
% \cite[e.g.,][]{melnikov2020, hansen2020a,stankova2019a}, as ``(...) the
% ability to anticipate subsequent cancer cell responses, provides a
% critical opportunity to obtain more favorable outcomes by steering and/or
% limiting the cancer cell’s resistance strategies'' \cite[]

% are used in adaptive therapy \cite{whatever}

% form basis of SMART designs in individualized therapy
% could be used in SMART designs \cite{whatever---los dos libros}
% \cite{tsiatis2020,chakraborty2013}
% (yo: check los dos zotero dirs, and look for conditional on genotypes)

% \cite{melnikov2020}: inhibiting growth in bacteria

% \cite{hansen2020a}: adaptive therapy and cpompetitive supression: gral ref

% \cite{gluzman2020}: conditional treatment

% \cite{west2019}

% Motivated by the idea that information on the prospective advance of the
% disease, even if limited, can be of critical importance for the design of
% treatment strategies and can improve therapeutic outcomes
% \cite[p.~243i]{palmer2013}

% Information on the prospective advance of the disease, even if limited,
% can be of critical importance for the design of treatment strategies
% \idea{references, as in paper with Claudia}. \todo{YES! DO!}
% \idea{[Relate predicting next genotype to adaptive therapy ---tarea para
%   Ramon]}.

% Tumor progression is governed by a combination of factors such as mutation, genetic drift or clonal selection
% \cite{greaves_evolutionary_20152015,lipinski_cancer_2016,williams2018}, as well as non-genetic
% variability \cite{Brock2009} and the tumor microenvironment
% \cite{Albini2007}. These have important stochastic components that
% fundamentally limit predictability. The ability
% to understand and predict evolutionary dynamics in cancer would allow for
% better prognosis and treatment. Investigating the mutational profiles of
% evolving clones within a tumor is possible by means of high-throughput
% genome sequencing, but the use of accurate models of tumor evolution is
% still required to unveil the true paths of cancer progression
% \cite{McPherson2017}. This is not of exclusive relevance to cancer but
% also to problems such as the evolution of antibiotic resistance
% \cite{toprak2012} or virulence \cite{Day2004}.

% % what are cancer progression models?
% Cancer Progression Models (CPMs) identify restrictions in the order of
% accumulation of mutations within a tumor using cross-sectional data. They
% encode all the possible mutational trajectories that are compatible with
% the restrictions (figure \ref{fig:fig1}A-B), which can serve to narrow the
% search for the true distribution of tumor progression pathways. However
% their performance to infer these complete evolutionary trajectories, from
% the root genotype with no driver mutations to a final, fixated genotype,
% is limited \cite{diaz-uriarte2019a,hosseini2019a}; this has been
% attributed to various factors, such as violations of the underlying
% assumptions of the models and the difficulty to acquire large, unbiased
% data sets to power them \cite{diaz-uriarte2019a}.

% what are we studying?
% In practice, cancer can only be first detected once it has already
% advanced to some extent.

% Detection of a specific tumor composition at a given point in time
% automatically discards all mutational pathways that do not visit the
% observed state, reducing the amount of possible evolutionary trajectories
% that can be followed from then on and thus potentially facilitating the
% task of predicting further tumor progression.\todo{is it here where we
%   explain how this differs from work with Claudia?} In this work we
% address the question of how accurately CPMs can predict short-term cancer
% evolution after detection. Rather than analyzing full evolutionary
% trajectories, we focus on the partial paths from an observed tumor state
% to the state that comes consecutively after.

The question of what genotype comes next, or what is the state that comes
consecutively after, requires that we carefully define the meaning of
``consecutive'' states in this context. If cancer progresses under
conditions of strong selection and weak mutation (SSWM)
\cite{gillespie1983}, only evolutionary trajectories of increasing fitness
are possible, and the evolution of a new, higher-fitness clone results in
rapid expansion and effective exclusion of previous ones before any new
mutations take place. In this situation, there is a well defined sequence
of genotypes with progressively increasing fitness in the path to tumor
fixation, with the most abundant clone at every time being a direct
descendant of a previously most abundant one itself (figure
\ref{fig:fig1}C). But, in cancer, large population sizes and high mutation
rates can prevent successful clonal sweeps \cite{nichol2019a, zhao2016} so
that instead of SSWM, clonal interference and stochastic tunneling
\cite{devisser_empirical_2014,szendro_predictability_2013,sniegowski_beneficial_2010,
  wodarz_dynamics_2014} become common: multiple clones can coexist at
significant fractions for prolonged periods of time, and new
higher-fitness clones can evolve from low fitness parents that represented
a minority of the population
(Fig~\ref{fig:fig1}D). % As a consequence, genotypes with a large
% number of mutations %high mutational load
% can already be present at the time of sampling yet go undetected if their
% relative abundance is small (figure
% \ref{fig:fig1}D). % Deviations from SSWM
% are known to undermine the predictive power of CPMs
% \cite{diaz-uriarte2018,diaz-uriarte2019a}, as the model underlying most
% CPMs seems to assume SSWM (see Discussion). This assumption, however, is
% not always stated
% explicitly.
% , but it is explicit, for example, in Hosseini et al.'s
% \cite{hosseini2019a} use of CPMs for predicting tumor evolution (see
% \nameref{discussion}).

To formulate the question of ``what genotype comes next'' 
%To accurately raise a question that does not rely on SSWM to be well
%defined,
so that it does not depend on the SSWM assumption we designate an
\textit{observed} genotype as that which is most abundant in the tumor at
the time of sampling. On the other hand, a Line of Descent (LOD) is
defined as a sequence of parent-child genotypes from the wild-type to the
(possibly local) 
fitness maximum where fixation occurs \cite{szendro_predictability_2013}.
By construction, assuming that mutations accumulate one by one and cannot
be lost, each genotype in the LOD has all the mutations of the previous
one plus exactly one more (as it directly descends from it). The LOD has
no more than one genotype with a given number of mutations, whereas in
general two or more observable genotypes could share the same mutation
count (e.g. in figure \ref{fig:fig1}D both the red and green genotypes
have one mutated driver and are observable). If the evolutionary process
does not conform to SSWM, not all genotypes in the LOD are necessarily
observable. From a clinical perspective, it is relevant to use the
information that is detectable to unveil the underlying dynamics of cancer
evolution, encoded in the LOD. In other words, we want to make a
prediction regarding \textit{what comes next in the LOD} based on the
\textit{observation} of a genotype. We thus formulate the following
general question: \textit{Given that a genotype with $\mathrm{n}$
  mutations has been observed, what genotype with $\mathrm{n}+1$ mutations
  is in the LOD?} Formulated like this, the question is unambiguous and
determined. Under SSWM conditions it can be expressed simply as:
\textit{given that a certain genotype has been observed, what genotype
  comes next in the path to tumor fixation?} However, it is important to
emphasize that even if there is no SSWM the more general question remains
well defined.

% how do we do it?

To address this question, we use a collection of evolutionary simulations
from which we obtain the true paths of tumor progression. For every
simulated process we keep track of % what genotypes represent the majority
% of the population as the tumor progressed (i.e., we list 
the most frequent genotypes. Then for each one of them we ask what
genotype has exactly one more mutation while also being a direct ancestor
of the ultimately fixated clone. Averaging over all simulations we obtain
a set of frequencies which represent the conditional probabilities that we
are interested in. We then examine five CPMs: Mutual Hazard Networks (MHN)
\cite{schill2020}, Conjunctive Bayesian Networks (CBN)
\cite{Gerstung2009,Gerstung2011,montazeri_large-scale_2016}, Oncogenetic
Trees (OT) \cite{Desper1999JCB,Szabo2008}, CAncer PRogression Inference
(CAPRI) \cite{capri_bioinformatics,capri_pnas} and CAncer PRogression
Extraction with Single Edges (CAPRESE) \cite{caprese_2014}, some of which
have more than one variant (giving a total of 13 different methods). We
provide these methods with cross-sectional samples taken from the
evolutionary simulations and compare their predictions with the true
probabilities. We explore the factors that affect performance, focusing on
those factors that depend on the properties of the observed genotype. We
want to understand whether CPMs can be used to predict short-term
evolution; for this question, the key response variable is the best
possible performance over all CPMs considered. Although we do address
method choice as a secondary question, this is not our main objective 
\textit{per se}. We discuss the consequences of our results for the analysis of
cancer data by analyzing 25 cancer data sets.

% This work differs from earlier work \cite{diaz-uriarte2019a,hosseini2019a}
% in that we examine a completely different question: previous work dealt
% with predicting full evolutionary trajectories, but here we examine the
% short-term prediction question \textit{Given that a genotype with
%   $\mathrm{n}$ mutations has been observed, what genotype with
%   $\mathrm{n}+1$ mutations is in the LOD?}, which can colloquially be
% summarized as ``What genotype comes next''. In addition, we add a recent
% new method, Mutual Hazard Networks (MHN) \cite{schill2020}, which is very
% different from the rest of the CPMs considered here. Moreover, we are
% examining genotype-specific factors in addition to global fitness
% landscape characteristics: in contrast to the more general, non-local
% descriptors that may not apply to the totality of the fitness
% landscape/evolutionary process, these genotype-specific factors can
% provide more detailed insight regarding the requirements for the methods
% to be reliable in what are arguably the more relevant questions for
% individualized and precision medicine.

% Why does this matter? Predicting important. In fact,

% experimental evolution studies in bacteria and yeast have allowed to
% partially disentangle the role of population size, selection, and
% mutational heterogeneity in parallel evolution \cite{bailey2017} and,
% thus, indirectly on the issues addressed here; using these results for
% cancer, however, would require much more detailed information (e.g.,
% effective population sizes) than is often available.

% So, what can we do with experimental data from cancer? This we examine in
% the discussion blablaba

% \idea{We probably need to make it explicit how it differs from previous
%   work, in particular Diaz-Uriarte and Vasallo.  If we end up doing what I
% think about comparing JS of methods and using that for real data sets, an
% additional comment is ``and we provide specific, practical advice that can
% be followed with the output of methods applied on real data'' ---or
% something of that sort}.
% \idea{Relate to this: Detection of a specific tumor composition at a given point in
% time automatically discards all mutational pathways that do not visit the
% observed state, reducing the amount of possible evolutionary trajectories
% that can be followed from then on and thus potentially facilitating the
% task of predicting further tumor progression. And relate to Bank}

% \idea{Say something about trying to identify, from method output itself,
%   when good performance}

% \idea{We use MHN. First to review that method. We use CPMs, but also
%   include MHN, a newly developed approach that blablab}

% \idea{This is KEY: we do what really matters: next genotype. It really
%   matters because: a) global predictions suck and are generally doomed
%   except maybe to estimate predictability; b) global predictions can
%   not benefit from local features; c) newxt is what matters for therapy
%   and adaptive treatment}.

\section{Methods} \label{methods}

\subsection{Overview}
\label{sec:overview}

To assess if we can use CPMs for short-term predictions of evolutionary
processes, we have compared predicted and true answers to the question
\textit{Given that a genotype with $\mathrm{n}$ mutations has been
  observed, what genotype with $\mathrm{n}+1$ mutations is in the LOD?}
under an extensive set of simulated scenarios. First, we simulated a large
number of tumor evolution trajectories under scenarios with varying
departures from the assumptions of CPM methods; next, we obtained a large
number of simulated tumor evolution trajectories.  Next, we emulated
cross-sectional sampling of the simulations, with varying sampling schemes
and sample sizes. These steps are described in section
\nameref{methods:simulsandsampling}.  The transition probabilities that
provide the true answer to the question \textit{Given that a genotype with
  $\mathrm{n}$ mutations has been observed, what genotype with
  $\mathrm{n}+1$ mutations is in the LOD?}  is explained in section
\nameref{sec:methods:sim}. To obtain the CPM predictions we analyzed each
of the sampled data sets with a total of 13 different CPM variants and
obtained the predictions as explained in section
\nameref{methods:cpm}. Next, we measured how close the predictions were to
the truth (section \nameref{sec:js}) and, finally, we analyzed the major
factors that affect the similarity between predictions and truth (section
\nameref{sec:linear-mixed-model}).  To examine the implications of our
results for the analyses of cancer data, we have analyzed 25 cancer data
sets, described in section \nameref{sec:datasets}.

\subsection{Simulated evolutionary processes and sampling} \label{methods:simulsandsampling}

Simulated data were taken from \cite{diaz-uriarte2019a}, where complete
details are provided. In summary, simulations were conducted under
evolutionary scenarios that differed in the number of genes, the type of
fitness landscape, the initial population size, and the mutation
rates. Landscapes were of either 7 or 10 genes. Fitness landscapes were of
three types: representable, local maxima or Rough Mount Fuji (RMF). The
three types of fitness landscapes incorporate increasing departures from
the assumptions of most CPMs: for the representable fitness landscapes a
DAG of restrictions exists with the same accessible genotypes and
accessible mutational paths; for local maxima fitness landscapes the set
of accessible genotypes can be represented by a DAG of restrictions, but
there are local fitness maxima and the fitness graph has missing paths
\cite{diaz-uriarte2019a}; the genotype with all genes mutated might or
might not be the genotype with largest fitness. The RMF fitness landscapes
\cite{neidhart_adaptation_2014, devisser_empirical_2014,fragata2019}
usually have multiple local fitness maxima and considerable reciprocal
sign epistasis so not even the set of accessible genotypes can be
represented by a DAG of restrictions. On the above fitness landscapes
evolutionary simulations were run. Three different populations sizes
($2\times10^3$, $5\times10^4$ or $1\times10^6$) and two mutation rate
regimes (mutation rate fixed at $10^{-5}$ or uniformly distributed in the
log scale between $0.2\times10^{-5}$ and $5\times10^{-5}$) were used to
generate variability with respect to SSWM. For each combination of the
above factors, 35 independent replicates were obtained for a total of 1260
landscapes (35 replicates $\times$ 2 number of genes $\times$ 3 types of
fitness landscapes $\times$ 3 initial population sizes $\times$ 2 mutation
rate regimes). Then, for each one of them, 20000 independent evolutionary
processes were simulated until fixation of one of the genotypes at a
fitness maximum (local or global). The complete history of the simulations
was stored so that the true LOD for each simulation was known. Each set of
simulations was then sampled under three detection regimes: with equal
probability with respect to the size distribution (\textit{uniform}
sampling), sampling predominantly early in the tumor progression process
(i.e., enriched in \textit{small} tumors), sampling predominantly late in
the tumor progression process (i.e., enriched in \textit{large}
tumors). When \textit{sampling}, we identified the most abundant genotype
and its mutated loci at the chosen time, so the sampling process is
consistent with how we defined \textit{observable} genotypes earlier. No
sampling error is considered. The full sample consists of three binary
matrices of observations (one per detection regime), each with 20000 rows
(one observation per simulation) and as many columns as driver genes. Each
element of the matrix is either 1 or 0 (true or false) depending on which
genes were mutated in which observations. For each detection regime, the
complete sample was used to generate five smaller, non-overlapping splits
of size 50, 200 or 4000. This produced a total of 56700 data sets (1260
fitness landscapes $\times$ 3 detection regimes $\times$ 3 sample sizes
$\times$ 5 splits).

\subsection{Transition probabilities from evolutionary simulations}
\label{sec:methods:sim}

Let us consider a matrix $\mathbf{P}$ where rows correspond to observable
genotypes and columns to genotypes of the LOD. The element $(i,j)$ of the
matrix, denoted as $p_{ij}$, represents the conditional probability that
the genotype $j$ (with $n+1$ mutations) is an element of the LOD given the
genotype $i$ (with $n$ mutations) has been observed. In the absence of
back mutations and when mutations accumulate one by one, $p_{ij}$ is
defined only if $i$ and $j$ are exactly one mutation away from each other.
Note that the genotype $j$ (element of the LOD and with $n+1$ mutations)
need not be a direct descendant of $i$ if SSWM does not hold. We built a
transition matrix $\mathbf{P}$ for every fitness landscape that we
generated by examining each of the 20000 evolutionary processes that were
simulated in the corresponding landscape, listing all observable genotypes
(i.e. those that represented the majority of the population at some point
during the simulated evolutionary process) and extracting the LOD of every
process tracking back the ancestors of the fixated genotype (definition
and details in \cite{szendro_predictability_2013}; using implementation in
\cite{diaz-uriarte2017}).  Each element $p_{ij}$ can thus be interpreted
as the fraction of the times that $j$ was in the LOD given that an
observation of $i$ had been made. See further details in Supplementary
Material, section \ref{S-supp-cpm-details}.

% \subsubsection{Special cases}\label{sec:special}

% When the end of the evolutionary process is reached there are no further
% transitions to genotypes of the LOD. Thus, matrix $\mathbf{P}$, which
% provides the true probability distributions for the question \textit{what
%   genotype of the LOD has $\mathrm{n}+1$ mutations iven that a genotype
%   with n mutations has been observed?} needs to add to additional states
% that correspond to two distinct special cases. When the observed genotype
% is the final, fixated one, we have simply reached the end of the
% evolutionary process; we handle this by adding to the $\mathbf{P}$ an
% ``end" situation (denoted as $p_{i,end}$). If SSWM conditions are not
% satisfied, however, observable genotypes need not be a part of the LOD so
% that there could be observable genotypes with more mutated loci than, or
% as many mutated loci as, the final genotype that gets fixated. In other
% words, it is possible that a clone with $m \geq n$ mutations establishes
% in the tumor for a period of time even if eventually another clone with
% $n$ mutations is the one that fixates.  We handle this by adding to the
% $\mathbf{P}$ a ``none" situation ($p_{i,none}$). Downstream analyses are
% performed including these two probabilities ($p_{i,end}, p_{i, none}$)
% (and row normalization was actually carried out after including $end$ and
% $none$).

% % \subsection{Transition probabilities from evolutionary simulations}
% % \label{methods:sim}

% % Let us consider a matrix $\mathbf{P}$ where rows correspond to observable
% % genotypes and columns to genotypes of the LOD. The element $(i,j)$ of the
% % matrix, denoted as $p_{ij}$, represents the conditional probability that
% % the genotype $j$ (with $n+1$ mutations) is an element of the LOD given the
% % genotype $i$ (with $n$ mutations) has been observed. Naturally, $p_{ij}$
% % is defined only if $i$ and $j$ are exactly one mutation away from each
% % other. Note that, given how our question is formulated, all the mutations
% % in $i$ need to also be present in $j$ only in SSWM, when $i$ would
% % necessarily be the parent of $j$. But in a more general scenario (e.g.,
% % when SSWM does not hold) it is possible that a genotype $i$ with $n$
% % mutations is observed despite not belonging to the LOD. Therefore the
% % genotype $j$ (element of the LOD and with $n+1$ mutations) need not be a
% % direct descendant of $i$. The only condition for $p_{ij}$ to be defined is
% % that $j$ has exactly one more mutation than $i$:

% % \begin{equation}
% % p_{ij} \in \mathbb{R} \iff n_{mut}(j) = n_{mut}(i) + 1
% % \label{eq:pij}
% % \end{equation} 
% % \idea{En realidad, no es $\mathbb{R}$; es $[0, 1]$, puesto que podemos
% %   definirlo como 0. }

% % We built a matrix $\mathbf{P}$ for every fitness landscape that we
% % generated. To obtain the elements $p_{ij}$ of one of them, we ran through
% % all 20000 evolutionary processes that were simulated in the corresponding
% % landscape. In each case, we first listed all observable genotypes
% % (i.e. those that represented the majority of the population at some point
% % during the simulated evolutionary process). Then we extracted the LOD of
% % every process tracking back the ancestors of the fixated
% % genotype (see definition and details in \cite{szendro_predictability_2013}; using implementation in \cite{diaz-uriarte2017}).
% % We identified the genotypes in the LOD with one mutation more than the
% % observable ones, and added 1 to the corresponding entries of
% % $\mathbf{P}$. Finally, we row-normalized $\mathbf{P}$ to meet the
% % normalization condition $\sum_{j} p_{ij} = 1$.

% % % \begin{equation}
% % % \sum_{j} p_{ij} = 1
% % % \label{eq:pij-norm}
% % % \end{equation}\todo{Added this equation, OK?}
% % %
% % Each element $p_{ij}$ can thus be interpreted as the fraction of the times that $j$ was in the LOD given that an observation of $i$ had been made. For a large number of simulations, this fraction represents a good approximation of the true probabilities.

% % \subsubsection{Special cases}\label{sec:special}

% % If SSWM conditions are not satisfied, observable genotypes need not be a part of the LOD and not all genotypes in the LOD are necessarily observable. A particular case of this are observable genotypes with more mutated loci (or just as many) than the final one that gets fixated: it is possible that a clone with $m \geq n$ mutations establishes in the tumor for a period of time even if eventually another clone with $n$ mutations is the one that fixates. This represents a special scenario for our purpose: the answer to out question (\textit{what genotype of the LOD has $\mathrm{n}+1$ mutations?}) is technically undefined if the LOD stops at $n$ mutations. We identify two scenarios: \textit{a)} the observed genotype is simply the final, fixated one, and \textit{b)} the observed genotype has transiently established as the most abundant, but fixation will occur at a different one (with less mutations) evolving through a different pathway.
% % \iffalse

% % \begin{itemize}
% % \item[\textit{a)}] The observed genotype is simply the final, fixated one.
% % \item[\textit{b)}] The observed genotype has transiently established as the most abundant, but fixation will occur at a different one (with less mutations) evolving through a different pathway.
% % \end{itemize}
% % %
% % \fi
% % We refer to \textit{a)} as an ``end" situation (we have simply reached the end of the evolutionary process), and to \textit{b)} as a ``none" situation (the only possible answer to our question is ``there is no such a genotype"). In practice, we handle these cases by attaching two additional columns to $\mathbf{P}$. These contain the probabilities that an observation of a genotype $i$ is made in an ``end" situation (denoted as $p_{i,end}$) or a ``none" situation ($p_{i,none}$). Downstream analyses are performed including these probabilities.

% % \idea{[Move all this chunk (or even almost everything after \ref{eq:pij})
% %   to the Supplementary? I'm worried it may break the flow of the
% %   paper. Would it make sense to simply write a super-simplified methods
% %   with just enough information to follow the reasoning, and move
% %   everything else to the supplementary?]}\todo{something needs to be left
% %   here, maybe all. Not sure yet how much, though}

\subsection{Transition probabilities from CPMs} \label{methods:cpm}

%  \idea{Explain how we go from the output of a CPM (DAG of restrictions,
%    but not in the time-discretized cases?)}\todo{in Diaz-Uriarte and
%    Vasallo (except we stop at the transition matrix between genotyps stage
%    ---I'll write it)}
% \idea{Notation: matrix of predictions from a CPM should be named
%   $\mathbf{\widehat{P}}$, and its elements $\hat{p}_{ij}$. The hat
%   indicates an estimator, which is precisely what CPMs provide: an
%   estimation of the true probabilities $p_{ij}$.}

General overviews of all the methods used here, except for MHN, are
available in \cite{beerenwinkel_computational_2016, Beerenwinkel2014,
  diaz-uriarte2019a,diaz-uriarte2018} and detailed descriptions are given
in the original references for each procedure (MHN \cite{schill2020}, CBN
\cite{Gerstung2009, Gerstung2011}, MCCBN
\cite{montazeri_large-scale_2016}, OT \cite{Szabo2008, Desper1999JCB},
CAPRI \cite{capri_bioinformatics, capri_pnas}, CAPRESE
\cite{caprese_2014}). Briefly, CBN, MCCBN, OT, CAPRESE, and CAPRI try to
identify restrictions in the order of accumulation of mutations from
cross-sectional data. CAPRESE and OT code these restrictions using trees
(i.e., under these two methods each event, such as a mutation, can
immediately depend on at most one other event); although the output of OT
and CAPRESE is often similar, the procedures are different (in OT weights
along edges can be directly interpreted as probabilities of transition
\cite{Szabo2008}, whereas CAPRESE uses a probability raising notion of
causation based on Suppes' probabilistic causation). CBN, MCCBN, and CAPRI
code these restrictions using DAGs (i.e., each event can depend on
multiple other events).  CBN and MCCBN are very closely related methods:
the CBN version we use is H-CBN as described in \cite{Gerstung2009,
  Gerstung2011}, and MCCBN is an implementation of the CBN model that uses
a Monte Carlo Expectation-Maximization (EM) algorithm for fitting the CBN
model \cite{montazeri_large-scale_2016} (instead of the simulated
annealing nested within EM of H-CBN); in contrast, CAPRI tries to identify
probability raising in the framework of Suppes' probabilistic
causation. Common to most of these methods is a model of deterministic
dependencies \cite{schill2020}, a model of accumulation of mutations where
an event (a mutation) can only occur if all its dependencies are satisfied
(though different methods allow for small error deviations from this
requirement); CAPRI and CAPRESE try to recover ``probability raising''
relations and, strictly, their trees/DAGs do not encode deterministic
dependencies but will be treated as doing so in this paper, since
obtaining probabilities of transition from them is not possible otherwise
---see details in section \nameref{S-supp-trans-prob} in the Supplementary
Material.

For all of these methods, since a DAG of restrictions determines a fitness
graph (see Fig.~\ref{fig:fig1} A-B and
\cite{diaz-uriarte2018,diaz-uriarte2019a}), we can obtain the set of
possible descendants for a genotype directly from this fitness graph.  For
CAPRESE and CAPRI all transitions were set as equiprobable (as
probabilities of transition are not available) but for CBN, MCCBN, and OT
we can compute the probabilities of transition between genotypes. CBN and
MCCBN return the parameters of the waiting time to occurrence of each
mutation given its restrictions are satisfied which allows us to compute
transition probabilities between genotypes using competing exponentials
(see \cite{montazeri_large-scale_2016, hosseini2019a, diaz-uriarte2019a})
and a similar procedure can be used with OT by using the edge weights
(though, strictly, the OT model, being an untimed oncogenetic tree does
not return these probabilities ---see \cite{diaz-uriarte2019a}). Mutual
Hazard Networks (MHN) \cite{schill2020} differ from the previous methods
in that events are modeled by a spontaneous rate of fixation and a
multiplicative effect each of these events can have on other events (i.e.,
pairwise interactions), which allows it to model both enabling and
inhibiting dependencies. The output of MHN is a transition rate matrix; as
for CBN and MCCBN, we can obtain the probability of transition to each
descendant genotype, given a transition, using competing exponentials from
the transition rate matrix (see equation 2 and Fig.~2 in
\cite{schill2020}). Details on software and computation of transition
probabilities are provided in the Supplementary Material, sections
\nameref{S-supp-cpm-details} and \nameref{S-supp-trans-prob}.

% \subsubsection{Time-discretized CPMs}\label{sec:td}

None of the CPM models considered here allow us to accommodate local
maxima: under all of the models, the genotype with all genes mutated is
always reached with probability 1 as time goes to infinity (this is also
the case for MHN, even if it can include inhibiting effects). To overcome
this limitation without making specific additional assumptions about the
sampling process, and for MHN, CBN, and MCCBN, we have used the
uniformization method \cite{grassmann1977,vandijk2018} (see also
\cite[Fig.\ 6]{schill2020}) to approximate the continuous-time Markov
chain ($\mathbf{Q}$) by a discrete time Markov chain 
(with transition matrix $= \mathbf{I} + \frac{1}{\gamma}\mathbf{Q}$, where
$\gamma = \max(|q_{ii}|)$). We will interpret the diagonal entries of the
time-discretized transition matrix as a lower bound on the probability
that the given genotypes behave as local maxima. The output from CBN, MHN,
and MCCBN after time discretization will be referred to as CBN\_td,
MHN\_td, and MCCBN\_td, respectively, and we will differentiate this
\textbf{time-discretized} results from those obtained, as explained above,
using competing exponentials. We will refer to these methods as the
\textbf{TD} methods (for time-discretized) whereas their counterparts,
CBN, MHN, MCCBN, obtained via competing exponentials on the rate
matrix, woll be referred below as the CE methods. OT, CAPRESE, and CAPRI do not
belong to any of those two groups, though none of them can accommodate
local maxima either; thus, when we refer to the ``non-TD'' methods, that
includes MHN, CBN, MCCBN (in their ``non-TD'' versions) as well as OT,
CAPRESE, and CAPRI.
%%CEr
%(``CE'' in Figure \ref{fig_biol_comp}).

% \idea{We should justify that much better. I don't see how. And first
%   sentence is the same as saying that the limiting distribution is
%   $\pi_{gentoypeWithAllGenesMutated} = 1$. Not sure that clarifies things
% for most readers.}

In summary, a total of 13 methods have been used: CBN, in three variants
(CBN, with transition probabilities using competing exponentials
conditional on transition; CBN\_uw, obtained from the former by using
equiprobable transitions; CBN\_td, the time-discretized version of CBN);
MCCBN, again in three variants (MCCBN; MCCBN\_uw; MCCBN\_td);
MHN in two variants (MHN and MHN\_td); OT in variants OT and OT\_uw
(equiprobable transitions); CAPRESE; CAPRI, in two variants, CAPRI\_AIC,
CAPRI\_BIC (which differ on the penalty used, AIC or BIC). Each of these
methods returns a matrix $\mathbf{\hat{P}}$ of predicted transition
probabilities.

% \idea{MHN: give a few details of how we used it. Basically, default with
%   parameters for balbla}\todo{explain something about the All others too?
%   CBN, OT, etc? It will
% be the fourth time I do it! Or refer to previous papers and just add a
% sentence for MHN? -- I would definitely just add the references, maybe a short general explanation of "traditional" CPMs and then add a couple more sentences for MHN (and maybe time-discretized versions of CBN and MCCBN)}

% \idea{misra: discuss how it is similar or different. Or in the intro. Or
%   in the methods. Who cites Misra?. Misra looks similar, but it gives a
%   0/1 decision, and has a parameter, threshold, `` (any number between 0
%   and 1), smaller numbers give you more detail (give a number between 0
%   and 1 that specifies the level of detail in output path file, this file
%   will have a .dot extension. We used 0.3 in the manuscript)'' And Misra
%   seems to only find ancestor of existing genotypes. So genotypes that are
%   not present (at least with many mutations) are never predicted to
%   exist. It is a somewhat different method, with difficult to use code,
%   and poorly explained arguments. Its output are not really probabilities
%   of paths.  }

\subsection{Quantification of similarity between predicted and true
  transition probabilities} \label{sec:js}

For every observed genotype $i$ we have compared the vector of true
probabilities $\mathbf{p}_i$, from the $i$th row of $\mathbf{P}$ with the
analogous vector of predictions from a CPM $\mathbf{\hat{p}}_i$ returned
from the $i$th row of $\mathbf{\hat{P}}$ for each CPM. We measured the
similarity between these two vectors using the Jensen-Shannon distance
($JS$) \cite{crooks2017}, the square root of the Jensen-Shannon divergence
\cite{lin1991}. The Jensen-Shannon divergence is a symmetrized
Kullback-Leibler divergence between two probability distributions that is
0 when the two distributions are identical and reaches its maximum value
of 1 if the two distributions do not overlap (we used log of base 2). See
further details in Supplementary Material, section \ref{S-supp-js}.

% Consider the vector of true probabilities
% $\mathbf{p}_i=\left(p_{ia},p_{ib},\cdots,p_{iz}\right)$, where
% $a,b,\cdots,z$ are all genotypes with one mutation more than $i$: this is
% nothing but the $i$-th row of the matrix $\mathbf{P}$ (excluding the
% elements that do not satisfy $n_{mut}(j) = n_{mut}(i) + 1$). Consider also
% the analogous vector of predictions from a CPM $\mathbf{\hat{p}}_i$. We
% measured the similarity between these two vectors using the Jensen-Shannon
% distance ($JS$) \cite{crooks2017}, the square root of the Jensen-Shannon
% divergence \cite{lin1991}. The Jensen-Shannon divergence is a symmetrized
% Kullback-Leibler divergence between two probability distributions that is
% 0 when the two distributions are identical and reaches its maximum value
% of 1 if the two distributions do not overlap (we used log of base 2); it
% is defined even if the two distributions do not have the same sample space
% (i.e., even if $p_i \ne 0$ and $q_i = 0$ or viceversa). (And the square
% root of the Jensen-Shannon divergence is a metric between probability
% distributions \cite{crooks2017})

% \idea{CHECK: ``excluding the undefined elements that do not satisfy the
%   condition in equation ref{eq:pij}'' : En realidad, no need to exclude
%   them; just set them to 0. This is equivalent. Pero dejo el ``excluding
%   the elements that do not satisfy $n_{mut}(j) = n_{mut}(i) + 1$'' que
%   juega el mismo papel}.

% Using the Jensen-Shannon distance, the square root
% of the Jensen-Shannon divergence is a metric between probability
% distributions.

% The Jensen-Shannon distance between them is defined as \cite{crooks2017, lin1991}.

% \begin{equation}
% JS\left(\mathbf{p}_i,\mathbf{\hat{p}}_i\right) = \sqrt{\frac{1}{2}d_{KL}\left(\mathbf{p}_i,\mathbf{q}\right) + \frac{1}{2}d_{KL}\left(\mathbf{\hat{p}}_i,\mathbf{q}\right)}
% \label{eq:js}
% \end{equation}
% %
% where we have used the Kullback–Leibler divergence, defined for two arbitrary vectors $\mathbf{x}=\left(x_1,x_2,\cdots\right)$ and $\mathbf{y}=\left(y_1,y_2,\cdots\right)$ as

% \begin{equation}
% d_{KL}\left(\mathbf{x},\mathbf{y}\right) = \sum_i x_i \log_2 \frac{x_i}{y_i}
% \label{eq:kl}
% \end{equation}
% %
% $JS$ is bounded between 0 and 1, is minimal when $\mathbf{p}_i$ and
% $\mathbf{\hat{p}}_i$ are identical and largest when they are maximally
% different. \idea{We opted for this metric because... and it is defined
%   even when pi or qi are 0 (?)}

% \idea{Move this entire chunk to the supplementary and just leave a sentence like ``to quantify the concordance between the true probabilities and the CPMs predictions, we used the Jensen-Shannon distance $JS$. This is a measure of disparity across probability distributions which is zero when the two are identical and has an upper bound of 1 for maximal dissimilarity (Supplementary Methods). We opted for this metric because... (?)"}

%% No longer relevant
% \subsubsection*{Averaging statistics with weights}
% \idea{Maybe move this to the supplementary and just leave a sentence here like "for those analyses in which the row-wise (genotype-wise?) statistics were averaged across the entire matrix (across all genotypes?), we gave each row (each genotype?) a weight according to the frequency with which it appeared in the corresponding sample provided to the CPMs (Supplementary Methods)"}

% \subsubsection{Mismatches in genotype observability and null model}
% \label{sec:mismatch_null}

% In the simulations, the most common genotype at any of the sampling times
% is always observed. But not all genotypes are always accessible (and,
% thus, observable) for all CPMs; if a genotype $j$ is not observable
% according to a CPM, all $\hat{p}_{jk}$ are undefined for all $k$ (no
% prediction can be made regarding what comes after $j$, since $j$ is not
% even observable). How do we compare the true probabilities ($p_{ij}$) with
% the estimations provided by the CPM ($\hat{p}_{ij}$) in these cases? If
% the genotype under consideration ($i$) is not observable in the
% simulations we simply exclude it from further analyses: since it is not
% going to be observed in practice, it is irrelevant what the CPM predicts.
% If it is observable in the simulations but not deemed so by the CPM, since
% no prediction is provided we use that of a simple null. Under this simple
% null model, all transitions from genotype $i$ to all genotypes with
% $n + 1$ mutations plus the $end$ and $none$ states are considered
% equiprobable.  If the genotype is observable according to both simulations
% and CPM, we calculate the Jensen-Shannon distance normally as explained
% earlier.  See Supplementary Material, section
% \ref{S-sm-mismatches} for further details.

% % model instead ($\mathbf{p}^0_i$). See Supplementary Material, section
% % \ref{S-sm-mismatches} for further details.

% % Else, if the genotype is observable according to both
% % simulations and CPM, we calculate the Jensen-Shannon distance normally as
% % explained earlier.

\subsection{Factors that affect predictive performance: linear mixed-effects model trees}
\label{sec:linear-mixed-model}

To examine the relevance of the different factors on the performance of
methods we used linear mixed-effects model trees. These are an extension
of recursive partitioning (or tree-based) methods, where observations are
first split repeatedly according to the predictor variables, which play
the role of partitioning variables, so that the dependent variable becomes
more homogeneous within each node; the linear mixed-effects model allows
for the addition of random effects \cite{fokkema2018, fokkema2020}. In our
models, the dependent variable is the minimal JS over all methods (for
each genotype by replicate by id by detection regime by sample size).  The
possible partitioning variables include characteristics that are fitness
landscape- and evolutionary dynamics-specific, sampling characteristics,
genotype (within fitness landscape)-specific, and genotype (within fitness
landscape) by sampling (sample size and detection regime) by
replicate-specific variables.  The fitness landscape- and evolutionary
dynamics-specific variables are % total number of
% loci or genes (numGenes in the figures),
SSWM regime (\texttt{SSWM} in the figures, measured as the average frequency of the
most frequent genotype), %freq\_most\_freq\_mean\_no450}
$\gamma$ (\texttt{gamma} ---this is not the same gamma as used in the
time-discretization of transition rate matrices) a measure of the amount
of epistasis in the fitness landscape, defined as ``the correlation of
fitness effects of the same mutation in single-mutant neighbors''
\cite{ferretti_measuring_2016} or, equivalently, ``the correlation in
fitness effects between genotypes that only differ by 1 locus, averaged
across the [fitness] landscape'' \cite{brouillet_magellan:_2015}, fraction
of pairs of loci with reciprocal sign epistasis (\texttt{epistRSign}), and number
of observed peaks in the fitness landscape (\texttt{numObservedPeaks}). The
sampling characteristics are detection regime (\texttt{detect} in the figures) and
sample size (\texttt{sample\_size}) ---recall that each fitness landscape was
sampled over all nine combinations of detection regime and sampling
size. The genotype (within fitness landscape)-specific variables are
number of mutations (\texttt{nMut} in the figures), fitness rank (\texttt{fitnessRank},
where the genotype of maximum fitness in a landscape has fitnessRank 1),
and the proportion of times the genotype is a (possibly local) fitness
maximum (\texttt{propLocalMax}) . The genotype (within fitness landscape) by
sampling by replicate specific variables are proportion of
times that genotype was observed in the sample (\texttt{observedProp}), and
difference between the observed proportion and the true frequency of that
genotype in the samples (where the true frequency is the frequency over
the 20000 samples) (\texttt{diff\_obs\_prop}). Random effects for the global model
are ID and, nested within ID, the crossed random effects of genotype and
replicate. As the meaning of the number of mutations, nMut, is different
in models with 7 and 10 genes, we fitted separate models for simulations
with 7 and 10 genes.
Models were fitted using the R package ``glmertree''
\cite{fokkema2018}. The Bonferroni-corrected significance level for
node-splitting was set at 0.01. We specified a minimal (weighted) size in
leaves (or terminal nodes) of 1\%; as the returned trees had over 50
leaves in all cases, to allow for interpretability we pruned the resulting
tree by recursively merging (starting from the leaves) all children node
with a fitted minimal JS $>$ 0.0677 (which corresponds to the 20\% best JS
---see Table \ref{S-tab_q_all} in Supplementary Material). In addition, we
merged all children with a fitted minimal JS that differed by less than
0.0677/4 (so as to collapse good performing nodes with minor
differences). Further details about the fitting procedure are provided in
the Supplementary Material section \ref{S-sec:data-proc-lmertree}.

% The possible partitioning variables include characteristics that are
% fitness landscape- and evolutionary dynamics-specific, sampling
% characteristics, genotype (within fitness landscape)-specific, and
% genotype (within fitness landscape) by sampling (sample size and detection
% regime) by replicate-specific variables. Random effects for the global
% model are ID and, nested within ID, the crossed random effects of genotype
% and replicate.  Models were fitted using the R package ``glmertree''
% \cite{fokkema2018}. The Bonferroni-corrected significance level for
% node-splitting was set at 0.01. We specified a minimal (weighted) size in
% leaves (or terminal nodes) of 1\%; as the returned trees had over 50
% leaves in all cases, to allow for interpretability we pruned the resulting
% tree by recursively merging (starting from the leaves) all children node
% with a fitted minimal JS $>$ 0.0677 (which corresponds to the 20\% best JS
% ---see Table \ref{S-tab_q_all} in Supplementary Material). In addition, we
% merged all children with a fitted minimal JS that differed by less than
% 0.0677/4 (so as to collapse good performing nodes with minor
% differences). Further details about the fitting procedure are provided in
% the Supplementary Material section \ref{S-sec:data-proc-lmertree}.

\subsection{Cancer data sets}
\label{sec:datasets}

We have used a total of 25 data sets. Twenty two of them were used in
\cite{diaz-uriarte2019a}. These 22 data sets include six different cancer
types (breast, glioblastoma, lung, ovarian, colorectal, and pancreatic
cancer), use different types of features (nonsynonymous somatic mutations,
copy number alterations, or both) analyzed in terms of pathways,
functional modules, genes, gene events, or mutations (yielding from 3 to
192 different features), and have samples sizes from 27 to 594; the
original sources are \cite{bamford2004, TCGA:2008-g, Jones2008,
  parsons2008b, Wood2007, Brennan:2013, ding2008, TCGA2011, Knutsen:2005,
  Piazza:2013, cgan2012} and all have been used previously in CPM research
\cite{Attolini2010a, Gerstung2011, Cheng2012, Misra2014, caprese_2014,
  capri_bioinformatics, capri_pnas, diaz-uriarte2019a}; for further
details about those 22 data sets, see the supplementary material (file
S5\_Text) in \cite{diaz-uriarte2019a}. We have also used
three CGH data sets previously used in \cite{Gerstung2009, schill2020}
that were originally obtained from the Progenetix database
\cite{baudis2001}. These are \texttt{Breast-CGH}, from 871 breast cancer
patients with 10 CGH alterations, \texttt{Renal-CGH}, from 251 renal cell
carcinoma with 12 CGH alterations, and \texttt{Colon-CGH}, from 570
colorectal cancers with 11 CGH alterations. These three data sets were
obtained from the Supplementary data of \cite{schill2020}. Thus, these 25
cancer data sets constitute a wide, representative example of data to
which researchers have applied or might want to apply CPMs. We have
analyzed all the data sets with all the methods considered; because
computing time increases nonlinearly with number of features for some
methods, and in particular CBN (and, to a lesser extent, MHN), have
difficulties with 16 or more features, for the data sets with more than 15
features, we have used the common procedure \cite{Diaz-Uriarte2015} of
selecting the 15 most frequent features.

%% Example of issues of MHN with many features:
%% example-MHN-issues-with-many-features.Rout
%% It is not the rates, but they provide no code to compute the transition
%% between genotypes, and that is where things fail for me.

% Using JS we computed the similarity for the prediction of the next
% genotype between pairs of methods for all genotypes that were observed at
% least once in the samples. % (These are not necessarily all possible
% % genotypes; moreover, of the methods considered, MHN gill generate
% % predictions for all possible genotypes, but for all other methods the
% % accessible genotypes is generally a much smaller set of genotypes which
% % need not be identical to the set of observed genotypes).
% Based upon the results of the simulations, we will focus, for conciseness,
% on the similarities between MHN and CBN as these are the methods than can
% model more complex dependency relationships between mutations and show
% better performance.

% \idea{Explain variables in figure legends; see the figure
%   legend itself}

% Finally, we have used the

% Glioblastoma: used by \cite{schill2020}. 261 patients, with 486 point
% mutations (M), amplifications (A), or deletions (D). But this data set was
% also used in \cite{cristea_pathtimex:_2016} (``pathTiMEx: Joint Inference of Mutually Exclusive Cancer
% Pathways and Their Progression Dynamics SIMONA CRISTEA, 1'') where they
% say ``consisting of point mutations and copy number aberrations''. But all
% come from the TCGA. Which is the source for GBM CNA.

% So possibly overlap.

% \begin{figure}[H]
%   \centering
%   \includegraphics[width=14.5cm,keepaspectratio]{figs/figexp-crop}
%   \internallinenumbers
%   \caption[Figure title]{\textbf{Figure title.} Figure caption. \idea{Creo    que esto ya no es relevante? FIXME}} \label{fig:figexp}
% \end{figure}

\section{Results}\label{results}

% \subsection{Predictive performance on simulated data}%Linear mixed-effects
% % model trees}
% \label{sec:linear-mixed-model-results}

Figure \ref{fig:7_spec} shows the liner mixed-effects model tree for
fitness landscapes with seven genes under the uniform detection regime and
with the largest sample size (4000). In the Supplementary Material
(Fig.~\ref{S-fig:10_spec}, section
\ref{S-suppl-linear-mixed-model-results}) we show the results for 10
genes. Figures \ref{S-fig:7_ALL} and \ref{S-fig:10_ALL}, in the
Supplementary Material, present the models fitted to all detection regimes
and sample sizes. In these figures we can see the relevance of detection
regime and sample size: the smallest JS were generally achieved under
the uniform detection regime and with the largest sample size (4000) (see
also Supplementary material, Tables \ref{S-tab_minjs} to
\ref{S-tab_q_rmf}).  
% \ref{S-tab_minjs_fl_mean}, \ref{S-tab_minjs_fl_median}, and Tables
% \ref{S-tab_q_all} to \ref{S-tab_q_rmf}).
Moreover, there were
interactions between detection regime and sample size on the one hand and
genotype-specific and fitness landscape-specific variables on the other;
these interactions can make it harder to understand the relevance of
fitness landscape-specific and genotype-specific variables on the quality
of predictions. Figures \ref{fig:7_spec} and \ref{S-fig:10_spec} isolate
the effects of fitness landscape and genotype-specific variables, which is
one of the main objectives of this paper.

[Fig.~\ref{fig:7_spec} about here]

Best achievable performance is generally larger (i.e., JS is lower) in the
representable than in the local maxima and RMF fitness landscapes, as can
be seen in node 4 of Fig.~\ref{fig:7_spec} (see also Tables
\ref{S-tab_minjs_fl_mean}, \ref{S-tab_minjs_fl_median}). Moreover, gamma,
reciprocal sign epistasis, and number of observed peaks, three of the
variables that play important roles in the fitted trees (e.g., nodes 5,
11, 14, 15, and 23 in Fig.~\ref{S-fig:7_ALL}; node 5 in
Fig.~\ref{S-fig:10_spec}; nodes 12 and 15 in Fig.~\ref{S-fig:10_ALL}),
allow to perfectly separate the RMF from the rest of the fitness
landscape, and to separate most local maxima fitness landscapes from the
representable ones (Fig.~\ref{S-fig:gamma_rsign_obs_peaks} in the
Supplementary Material). The results indicate, though, that highly
specific fitness landscape characteristics (fitness landscape itself,
gamma, reciprocal sign epistasis, number of observed peaks) show complex
interactions with genotype-specific characteristics. For example, in
Fig.~\ref{fig:7_spec}, fitness landscape plays a role in node 4 only for
genotypes with small fitness rank (i.e., large relative fitness of a
genotype compared to the rest of the genotypes in the fitness landscape)
and that are rarely local maxima. Likewise, number of observed peaks in
node 19, which would separate representable from local maxima and RMF
landscapes, is relevant for the one-before-last genotype. Similar
interactions between genotype-specific and fitness landscape-specific
characteristics can be seen in the figures in the Supplementary Material
(Fig.~\ref{S-fig:7_ALL}: node 23, child of genotype-specific nodes 1 and
19; Fig.~\ref{S-fig:10_ALL}: nodes 15 and 12, mediated by
genotype-specific nodes 1, 11, and 14; Fig.~\ref{S-fig:10_spec}: node 5,
affected and which affects ancestor and descendant nodes of
genotype-specific characteristics). A general observation is that deviations
from SSWM lead to poorer predictive ability; these effects can depend both on
general fitness landscape characteristics (e.g., node 5 in Fig.~\ref{fig:7_spec})
and genotype-specific characteristics (e.g., node 21 in Fig.~\ref{S-fig:7_ALL}
and node 19 in Fig.~\ref{S-fig:10_ALL} in Supplementary Material).

Not all genotype-specific characteristics had a relevant effect. Observed
proportion of a genotype and the difference between its observed proportion and
true frequency in the sample are variables that do not appear in the final trees
(although the former did appear in the models before prunning). Number of
mutations had a non-linear effect: genotypes with either none or close to
maximum number of mutations lead to better performance, whereas genotypes with
intermediate number of mutations had worse performance (e.g., nodes 1 and 14 in
Fig.~\ref{fig:7_spec}, nodes 1, 3, 6 in Fig.~\ref{S-fig:10_spec}, nodes 2 and 6
in Fig.~\ref{S-fig:10_ALL}). This indicates that prediction becomes harder when
the number of potential next genotypes increases, which happens at intermediate
numbers of mutations.
%(\idea{rm this? Note that this pattern
  % is present for all methods, both the MHN methods, that always provide
  % $\hat{p}_{ij}$, as well as the rest of the methods where, as explained
  % in section \ref{sec:mismatch_null}, we used a simple equiprobable null
  % model when the observed genotype was not among the accessible genotypes:
  % see Supplementary Material \ref{S-JS-num-mut}}).
Smaller fitnessRank
(i.e., large relative fitness of a genotype) was generally associated to
better predictions (e.g., nodes 9 and 15 in Fig.~\ref{fig:7_spec}; nodes
12, 13, and 4 in Fig.~\ref{S-fig:10_spec}) but, as with most other
characteristics, the effect was strongly affected by other variables
(node 1 in Fig.~\ref{S-fig:7_ALL}, nodes 11 and 21 in
Fig.~\ref{S-fig:10_ALL}). Note that genotypes of fitness rank 1 are
necessarily fitness maxima.

In fact, how often a genotype is a local fitness maximum (propLocalMax) not only
affected performance but also which methods were better. This is apparent in
node 3 in Fig.~\ref{fig:7_spec}, node 2 in Fig.~\ref{S-fig:10_spec}, node 19 in
Fig.~\ref{S-fig:7_ALL}, and node 1 in Fig.~\ref{S-fig:10_ALL}: when a genotype
was frequently (over 20 to 50\%, depending on the data set) a local maximum, the
TD methods lead to much better performance that the other methods. Moreover,
once the frequency of local maxima was large, making it even larger lead to
better performance (node 11 in Fig.~\ref{fig:7_spec}; nodes 14, 17, and 19 in
Fig.~\ref{S-fig:10_spec}; node 19 in Fig.~\ref{S-fig:7_ALL}; node 14 in
Fig.~\ref{S-fig:10_ALL}). This phenomenon can also be looked at from the
perspective of global fitness landscape characteristics: an increase in the
number of local fitness peaks (numObservedPeaks) can lead to the TD methods
being better than their CE counterparts (node 19 in Fig.~\ref{fig:7_spec}) or to
the TD methods improving their performance (node 15 in
Fig.~\ref{S-fig:10_ALL}).This is also observed with increasing reciprocal sign
epistasis (episRSign ---nodes 5, 11, 14 in Fig.~\ref{S-fig:7_ALL}), a parameter
that can almost perfectly separate the RMF and local maxima landscapes from the
representable ones (see Fig.~\ref{S-fig:gamma_rsign_obs_peaks}). Note that
epistRSign no longer plays a role in a switch between methods in node 23, since
this node only splits genotypes in fitness landscapes that cannot be
representable (genotypes that have a probability of being local maxima $> 0$
even when their fitness rank is $> 1$).

Scenarios with very good TD performance (JS less than approximately 0.1)
comprise, then, between 1 and 5\% of the cases (4\% in nodes 10 and 14 in
Fig.~\ref{fig:7_spec}; 5\% in nodes 16, 20, and 21 in Fig.~\ref{S-fig:10_spec};
1\% in node 24 of Fig.~\ref{S-fig:7_ALL}; 2\% in nodes 18 and 22 of
Fig.~\ref{S-fig:10_ALL}; see also Tables~\ref{S-tab_q_all} to
\ref{S-tab:xtabs_ce1}). In these cases, MHN\_td was generally slightly better
performing than CBN\_td (and MCCBN\_td was generally the worse of the TD
methods). Within the non-TD methods, when they could do well, most methods had
roughly similar performance (except for CAPRI, that was normally the worse
performer), with a slight advantage of CBN and MHN. CBN was generally among the
best performers, and scenarios where MHN was the single best method were rare
(but see node 20 in Fig.~\ref{fig:7_spec}). For methods with both a weighted and
an unweigthed version (CBN\_uw, MCCBN\_uw, OT\_uw) using weights never lead to
worse performance and sometimes improved it. As with the TD methods, the CE
methods were consistently very good only in limited scenarios (see also
Tables~\ref{S-tab_q_all} to \ref{S-tab:xtabs_ce1}. For CE methods these were
nodes 7 and 20 in Fig.~\ref{fig:7_spec} which comprised about 7\% of the
(weighted) observations, and slightly worse in nodes 6 and 16 ---where MHN's
performance was surprisingly poor--- that comprised another 6\% of the
observations. Similar patterns are seen in Figures \ref{S-fig:10_spec} (node 23,
with 5\% of observations), \ref{S-fig:7_ALL} (node 17, and with worse
performance 6, 12, 16, comprising overall 6\% of observations) and
\ref{S-fig:10_ALL} (node 10 and, with worse performance, 9, both with 1\% of the
observations).

% In summary, then, there were complex interactions between fitness
% landscape-specific characteristics, sampling, and genotype-specific
% characteristics that affected the best achievable performance; for
% scenarios where the best achievable performance was good, these three
% factors also affected which methods were the best. In general, the best
% performance was seen with large sample sizes, a uniform detection regime,
% with representable fitness landscapes and under a SSWM regime. Genotypes
% with maximum or close to maximum fitness (i.e., small fitness rank) with
% either 0 or close to the maximum number of mutations, are those for which
% predictions tend to be better. In terms of method choice, and when the
% minimal achievable JS can be small (i.e., where it matters) there were
% sharply contrasting differences between the 13 methods examined: either
% the TD methods (those that can model local maxima) MHN\_td and CBN\_td
% both do a very good or both do a very bad job, with MHN and CBN presenting
% the opposite pattern (generally both a very bad job or a very good job).
% If the genotype can be a local maximum, then the more often it is a local
% maximum, the better the prediction, and when a genotype is frequently a
% local maximum a simple approach like the TD approach with CBN or MHN can
% perform well.

% \idea{some of the above repeat in first paragraph of discussion}

\section{Discussion: implications for the analysis of biological data sets} \label{discussion}

Can we use cancer progression models (CPMs) to predict the short-term
evolution of a tumor? Informally, we could word this question as ``what
genotype comes next in the path to tumor fixation?''. More precisely, we
can ask whether, conditional on us observing a particular genotype with
$\mathrm{n}$ mutations as the most abundant genotype in a tumor, we can
predict what genotype with $\mathrm{n}+1$ mutations is in the Line of
Descent (LOD) to the fitness maximum (where fixation occurs). We have
examined the performance of 13 different CPM procedures for predicting
short-term evolution, focusing mainly on the best achievable performance
over all 13 methods (i.e., on how close to the truth are the predictions
of short-term evolution, regardless of method) and, secondarily, on the
choice of method. The analysis of simulated data indicated that it was
possible to provide predictions of short-term evolution that closely
matched the true probabilities of short-term evolution only under some
specific combinations of genotype characteristics and fitness landscape
characteristics. And when  good performance was possible, no method or
set of methods was consistently superior, making choice of best method
also dependent on genotype and fitness landscape properties.

%\change{And when  good performance was possible, no method or
%set of methods was consistently superior, making choice of best method
%dependent on genotype and fitness landscape characteristics.}{These characteristics
%also determined the choice of the best method or group of methods, since even in
%the cases where good performance was possible no specific method was consistently
%superior.}\textcolor{orange}{Hummm... no estoy seguro del rewording: it misses the
%emphasis en la dependencia de la decisión. Y el empezar la frase con el ``And''
%era para añadir ``dramatismo'' a la cosa} % Application
%% of the best performing methods to 25 biological data sets showed wide
%% variability in predictions and the difficulty of choosing the specific set
%% of predictions to trust because of the lack of detailed information about
%% genotype and fitness landscape characteristics.

% With the simulated data (i.e., data for which we can measure the deviation
% of predictions from the truth) b
Best achievable performance was affected by
complex interactions between fitness landscape-specific
characteristics% (type of fitness landscape and associated variables gamma ---correlation
% of fitness effects in single-mutant neighbors---, reciprocal sign
% epistasis, and number of peaks in the landscape)
, deviations from the
strong selection and weak mutation (SSWM) regime, sampling%  (sample size
% and whether tumors are sampled early, late, or uniformly during tumor
% evolution)
, and genotype-specific characteristics%  (fitness rank, number of
% mutations, and probability of being a local fitness maximum)
.  The
best performance was seen with large sample sizes, a uniform detection
regime, with representable fitness landscapes and under a SSWM
regime.  % In general (see
% Fig.~\ref{fig:7_spec} and Supplementary Material
% Figs.~\ref{S-fig:10_spec}, ~\ref{S-fig:7_ALL}, ~\ref{S-fig:10_ALL})
Genotypes with maximum or close to maximum fitness (i.e., small fitness
rank) with either 0 or close to the maximum number of mutations, are those
for which predictions tend to be better.  For scenarios where the best
achievable performance was good (i.e., where it matters), these three
factors also affected which methods were the best. There were sharply
contrasting differences between methods, in particular, methods
that can model local maxima ---the TD methods MHN\_td and CBN\_td--- tended to
either both
perform very well or both perform very bad. The corresponding two methods
that do not model local maxima ---MHN and CBN--- generally present the opposite
pattern, performing well when the TD methods do not and vice-versa.

Since our focus are predictions conditional on the observed genotype, we
can also summarize the results by dividing genotypes into three groups: A)
Genotypes for which using the TD methods lead to very good predictions (JS
$< 0.1$); B) Genotypes for which the CE methods can provide very good
predictions (JS $< 0.1$); C) The remaining cases, which constitute the
majority of the cases, where performance was generally poor (JS $> 0.25$).
Group A consists of genotypes that are more often than not local maxima
(probability of being a local maximum $>0.5$) and of large fitness (small
fitnessRank) or that are often local maxima (probability of being a local
maximum $>0.25$) while deviations from SSWM are minor. Fitness
landscape-specific characteristics that make these genotypes 
more likely to be observed
are large number of local fitness peaks and large reciprocal sign epistasis.
For these genotypes, the TD methods can
correctly predict being at the end of the evolutionary process, something
that the remaining methods cannot account for, which explains the very
poor performance of the non-TD methods in these cases (JS $> 0.75$). In
group B are genotypes of large relative fitness, no mutations or with all
except one loci mutated, in evolutionary scenarios with SSWM and smooth
fitness landscapes (little reciprocal sign epistasis, large gamma) and few
local peaks (less than 3). For these genotypes, the TD methods make very
poor predictions (JS $> 0.75$) while MHN and CBN do particularly good jobs, but
CAPRESE and OT, two methods that can only model simpler patterns of
dependencies, generally perform well too. In the remaining cases (group
C), no method can provide good predictions. Assessing whether we can
achieve good performance and, if so, deciding
which method
to use, requires detailed information about the genotype, the fitness
landscape, and the evolutionary dynamics. For some small subset of
scenarios performance could be very good but it could also be dismal if we
use the incorrect method.

% \idea{Somewhere, maybe in x1 or in x2, reintroduce the idea of two
%   Flandscape traits that give advantage to TD:
% ``an increase in the number of local fitness peaks (numObservedPeaks):
% this can lead to the TD methods being better than their CE counterparts''
% and ``increasing reciprocal sign epistasis
% (epistRSign) can lead to the TD methods performing better than their CE
% counterparts'': }

% A) Genotypes for which using the TD methods lead to very good predictions
% (JS $< 0.1$), which are genotypes that are more often than not local
% maxima (probability of being a local maximum $>0.5$) and of large fitness
% (small fitnessRank) or that are often local maxima (probability of being a
% local maximum $>0.25$) while deviations from SSWM are minor; the TD
% methods can correctly predict being at the end of the evolutionary
% process, something that the remaining methods cannot account for, which
% explains the very poor performance of the non-TD methods in these cases
% (JS $> 0.75$). B) Genotypes for which the CE methods can provide very good
% predictions (JD $< 0.1$), which are genotypes of large relative fitness,
% no mutations or with all except one loci mutated, in evolutionary
% scenarios with SSWM and smooth fitness landscapes (little reciprocal sign
% epistasis, large gamma) and few local peaks (less than 3); for these
% genotypes, the TD methods make very poor predictions (JS $> 0.75$); MHN
% and CBN do particularly good jobs, but CAPRESE and OT, two methods that
% can only model simpler patterns of dependencies, generally perform well
% too. C) The remaining cases, which constitute the majority of the cases,
% where performance was generally poor (JS $> 0.25$). Therefore, assessing
% that we can achieve good performance and, if we can achieve it, deciding
% which method to use, requires detailed information about the genotype, the
% fitness landscape, and the evolutionary dynamics; for some small subset of
% scenarios performance could be very good but it could also be dismal if we
% use the incorrect method.

Our results for short-term predictions highlight variables that have been
found to be relevant for long-term prediction of the complete evolutionary
paths.  Sampling regime and sample size have previously been found
relevant by \cite{diaz-uriarte2019a,hosseini2019a} (see also
\cite{Diaz-Uriarte2015, diaz-uriarte2018} for the effect of sampling regime
in inferring the DAGs of restrictions \textit{per se}). Likewise,
evolutionary scenarios with clonal interference in rugged fitness
landscapes (small gamma, large reciprocal sign epistasis, large number of
local fitness maxima) lead to poorer performance of CPMs compared to SSWM
regimes in single-peaked fitness landscapes
\cite{diaz-uriarte2019a,hosseini2019a}. % and the SSWM
% assumption is explicit in how \cite{hosseini2019a} use CBN for predicting
% tumor evolution.
The role of smooth fitness landscapes for CPM's predictive ability is
directly related to the CPM assumption that final fixation occurs in the
genotype with all loci mutated and their inability to model reciprocal
sign epistasis \cite{diaz-uriarte2018} (this problem also affects MHN,
specially if there are higher-order epistatic
interactions). Interestingly, these are the fitness landscapes where the
TD methods (which allow us to capture local maxima) perform better than
the CE methods. The role of SSWM for CPMs has been discussed before by
\cite{Sprouffske2011, diaz-uriarte2019a} and \cite{hosseini2019a} used it
explicitly for predicting tumor evolution. In fact, all of the methods
considered here implicitly assume SSWM mainly for two reasons. First, CPMs
are generally applied to cross sectional data from bulk sequencing (not
single-cell sequencing) without undergoing any deconvolution step
(i.e. without attempting to infer the clonal composition of the sequenced
tumor), which reflects the implicit assumption that an actual existing
genotype can be obtained from bulk sequencing (see
\cite{alves_multiregional_2017} for discussion of how this assumption can
fail with bulk samples as we collapse over existing genotypes); if this
assumption does not hold, the question ``what genotype comes next'' is
undefined because we are conditioning on a non-existing genotype.  Second,
the observed genotypes are considered true steps in the path to tumor
fixation, thus neglecting the possibility of tunneling effects
\cite{devisser_empirical_2014,szendro_predictability_2013,sniegowski_beneficial_2010}. These
two assumptions are true if SSWM holds (see Fig.~\ref{fig:fig1}, and note
that the reverse is not necessarily true), in which case the most fit
genotype at any given time is effectively the only one in the population,
thus the only one observed, and an ancestor of the ultimately fixated
one. Even when CPMs specifically include error estimations
(e.g. \cite{Gerstung2009}), these typically refer to sequencing error but
not necessarily to deviations from SSWM.  And our reformulation of the
question ``what genotype comes next'' as we did in section
\nameref{sec:intro} % and \nameref{sec:special}
allows us to deal with the second violation, but of course not with the
first. In our analysis of the simulated data the first assumption holds
(we sample true, existing genotypes); yet, our results show that using
CPMs to predict short-term evolution when the second assumption does not
hold resulted in poorer performance. It must be emphasized, though, that
the above comments refer to the effects of deviations from SSWM on the
performance of the methods considered, not on predictability itself (the
role of SSWM on predictability itself is more nuanced: see, e.g.,
\cite{ferretti2018,devisser_empirical_2014,bailey2017,
  szendro_predictability_2013,zhao2016, bank_predictability_2016}).

% in the limit (if
% only one mutation can improve fitness), evolutionary trajectories can even
% be deterministic \cite{ferretti2018}, but deviations from SSWM can
% actually improve predictability, such as those brought about by moderately
% large (but not too large) population sizes \cite{zhao2016,
%   devisser_empirical_2014, szendro_predictability_2013} which can increase
% parallel evolution and the role that selection plays, as shown in
% experimental evolution studies in yeast and bacteria \cite{bailey2017}.

In addition to the above factors, which affect both short- and long-term
predictions, our results highlight that when we want to make conditional,
short-term predictions, features of the genotypes for which we make the
predictions can be as relevant as, or more relevant than, fitness
landscape-specific features (see also \cite{bank_predictability_2016,
  visser2018,ferretti2018} for a comparison of short-term vs.\ long-term
predictability). We have already mentioned the combined importance of
fitness rank and probability of being a local maximum, and the latter,
together with reciprocal sign epistasis, illustrate factors that can
increase short-term predictability while decreasing global-path
predictability. The role of number of mutations, where predictions are
worse from genotypes with intermediate numbers of mutations, can be
understood as a consequence of the inverted U-shape relationship between
the number of entries that can be non-zero in $\mathbf{P}$ and the number
of mutations in a genotype (${N}\choose{n + 1}$: see section
\nameref{sec:methods:sim}): the problem of prediction becomes harder at
intermediate numbers of mutations.  The role of genotype-specific factors
is a noteworthy feature because it suggests that CPMs might be leveraged
to provide short-term predictions even when global, long-term predictions
are difficult because fitness landscape- and evolutionary model-specific
assumptions are violated.

% So, if the genotype can be a local maximum, a simple approach like the
% TD approach can perform well, and it will perform better the more often
% the genotype is a local maximum.

%% Simplifying the results, we can divide genotypes in three groups.

%% \idea{EH! Some of that could be the result of the methods we use!!! Rewrite}

What are the implications of the above results for the analysis of cancer
data? We have compared the predictions from the best performing methods on
25 biological data sets: even if we do not know the true transition
probabilities, we can examine the consequences of assuming that we choose
one method when another method is the one that gives the best (or even
perfect) predictions. This is what we do in
Fig.~\ref{fig:biol_boxplots-3-comps}. The left panel
of Fig.~\ref{fig:biol_boxplots-3-comps}A,
for example, shows
the difference in predictions (measured using JS) if the method that gave
perfect predictions were CBN but we had used MHN (or viceversa).  If we
found that, for a given genotype, two or more methods gave similar
predictions (i.e. comparing their predicted probabilities yielded a small
JS) we could argue that they could all be potentially providing good
results. On the other hand,
very different predictions would indicate that at least one of them
(if not all) is failing, and in that case
there are no solid grounds to choose one or the
other since we lack access to the underlying factors that affect
performance.  As can be seen, for most data sets there is wide
variability in the consequences of using the wrong method, and often the
difference in predictions is very large, specially if we use a TD vs.\ a
non-TD method (MHN vs.\ MHN\_td, right-most panel). 
The high variability between method
predictions shown in Fig.~\ref{fig:biol_boxplots-3-comps}A ---even across genotypes 
within a same dataset--- highlights the importance
of conditioning predictions to the observed genotype, as different methods can make
very similar or substantially different predictions depending on it.
Particularly in the latter case, if any of the methods was indeed providing accurate
predictions, determining which one is the correct one would depend on the properties
of the observed genotype.

[Figs.~\ref{fig:biol_boxplots-3-comps} and \ref{fig:biol_freqs} about
here]

However, in general we do not have access to the underlying parameters
  that would allow us to make an informed decision on method (or method family)
  choice. So how do the above results contribute to the applied analysis of actual
  cancer data? Fig.~\ref{fig:biol_boxplots-3-comps}B shows the dissimilarity
between the predictions of two specific CE methods (CBN and MHN), two specific
TD methods (CBN\_td and MHN\_td), and two methods from different groups (MHN and
MHN\_td).
Many genotypes accumulate in the area of the plot where the predictions of the
two CE methods, as well as those of the two TD methods, are very different, with variable
similarity between CE and TD predictions. But a few genotypes escape this trend.
Looking back at the simulations, we have seen that there
is a sharp difference in behavior of TD and CE methods when the best
achievable performance is good; and necessarily CBN and MHN (or CBN\_td
and MHN\_td) can only both make correct predictions if both are making
similar predictions. This same pattern ---predictions
of methods within the same
  group are similar and predictions of methods from different groups are not--- is
  observed in the biological data for a small subset of genotypes.
  What about limiting our interpretation to those genotypes?
This approach by itself will most likely yield only modest improvements
in performance (see Tables
\ref{S-tab:xtabs_td0} to \ref{S-tab:xtabs_ce1} showing that, for TD, the
proportion of cases with very good TD performance could at most increase
from an unconditional 0.04 to a conditional 0.06, and the proportion of
cases with very good performance of CE from an unconditional 0.04 to a
conditional 0.16).
%But there is a more fundamental problem illustrated in Fig.~\ref{fig:biol_freqs}.
Fig.~\ref{fig:biol_freqs} shows the fraction of genotypes,
for each data set, that meet fairly stringent conditions of similarity
either within the CE methods, i.e. between CBN and MHN, or within the TD
methods, i.e. between CBN\_td and MHN\_td (see Supplementary Material,
figures in Section \ref{S-sm-indiv-biol-data-plots}, for similarity for
each individual genotype for each data
set% ; note that these patterns are not related to
% sample size ---see Fig.~\ref{S-fig:biol_freqs_sampsize}
).  For 11 of the
data sets, there are more than 10\% of the genotypes that fall in either
of these categories: which among the predictions that fulfill those
similarity conditions should we trust? The difficulty of this choice is
even more dramatic when the same genotype simultaneosuly shows large similarity
within TD methods, large similarity within CE methods, and large
dissimilarity between the CE and TD methods, as indicated by the gray
points in Fig.~\ref{fig:biol_freqs}. In these cases, predictions between families of
methods are very different and yet method (or method family) choice would
require fitness landscape-specific, evolutionary model-specific and
genotype-specific information. % Finally, the question ``what genotype comes
% next'' could actually be undefined if the genotypes for which we ask the
% question do not really exist but, rather, are an artifact from using bulk
% sequencing and collapsing over genotypes.

  Even if this information is generally unknown in practice,
  our results provide a foundation to narrow down the
  search for genotypes for which good performance is possible.
  Furthermore, we have demonstrated that we do not need complete
  information on the general properties of the fitness landscape; instead, even partial
  information on those specific genotypes could serve to make informed decisions
  on method choice. Gathering information on the properties of individual
  genotypes is, in principle, a more reasonable objective than obtaining a full
  description of the complete fitness landscape for every cancer type.

%\idea{This previous paragraph could maybe be put in a more positive perspective?
%Yes, it is true that even when CE predictions are similar + TD predictions are
%similar + TD vs CE predictions are dissimilar, we do not have access to the 
%underlying parameters that would allow us to make an informed decision on method
%(or method family) choice. But at least we are seeing the same pattern as in the
%simulations for a subset of genotypes: predictions of
%methods within the same group are similar and predictions of methods from different
%groups are not (maybe there is a more clear way to show this in the figure?).
%We know which specific genotypes follow this pattern, so we
%know that those are the genotypes for which potentially at least one method (or family)
%could be providing good results. So we actually know which genotypes we should be
%looking at.}
%
%\idea{And sure, we do not have access to the parameters that would allow us to know which
%specific method we should be using.
%But that "is not our problem": we know what conditions need to be met to
%achieve good performance, and we know that when good performance is possible, one
%family of methods does well and the other does not. We know which genotypes specifically
%satisfy this behavior, so we can narrow down the search for genotypes for which
%good performance is possible.
%Furthermore, we have demonstrated that we do not need
%complete information on the general properties of the fitness landscape; instead, even
%partial information on those specific genotypes could serve to make informed
%decisions on method choice. Gathering information on the properties of specific genotypes 
%could, in principle, be a more realistic goal than obtaining a full description of
%the complete fitness landscape for each cancer type. Even if "how to do this" is beyond
%the scope of our work, our findings provide a foundation for "what we should be looking
%for".}

%\textcolor{orange}{Given that we do not have access to the underlying parameters
%  that would allow us to make an informed decision on method (or method family)
%  choice, how do the above results contribute to the applied analysis of actual
%  cancer data? In the biological data sets, we see the same pattern as in the
%  simulations for a subset of genotypes: predictions of methods within the same
%  group are similar and predictions of methods from different groups are not.
%  We can identify which specific genotypes follow this pattern, so we know that those
%  are the genotypes for which at least one method (or method family) could potentially
%  be providing good results. This provides a foundation to narrow down the
%  search for genotypes for which good performance is possible.}
%
%  \textcolor{orange}{Furthermore, we have demonstrated that we do not need complete
%  information on the general properties of the fitness landscape; instead, even partial
%  information on those specific genotypes could serve to make informed decisions
%  on method choice. Gathering information on the properties of specific
%  genotypes could, in principle, be a more realistic goal than obtaining a full
%  description of the complete fitness landscape for each cancer type. Even if
%  "how to do this" is beyond the scope of our work, our findings provide a
%  foundation for "what we should be looking for".}

\subsection{Conclusion}
\label{sec:conclusion}

CPMs could, under very specific combinations of genotype characteristics
and fitness landscape characteristics, be used to obtain good predictions
of the short-term evolution of a tumor, even when long-term predictions
are not possible because of violations of assumptions of CPM
procedures. But method choice and assessing if the predictions obtained
are to be trusted, for example to guide therapeutic decisions, requires
detailed knowledge which is not available with empirical data. More
generally, our work shows the promises and practical difficulties of using
cross-sectional data in an evolutionary context to guide individualized
therapies, even when cross-sectional data recapitulate the equivalent of
many repeated evolutionary processes.  % , lack of detailed knowledge about key
% parameters and variables do not allow direct application of the insights
% gained from repeated evolution experiments in other systems.
Exploiting the potential of these methods will require further work to, first,
examine the interpretation of their output and the consequences of their
usage when key assumptions do not hold; second,
identify if the
methods' output, by itself or in combination with data set features, can
provide indicators of performance; and third, develop
strategies to determine the characteristics of individual genotypes in real tumors and adapt
method use consequently.
For all three tasks, it will be
important to make explicit the relationship between assumptions with
respect to the evolutionary model (e.g., SSWM), fitness landscape
characteristics (e.g., local maxima and reciprocal sign epistasis),
sampling regimes (late vs.\ early tumors), and predictions, all in the
context of well defined concepts such as lines of descent and for
specified tasks such as interventions.

% \begin{figure}[!h]
%   \centering
%   \includegraphics[width=15.5cm,keepaspectratio]{figs/biol-data_js-vs-nMut_multi-glmm.pdf}
%   \caption{\textbf{Similarity of predictions between methods and relative
%       number of mutations of genotypes.} Y-axis: similarity, measured with
%     JS, of predictions of next genotype between different methods. X-axis:
%     the number of mutations of the source genotype relative to the total
%     number of genes (or loci that could be mutated) for the give data
%     set. Different colors identify different data sets.  Upper panel, TD
%     vs.\ TD: average similarity between CBN\_td and MHN\_td and MCCBN\_td
%     and MHN\_td (``TD'' stands for ``time-discretized''). Middle panel:
%     average similarity between CBN and MHN and MCCBN and MHN (``CE'':
%     competing exponentials). Lower panel: average similarity of CBN and
%     CBN\_td, MCCBN and MCCBN\_td, and MHN and MHN\_td (i.e., between the
%     CE and TD versions of each method).  Lines shown represent the output
%     from a generalized linear mixed effects model for a beta-distributed
%     response (JS); we used natural cubic splines to model the relationship
%     with relative number of mutations of genotypes in both the fixed and
%     random effects (data set id) components. The black thick line shows
%     the fixed effect fit (and the colored, thin, lines the data
%     set-specific fits ---i.e., including the random effects). Models
%     fitted using glmmTMB \cite{brooks2017}. See further details in the
%     Supplementary Material (``Model fits to similarity of predictions
%     between'').}

%   \label{fig_biol_comp} \end{figure}

\section{Acknowledgments}
% \idea{Ask Achim Zeileis and Rudolf Schill}

We thank A.~Zeileis for answers about the use of the glmertree package and
R.~Schill for answers about the use of MHN, and R.~Guantes and C.~Lazaro-Perea
for comments on the ms.
%% assistance me parecía too much. Fueron dos respuestas.

\section{Funding}
Supported by BFU2015-67302-R (MINECO/FEDER, EU) and
PID2019-111256RB-I00/AEI/10.13039/501100011033 
to RDU. JDC supported by
PEJD-2018-POST/BMD-8960 from Comunidad de Madrid.

\section{Data \& code availability}

Code for the analyses in this article is available at
\url{https://github.com/rdiaz02/what_genotype_next}.
%% \url{github.com/jdiazc9/what-genot-next}.

% Details and code for the simulations available from the Supplementary
% Material of Díaz-Uriarte and
% Vasallo, 2019 \cite{diaz-uriarte2019a}. Code for the analyses in this
% article is available at \idea{\url{github.com/jdiazc9/oncolab}: change
%   that to the definitive one}.

% \end{document}

% REFERENCES
{\small
  %%\bibliography{refs}
  \bibliography{extracted-draft}
\bibliographystyle{mybibwithurl} %% so that Crooks URL always shows
%%\bibliographystyle{mystyle} % choose this for: "This is some text in the paper blablabla [1-3], this is how it is cited [4]."
% \bibliographystyle{apalike} % choose this for: "This is some text in the paper blablabla [Author1, 2009, Author2 and Author3, 2016], this is how it is cited [Author4 et al., 2012]."
}

\clearpage

\section{Figures}
\clearpage
% \todo{Hay que quitar el trim de las figuras. Deberían ser figuras que
%   están ya cropped ---y que serán las que enviemos}

%%\idea{We need to add an MHN and change the figure accordingly}
% \idea{el panel B deberia tener flechas que indiquen que se mueve hacia la
%   derecha (i.e., ganando mutaciones)}

\begin{figure}[H]
	\centering
	% \fcolorbox{white}{white}{\includegraphics[width=0.8\linewidth,trim={4cm
        %     6.5cm 4cm 5.5cm}]{figs/final-layout/fig1}}
        \includegraphics[width=0.99\linewidth,keepaspectratio]{figs/fig1-crop-witharrows-compact}
	\internallinenumbers
	\caption[Inferring evolutionary trajectories using Cancer
        Progression Models.]{\textbf{Inferring evolutionary trajectories
            using Cancer Progression Models.} \textbf{A.} Example of a
          directed acyclic graph (DAG) of restrictions in the order of
          accumulation of mutations, analogous to the ones produced by
          most CPMs, such as CBN, MCCBN, OT, CAPRESE, and CAPRI. Given
          four driver genes (numbered 1 to 4), the DAG indicates the
          constraints in the order of accumulation of mutations
          (arrows). Here, a mutation in the 4th loci is conditioned to the
          2nd and 3rd ones having been previously mutated.  MHN uses a
          network to encode both promoting and inhibiting effects between
          pairs of events; the network could be identical to the one in A
          (meaning that mutations 2 and 3 have a promoting effect on
          mutation 4), but it could also include inhibiting effects; in
          contrast to DAGs of restrictions, MHNs do not denote
          deterministic dependencies.  \textbf{B.} Restrictions in the
          order of accumulation of mutations limit the amount of available
          evolutionary
          trajectories and stochastic dependencies change the
          probabilities of trajectories.  In this network, nodes represent
          different genotypes of four driver genes and edges connect those
          that are a single mutation away. Genotype names are assigned
          according to which genes are mutated (1) or not mutated
          (0). Without restrictions many mutational pathways are possible
          (any route through both solid and/or dashed edges, from left to
          right).  However, only a subset of genotypes satisfy the
          constraints in panel A (solid nodes); thus, only trajectories
          connecting these are predicted to be accessible pathways (solid
          edges). For example, given a specific genotype (1000, red node),
          restrictions in the mutational order limit the possible clones
          that can evolve from it consecutively after (red edges). With
          CBN, MCCBN, and OT, in addition to the set of accessible
          pathways, it is possible to obtain transition probabilities
          along the black lines.  Under the MHN model, as there are no
          deterministic dependencies, all transitions (to genotypes with
          one additional mutation) are possible, but transitions affected
          by inhibitory stochastic dependencies will have very low
          probability (in this simplified depiction they would be the gray
          lines) and those affected by promoting stochastic dependencies
          will have higher probability (black lines); MHN provides
          transition probabilities for all the lines in the figure.
          \textbf{C.}  Example of cancer progression with four driver
          genes under SSWM conditions. Progression towards the fitness
          maximum occurs through a defined sequence of parent-child
          genotypes, each one of them establishing as the most abundant
          rapidly after evolving. \textbf{D.} Example of cancer
          progression with four driver genes under conditions of no
          SSWM. Here, the sequence of parent-child genotypes from the
          wild-type with no driver mutations to the final one (i.e. the
          Line Of Descent or LOD) is: 0000 $\rightarrow$ 1000
          $\rightarrow$ 1010 $\rightarrow$ 1110. Notice that in the most
          general case, the ultimately fixated genotype need not be the
          one with all driver mutations (1111).  Some genotypes can be
          most abundant in the tumor during prolonged periods of time
          despite not being part of the LOD (0010, in green), and
          genotypes that do belong to the LOD may never be most abundant
          (1010, in blue). In all models and figures we do assume that
          there are no back mutations and that crossing valleys in the
          fitness landscape using a single multi-mutation step is not
          possible. }\label{fig:fig1}
\end{figure}

\begin{landscape}
  \begin{figure}[!h]
    \vspace*{-45pt}
%  \centering
\includegraphics[width=22.5cm,keepaspectratio]{figs/weighted-7-specific-no-split-01-sqrt-all_methods-js.pdf}
  \internallinenumbers
  \caption{\textbf{Linear mixed-effects model tree fitted to data from
      fitness landscapes with 7 loci, sample size of 4000 and uniform
      sampling.}  Internal nodes show the partitioning variables and the
    values that lead to each partition; light blue denotes the values or
    categories that lead down the left split and salmon the values or
    categories that lead down the right split (partitioning criteria are
    also shown in the edges leading down a node). Leaves or terminal nodes
    show (weighted) boxplots of the dependent variable for the
    observations that fall in that leave and, as label, the
    name of the node and the sample size (sum of weighted observations and
    \% of the total data set for the fit). The dot plots underneath the
    boxplots show the (weighted) mean JS for each method for that leave.}
  \label{fig:7_spec} \end{figure}
% \restoregeometry
\end{landscape}

\begin{figure}[!h]
  \centering
  %\includegraphics[width=15.5cm,keepaspectratio]{figs/boxplots-three-comps_all_datasets_combined.pdf}
  \includegraphics[width=15.5cm,keepaspectratio]{figs/biol_data_box_scatter.pdf}

  \caption{\textbf{Similarity between predictions of three pairs of
      methods for the 25 biological data sets.} For each data set, the JS
    divergence between the predictions of a pair of methods is represented in
    \textbf{A.}
    each of the three panels or \textbf{B.} each of the three axes. 
    Dots correspond to all genotypes
    for which we computed predictions, which are the genotypes observed in the
    samples, except for the genotype with all genes mutated if it was observed.
    In parentheses,
    after the data set, we indicate the total sample size (n) and number of genes (g).}

  \label{fig:biol_boxplots-3-comps} \end{figure}

\begin{figure}[!h]
  \centering
  \includegraphics[width=15.5cm,keepaspectratio]{figs/fraction_CE_TD_similar_all_datasets.pdf}

  \caption{\textbf{Percentage of genotypes with similar TD and similar CE
      predictions for the 25 biological data sets.} For each data set, the
    orange points show the percentage of genotypes in which predictions
    had both similar CE ($JS_{CBN, MHN} \le 0.0677$) and different CE-TD
    ($(1/2)\ (JS_{CBN, CBN\_td} + JS_{MHN, MHN\_td}) \ge 0.7$),  the
    blue dots those with similar TD ($JS_{CBN\_td, MHN\_td} \le 0.0677$)
    and different CE-TD, and the gray points the percentage of genotypes
    that had similar CE, similar TD, and different CE-TD. Values shown are the percentage relative to the
    total number of genotypes for which we computed predictions, which are
    the genotypes observed in the samples, except for the genotype with
    all genes mutated,
    if it was observed. In parentheses, after the data set, we indicate
    the total number of genotypes observed (GT).  % For a given data set, genotypes that have boot similar CE and
    % similar TD contribute to both counts.  Panels B) and C): scatter plots
    % of Similar TD and Similar CE, shown in panel A, against sample
    % size.
  }

  \label{fig:biol_freqs} \end{figure}

\end{document}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%% Old results stuff

% So: when we can achieve good performance depends crucially on
% balblaba. (Give one or two). And when we should used one type of method
% vs. other too (blablba)

% How is this relevant for real data sets, for which we know nothing of
% those characts? We have examined if this disyuntive could actually be
% relevant. And it is. Blabla.

% % \subsection{Results trees, II} \label{res2}

% % \begin{enumerate}
% % \item If prop local max is large, then a simple approach like TD works
% %   (CBN and MHN)
% %   % \begin{enumerate}
% %   % \item Fig 2 node 10 (2\% ), to lesser degree nodes 12, 13 (2 and 3
% %   %   perc.)
% %   % \item Suppl Fig 3: nodes 19 (and 20 and 21: 1 and 2 percent), node 16 (2
% %   %   perc) and 15 (1) and 22 (5 \% ) to lessser degree
% %   % \item Even in less than ideal situations that is the case. Suppl Fig 1:
% %   %   nodes 24 and 25 (with 1 and 5 \% )
% %   % \item Suppl Fig 2, nodes 18 and 22 (1 and 1 \% )
% %   % \item Lesser degree 16, 17, 23 (with 2, 1, 1,\% ): ehh??
% %   % \end{enumerate}
% % \item What about CE?
% %   \begin{enumerate}
% %   \item Cases where MHN is really the single best method are rare
% %   \item When CE can do well (JS < 0.2) most methods about equially good
% %     (except CAPRI), though CBN generally best (but recall simul scenarios
% %     might favor CBN). So when good CE performance, then pretty much any of
% %     the reasonable methods (all except CAPRI) will work. OT not the best,
% %     and MHN sometimes not best.
% %   \item CE is consistently gopod only in limited cases
% %     \begin{enumerate}
% %     \item Fig2: nodes 6, 7, 16, 20 (with 4, 1, 2, 6\% ); in node 16 MHN is
% %       poor
% %     \item Suppl Fig 1: Maybe node 6, 12, 16, 17 (with 1, 1, 2, 2\% data )
% %     \item Suppl Fig 2: nodes 10 and maybe 9 (sum of 2\% )
% %     \item Suppl Fig 3: node 23 (5\% ). OJO AQUI: probablemente escamotea
% %       que tenemos mezcla de buenos TD y buenos CE, como el nodo 19 en la
% %       figura del paper.
% %     \end{enumerate}
% %   \end{enumerate}
% % \end{enumerate}

% \subsection{Results: connect simuls-biol} \label{biol1}

% \begin{enumerate}
% \item If TD is good, the td methods are similar
% \item If CE is good, the CE methods are similar
% \item The converse is, of course, not true
% \item Particularly the case if TD and CE are very different.
% \item Refer to section ``Categorization of similarities between methods and
%   performance'' in suppl mat (i.e., use that info judiciously in the
%   paper, citing key numbers)
% \item Thus, we analyze biol data and focus on similar TD, similar CE,
%   dissimilar CE-TD: how often is this seen in the data?
% \end{enumerate}

% The fitness-landscape specific variables are

% - SSWM regime
% -  $\gamma$
% - (epistRSign)
% - numObservedPeaks

% - detection regime
% - sample\_size)

% - nMut
% - fitnessRank
% - propLocalMax

% - observedProp
% - diff\_obs\_prop

% epistRSign can flip which method is best: Node 11 in
% Fig.~\ref{S-fig:7_ALL}; smaller is better (also node 23 in the figure)

% numObservedPeaks can flip which method is best: Node 19 in
% Fig.~\ref{fig:7_spec}

% , ; notice that the effect of fitness landscape type
% is not always apparent in these trees, but three of the variables included
% in the fits, and that appear in the fitted trees, gamma, reciprocal sign
% epistasis, and number of observed peaks,

% The fitted trees show (see supporting information, Figures \todo{blabla})
% show that \todo{blabla: sample size, detection regime}. The effect of
% fitness landscape type is not \todo{always, often} apparent in these
% trees, but three of the variables included in the fits, and that appear in
% the fitted trees, gamma, reciprocal sign epistasis, and number of observed
% peaks, allow to perfectly separate the RMF from the rest of the fitness
% landscape, and to separate most local maxima fitness landscapes from the
% representable ones (see figure \todo{gamma-rsign-obs-peaks-FL.pdf} in the
% supplementary material).

% The results for the tree fitted to seven genes, uniform, sample size =
% 4000, show \todo{blablabla}.

% Use the 7 genes, specific.

% number of peaks, ruggeddness of landscape matters: see node 28 vs. 29. Not
% unlike the pattern of 14 (here max fitness, so can only jump to end). node
% 28 also kinda matches node 21 in the 10 genes.  ode 28 is smooth, so it
% jumps to 7 mutations, it is often the single option. Note also node
% 15. And note that these cases add up to close to 8\% of observations

% Very obvious two different groups of methods, that, where it matters,
% either do a very good or a very bad job (with the other group doing a very
% bad or very good job)

% To prevent detection regime and sample size from hindering
% interpretation, we have fitted the models to the best performing
% conditions, sample size of 4000 and uniform detection regime.

% to prevent the effects of sample size and detection regime from
% obscuring key patterns.

% Our main objective is understanding the relevance of fitness
% landscape-specific and genotype-specific variables for on the quality of
% predictions. Thus, we have fitted the models to the best performing
% conditions, sample size of 4000 and uniform detection regime, to prevent
% the effects of sample size and detection regime from obscuring key
% patterns.

% The linear mixed-effects model trees fitted to the complete data sets for
% 7 and 10 genes showed that sample size and detection regime had a strong
% effect on performance (see Supplementary Material, Figures fig\_10\_ALL
% and fig\_7\_ALL), with better performance achieved under the uniform
% detection regime and with the largest sample size (4000) (see also Table
% tab\_minjs in Supplementary material); there are interactions,
% blablaba. We focus on something else.

% For both 7 and 10 genes, JS was smallest in the uniform detection regime
% with sample size of 4000 \todo{give numbers; do we even need this?}.

% \begin{figure}[!h]
%   \centering
%   \includegraphics[width=15.5cm,keepaspectratio]{figs/fraction_CE_TD_similar_all_datasets_combined.pdf}

%   \caption{\textbf{Percentage of genotypes with Similar TD or Similar CE
%       predictions for the 25 biological data sets.} A) For each data set,
%     the orange dots show the percentage of genotypes in which predictions
%     had both similar CE
%     ($(1/3) \ (JS_{CBN, MHN} + JS_{CBN, MCCBN} + JS_{MCCBN, MHN}) \le
%     0.0677$) and Different CE-TD
%     ($(1/2)\ (JS_{CBN, CBN\_td} + JS_{MHN, MHN\_td}) \ge 0.7$), and the
%     blue dots those with similar TD ($JS_{CBN\_td, MHN\_td} \le 0.0677$)
%     and Different CE-TD. Values shown are the percentage relative to the
%     total number of genotypes for which we computed predictions, which are
%     the genotypes observed in the samples, except for the last genotype,
%     if it was observed. In parentheses, after the data set, we indicate
%     the total number of genotypes observed (GT).  For a given data set, genotypes that have boot similar CE and
%     similar TD contribute to both counts.  Panels B) and C): scatter plots
%     of Similar TD and Similar CE, shown in panel A, against sample
%     size. }

%   \label{fig_biol_freqs} \end{figure}

%% FIXME Place in Suppl Mat
% \begin{figure}[!h]
%   \centering
  % \includegraphics[width=15.5cm,keepaspectratio]{figs/fraction_CE_TD_similar_all_datasets_sample_size.pdf}
  % \caption{\textbf{Percentage of genotypes with Similar TD or Similar CE
  %     predictions for the 24 biological data sets: relationship to sample size.} For each data set,
  %   the orange dots show the percentage of genotypes in which predictions had both similar CE
  %   ($(1/3) \ (JS_{CBN, MHN} + JS_{CBN, MCCBN} + JS_{MCCBN, MHN}) \le
  %   0.2$) and Different CE-TD ($(1/2)\  (JS_{CBN, CBN\_td} + JS_{MHN,
  %     MHN\_td}) \ge 0.6$), and the blue dots those with similar TD
  %   ($JS_{CBN\_td, MHN\_td} \le 0.2$) and Different CE-TD. Values shown
  %   are the percentage relative to the total number of genotypes for which we
  %   computed predictions, which are  the genotypes observed in the
  %   samples; the total number of genotypes observed is indicated to the
  %   right of the data set name as ``(n = x)''. For a given data set, the
  %   sum of the two numbers can be greater that 100, since genotypes that
  %   have boot similar CE and similar TD contribute to both counts.
  %    }

  % \label{fig_biol_freqs_sampsize} \end{figure}

%% From analyze-biol-data-3.R

% Figures \ref{S-fig:7_ALL} and \ref{S-fig:10_ALL} show the models
% fitted to all detection regimes and sample sizes; these figures showed the
% relevance of detection regime and sample size, with the smallest JS were
% generally achieved under the uniform detection regime and with the largest
% sample size (4000) (see also Supplementary material, Table
% \ref{S-tab_minjs} in section \nameref{S-suppl_minimal_js_ssize_det}).
% Moreover, there were interactions between detection regime and sample size
% on the one hand and genotype-specific and fitness landscape-specific
% variables. These interactions can make it harder to understand the
% relevance of fitness landscape-specific and genotype-specific variables on
% the quality of predictions,

% . In the Supplementary Material
% (Fig.~\ref{S-fig:10_ALL}, section
% \ref{S-suppl-linear-mixed-model-results}) we show the results for 10
% genes.  These figures showed the relevance of detection regime and sample
% size, with the smallest JS were generally achieved under the uniform
% detection regime and with the largest sample size (4000) (see also
% Supplementary material, Table \ref{S-tab_minjs} in section
% \nameref{S-suppl_minimal_js_ssize_det}). Moreover, there are interactions
% between detection regime and sample size on the one hand and
% genotype-specific and fitness landscape-specific variables. These
% interactions can make it harder to understand the relevance of fitness
% landscape-specific and genotype-specific variables on the quality of
% predictions, which is one of the main objectives of this paper. Thus, we
% have fitted the linear mixed-effects model trees under the best performing
% conditions, sample size of 4000 and uniform detection regime, to isolate
% the effects of fitness landscape and genotype-specific variables; results
% are shown in Figure \ref{S-fig:7_spec} for fitness landscapes with seven
% genes, and in the Supplementary Material (Figure fig\_10\_spec) for
% fitness landscapes with 10 genes.

% The smallest JS were generally achieved under the uniform detection regime
% and with the largest sample size (4000) (see Supplementary material,
% ``Minimal JS for combinations of number of genes, sample size, and
% detection region'') %%Table tab\_minjs in
% a pattern that can also be observed in the linear
% mixed-effects model trees fitted to the complete data sets for 7 and 10
% genes (Supplementary Material, ``Linear mixed-effects model trees for
% complete data''). These figures also showed the presence of interactions
% between these two variables and other genotype-specific and fitness
% landscape-specific variables. These interactions can make it harder to
% understand the relevance of fitness landscape-specific and
% genotype-specific variables on the quality of predictions, which is one of
% the main objectives of this paper. Thus, we will focus below on models
% fitted under the best performing conditions, sample size of 4000 and
% uniform detection regime, to isolate the effects of fitness landscape and
% genotype-specific variables; results are shown in Figure \ref{S-fig:7_spec}
% for fitness landscapes with seven genes, and in the Supplementary Material
% (Figure fig\_10\_spec) for fitness landscapes with 10 genes.

%  of this same model nor
% in node 12 in Fig.~\ref{S-fig:10_ALL} since those nodes split genotypes landscapes
% that are not representable ().

% Ojo: nodo 19 does not split con los de 1 peak, pero es 1 o 2, on the left!
% Some of these results are easy to understand. For example, in
% Fig.~\ref{fig:7_spec} the relevance of numObservedPeaks in node 19 for
% determining whether TD methods are better than their CE counterparts. Node
% 19 splits genotypes with 6 mutations (in fitness landscapes of 7
% genes). Now, if there is only peak, we are in a representable fitness
% landscape, and the next transition will necessarily be to the genotype
% with 7 mutations; the TD methods, that can predict that the genotype with
% 6 mutations is a local maximum will therefore fail. NOPE not that simple:
% a) this is 1 or 2 on the left, so not just representable; b) notice
% CAPRI's behavior.

% \idea{Ya no tengo nada claro que mostrar la figura que mostramos tenga
%   sentido. Podríamos mostrar solo la Suppl 1 y Suppl 2, en el main text.
%   No tengo claro que esté bien justificado (por que centrarse en los
%   mejores?) ni que ayude a clarificar la discusion. O lo ayudaría al
%   revés: mostar el Suppl Fig 1 en resultados, lo demças en suppl mat}

% \idea{La discusion podria ser muy distinta: complex interactions; SSWM,
%   rank, and other genotype specific matter. Done}

% They can be very
% different, but we still would need to choose. How? Using info we do not
% have about SSWM, fitness landscape, etc.

% Fig.~\ref{fig:biol_freqs} shows blablba \idea{add to that
%   plots those genotype where all three conditions are true; } So the CE
% can be similar, the TD can be similar: should we use them? When we observe
% similar CE and good performance depends on features we do not see; ditto
% for TD. And, even more dramatially, what if the same genotype shows both
% patterns?  They can be very different, but we still would need to
% choose. How? Using info we do not have about SSWM, fitness landscape, etc.

% since Fig.~\ref{fig:biol_freqs} based on the necessarily true conclusion
% that CBN and MHN can only both make correct predictions if both are making
% similar predictions, and similarly for CBN\_td and MHN\_td. Reversing the
% conditioning, however, does not guarantee a large improvement in
% performance.

% Using this simpleminded approach, we can in fact improve the
% performance, relative to baseline levels, if we condition on similar CE
% (or similar TD) but different CE-TD, but that improvement is rather modest
% \idea{blablaba ``Categorization of similarities between methods and
%   performance'' in suppl mat, and this is overestimate of performance,
%   etc, etc}. We can try to restrict the analysis to such
% conditions. Fig.~\ref{fig:biol_freqs} shows blablba \idea{add to that
%   plots those genotype where all three conditions are true; } So the CE
% can be similar, the TD can be similar: should we use them? When we observe
% similar CE and good performance depends on features we do not see; ditto
% for TD. And, even more dramatially, what if the same genotype shows both
% patterns?  They can be very different, but we still would need to
% choose. How? Using info we do not have about SSWM, fitness landscape, etc.

% \subsection{Results biol} \label{biol01}

% \begin{enumerate}
% \item In figure  \ref{fig:biol_freqs}
% \item Details for each data set (similarity CE, similarity TD, difference
%   CE-TD) in suppl mat (``Similarity of predictions for all genotypes for the biological
%   data sets'').
% \item Many cases where a large percentage genotypes very simlar TD or very
%   similar CE and very different CE to TD
% \item Wide variation among data sets
% \item Not related to sample size (panels B and C)
% \item Some of those preds could correspond to good predictions but telling
%   would require knowing more.
% \end{enumerate}

\section{Como hacer el conectar con datos biol: miscel, other ideas, what
  not to do, etc}
\label{sec:como-hacer-el}

Some options

\begin{enumerate}
\item Classify all output from simuls in four groups
  \begin{itemize}
  \item CE w.r.t. real approx 0, TD approx 1
  \item TD w.r.t. real approx 0, CE approx 1
  \item All $> 0.2$
  \item All $< 0.2$
  \end{itemize}
  \begin{itemize}
  \item But that sucks because we have a multi-class classification
    problem and that is a PITA with mixed effects
    models, etc.
\item Moreover, this
  would be circular thing (as it is based on previously looking at the
  data with the other trees). 
\item And of course we don't have anything similar with real data.

  \end{itemize}
\item Look at similarities between methods. What we didd with the other
  trees. That nothing came out, is it a result? Nope, since this is
  actually dealing with another issue. And it could be explored in much
  more detail. This ain't the point of the paper.
\item Focuse on some transitions/cases: like the genotype before
  last. E.g., node 19, 20, 21 in Fig 2. Pero eso es coger la cosas por los
  pelos. Compatre with node 23 in Suppl Fig 3. 
\item Try to do a principled search, over biological data, of regions with
  CE vs TD huge, within-CE small and within-TD small.

  Tghis requires defining things in a principled way. So the whole mess of
  a PCA if descriptive, PLS-like for predictive, etc. And we are back,
  like above, in another issue. Not what we are dealing with.
\item Focus on the key idea: choice of method/quality of predictions ought
  to depend on details of evol model etc that we ignore. Show how this
  matters. Show that it matters (Fig 2 of the paper) and show that there
  are many cases in biol data that likely fall there. \textbf{This is the
    connecting logic}.

  Emphasize: \textbf{we are not focusing on choice of single best method,
    just in seeing what is possible. Choice of best is a different
    question}.

  In results go over what is in \ref{res2}. Then, say:
  \begin{enumerate}
  \item If TD is good, the td methods are similar
  \item If CE is good, the CE methods are similar
  \item The converse is, of course, not true
  \item Refer to section ``Categorization of similarities between methods and
  performance'' in suppl mat (i.e., use that info judiciously in the
  paper, citing key numbers)
\item Now say: we will focus on cases in the biological data where the CEs
  are similar and the TDs are similar, as those can be good.
\item In the biological data, show statistics, for the data sets, of how
  often we have similar TD + dissimular CE-TD and similar CE + dissimilar
  CE-TD. Comment that ``choosing the best method and, thus, choosing
  predictions depends on key featuyres fo the evol model''. This is figure \ref{fig:biol_boxplots-3-comps}

  Show figure where it can be seen that we have used a whole bunch of data
  sets.

  We are done.

  \end{enumerate}

  \item Es tentador hacer más modelos de árboles con solo los datos
    simulados con
    buena predicción (JS $le 0.2$), y hacer modelo logístico para separar:
    ``los mejores son los TD'' vs.\ ``los mejores son los CE''. Pero esto:
    \begin{itemize}
    \item Es una pregunta distinta y el paper already too long
    \item Añade más tiempo de análisis.
    \item No está claro cómo se conecta con los datos biológicos.
    \item ¿Merece la pena?
    \end{itemize}

\item Es tentador hacer, con los datos simulados, una figura del tipo de
  \ref{fig:biol_freqs}. Dividiendo por type of landscape y por sampling
  type y quizas sample size. Pero:
  \begin{enumerate}
  \item Más tiempo análisis
  \item Alarga paper
  \item Delicate issue de los pesos
  \item Mejora la conexion pero no sé si merece la pena
  \end{enumerate}

\end{enumerate}

\idea{Do not forget to mention inherent variability of CBN (and MCCBN)
  which is very noticeable when relatively large number of variables, as
  has been noted before (Diaz-Uriarte and Vasallo, 2019; Hosseini et al.,
  2019). Ellaborate on this, etc, etc. Hosseini develops his strategy for
  this. With real data might one several runs for predicting genotye? etc,
etc. To be explored in further papers.}

\idea{Beware: we do not want anyone asking for cross-validatino of
  performance estimates! If they ask, we will say that outside of the
  scope, and a moot point anyway. Suppl mat tables present upper limits on
what is already modest performance}

Arriba en resultados explicar porqué uno se queda con cualquier genotipo
para el que hay rpedcciones de al menos 3. Uno de esos 3 siempre es MHN.

%%%%%%%%%%%%%%%%%%%%% Former biological data stuff

%  (note that being local maxima is often associated with having
% large relative fitness ---i.e., small fitness rank--- and local maxima
% genotypes are more common in fitness landscapes with a larger number of
% local fitness maxima)

% What those results emphasize, then, is that predicting if performance will
% be acceptable, and what method to use to achieve that performance, depends
% on features of the fitness landscape, the evolutionary process, and the
% genotype. This highlights the difficulties of interpreting the

The results from the simulated data have direct implications for the
interpretation of the application of these methods to biological data. As
Fig.~\ref{fig:biol_boxplots-3-comps} shows, there are often very large
differences in the predictions of a TD and a non-TD method (MHN vs.\
MHN\_td, right-most panel), something we could observe if we were dealing
with genotypes of types A) or B) above. But we have no solid grounds to
choose one method over the other, as that choice depends on information
about the evolutionary model and the genotype itself that we
ignore. Fig.~\ref{fig:biol_freqs} emphasizes this point: a tempting
approach when analyzing data would be to focus interpretation on genotypes
for which we see very dissimilar CE-TD predictions and at the same time
either similar TD or similar CE (or both).  This conditioning, by itself,
is unlikely to yield relevant improvements in performance (see Tables
\ref{S-tab:xtabs_td0} to \ref{S-tab:xtabs_ce1}). And it leaves unanswered
the question of which among the predictions that fulfill those similarity
conditions we should trust. The difficulty of this choice is even more
dramatic when the same genotype shows both large similarity within TD
methods, large similarity within CE methods, and large dissimilarity
between the CE and TD methods, as indicated by the gray points in
Fig.~\ref{fig:biol_freqs}: predictions between families of methods are
very different and yet choice of family of method would require fitness
landscape-specific and genotype-specific information.

\subsection{Analysis of biological data sets} \label{sec:biol-data}

\idea{algunas cosas de debajo repetir o enfatiza más en discusión y menos aquí}

% Do the above results matter for analysis of biological data? Yes they
% do.
When analyzing biological data, even if we do not know the true transition
probabilities, we can examine the consequences of assuming that we choose
one method when another method is the one that gives the best (or even
perfect) predictions. This is what we do in
Fig.~\ref{fig:biol_boxplots-3-comps}, where we use the four best
performing methods to illustrate the consequences of choosing the wrong
method. The left panel, for example, shows the difference in predictions
(measured using JS) if the method that gave perfect predictions were CBN
but we had used MHN (or viceversa).  As can be seen, for most data sets,
there is wide variability in the consequences of using the wrong method,
and often the difference in predictions is very large, specially if we use
a TD vs.\ a non-TD method (MHN vs.\ MHN\_td, right-most
panel).% And most of the time we have no
% solid grounds to choose one method over the other, as that choice depends
% on information about the evolutionary model and the genotype itself that
% we ignore.

% \idea{}The 25 biol data sets show ranges of variation
% comparable to those from the simulated data \idea{DO THIS PLOT for suppl
%   mat}.

\idea{algunas cosas de debajo mover a discusion: ``more dramatic'', blablablba}

% A naive approach might try to leverage on the similarity between methods, 
% as we have seen % :
% Fig.~\ref{fig:7_spec}, as well as Figures~\ref{S-fig:10_spec},
% \ref{S-fig:7_ALL} and \ref{S-fig:10_ALL}, indicate
We have seen, with the simulated data, that there is a sharp difference in
behavior of TD and CE methods when the best achievable performance is
good.  And necessarily CBN and MHN can only both make correct predictions
if both are making similar predictions; likewise, CBN\_td and MHN\_td can
only both make correct predictions if both are making similar prediction.
A simple approach could try to limit analyses to genotypes that show very
distinct CE-TD patterns and either  similar CE patterns, or similar
TD patterns. Note that conditioning in these similarity patterns could
bring at most very modest improvements in performance (see Tables
\ref{S-tab:xtabs_td0} to \ref{S-tab:xtabs_ce1}).
Fig.~\ref{fig:biol_freqs} shows the fraction of genotypes, for each data
set, that meet fairly stringent conditions of similarity either within the
CE methods (i.e., between CBN and MHN) or within the TD methods (i.e.,
between CBN\_td and MHN\_td). (See Supplementary Material, figures in
Section \ref{S-sm-indiv-biol-data-plots}, for similarity for each
individual genotype for each data set; note that these patterns are not
related to sample size ---see Fig.~\ref{S-fig:biol_freqs_sampsize}). There
is wide variation among biological data sets for the fraction of genotypes
that meet these criteria, but for several of the data sets, there are more
than 20\% of the genotypes that fall in either of these categories,
including genotypes for which all three criteria (dissimilar CE-TD,
similar TD, similar CE) are met.

% A naive approach might try to leverage on the similarity between methods, 
% as we have seen % :
% % Fig.~\ref{fig:7_spec}, as well as Figures~\ref{S-fig:10_spec},
% % \ref{S-fig:7_ALL} and \ref{S-fig:10_ALL}, indicate
% that there is a sharp
% difference in behavior of methods when the best achievable performance is
% good. And necessarily CBN and MHN can only both make correct predictions
% if both are making similar predictions; likewise, CBN\_td and MHN\_td can
% only both make correct predictions if both are making similar prediction.
% But, by itself, conditioning in these similarity patterns within CE
% methods and within TD methods could bring at most very modest improvements
% in performance (see Tables \ref{S-tab:xtabs_td0} to
% \ref{S-tab:xtabs_ce1}). Fig.~\ref{fig:biol_freqs} shows the fraction of
% genotypes, for each data set, that meet fairly stringent conditions of
% similarity either within the CE methods (i.e., between CBN and MHN) or
% within the TD methods (i.e., between CBN\_td and MHN\_td). (See
% Supplementary Material, figures in Section
% \ref{S-sm-indiv-biol-data-plots}, for similarity for each individual
% genotype for each data set; note that these patterns are not related to
% sample size ---see Fig.~\ref{S-fig:biol_freqs_sampsize}). There is wide
% variation among biological data sets, and for several of the data sets,
% there is a large fraction of genotypes with very simlar TD or very similar
% CE and very different CE to TD. Which among these predictions could we
% trust?  The difficulty of this choice is even more dramatic when the same
% genotype shows both large similarity within TD methods, large similarity
% within CE methods, and large dissimilarity between the CE and TD methods,
% as indicated by the gray points in Fig.~\ref{fig:biol_freqs}: predictions
% between families of methods are very different and yet choice of family of
% method would require fitness landscape-specific and genotype-specific
% information, as we have seen in Fig.~\ref{fig:7_spec}, as well as
% Figures~\ref{S-fig:10_spec}, \ref{S-fig:7_ALL} and \ref{S-fig:10_ALL}.

\subsubsection*{SSWM}

Deviations from SSWM are known to undermine the predictive power of CPMs \cite{diaz-uriarte2018,diaz-uriarte2019a}. The model underlying most CPMs seems to implicitly assume SSWM mainly for two reasons: \textit{a)} CPMs are powered with cross sectional data that do not undergo any deconvolution step (i.e. inference of clonal composition of the sequenced tumor), implying that a unique genotype\todo{``clone genotype"?} can be obtained from bulk sequencing, and \textit{b)} those observed clones are considered true steps in the path to tumor fixation, thus neglecting the possibility of tunneling effects. The two are true only if there is SSWM, in which case the most fit genotype at any given time is effectively the only one in the population and a direct ancestor of the ultimately fixated one. Even when CPMs specifically include error estimations (e.g. in Gerstrung et al., 2009\cite{Gerstung2009}), these typically refer to sequencing error but not necessarily to deviations from SSWM.

\iffalse
Most CPMs assume that cancer progresses under conditions of strong
selection and weak mutation (SSWM).\cite{hosseini2019a}\todo{I do not
find where Hosseini et al say that. Reword along:}
\idea{The model underlying most CPMs seems to
  implicitly assume SSWM: poorly written now, but a couple of key ideas:
  observed data are used directly, without any deconvolution step, which
  seems to indicate that even bulk sequencing can obtain a ``clone
  genotype''; this would only work if SSWM; and then those observed clone
  genotypes are considered true steps in the path to the maximum (i.e., no
  tunneling); note that the error included in models, e.g., Gerstung et
  al., 2009, is designed to capture sequencing error, but not necessarily
  deviations from SSWM ---slightly more subtle in MCCBN, though. Reword
  this properly, move stuff out to discussion, and leave here a simple
  sentence. One part of that sentence could be ``This assumption is rarely
  stated explicitly, but it is explicit in how \cite{hosseini2019a} use
  CBN for predicting tumor evolution.''  } \idea{Recheck MHN paper for
  SSWM}.
\fi

\idea{  One part of that sentence could be ``This assumption is rarely
  stated explicitly, but it is explicit in how \cite{hosseini2019a} use
  CBN for predicting tumor evolution.''  } \idea{Recheck MHN paper for
  SSWM}.

this number has an inverted U-shape with the
number of mutated loci (for a genotype $i$, the number of non-zero
$p_{i,j}$ is ${N}\choose{n + 1}$, where $n$ is the number of mutations of
the genotype and $N$ the total number of loci).

When we are interested in predicting the complete evolutionary path the
only features that matter are global features.
 But when we want to make conditional, short-term
predictions, features of the genotypes for which we make the conditional
predictions can be as relevant as, or more relevant than, fitness
landscape-specific features \idea{citar banks y quizas otros}. This is a
noteworthy difference because it suggests that CPMs could be leveraged to
provide short-term predictions even when global, long-term predictions are
difficult because fitness landscape-specific assumptions are violated.

Our results indicate that,
were we to know a genotype can be a local maximum, specially if SSWM
holds, then we could make

Los resultados de que a veces algunos tipos de métodos dan valores muy
similares enter sí, pero muy distintos a otros, no son solo cosa de las
simulaciones sino tb. en datos empírciso; ojo entonces con la
interpretación porque depende de factores que no conocemos.  \idea{plot el
  MHN vs MHN\_td para simulados?  sea como
  \ref{fig:biol_boxplots-3-comps}?}

\idea{remember that the local
maxima scenarios could slightly favor CBN, since the local maxima
landscapes are built as deviations from the CBN, and in part from OT/CAPRESE ).Also representable? What is the model for MHN?}

\idea{Null model no está. Either mention in results, when we mention
  results for number of mutations. Or mention here. I think that to keep
  ms short, null model figures should go in suppl mat, unless null model
  becomes a more important part of the ms.}

% \begin{figure}[H]
%   \centering
%   \fcolorbox{black}{white}{\includegraphics[width=0.8\linewidth,trim={4.5cm 10cm 4cm 9cm}]{figs/figexp}}
%   \internallinenumbers
%   \caption[Figure title]{\textbf{Figure title.} Figure caption. \idea{Esta
%     figura tapa texto encima}}
%   \label{figexp}
% \end{figure}

% The effects of the sampling and
% fitness landscape characteristics in predicting short-term evolution (as
% summarized in the linear mixed-effects trees (Fig.~\ref{fig:7_spec} and
% Supplementary Material Figs.~\ref{S-fig:10_spec}, ~\ref{S-fig:7_ALL},
% ~\ref{S-fig:10_ALL}) are, thus, not surprising. What is noteworthy is that
% local, genotype-specific features, i.e., features of the genotypes for
% which we make the conditional predictions, can be much more relevant than
% fitness landscape-specific features \idea{citar banks y quizas
%   otros}. These suggests that CPMs could be leveraged to provide
% short-term predictions even when global, long-term predictions are
% difficult: when we are interested in predicting the complete path
% evolutionary path, of course, the only features that matter are global
% features and these can often lead to poor performance
% \cite{diaz-uriarte2019a}. Our results here, however, indicate that good
% conditional predictions are possible even if fitness landscape-specific
% assumptions are violated.

% \idea{NO! Reword as}

% The effects of the sampling and
% fitness landscape characteristics in predicting short-term evolution (as
% summarized in the linear mixed-effects trees (Fig.~\ref{fig:7_spec} and
% Supplementary Material Figs.~\ref{S-fig:10_spec}, ~\ref{S-fig:7_ALL},
% ~\ref{S-fig:10_ALL}) are, thus, not surprising.

#####################   SSWM

it is widely accepted that intra-tumor heterogeneity is
common in cancer \cite{mcgranahan_clonal_2017, hinohara2019, zhao2016},
which can be partly due to deviations from SSWM arising from clonal
interference and stochastic tunneling
\cite{devisser_empirical_2014,szendro_predictability_2013,sniegowski_beneficial_2010};
for example, in multiple myeloma ``heteroeneoys clonal mixtures with
shifting predominant clones'' have been described
\cite[][p.~1067]{keats2012} %% could give example of Goto et al., 2018
                            %% lung cancer
suggesting the existence of non-SSWM regimes,
and stochastic tunneling in
cancer has, itself, been the focus of theoretical work
\cite[e.g.,][]{Ashcroft2015,wodarz_dynamics_2014}.
% deviations from SSWM and phenomena such as clonal
% interference and stochastic tunneling (due to rugged fitness landscapes
% where epistasis is common) are frequent
% \cite{devisser_empirical_2014}\todo{more refs; in fact, this is not a
%   reference about cancer}. \idea{search for refs of SSWM and tunneling in
%   cancer; there is one by Michor and see also the book by Wodarz and
%   komarova}.

it is widely accepted that intra-tumor heterogeneity is
common in cancer \cite{mcgranahan_clonal_2017, hinohara2019, zhao2016},
which can be partly due to deviations from SSWM arising from clonal
interference and stochastic tunneling
\cite{devisser_empirical_2014,szendro_predictability_2013,sniegowski_beneficial_2010};
for example, in multiple myeloma ``heteroeneoys clonal mixtures with
shifting predominant clones'' have been described
\cite[][p.~1067]{keats2012} %% could give example of Goto et al., 2018
                            %% lung cancer
suggesting the existence of non-SSWM regimes,
and stochastic tunneling in
cancer has, itself, been the focus of theoretical work
\cite[e.g.,][]{Ashcroft2015,wodarz_dynamics_2014}.
% deviations from SSWM and phenomena such as clonal
% interference and stochastic tunneling (due to rugged fitness landscapes
% where epistasis is common) are frequent
% \cite{devisser_empirical_2014}\todo{more refs; in fact, this is not a
%   reference about cancer}. \idea{search for refs of SSWM and tunneling in
%   cancer; there is one by Michor and see also the book by Wodarz and
%   komarova}.

When we move away from SSWM, multiple clones can coexist at significant
fractions for prolonged periods of time, and new higher-fitness clones can
evolve from low fitness parents that represented a minority of the
population. Genotypes with a large number of
mutations %high mutational load
can already be present at the time of sampling yet go undetected if their
relative abundance is small (figure
\ref{fig:fig1}D). % Deviations from SSWM
% are known to undermine the predictive power of CPMs
% \cite{diaz-uriarte2018,diaz-uriarte2019a}, as the model underlying most
% CPMs seems to assume SSWM (see Discussion). This assumption, however, is
% not always stated
% explicitly.
% , but it is explicit, for example, in Hosseini et al.'s
% \cite{hosseini2019a} use of CPMs for predicting tumor evolution (see
% \nameref{discussion}).

Widely accepted that ITH common in cancer \cite{mcgranahan_clonal_2017}
and there are examples that suggest presence of non-SSWM regimes
\cite{keats2012} with ``heteroeneoys clonal mixtures with shifting
predominant clones'' and stochastic tunneling has, itself, been the focus
of theoretical work \cite{Ashcroft2015}

\cite{bauer2015}: yeap, cite this for predict in
non-SSWM

###############################################################

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2021-04-14 draft/supp.tex
supplementary material created and some methods added

\documentclass[a4paper,10pt]{article}

% packages
\usepackage[utf8]{inputenc} % allows usage of spanish special characters
\usepackage[spanish,english]{babel} % english dictionary for proper hyphenation
\usepackage{amsmath} % math expressions
\usepackage{txfonts} % nice fonts 
\usepackage{authblk} % author affiliations
\usepackage{graphicx} % images
\usepackage{float} % image positioning
\usepackage{multirow} % allows merging cells in tables
\usepackage{mathtools} % nice matrices
\usepackage{rotating} % rotated text
\usepackage{caption} % customization of figure/table captions
\usepackage{subcaption} % collages of multiple images
%% \usepackage[hidelinks]{hyperref} % hyperlinks
\usepackage{hyperref} % hyperlinks
\usepackage{nameref} % cross-references between draft and supplementary material
\usepackage{zref-xr,zref-user} % more cross-references formatting
\usepackage[square,numbers,sort&compress]{natbib}
\usepackage{xcolor} % text color
\usepackage[margin=3cm]{geometry} % margins

\usepackage{pdflscape} %% Landscape
\usepackage{dcolumn}
\usepackage{booktabs}
\usepackage{savesym}
\savesymbol{iint}
\restoresymbol{TXF}{iint}
%% \usepackage{pdfpages} %% to include a pdf file

% customization
\setcounter{Maxaffil}{0} % affiliations
\renewcommand\Affilfont{\itshape\small} % style of affiliations text
\makeatletter\renewcommand{\@biblabel}[1]{#1.}\makeatother % change [number] for number. in reference list
\newcommand{\idea}[1]{\textcolor{red}{#1}}
\newcommand{\mr}{\textcolor{red}{\textbf{[missing ref(s)]}}}
\newcommand{\Burl}[1]{\url{#1}}
\zxrsetup{toltxlabel} % cross references: draft & suppl mat
\zexternaldocument*[M-]{draft}
\hypersetup{
  colorlinks = true,
  citecolor=  black,
  linkcolor = {blue},
  filecolor = cyan % controls color of external ref, if used
}
\renewcommand{\thefigure}{\textbf{S\arabic{figure}}} % preceed the Suppl Mat figures with S
\renewcommand{\thetable}{\textbf{S\arabic{table}}} % preceed the Suppl Mat tables with S
\renewcommand{\theequation}{S\arabic{equation}} % preceed the Suppl Mat equations with S
\captionsetup[figure]{font=small, labelfont={bf}, labelformat={default},
   labelsep=period, name={Fig.}} % custom figure captions
\captionsetup[table]{font=small, labelfont={bf}, labelformat={default},
   labelsep=period, name={Table}} % custom table captions

% title, authors, affiliations
\title{Supplementary Material for ``Provisional Title'' }
\author[1,2]{Authors}
\author[1,$\dagger$]{Álvaro Sánchez}
\affil[1]{Department of Ecology \& Evolutionary Biology and Microbial
  Sciences Institute, Yale University, New Haven, CT, USA}
\affil[2]{Other affiliations...}
\affil[$\dagger$]{To whom correspondence should be addressed: \normalfont alvaro.sanchez@yale.edu}
\date{}

\newcommand\legendtreefigs{Internal nodes show the partitioning variables   and the values that lead to each partition; light blue denotes the   values or categories that lead down the left split and salmon the values   or categories that lead down the right split (partitioning criteria are   also shown in the edges leading down a node). Leaves or terminal nodes   show (weighted) boxplots of the dependent variable for the observations   that belong that fall in that leave and, as label, the name of the node   ax_3,nd the sample size (sum of weighted observations and \% of the total   data set for the fit). The dot plots underneath the boxplots show the   mean JS for all the methods for that leave.}

\begin{document}

\maketitle
\tableofcontents
\newpage

\section{Supplementary Methods}\label{supp-methods}

\subsection{Data processing}\label{supp-methods:data-proc}

Thresholds for vector similarity in ESV space, etc.

\subsubsection{Special cases}\label{supp-methods:data-proc:special}

Blablabla.

\subsection{Inferring species abundance from sequencing data}
\label{supp-methods:species-from-seq}

Consider a community of $N$ species with relative abundances represented
by the components of a vector $\mathbf{x} = \left( x_1, x_2, \dotsc,
x_N\right)$ such that $x_i$ is the relative abundance of the $i$-th species.
Naturally, the conditions
%
\begin{equation}
\sum_i x_i = 1
\label{supp-eqn:normx_1}
\end{equation}
%
and
%
\begin{equation}
0 \leq x_i \leq 1 \; \textrm{for all} \; i
\label{supp-eqn:normx_2}
\end{equation}
%
are satisfied.

Performing 16S rRNA gene sequencing on such a community yields a list of
sequences (exact sequence variants or ESVs) and their respective abundance.
Normalizing by the total number of sequences, we can obtain a vector $\mathbf{s}
= \left( s_1, s_2, \dotsc \right)$ that also satisfies the normalization
conditions of equations \ref{supp-eqn:normx_1} and \ref{supp-eqn:normx_2}.
Only if there are as many ESVs as species $\mathbf{s}$ has length $N$
(the same as as $\mathbf{x}$), but in the most general case it is possible that
a) multiple species share a same ESV and/or b) species carrying multiple copies
of the 16S rRNA gene have different sequences for each copy\mr. We denote
the length of $\mathbf{s}$ as $M$, which need not be equal to $N$.

By sequencing each species individually, we can build a $M \times N$ 
matrix $\mathbf{Q}$ such
that the element in the $(i,j)$ position, $q_{ij}$, represents the frequency of
the $i$-th ESV when the $j$-th species is sequenced. $\mathbf{Q}$ can be used
to determine the fraction of sequences of a given ESV that are obtained from
sequencing a community with any arbitrary composition. In other words,
$\mathbf{Q}$ maps $\mathbf{x}$ to $\mathbf{s}$:

\begin{equation}
\mathbf{Q} \cdot \mathbf{x} = \mathbf{s}
\label{supp-eqn:Qxs}
\end{equation}

Note that equation \ref{supp-eqn:Qxs} is only true if all species have the same
number of copies of the 16S rRNA gene. Otherwise, species with high copy numbers
will yield more sequences, thus leading to an over-representation and vice-versa.
Information on the 16S copy number for every species in the communities would be
required to overcome this limitation, but this is not feasible when dealing with
highly diverse communities. For our purpose in this
work, we will need to assume that the copy numbers of the 16S are relatively
conserved across species and equation \ref{supp-eqn:Qxs} holds. Additionally,
we have sequencing data from a subset of the species in our
communities, but not from all of them. This means that we cannot build a complete
$\mathbf{Q}$ matrix. To address this, we first identify every ESV obtained
from community sequencing that we cannot map back to (at least) one of the
species for which we have single-species sequencing information. We then
assume that each
of those ESVs maps uniquely to a single species. This gives us a $\mathbf{Q}$
matrix of the form:

\begin{equation}
\mathbf{Q} = \; \; \; \; \; \begin{matrix}
 &

\begin{matrix}
\; \; \;
\rotatebox{90}{\scriptsize species 1} \; \; \; & 
\rotatebox{90}{\scriptsize species 2} \; \; \; &
\rotatebox{90}{\scriptsize species 3} \; \; &
\cdots \; \; &
\rotatebox{90}{\scriptsize species $N-2$} \! \! & 
\rotatebox{90}{\scriptsize species $N-1$} \! \! &
\rotatebox{90}{\scriptsize species $N$}
\end{matrix}

\\

\begin{matrix*}[r]
\rotatebox{0}{\scriptsize sequence 1} \\
\rotatebox{0}{\scriptsize sequence 2} \\
\rotatebox{0}{\scriptsize sequence 3} \\
\vdots \; \; \; \\
\rotatebox{0}{\scriptsize sequence $M-2$} \\ 
\rotatebox{0}{\scriptsize sequence $M-1$} \\
\rotatebox{0}{\scriptsize sequence $M$}
\end{matrix*}

&

\begin{pmatrix}
q_{11} & q_{12} & q_{13} &  & 0 & 0 & 0 \\
q_{21} & q_{22} & q_{23} & \cdots & 0 & 0 & 0 \\
q_{31} & q_{32} & q_{33} &  & 0 & 0 & 0 \\
 & \vdots &  & \ddots  &  &  &  \\
0 & 0 & 0 &  & 1 & 0 & 0 \\ 
0 & 0 & 0 &  & 0 & 1 & 0 \\ 
0 & 0 & 0 &  & 0 & 0 & 1 
\end{pmatrix}

\end{matrix}
\label{supp-eqn:Q_modified}
\end{equation}
%
While a one-to-one mapping between species and ESVs may not be true in all cases,
in practice the ESVs for which we make this assumption have low relative
abundances in the communities, minimizing the impact of any potential artifact.

Having built our $\mathbf{Q}$ matrix according to equation
\ref{supp-eqn:Q_modified}, obtaining species abundances from sequencing data
is relatively straigthforward: we need to find the $\mathbf{x}$ that satisfies
equation \ref{supp-eqn:Qxs} for a given $\mathbf{s}$. However, in some cases
sequencing noise can introduce deviations in $\mathbf{s}$ that make it so the
vector $\mathbf{x}$ that solves equation \ref{supp-eqn:Qxs} does not meet the
boundary conditions in equation \ref{supp-eqn:normx_2}, i.e., some $x_i$ may be
negative or greater than 1. These cases are obviously problematic, but we can
avoid them by accounting for potential deviations in $\mathbf{s}$. We do this
by solving the following nonlinear optimization problem. We want to obtain an estimate of the true community composition $\mathbf{x}$
(that we will denote as $\mathbf{\hat{x}}$) so that the product
$\mathbf{Q} \cdot \mathbf{\hat{x}}$ is as close as possible to $\mathbf{s}$ while
having $\mathbf{\hat{x}}$ satisfy the normalization condition in equation
\ref{supp-eqn:normx_1} and the boundary conditions in equation
\ref{supp-eqn:normx_2}. For every possible $\mathbf{\hat{x}}$, we define a vector
$\mathbf{\epsilon} = \mathbf{\epsilon} \left( \mathbf{\hat{x}} \right) = 
\left( \epsilon_1, \epsilon_2, \cdots, \epsilon_M \right)$ as

\begin{equation}
\mathbf{\epsilon} \left( \mathbf{\hat{x}} \right) =
\mathbf{Q} \cdot \mathbf{\hat{x}} - \mathbf{s}
\end{equation}
%
and a function $f \left( \mathbf{\hat{x}} \right)$ as

\begin{equation}
f \left( \mathbf{\hat{x}} \right) = \sum_i \epsilon_i^2
\end{equation}
%
We also define a function $h \left( \mathbf{\hat{x}} \right)$ as

\begin{equation}
h \left( \mathbf{\hat{x}} \right) = 1 - \mathbf{\hat{x}}
\end{equation}
%
The best estimation for the species composition of a community will be the
$\mathbf{\hat{x}}$ that minimizes $f \left( \mathbf{\hat{x}} \right)$ while
satisfying the normalization condition (eq. \ref{supp-eqn:normx_1}) and the
boundary conditions (eq. \ref{supp-eqn:normx_2}). We solve
this problem using the \texttt{nloptr} package for R\mr, passing it
$f \left( \mathbf{\hat{x}} \right)$ as the function to minimze and
$h \left( \mathbf{\hat{x}} \right) = 0$ as an equality constraint, as well as
0 and 1
as the lower and upper bounds for the entries of $\mathbf{\hat{x}}$. Further
details can be found in the code for the analysis (see section 
\nameref{supp-datacode} of this Supplementary Material).

\clearpage

\section{Data \& code availability}
\label{supp-datacode}

Data and code for the analyses in this article, as well as code for the
\texttt{community-simulator} package updated with the new functionalities,
is available at \url{https://github.com/jdiazc9/coalescence}.

\clearpage

\section{Supplementary Results}\label{supp-results}

Blabla

\subsection{Linear mixed-effects model trees}
\label{suppl-linear-mixed-model-results}

Blabla

\clearpage

\section{Supplementay Figures}\label{supp-figs}

Blablabla.

\clearpage

% references
{\small
  %% \bibliography{refs}
  \bibliography{extracted-supp}
% \bibliographystyle{unsrt}
%% \bibliographystyle{mystyle}
\bibliographystyle{mybibwithurl}
}

\end{document}

### 2021-04-15 draft/supp.tex
aesthetic changes (for consistency and clarity of math notations)

@@
 \usepackage{amsmath} % math expressions
+\usepackage{upgreek} % upright greek letters for math
 \usepackage{txfonts} % nice fonts 
@@
 \usepackage{subcaption} % collages of multiple images
-%% \usepackage[hidelinks]{hyperref} % hyperlinks
 \usepackage{hyperref} % hyperlinks
@@
 conditions of equations \ref{supp-eqn:normx_1} and \ref{supp-eqn:normx_2}.
-Only if there are as many ESVs as species $\mathbf{s}$ has length $N$
-(the same as as $\mathbf{x}$), but in the most general case it is possible that
-a) multiple species share a same ESV and/or b) species carrying multiple copies
+Only if there are as many ESVs as species does $\mathbf{s}$ have length $N$
+(the same as $\mathbf{x}$), but in the most general case it is possible that
+\textit{a)} multiple species share a same ESV and/or \textit{b)} species carrying
+multiple copies
 of the 16S rRNA gene have different sequences for each copy\mr. We denote
@@
 equation \ref{supp-eqn:Qxs} for a given $\mathbf{s}$. However, in some cases
-sequencing noise can introduce deviations in $\mathbf{s}$ that make it so the
+sequencing error can introduce deviations in $\mathbf{s}$ that make it so the
 vector $\mathbf{x}$ that solves equation \ref{supp-eqn:Qxs} does not meet the
@@
 avoid them by accounting for potential deviations in $\mathbf{s}$. We do this
-by solving the following nonlinear optimization problem. We want to obtain an estimate of the true community composition $\mathbf{x}$
-(that we will denote as $\mathbf{\hat{x}}$) so that the product
-$\mathbf{Q} \cdot \mathbf{\hat{x}}$ is as close as possible to $\mathbf{s}$ while
+by solving the following nonlinear optimization problem: we want to obtain an
+estimate of the true community composition
+(we will denote the estimate as $\mathbf{\hat{x}}$) so that the product
+$\mathbf{Q} \cdot \mathbf{\hat{x}}$ is as close as possible to the $\mathbf{s}$
+obtained from sequencing while
 having $\mathbf{\hat{x}}$ satisfy the normalization condition in equation
 \ref{supp-eqn:normx_1} and the boundary conditions in equation
-\ref{supp-eqn:normx_2}. For every possible $\mathbf{\hat{x}}$, we define a vector
-$\mathbf{\epsilon} = \mathbf{\epsilon} \left( \mathbf{\hat{x}} \right) = 
-\left( \epsilon_1, \epsilon_2, \cdots, \epsilon_M \right)$ as
+\ref{supp-eqn:normx_2}. For every possible $\mathbf{\hat{x}}$, we can define a vector
+$\boldsymbol{\upvarepsilon} = \boldsymbol{\upvarepsilon} \left( \mathbf{\hat{x}} \right) = 
+\left( \varepsilon_1, \varepsilon_2, \cdots, \varepsilon_M \right)$ as
 
 \begin{equation}
-\mathbf{\epsilon} \left( \mathbf{\hat{x}} \right) =
+\boldsymbol{\upvarepsilon} \left( \mathbf{\hat{x}} \right) =
 \mathbf{Q} \cdot \mathbf{\hat{x}} - \mathbf{s}
@@
 \begin{equation}
-f \left( \mathbf{\hat{x}} \right) = \sum_i \epsilon_i^2
+f \left( \mathbf{\hat{x}} \right) =
+\left| \boldsymbol{\upvarepsilon} \left( \mathbf{\hat{x}} \right) \right| ^2 =
+\sum_i \varepsilon_i^2 \left( \mathbf{\hat{x}} \right)
 \end{equation}
 %
-We also define a function $h \left( \mathbf{\hat{x}} \right)$ as
+We can also define a function $\mathbf{h} \left( \mathbf{\hat{x}} \right)$ as
 
 \begin{equation}
-h \left( \mathbf{\hat{x}} \right) = 1 - \mathbf{\hat{x}}
+\mathbf{h} \left( \mathbf{\hat{x}} \right) = 1 - \mathbf{\hat{x}}
 \end{equation}
@@
 this problem using the \texttt{nloptr} package for R\mr, passing it
-$f \left( \mathbf{\hat{x}} \right)$ as the function to minimze and
-$h \left( \mathbf{\hat{x}} \right) = 0$ as an equality constraint, as well as
-0 and 1
+$f \left( \mathbf{\hat{x}} \right)$ as the function to minimize,
+$\mathbf{h} \left( \mathbf{\hat{x}} \right) = 0$ as an equality constraint,
+and 0 and 1
 as the lower and upper bounds for the entries of $\mathbf{\hat{x}}$. Further
@@
 
-\subsection{Linear mixed-effects model trees}
-\label{suppl-linear-mixed-model-results}
+\subsection{Example subsection}
+\label{supp-results:example}
 
@@
 % references
-{\small
+%{\small
   %% \bibliography{refs}
-  \bibliography{extracted-supp}
+%  \bibliography{extracted-supp}
 % \bibliographystyle{unsrt}
 %% \bibliographystyle{mystyle}
-\bibliographystyle{mybibwithurl}
-}
+%\bibliographystyle{mybibwithurl}
+%}
 

### 2021-04-15 draft/supp.tex
more supp methods added

@@
 \usepackage[margin=3cm]{geometry} % margins
+\usepackage[nottoc,numbib]{tocbibind} % add references to table of contents
 
@@
 \makeatletter\renewcommand{\@biblabel}[1]{#1.}\makeatother % change [number] for number. in reference list
+\addto{\captionsenglish}{\renewcommand{\bibname}{Supplementary References}}
 \newcommand{\idea}[1]{\textcolor{red}{#1}}
@@
    labelsep=period, name={Table}} % custom table captions
+\hypersetup{citecolor={blue}}
 
@@
 \title{Supplementary Material for ``Provisional Title'' }
-\author[1,2]{Authors}
-\author[1,$\dagger$]{Álvaro Sánchez}
-\affil[1]{Department of Ecology \& Evolutionary Biology and Microbial
-  Sciences Institute, Yale University, New Haven, CT, USA}
-\affil[2]{Other affiliations...}
+\author[1,2,3]{Authors}
+\author[1,2,$\dagger$]{Álvaro Sánchez}
+\affil[1]{Department of Ecology \& Evolutionary Biology,
+Yale University, New Haven, CT, USA}
+\affil[2]{Microbial Sciences Institute,
+Yale University, New Haven, CT, USA}
+\affil[3]{Other affiliations...}
 \affil[$\dagger$]{To whom correspondence should be addressed: \normalfont alvaro.sanchez@yale.edu}
@@
 
-Thresholds for vector similarity in ESV space, etc.
+\subsubsection{Examining community composition}
+\label{supp-methods:data-proc:community-composition}
+
+We first analyze the composition of the communities assembled in each carbon
+source by examining the relative abundances of the exact sequence variants
+(ESVs) obtained from 16S rRNA gene sequencing. We define a vector $\mathbf{s}
+= \left( s_1, s_2, \cdots, s_M \right)$ such that $s_i$ represents the relative
+abundance
+of the $i$-th ESV, being $M$ the total number of unique ESVs. We assembled
+a total of 16 communities, 8 in glutamine
+and 8 in citrate (see Methods \mr), with 3 biological replicates of each. We
+start
+by comparing the composition of replicate communities in terms of their ESV
+relative abundances. To do this, we measure the distance $d$ between two
+communities (with ESV abundances $\mathbf{s}$ and $\mathbf{s}'$ respectively) as
+
+\begin{equation}
+d \left( \mathbf{s}, \mathbf{s}' \right) =
+\sqrt{\sum_i{\left(s_i - s'_i\right)^2}}
+\label{supp-eqn:esvdist}
+\end{equation}
+%
+We compute all the distances across replicates of the same community, obtaining
+a set of 48 distances (16 communities $\times$ 3 pairwise distances between the
+three replicates) that we will denote as $\left\{d\right\}$
+
+\begin{equation} 
+\left\{d\right\} = 
+\left\{ d\left(\mathrm{community} \; i \; \mathrm{replicate} \; j,
+\mathrm{community} \; i \; \mathrm{replicate} \; j' \right) \;
+\text{for all $i$, $j$, $j' \neq j$} \right\}
+\end{equation}
+%
+We then define a threshold $d_T$ as
+
+\begin{equation}
+d_T = \mathrm{Q}4 \left( \left\{d\right\} \right) + 
+\mathrm{IQR} \left( \left\{d\right\} \right)
+\label{supp-eqn:threshold}
+\end{equation}
+%
+Where Q4 and IQR represent the fourth quantile and the interquartile range
+respectively. Note that this threshold is analogous to the one used in
+standard boxplots to identify outliers.
+
+Finally, we discard the samples that are at a distance larger than $d_T$ from
+all other replicates of the same community. These are replicate 2 of community
+1 in glutamine and replicate 2 of community 8 in citrate (see Figure S1 \mr).
+The remaining replicates were used in further analyses. We took the average
+ESV abundance across replicates when indicated.
 
@@
 multiple copies
-of the 16S rRNA gene have different sequences for each copy\mr. We denote
+of the 16S rRNA gene have different sequences for each copy \cite{Vetrovsky2013}. We denote
 the length of $\mathbf{s}$ as $M$, which need not be equal to $N$.
@@
 %
-We can also define a function $\mathbf{h} \left( \mathbf{\hat{x}} \right)$ as
+We can also define a function $h \left( \mathbf{\hat{x}} \right)$ as
 
 \begin{equation}
-\mathbf{h} \left( \mathbf{\hat{x}} \right) = 1 - \mathbf{\hat{x}}
+h \left( \mathbf{\hat{x}} \right) = 1 - \sum_i\hat{x}_i
 \end{equation}
@@
 boundary conditions (eq. \ref{supp-eqn:normx_2}). We solve
-this problem using the \texttt{nloptr} package for R\mr, passing it
+this problem using the \texttt{nloptr} package for R \cite{nloptr}, passing it
 $f \left( \mathbf{\hat{x}} \right)$ as the function to minimize,
-$\mathbf{h} \left( \mathbf{\hat{x}} \right) = 0$ as an equality constraint,
+$h \left( \mathbf{\hat{x}} \right) = 0$ as an equality constraint,
 and 0 and 1
-as the lower and upper bounds for the entries of $\mathbf{\hat{x}}$. Further
+as the lower and upper bounds for the entries of $\mathbf{\hat{x}}$, and using
+the augmented Lagrangian algorithm \cite{auglag1,auglag2}. Further
 details can be found in the code for the analysis (see section 
@@
 % references
-%{\small
-  %% \bibliography{refs}
-%  \bibliography{extracted-supp}
-% \bibliographystyle{unsrt}
-%% \bibliographystyle{mystyle}
-%\bibliographystyle{mybibwithurl}
-%}
+\bibliographystyle{mystyle}
+%\bibliographystyle{unsrt}
+\bibliography{refs}
 

### 2021-04-16 draft/supp.tex
typos

@@
 \begin{equation}
-d_T = \mathrm{Q}4 \left( \left\{d\right\} \right) + 
-\mathrm{IQR} \left( \left\{d\right\} \right)
+d_T = \mathrm{Q}3 \left( \left\{d\right\} \right) + 
+1.5 \; \mathrm{IQR} \left( \left\{d\right\} \right)
 \label{supp-eqn:threshold}
@@
 %
-Where Q4 and IQR represent the fourth quantile and the interquartile range
+Where Q3 and IQR represent the third quartile and the interquartile range
 respectively. Note that this threshold is analogous to the one used in
@@
 
-\subsubsection{Special cases}\label{supp-methods:data-proc:special}
+\subsubsection{Characterizing dominant species}\label{supp-methods:data-proc:doms}
 
-Blablabla.
+As described in the main text Methods \mr, communities were plated and the most
+abundant species (as determined by counting of colony morphotypes) was isolated
+and allowed to regrow in monoculture prior to sequencing. We then characterize
+the isolated species to, first, make sure that they were dominant in the communities
+they were isolated from, and second, identify those instances where two or more
+communities share the same dominant.
+
+Since we assembled 16 communities, we have 16 isolates. We will denote as
 

### 2021-04-16 draft/supp.tex
more supp

--- 2021-04-15 draft/supp.tex
@@
 \usepackage{multirow} % allows merging cells in tables
+\usepackage{makecell} % more table customization
 \usepackage{mathtools} % nice matrices
@@
 \usepackage[square,numbers,sort&compress]{natbib}
-\usepackage{xcolor} % text color
+\usepackage[table,xcdraw]{xcolor} % text and table background colors
 \usepackage[margin=3cm]{geometry} % margins
@@
 \hypersetup{citecolor={blue}}
+\definecolor{lightgray}{gray}{0.9}
 
@@
 1 in glutamine and replicate 2 of community 8 in citrate (see Figure S1 \mr).
-The remaining replicates were used in further analyses. We took the average
+The remaining replicates were used in further analyses, taking the average
 ESV abundance across replicates when indicated.
 
-\subsubsection{Special cases}\label{supp-methods:data-proc:special}
+\subsubsection{Characterization of dominant species}
+\label{supp-methods:data-proc:doms}
 
-Blablabla.
+As described in the main text Methods \mr, generationally stabilized communities
+were plated and the most abundant species
+(determined by counting of colony morphotypes) was isolated
+and allowed to regrow. We grew three biological
+replicates of each isolated species in monoculture and then performed
+16S rRNA gene sequencing. We characterize
+the isolated species to, first,
+determine whether multiple communities share the same dominant
+and, second, to
+make sure that the dominants were correctly identified.
 
-\subsection{Inferring species abundance from sequencing data}
+Analogous to what we did in the section above
+(\nameref{supp-methods:data-proc:community-composition}), we start by estimating
+the sequencing error by comparing the distances between all pairs of replicates.
+We obtain an ``error threshold'' like we did in equation \ref{supp-eqn:threshold},
+and we discard replicates that fall at a larger distance
+from all other replicates of the same isolate. This eliminates one replicate
+from the monocultures of the isolates from glutamine communities 2 and 3, and
+an additional replicate from citrate community 1.
+
+We then compare each isolate with
+one another. The distance between the vectors of ESV abundances of two
+different isolates being below our threshold indicates that both
+are the same species, or at the very least that we cannot
+distinguish them through 16S sequencing. These instances were removed from 
+the dominant pairwise competition analysis. We find that indeed some
+communities share
+the same (or undistinguishable) isolated dominant. This leaves us with 9 unique
+isolates.
+
+Table \ref{supp-tab:isolates} shows the relative ESV abundance of each unique
+isolate, as well as which communities they were isolated from.
+Table \mr shows the ESVs obtained from 16S sequencing
+of the monocultures. Note that three of the isolates partially share a same
+ESV; however, when growing them in pairwise competition there is always at least
+another ESV that is unique to only one of them.
+Further details can be found in the code for the analysis (see section 
+\nameref{supp-datacode} of this Supplementary Material).
+
+\begin{table}
+\scriptsize
+\centering
+\rowcolors{2}{lightgray}{white}
+
+\begin{tabular}{lccccccccc}
+ & \makecell{Isolate 1 \\ G-1, C-8} 
+ & \makecell{Isolate 2 \\ G-2, G-5, C-3, C-5 }
+ & \makecell{Isolate 3 \\ G-3 }
+ & \makecell{Isolate 4 \\ G-4, G-6 }
+ & \makecell{Isolate 5 \\ G-7, G-8 }
+ & \makecell{Isolate 6 \\ C-1 }
+ & \makecell{Isolate 7 \\ C-2, C-7 }
+ & \makecell{Isolate 8 \\ C-4 }
+ & \makecell{Isolate 9 \\ C-6 } \\ \cline{2-10} 
+\multicolumn{1}{l|}{Sequence 1} &  & 1 &  & 0.86 &  & 0.63 &  &  &  \\
+\multicolumn{1}{l|}{Sequence 2} &  &  &  &  &  &  & 1 &  &  \\
+\multicolumn{1}{l|}{Sequence 3} &  &  & 1 &  &  &  &  &  &  \\
+\multicolumn{1}{l|}{Sequence 4} & 1 &  &  &  &  &  &  &  &  \\
+\multicolumn{1}{l|}{Sequence 5} &  &  &  &  &  &  &  & 1 &  \\
+\multicolumn{1}{l|}{Sequence 6} &  &  &  &  & 1 &  &  &  &  \\
+\multicolumn{1}{l|}{Sequence 7} &  &  &  & 0.14 &  &  &  &  &  \\
+\multicolumn{1}{l|}{Sequence 8} &  &  &  &  &  &  &  &  & 1 \\
+\multicolumn{1}{l|}{Sequence 9} &  &  &  &  &  & 0.37 &  &  &
+\end{tabular}
+
+\caption{\textbf{Species to ESVs map of the dominant species isolated from
+the communities.} For every isolate we indicate the community (or communities if
+they share dominants) from which it was obtained. G-$n$ and C-$n$
+($n=1 \; \text{to} \; 8$)
+denote the $n$-th community in glutamine and citrate respectively.
+Values in the table represent relative ESV abundances from 16S sequencing
+of monocultures.
+Empty entries of the table represent zeros.}
+\label{supp-tab:isolates}
+
+\end{table}
+
+\begin{table}
+\footnotesize
+\centering
+\rowcolors{0}{lightgray}{white}
+
+\begin{tabular}{ll}
+
+Sequence 1 &
+\makecell[l]{
+\tt GCAAGCGTTA ATCGGAATTA CTGGGCGTAA AGCGCGCGTA GGTGGTTTGT TAAGTTGGAT \\
+\tt GTGAAAGCCC CGGGCTCAAC CTGGGAACTG CATCCAAAAC TGGCAAGCTA GAGTACGGTA \\
+\tt GAGGGTGGTG GAATTTCCTG TGTAGCGGTG AAATGCGTAG ATATAGGAAG GAACACCAGT \\
+\tt GGCGAAGGCG ACCACCTGGA CTGATACTGA CACTGAGGTG CGAAAGCGTG GGG
+}
+
+\\
+
+Sequence 2 &
+\makecell[l]{
+\tt GCAAGCGTTA ATCGGAATTA CTGGGCGTAA AGCGCACGCA GGCGGTCTGT CAAGTCGGAT \\
+\tt GTGAAATCCC CGGGCTCAAC CTGGGAACTG CATTCGAAAC TGGCAGGCTA GAGTCTTGTA \\
+\tt GAGGGGGGTA GAATTCCAGG TGTAGCGGTG AAATGCGTAG AGATCTGGAG GAATACCGGT \\
+\tt GGCGAAGGCG GCCCCCTGGA CAAAGACTGA CGCTCAGGTG CGAAAGCGTG GGG
+}
+
+\end{tabular}
+
+\caption{\textbf{Exact Sequence Variants of the isolated dominant species.}
+Sequence numbering as in table \ref{supp-tab:isolates}.}
+\label{supp-tab:esvs}
+
+\end{table}
+
+\subsection{Inferring species abundances from sequencing data}
 \label{supp-methods:species-from-seq}
@@
 that the element in the $(i,j)$ position, $q_{ij}$, represents the frequency of
-the $i$-th ESV when the $j$-th species is sequenced. $\mathbf{Q}$ can be used
-to determine the fraction of sequences of a given ESV that are obtained from
+the $i$-th ESV when the $j$-th species is sequenced (see table
+\ref{supp-tab:isolates} and section \nameref{supp-methods:data-proc:doms} of
+this Supplementary Material).
+$\mathbf{Q}$ can be used
+to determine the fraction of sequences of a given ESV that are generated when
 sequencing a community with any arbitrary composition. In other words,

### 2021-04-20 data/analyzeData.R
able to switch between modes in ESV-to-species mapping

@@ -27,6 +27,9 @@ if(!dir.exists('plots')) dir.create('plots')
 display_plots <- TRUE
 save_plots <- TRUE
 
+# general options
+deconvolute_ESVs <- FALSE # should species abundance be estimated from ESV abundance?
+
 
 ### ----------------------------------------------------------------------
 ### LOAD AND FORMAT DATA
@@ -784,12 +787,51 @@ dominants[['Citrate']] <- list(isolate_2 = c('Community3','Community5'),
                                isolate_8 = 'Community4',
                                other = c('Community6','Community8'))
 
+### Now we have an additional decision to make. Isolates 2, 4 and 6 share a same
+### ESV (even though isolates 4 and 6 carry additional copies of the 16S rRNA
+### gene mapping to different ESVs). We can use the matrix in
+###    > isolates[['seq']]
+### to deconvolute ESV abundances into species abundances in those samples
+### containing isolates 2, 4 and 6 (see isolates_abundance_from_ESVs.R).
+### However, this procedure is not standard and nothing similar seems to have
+### been described in the literature. The standard approach would be to just
+### consider isolates 2, 4 and 6 as indistinguishable, which would give us:
+
+# should we deconvolute ESV abundance into species abundance? if not...
+if (!deconvolute_ESVs) {
+  
+  # unify isolates 2, 4 and 6 in deconvolution matrix
+  isolates[['seq']] <- isolates[['seq']][,
+                                         ! colnames(isolates[['seq']]) %in% c('isolate_4','isolate_6')]
+  
+  # merge isolates groups
+  for (cs in carbon_sources) {
+    isolates[['groups']][[cs]][['isolate_2']] <- c(isolates[['groups']][[cs]][['isolate_2']],
+                                                   isolates[['groups']][[cs]][['isolate_4']],
+                                                   isolates[['groups']][[cs]][['isolate_6']])
+    isolates[['groups']][[cs]][['isolate_4']] <- NULL
+    isolates[['groups']][[cs]][['isolate_6']] <- NULL
+  }
+  
+  # merge dominants
+  dominants[['Glutamine']] <- list(isolate_1 = 'Community1',
+                                   isolate_2 = c('Community2','Community4','Community6'),
+                                   isolate_3 = 'Community3',
+                                   isolate_5 = c('Community7','Community8'),
+                                   other = 'Community5')
+  dominants[['Citrate']] <- list(isolate_2 = c('Community1','Community3','Community5'),
+                                 isolate_7 = c('Community2','Community7'),
[…]

### 2021-04-21 draft/draft.tex
draft skeleton

@@
-\synctex=1
 \documentclass[a4paper,10pt]{article}
 
-%% TO-DO: Achim and Schill in acknowl
-%% TO-DO: link to repo in paper and suppl mat
-%% TO-DO: my new grant
-%% TO-DO: give URLs to, e.g, Crooks reference: mybibwithurl?
-
-%% For tracking changes
-%% https://tex.stackexchange.com/questions/65453/track-changes-in-latex
-
 % packages
-
-% %% txfonts breaks amsmath.
-% %% http://www.texfaq.org/FAQ-alreadydef
-
-\usepackage{gitinfo2}
-
-\usepackage{savesym}
-\usepackage{amsmath}
-\savesymbol{iint}
-\usepackage{txfonts}
-\restoresymbol{TXF}{iint}
-
 \usepackage[utf8]{inputenc} % allows usage of spanish special characters
 \usepackage[spanish,english]{babel} % english dictionary for proper hyphenation
-%\usepackage{txfonts} % nice fonts
+\usepackage{amsmath} % math expressions
+\usepackage{upgreek} % upright greek letters for math
+\usepackage{txfonts} % nice fonts 
 \usepackage{authblk} % author affiliations
@@
 \usepackage{float} % image positioning
+\usepackage{multirow} % allows merging cells in tables
+\usepackage{makecell} % more table customization
+\usepackage{mathtools} % nice matrices
+\usepackage{rotating} % rotated text
+\usepackage{caption} % customization of figure/table captions
 \usepackage{subcaption} % collages of multiple images
-%%\usepackage[hidelinks]{hyperref} % hyperlinks
-\usepackage{hyperref}
-% \usepackage{cite} % contraction of references, use
-%                   % \usepackage[superscript]{cite} for superscript
-%                   % citations
-%% cite does not follow the usual pre and post citation
+\usepackage{hyperref} % hyperlinks
+\usepackage{nameref} % cross-references between draft and supplementary material
+\usepackage{zref-xr,zref-user} % more cross-references formatting
 \usepackage[square,numbers,sort&compress]{natbib}
-\usepackage{xcolor} % text color
+\usepackage[table,xcdraw]{xcolor} % text and table background colors
 \usepackage[margin=3cm]{geometry} % margins
+\usepackage[nottoc,numbib]{tocbibind} % add references to table of contents
 \usepackage{lineno} % line numbers
-\usepackage{csquotes} % paragraph quotes
-\usepackage{multirow} % joining rows in tables
-\usepackage{array}
-\usepackage[normalem]{ulem}
-\usepackage{todonotes}
-%%\usepackage{amsmath}: something breaks. What?
+
 \usepackage{pdflscape} %% Landscape
-\usepackage{ulem} % strike-through text
+\usepackage{dcolumn}
+\usepackage{booktabs}
+\usepackage{savesym}
+\savesymbol{iint}
+\restoresymbol{TXF}{iint}
+%% \usepackage{pdfpages} %% to include a pdf file
 
-%% cross references with suppl mat
-\usepackage{nameref}
-\usepackage{zref-xr,zref-user}
-\zxrsetup{toltxlabel}
-\zexternaldocument*[S-]{supp}
+% customization
+\setcounter{Maxaffil}{0} % affiliations
+\renewcommand\Affilfont{\itshape\small} % style of affiliations text
+\makeatletter\renewcommand{\@biblabel}[1]{#1.}\makeatother % change [number] for number. in reference list
+\addto{\captionsenglish}{\renewcommand{\bibname}{References}}
+\newcommand{\idea}[1]{\textcolor{red}{#1}}
+\newcommand{\mr}{\textcolor{red}{\textbf{[missing ref(s)]}}}
+\newcommand{\Burl}[1]{\url{#1}}
+\zxrsetup{toltxlabel} % cross references: draft & suppl mat
+\zexternaldocument*[M-]{draft}
 \hypersetup{
@@
   linkcolor = {blue},
-  filecolor = cyan %% controls color of external ref, if used
+  filecolor = cyan % controls color of external ref, if used
 }
-
-% customization
-\setcounter{Maxaffil}{0} % affiliations
-\renewcommand\Affilfont{\itshape\small} % style of affiliations text
-\captionsetup[figure]{font=small, labelfont={bf}, labelformat={default}, labelsep=period, name={Fig.}} % custom figure captions
-\makeatletter\renewcommand{\@biblabel}[1]{#1.}\makeatother % change [number] for number. in reference list
-\newcommand{\idea}[1]{\textcolor{red}{#1}}
-\newcommand{\mr}{\textcolor{red}{\textbf{[missing ref(s)]}}}
-\newcommand{\change}[2]{\textcolor{red}{\sout{#1}}\textcolor{blue}{#2}}
-\linenumbers % add line numbers to whole doc
+\renewcommand{\thefigure}{\textbf{S\arabic{figure}}} % preceed the Suppl Mat figures with S
+\renewcommand{\thetable}{\textbf{S\arabic{table}}} % preceed the Suppl Mat tables with S
+\renewcommand{\theequation}{S\arabic{equation}} % preceed the Suppl Mat equations with S
+\captionsetup[figure]{font=small, labelfont={bf}, labelformat={default},
+   labelsep=period, name={Fig.}} % custom figure captions
+\captionsetup[table]{font=small, labelfont={bf}, labelformat={default},
+   labelsep=period, name={Table}} % custom table captions
+\hypersetup{citecolor={blue}}
+\definecolor{lightgray}{gray}{0.9}
 
 % title, authors, affiliations
-\title{Conditional prediction of consecutive tumor evolution using cancer
-  progression models: What genotype comes next?}
-%\title{Conditional prediction of short-term tumor evolution using cancer progression models}
-%\title{Prediction of consecutive tumor evolution using cancer progression models}
-%\title{Conditional prediction of upcoming tumor evolution using cancer progression models}
-%\title{Partial prediction of tumor evolutionary trajectories using cancer progression models}
-%\title{Using cancer progression models to predict upcoming tumor evolution}
-\author[1,2,3]{Juan Diaz-Colunga}
-\author[1,2,$\dagger$]{Ramon Diaz-Uriarte}
-\affil[1]{Dpt. of Biochemistry, School of Medicine, Universidad Autónoma de Madrid, Madrid, Spain}
-\affil[2]{Instituto de Investigaciones Biomédicas `Alberto Sols'
-  (UAM-CSIC), Madrid, Spain}
-\affil[3]{Department of Ecology \& Evolutionary Biology and Microbial
-  Sciences Institute, Yale University, New Haven, CT, USA}
-%\affil[3]{Dpt. of Ecology and Evolutionary Biology, Yale University, New Haven, Connecticut, USA}
-%\affil[4]{Microbial Sciences Institute, Yale University, New Haven, Connecticut, USA}
-\affil[$\dagger$]{To whom correspondence should be addressed: \normalfont ramon.diaz@iib.uam.es}
-%\date{}
+\title{Provisional Title}
+\author[1,2,3]{Authors}
+\author[1,2,$\dagger$]{Álvaro Sánchez}
+\affil[1]{Department of Ecology \& Evolutionary Biology,
+Yale University, New Haven, CT, USA}
+\affil[2]{Microbial Sciences Institute,
+Yale University, New Haven, CT, USA}
+\affil[3]{Other affiliations...}
+\affil[$\dagger$]{To whom correspondence should be addressed: \normalfont alvaro.sanchez@yale.edu}
+\date{}
+
+\newcommand\legendtreefigs{Internal nodes show the partitioning variables   and the values that lead to each partition; light blue denotes the   values or categories that lead down the left split and salmon the values   or categories that lead down the right split (partitioning criteria are   also shown in the edges leading down a node).
+Leaves or terminal nodes   show (weighted) boxplots of the dependent variable for the observations   that belong that fall in that leave and, as label, the name of the node   ax_3,nd the sample size (sum of weighted observations and \% of the total   data set for the fit).
+The dot plots underneath the boxplots show the mean JS for all the methods for that leave.}
 
@@
 
-% \date{\gitAuthorDate\ {\footnotesize (Release\gitRels: Rev:
-%      \gitAbbrevHash)}}
+\linenumbers
 
@@
 
-  Accurate prediction of tumor progression is key for adaptive
-  therapy and precision medicine. Cancer progression models (CPMs) can be
-  used to infer dependencies in mutation accumulation from cross-sectional
-  data and provide predictions of tumor progression paths. But their
-  performance when predicting the complete evolutionary paths is limited
-  by violations of assumptions and the size of available data
-  sets. Instead of predicting full tumor progression paths, we can focus
-  on short-term predictions, more relevant for diagnostic and therapeutic
-  purposes. Here we examine if five distinct CPMs can be used to answer
-  the question ``Given that a genotype with $\mathrm{n}$ mutations has
-  been observed, what genotype with $\mathrm{n}+1$ mutations is next in
-  the path of tumor progression'' or, shortly, ``What genotype comes
-  next''. Using simulated data we find that under specific combinations of
-  genotype and fitness landscape characteristics CPMs can provide
-  predictions of short-term evolution that closely match the true
-  probabilities, and that some genotype characteristics (fitness and
-  probability of being a local fitness maximum) can be much more relevant
-  than global features. Thus, CPMs can provide short-term predictions even
-  when global, long-term predictions are not possible because fitness
-  landscape- and evolutionary model-specific assumptions are violated.
-  When good performance is possible, we observe significant variation in
-  the quality of predictions of different methods. Genotype-specific and
-  global fitness landscape characteristics are required to determine which
-  method provides best results in each case.  Application of these methods
-  to 25 cancer data sets shows that their use is hampered by lack of the
-  information needed to make principled decisions about method choice and
-  what predictions to trust. Fruitful use of these methods for short-term
-  predictions requires adapting method's use to local genotype
-  characteristics and obtaining reliable indicators of performance; it
-  will also be necessary to clarify the interpretation of the method's results
-  when key assumptions do not hold.
+The abstract goes here.
 
@@
 
-%% bioniformatics:  A maximum of 150 words is recommended. https://academic.oup.com/bioinformatics/pages/instructions_for_authors
-%% PLoS Comp Biol: I think around 300
+\section{Introduction}\label{intro}
 
-%% Length of ms: For bioinformatics: Original papers: up to 7 pages, this
-%% is approx. 5000 words, not including figures,
-%%
-%% Do as pdftotext draft.pdf - | wc -w
-%% Maybe rm refs and figures before word counting
-%% Place \end{document} before refs; uncomment it.
-
-%% As of now, this is twice as long as it should be for Bioinformatics:
-%% 9400 words withouth figures.
-
-\section{Introduction}\label{sec:intro}
-
-Accurately predicting tumor evolution would allow for better prognosis and
-treatment. Unfortunately, the stochastic components in key factors that
-govern tumor progression (mutation, genetic drift or clonal selection
-\cite{greaves_evolutionary_2015,lipinski_cancer_2016,williams2018}, as
-well as non-genetic variability \cite{Brock2009} and the tumor
-microenvironment \cite{Albini2007}) fundamentally limit
-predictability. This is also common to other problems
-\cite{lassig_predicting_2017, losos2018} such as the evolution of
-antibiotic resistance \cite{toprak2012} or virulence \cite{Day2004}, where
-even small increases in predictive power could be of critical importance
-for the design of treatment strategies \cite{palmer2013}.
-
-% what are cancer progression models?
-Cross-sectional high-throughput data are widely available for many cancer
-types, and Cancer Progression Models (CPMs) \cite{Beerenwinkel2014,
-  beerenwinkel_computational_2016}, originally developed to identify
-restrictions in the order of accumulation of mutations in tumors
-using cross-sectional data, could thus be key tools for predicting cancer
-progression. CPMs (implicitly) encode all the possible mutational
-trajectories that are compatible with the restrictions and, for some
-methods, provide probabilities of different paths (figure
-\ref{fig:fig1}A-B). However their performance to infer the complete
-evolutionary trajectories, from the root genotype with no driver mutations
-to a final, fixated genotype, is limited
-\cite{diaz-uriarte2019a,hosseini2019a}; this has been attributed mainly to
-violations of the underlying assumptions of the models (fitness landscapes
-with a single fitness maximum in the genotype with all loci mutated,
-strong selection week mutation regime \cite{gillespie1983}) and the
-difficulty to acquire
-large, unbiased data sets to power them \cite{diaz-uriarte2019a}.
-
-Predicting the complete evolutionary path, however, might not be the most
-relevant objective. Detection of a specific tumor composition at a given
-point in time automatically discards all mutational pathways that do not
-visit the observed state, reducing the amount of possible evolutionary
-trajectories that can be followed from then on and thus potentially
-facilitating the task of predicting further tumor progression. In this
-work we address the question of how accurately CPMs can predict short-term
-cancer evolution after a genotype is detected: what is the state that
-comes consecutively after or ``what genotype comes next''. Even if the
-predictability of the complete path is low because of violations of method
-assumptions, genotype-specific features could lead to local improvements
-in short-term predictability \cite{bank_predictability_2016,
-  visser2018,ferretti2018}. Moreover, in contrast to the more general,
-non-local descriptors, these genotype-specific factors can provide more
-detailed insight regarding the requirements for the methods to be
-reliable. In fact, conditional predictions on the observed genotype are
-arguably much more relevant than complete evolutionary path predictions
-for both adaptive therapy \cite[e.g.,][]{melnikov2020,hansen2020a,
-  stankova2019a} and for SMART (sequential, multiple assignment,
-randomized trial) designs in dynamic treatment regimes for personalized
-medicine \cite{tsiatis2020,chakraborty2013}.
-
-[Fig.~\ref{fig:fig1} about here]
-
-% Rather than analyzing full evolutionary
-% trajectories, as has been done in other work
-% \cite{diaz-uriarte2019a,hosseini2019a}, we focus here on the question
-% ``what genotype comes next'', the partial paths from an observed tumor
-% state to the state that comes consecutively after.
-
-% This might, on the one hand, lead to improvements in predictive power,
-% even when the predictability of the complete path is low because of
-% violations of method assumptions: genotype-specific features could lead to
-% local improvements in short-term predictability
-% \cite{bank_predictability_2016, visser2018,ferretti2018}. On the other
-% hand, conditional predictions on the observed genotype are much more
-% relevant clinically and for treatment purposes in adaptive therapy
-% \cite[e.g.,][]{melnikov2020,hansen2020a, stankova2019a} and for SMART
-% (sequential, multiple assignment, randomized trial) designs in dynamic
-% treatment regimes for personalized medicine
-% \cite{tsiatis2020,chakraborty2013}.
-
-% , accurately predicting the next genotype could be among
-% the tools for the physician to become, in a game theory context, the
-% ``rational leader'' that anticipates and steers to limit cancer's
-% resistance strategies \cite{stankova2019a}; moreover, observed genotypes
-% and predicted next genotypes could be of use in SMART (sequential,
-% multiple assignment, randomized trial) designs in dynamic treatment
-% regimes for personalized medicine \cite{tsiatis2020,chakraborty2013}.
-
-% \cite[prenote][posnote]{hansen2020a}
-
-% accurately predicting the
-% next genotype would play a key role in adaptive therapy
-% \cite[e.g.,][]{melnikov2020, hansen2020a,stankova2019a}, as ``(...) the
-% ability to anticipate subsequent cancer cell responses, provides a
-% critical opportunity to obtain more favorable outcomes by steering and/or
-% limiting the cancer cell’s resistance strategies'' \cite[]
-
-% are used in adaptive therapy \cite{whatever}
-
-% form basis of SMART designs in individualized therapy
-% could be used in SMART designs \cite{whatever---los dos libros}
-% \cite{tsiatis2020,chakraborty2013}
-% (yo: check los dos zotero dirs, and look for conditional on genotypes)
-
-% \cite{melnikov2020}: inhibiting growth in bacteria
-
-% \cite{hansen2020a}: adaptive therapy and cpompetitive supression: gral ref
-
-% \cite{gluzman2020}: conditional treatment
-
-% \cite{west2019}
-
-% Motivated by the idea that information on the prospective advance of the
-% disease, even if limited, can be of critical importance for the design of
-% treatment strategies and can improve therapeutic outcomes
-% \cite[p.~243i]{palmer2013}
-
-% Information on the prospective advance of the disease, even if limited,
-% can be of critical importance for the design of treatment strategies
-% \idea{references, as in paper with Claudia}. \todo{YES! DO!}
-% \idea{[Relate predicting next genotype to adaptive therapy ---tarea para
-%   Ramon]}.
-
-% Tumor progression is governed by a combination of factors such as mutation, genetic drift or clonal selection
-% \cite{greaves_evolutionary_20152015,lipinski_cancer_2016,williams2018}, as well as non-genetic
-% variability \cite{Brock2009} and the tumor microenvironment
-% \cite{Albini2007}. These have important stochastic components that
-% fundamentally limit predictability. The ability
-% to understand and predict evolutionary dynamics in cancer would allow for
-% better prognosis and treatment. Investigating the mutational profiles of
-% evolving clones within a tumor is possible by means of high-throughput
-% genome sequencing, but the use of accurate models of tumor evolution is
-% still required to unveil the true paths of cancer progression
-% \cite{McPherson2017}. This is not of exclusive relevance to cancer but
-% also to problems such as the evolution of antibiotic resistance
-% \cite{toprak2012} or virulence \cite{Day2004}.
-
-% % what are cancer progression models?
-% Cancer Progression Models (CPMs) identify restrictions in the order of
-% accumulation of mutations within a tumor using cross-sectional data. They
-% encode all the possible mutational trajectories that are compatible with
-% the restrictions (figure \ref{fig:fig1}A-B), which can serve to narrow the
-% search for the true distribution of tumor progression pathways. However
-% their performance to infer these complete evolutionary trajectories, from
-% the root genotype with no driver mutations to a final, fixated genotype,
-% is limited \cite{diaz-uriarte2019a,hosseini2019a}; this has been
-% attributed to various factors, such as violations of the underlying
-% assumptions of the models and the difficulty to acquire large, unbiased
-% data sets to power them \cite{diaz-uriarte2019a}.
-
-% what are we studying?
-% In practice, cancer can only be first detected once it has already
-% advanced to some extent.
-
-% Detection of a specific tumor composition at a given point in time
-% automatically discards all mutational pathways that do not visit the
-% observed state, reducing the amount of possible evolutionary trajectories
-% that can be followed from then on and thus potentially facilitating the
-% task of predicting further tumor progression.\todo{is it here where we
-%   explain how this differs from work with Claudia?} In this work we
-% address the question of how accurately CPMs can predict short-term cancer
-% evolution after detection. Rather than analyzing full evolutionary
-% trajectories, we focus on the partial paths from an observed tumor state
-% to the state that comes consecutively after.
-
-The question of what genotype comes next, or what is the state that comes
-consecutively after, requires that we carefully define the meaning of
-``consecutive'' states in this context. If cancer progresses under
-conditions of strong selection and weak mutation (SSWM)
-\cite{gillespie1983}, only evolutionary trajectories of increasing fitness
-are possible, and the evolution of a new, higher-fitness clone results in
-rapid expansion and effective exclusion of previous ones before any new
-mutations take place. In this situation, there is a well defined sequence
-of genotypes with progressively increasing fitness in the path to tumor
-fixation, with the most abundant clone at every time being a direct
-descendant of a previously most abundant one itself (figure
-\ref{fig:fig1}C). But, in cancer, large population sizes and high mutation
-rates can prevent successful clonal sweeps \cite{nichol2019a, zhao2016} so
-that instead of SSWM, clonal interference and stochastic tunneling
-\cite{devisser_empirical_2014,szendro_predictability_2013,sniegowski_beneficial_2010,
-  wodarz_dynamics_2014} become common: multiple clones can coexist at
-significant fractions for prolonged periods of time, and new
-higher-fitness clones can evolve from low fitness parents that represented
-a minority of the population
-(Fig~\ref{fig:fig1}D). % As a consequence, genotypes with a large
-% number of mutations %high mutational load
-% can already be present at the time of sampling yet go undetected if their
-% relative abundance is small (figure
-% \ref{fig:fig1}D). % Deviations from SSWM
-% are known to undermine the predictive power of CPMs
-% \cite{diaz-uriarte2018,diaz-uriarte2019a}, as the model underlying most
-% CPMs seems to assume SSWM (see Discussion). This assumption, however, is
-% not always stated
-% explicitly.
-% , but it is explicit, for example, in Hosseini et al.'s
-% \cite{hosseini2019a} use of CPMs for predicting tumor evolution (see
-% \nameref{discussion}).
-
-To formulate the question of ``what genotype comes next'' 
-%To accurately raise a question that does not rely on SSWM to be well
-%defined,
-so that it does not depend on the SSWM assumption we designate an
-\textit{observed} genotype as that which is most abundant in the tumor at
-the time of sampling. On the other hand, a Line of Descent (LOD) is
-defined as a sequence of parent-child genotypes from the wild-type to the
-(possibly local) 
-fitness maximum where fixation occurs \cite{szendro_predictability_2013}.
-By construction, assuming that mutations accumulate one by one and cannot
-be lost, each genotype in the LOD has all the mutations of the previous
-one plus exactly one more (as it directly descends from it). The LOD has
-no more than one genotype with a given number of mutations, whereas in
-general two or more observable genotypes could share the same mutation
-count (e.g. in figure \ref{fig:fig1}D both the red and green genotypes
-have one mutated driver and are observable). If the evolutionary process
-does not conform to SSWM, not all genotypes in the LOD are necessarily
-observable. From a clinical perspective, it is relevant to use the
-information that is detectable to unveil the underlying dynamics of cancer
-evolution, encoded in the LOD. In other words, we want to make a
-prediction regarding \textit{what comes next in the LOD} based on the
-\textit{observation} of a genotype. We thus formulate the following
-general question: \textit{Given that a genotype with $\mathrm{n}$
-  mutations has been observed, what genotype with $\mathrm{n}+1$ mutations
-  is in the LOD?} Formulated like this, the question is unambiguous and
-determined. Under SSWM conditions it can be expressed simply as:
-\textit{given that a certain genotype has been observed, what genotype
-  comes next in the path to tumor fixation?} However, it is important to
-emphasize that even if there is no SSWM the more general question remains
-well defined.
-
-% how do we do it?
-
-To address this question, we use a collection of evolutionary simulations
-from which we obtain the true paths of tumor progression. For every
-simulated process we keep track of % what genotypes represent the majority
-% of the population as the tumor progressed (i.e., we list 
-the most frequent genotypes. Then for each one of them we ask what
-genotype has exactly one more mutation while also being a direct ancestor
-of the ultimately fixated clone. Averaging over all simulations we obtain
-a set of frequencies which represent the conditional probabilities that we
-are interested in. We then examine five CPMs: Mutual Hazard Networks (MHN)
-\cite{schill2020}, Conjunctive Bayesian Networks (CBN)
-\cite{Gerstung2009,Gerstung2011,montazeri_large-scale_2016}, Oncogenetic
-Trees (OT) \cite{Desper1999JCB,Szabo2008}, CAncer PRogression Inference
-(CAPRI) \cite{capri_bioinformatics,capri_pnas} and CAncer PRogression
-Extraction with Single Edges (CAPRESE) \cite{caprese_2014}, some of which
-have more than one variant (giving a total of 13 different methods). We
-provide these methods with cross-sectional samples taken from the
-evolutionary simulations and compare their predictions with the true
-probabilities. We explore the factors that affect performance, focusing on
-those factors that depend on the properties of the observed genotype. We
-want to understand whether CPMs can be used to predict short-term
-evolution; for this question, the key response variable is the best
-possible performance over all CPMs considered. Although we do address
-method choice as a secondary question, this is not our main objective 
-\textit{per se}. We discuss the consequences of our results for the analysis of
-cancer data by analyzing 25 cancer data sets.
-
-% This work differs from earlier work \cite{diaz-uriarte2019a,hosseini2019a}
-% in that we examine a completely different question: previous work dealt
-% with predicting full evolutionary trajectories, but here we examine the
-% short-term prediction question \textit{Given that a genotype with
-%   $\mathrm{n}$ mutations has been observed, what genotype with
-%   $\mathrm{n}+1$ mutations is in the LOD?}, which can colloquially be
-% summarized as ``What genotype comes next''. In addition, we add a recent
-% new method, Mutual Hazard Networks (MHN) \cite{schill2020}, which is very
-% different from the rest of the CPMs considered here. Moreover, we are
-% examining genotype-specific factors in addition to global fitness
-% landscape characteristics: in contrast to the more general, non-local
-% descriptors that may not apply to the totality of the fitness
-% landscape/evolutionary process, these genotype-specific factors can
-% provide more detailed insight regarding the requirements for the methods
-% to be reliable in what are arguably the more relevant questions for
-% individualized and precision medicine.
-
-% Why does this matter? Predicting important. In fact,
-
-% experimental evolution studies in bacteria and yeast have allowed to
-% partially disentangle the role of population size, selection, and
-% mutational heterogeneity in parallel evolution \cite{bailey2017} and,
-% thus, indirectly on the issues addressed here; using these results for
-% cancer, however, would require much more detailed information (e.g.,
-% effective population sizes) than is often available.
-
-% So, what can we do with experimental data from cancer? This we examine in
-% the discussion blablaba
-
-% \idea{We probably need to make it explicit how it differs from previous
-%   work, in particular Diaz-Uriarte and Vasallo.  If we end up doing what I
-% think about comparing JS of methods and using that for real data sets, an
-% additional comment is ``and we provide specific, practical advice that can
-% be followed with the output of methods applied on real data'' ---or
-% something of that sort}.
-% \idea{Relate to this: Detection of a specific tumor composition at a given point in
-% time automatically discards all mutational pathways that do not visit the
-% observed state, reducing the amount of possible evolutionary trajectories
-% that can be followed from then on and thus potentially facilitating the
-% task of predicting further tumor progression. And relate to Bank}
-
-% \idea{Say something about trying to identify, from method output itself,
-%   when good performance}
-
-% \idea{We use MHN. First to review that method. We use CPMs, but also
-%   include MHN, a newly developed approach that blablab}
-
-% \idea{This is KEY: we do what really matters: next genotype. It really
-%   matters because: a) global predictions suck and are generally doomed
-%   except maybe to estimate predictability; b) global predictions can
-%   not benefit from local features; c) newxt is what matters for therapy
-%   and adaptive treatment}.
-
-\section{Methods} \label{methods}
-
-\subsection{Overview}
-\label{sec:overview}
-
-To assess if we can use CPMs for short-term predictions of evolutionary
-processes, we have compared predicted and true answers to the question
-\textit{Given that a genotype with $\mathrm{n}$ mutations has been
-  observed, what genotype with $\mathrm{n}+1$ mutations is in the LOD?}
-under an extensive set of simulated scenarios. First, we simulated a large
-number of tumor evolution trajectories under scenarios with varying
-departures from the assumptions of CPM methods; next, we obtained a large
-number of simulated tumor evolution trajectories.  Next, we emulated
-cross-sectional sampling of the simulations, with varying sampling schemes
-and sample sizes. These steps are described in section
-\nameref{methods:simulsandsampling}.  The transition probabilities that
-provide the true answer to the question \textit{Given that a genotype with
-  $\mathrm{n}$ mutations has been observed, what genotype with
-  $\mathrm{n}+1$ mutations is in the LOD?}  is explained in section
-\nameref{sec:methods:sim}. To obtain the CPM predictions we analyzed each
-of the sampled data sets with a total of 13 different CPM variants and
-obtained the predictions as explained in section
-\nameref{methods:cpm}. Next, we measured how close the predictions were to
-the truth (section \nameref{sec:js}) and, finally, we analyzed the major
-factors that affect the similarity between predictions and truth (section
-\nameref{sec:linear-mixed-model}).  To examine the implications of our
-results for the analyses of cancer data, we have analyzed 25 cancer data
-sets, described in section \nameref{sec:datasets}.
-
-\subsection{Simulated evolutionary processes and sampling} \label{methods:simulsandsampling}
-
-Simulated data were taken from \cite{diaz-uriarte2019a}, where complete
-details are provided. In summary, simulations were conducted under
-evolutionary scenarios that differed in the number of genes, the type of
-fitness landscape, the initial population size, and the mutation
-rates. Landscapes were of either 7 or 10 genes. Fitness landscapes were of
-three types: representable, local maxima or Rough Mount Fuji (RMF). The
-three types of fitness landscapes incorporate increasing departures from
-the assumptions of most CPMs: for the representable fitness landscapes a
-DAG of restrictions exists with the same accessible genotypes and
-accessible mutational paths; for local maxima fitness landscapes the set
-of accessible genotypes can be represented by a DAG of restrictions, but
-there are local fitness maxima and the fitness graph has missing paths
-\cite{diaz-uriarte2019a}; the genotype with all genes mutated might or
-might not be the genotype with largest fitness. The RMF fitness landscapes
-\cite{neidhart_adaptation_2014, devisser_empirical_2014,fragata2019}
-usually have multiple local fitness maxima and considerable reciprocal
-sign epistasis so not even the set of accessible genotypes can be
-represented by a DAG of restrictions. On the above fitness landscapes
-evolutionary simulations were run. Three different populations sizes
-($2\times10^3$, $5\times10^4$ or $1\times10^6$) and two mutation rate
-regimes (mutation rate fixed at $10^{-5}$ or uniformly distributed in the
-log scale between $0.2\times10^{-5}$ and $5\times10^{-5}$) were used to
-generate variability with respect to SSWM. For each combination of the
-above factors, 35 independent replicates were obtained for a total of 1260
-landscapes (35 replicates $\times$ 2 number of genes $\times$ 3 types of
-fitness landscapes $\times$ 3 initial population sizes $\times$ 2 mutation
-rate regimes). Then, for each one of them, 20000 independent evolutionary
-processes were simulated until fixation of one of the genotypes at a
-fitness maximum (local or global). The complete history of the simulations
-was stored so that the true LOD for each simulation was known. Each set of
-simulations was then sampled under three detection regimes: with equal
-probability with respect to the size distribution (\textit{uniform}
-sampling), sampling predominantly early in the tumor progression process
-(i.e., enriched in \textit{small} tumors), sampling predominantly late in
-the tumor progression process (i.e., enriched in \textit{large}
-tumors). When \textit{sampling}, we identified the most abundant genotype
-and its mutated loci at the chosen time, so the sampling process is
-consistent with how we defined \textit{observable} genotypes earlier. No
-sampling error is considered. The full sample consists of three binary
-matrices of observations (one per detection regime), each with 20000 rows
-(one observation per simulation) and as many columns as driver genes. Each
-element of the matrix is either 1 or 0 (true or false) depending on which
-genes were mutated in which observations. For each detection regime, the
-complete sample was used to generate five smaller, non-overlapping splits
-of size 50, 200 or 4000. This produced a total of 56700 data sets (1260
-fitness landscapes $\times$ 3 detection regimes $\times$ 3 sample sizes
-$\times$ 5 splits).
-
-\subsection{Transition probabilities from evolutionary simulations}
-\label{sec:methods:sim}
-
-Let us consider a matrix $\mathbf{P}$ where rows correspond to observable
-genotypes and columns to genotypes of the LOD. The element $(i,j)$ of the
-matrix, denoted as $p_{ij}$, represents the conditional probability that
-the genotype $j$ (with $n+1$ mutations) is an element of the LOD given the
-genotype $i$ (with $n$ mutations) has been observed. In the absence of
-back mutations and when mutations accumulate one by one, $p_{ij}$ is
-defined only if $i$ and $j$ are exactly one mutation away from each other.
-Note that the genotype $j$ (element of the LOD and with $n+1$ mutations)
-need not be a direct descendant of $i$ if SSWM does not hold. We built a
-transition matrix $\mathbf{P}$ for every fitness landscape that we
-generated by examining each of the 20000 evolutionary processes that were
-simulated in the corresponding landscape, listing all observable genotypes
-(i.e. those that represented the majority of the population at some point
-during the simulated evolutionary process) and extracting the LOD of every
-process tracking back the ancestors of the fixated genotype (definition
-and details in \cite{szendro_predictability_2013}; using implementation in
-\cite{diaz-uriarte2017}).  Each element $p_{ij}$ can thus be interpreted
-as the fraction of the times that $j$ was in the LOD given that an
-observation of $i$ had been made. See further details in Supplementary
-Material, section \ref{S-supp-cpm-details}.
-
-% \subsubsection{Special cases}\label{sec:special}
-
-% When the end of the evolutionary process is reached there are no further
-% transitions to genotypes of the LOD. Thus, matrix $\mathbf{P}$, which
-% provides the true probability distributions for the question \textit{what
-%   genotype of the LOD has $\mathrm{n}+1$ mutations iven that a genotype
-%   with n mutations has been observed?} needs to add to additional states
-% that correspond to two distinct special cases. When the observed genotype
-% is the final, fixated one, we have simply reached the end of the
-% evolutionary process; we handle this by adding to the $\mathbf{P}$ an
-% ``end" situation (denoted as $p_{i,end}$). If SSWM conditions are not
-% satisfied, however, observable genotypes need not be a part of the LOD so
-% that there could be observable genotypes with more mutated loci than, or
-% as many mutated loci as, the final genotype that gets fixated. In other
-% words, it is possible that a clone with $m \geq n$ mutations establishes
-% in the tumor for a period of time even if eventually another clone with
-% $n$ mutations is the one that fixates.  We handle this by adding to the
-% $\mathbf{P}$ a ``none" situation ($p_{i,none}$). Downstream analyses are
-% performed including these two probabilities ($p_{i,end}, p_{i, none}$)
-% (and row normalization was actually carried out after including $end$ and
-% $none$).
-
-% % \subsection{Transition probabilities from evolutionary simulations}
-% % \label{methods:sim}
-
-% % Let us consider a matrix $\mathbf{P}$ where rows correspond to observable
-% % genotypes and columns to genotypes of the LOD. The element $(i,j)$ of the
-% % matrix, denoted as $p_{ij}$, represents the conditional probability that
-% % the genotype $j$ (with $n+1$ mutations) is an element of the LOD given the
-% % genotype $i$ (with $n$ mutations) has been observed. Naturally, $p_{ij}$
-% % is defined only if $i$ and $j$ are exactly one mutation away from each
-% % other. Note that, given how our question is formulated, all the mutations
-% % in $i$ need to also be present in $j$ only in SSWM, when $i$ would
-% % necessarily be the parent of $j$. But in a more general scenario (e.g.,
-% % when SSWM does not hold) it is possible that a genotype $i$ with $n$
-% % mutations is observed despite not belonging to the LOD. Therefore the
-% % genotype $j$ (element of the LOD and with $n+1$ mutations) need not be a
-% % direct descendant of $i$. The only condition for $p_{ij}$ to be defined is
-% % that $j$ has exactly one more mutation than $i$:
-
-% % \begin{equation}
-% % p_{ij} \in \mathbb{R} \iff n_{mut}(j) = n_{mut}(i) + 1
-% % \label{eq:pij}
-% % \end{equation} 
-% % \idea{En realidad, no es $\mathbb{R}$; es $[0, 1]$, puesto que podemos
-% %   definirlo como 0. }
-
-% % We built a matrix $\mathbf{P}$ for every fitness landscape that we
-% % generated. To obtain the elements $p_{ij}$ of one of them, we ran through
-% % all 20000 evolutionary processes that were simulated in the corresponding
-% % landscape. In each case, we first listed all observable genotypes
-% % (i.e. those that represented the majority of the population at some point
-% % during the simulated evolutionary process). Then we extracted the LOD of
-% % every process tracking back the ancestors of the fixated
-% % genotype (see definition and details in \cite{szendro_predictability_2013}; using implementation in \cite{diaz-uriarte2017}).
-% % We identified the genotypes in the LOD with one mutation more than the
-% % observable ones, and added 1 to the corresponding entries of
-% % $\mathbf{P}$. Finally, we row-normalized $\mathbf{P}$ to meet the
-% % normalization condition $\sum_{j} p_{ij} = 1$.
-
-% % % \begin{equation}
-% % % \sum_{j} p_{ij} = 1
-% % % \label{eq:pij-norm}
-% % % \end{equation}\todo{Added this equation, OK?}
-% % %
-% % Each element $p_{ij}$ can thus be interpreted as the fraction of the times that $j$ was in the LOD given that an observation of $i$ had been made.
-For a large number of simulations, this fraction represents a good approximation of the true probabilities.
-
-% % \subsubsection{Special cases}\label{sec:special}
-
-% % If SSWM conditions are not satisfied, observable genotypes need not be a part of the LOD and not all genotypes in the LOD are necessarily observable.
-A particular case of this are observable genotypes with more mutated loci (or just as many) than the final one that gets fixated: it is possible that a clone with $m \geq n$ mutations establishes in the tumor for a period of time even if eventually another clone with $n$ mutations is the one that fixates.
-This represents a special scenario for our purpose: the answer to out question (\textit{what genotype of the LOD has $\mathrm{n}+1$ mutations?}) is technically undefined if the LOD stops at $n$ mutations.
-We identify two scenarios: \textit{a)} the observed genotype is simply the final, fixated one, and \textit{b)} the observed genotype has transiently established as the most abundant, but fixation will occur at a different one (with less mutations) evolving through a different pathway.
-% % \iffalse
-
-% % \begin{itemize}
-% % \item[\textit{a)}] The observed genotype is simply the final, fixated one.
-% % \item[\textit{b)}] The observed genotype has transiently established as the most abundant, but fixation will occur at a different one (with less mutations) evolving through a different pathway.
-% % \end{itemize}
-% % %
-% % \fi
-% % We refer to \textit{a)} as an ``end" situation (we have simply reached the end of the evolutionary process), and to \textit{b)} as a ``none" situation (the only possible answer to our question is ``there is no such a genotype").
-In practice, we handle these cases by attaching two additional columns to $\mathbf{P}$.
-These contain the probabilities that an observation of a genotype $i$ is made in an ``end" situation (denoted as $p_{i,end}$) or a ``none" situation ($p_{i,none}$).
-Downstream analyses are performed including these probabilities.
-
-% % \idea{[Move all this chunk (or even almost everything after \ref{eq:pij})
-% %   to the Supplementary? I'm worried it may break the flow of the
-% %   paper. Would it make sense to simply write a super-simplified methods
-% %   with just enough information to follow the reasoning, and move
-% %   everything else to the supplementary?]}\todo{something needs to be left
-% %   here, maybe all. Not sure yet how much, though}
-
-\subsection{Transition probabilities from CPMs} \label{methods:cpm}
-
-%  \idea{Explain how we go from the output of a CPM (DAG of restrictions,
-%    but not in the time-discretized cases?)}\todo{in Diaz-Uriarte and
-%    Vasallo (except we stop at the transition matrix between genotyps stage
-%    ---I'll write it)}
-% \idea{Notation: matrix of predictions from a CPM should be named
-%   $\mathbf{\widehat{P}}$, and its elements $\hat{p}_{ij}$. The hat
-%   indicates an estimator, which is precisely what CPMs provide: an
-%   estimation of the true probabilities $p_{ij}$.}
-
-General overviews of all the methods used here, except for MHN, are
-available in \cite{beerenwinkel_computational_2016, Beerenwinkel2014,
-  diaz-uriarte2019a,diaz-uriarte2018} and detailed descriptions are given
-in the original references for each procedure (MHN \cite{schill2020}, CBN
-\cite{Gerstung2009, Gerstung2011}, MCCBN
-\cite{montazeri_large-scale_2016}, OT \cite{Szabo2008, Desper1999JCB},
-CAPRI \cite{capri_bioinformatics, capri_pnas}, CAPRESE
-\cite{caprese_2014}). Briefly, CBN, MCCBN, OT, CAPRESE, and CAPRI try to
-identify restrictions in the order of accumulation of mutations from
-cross-sectional data. CAPRESE and OT code these restrictions using trees
-(i.e., under these two methods each event, such as a mutation, can
-immediately depend on at most one other event); although the output of OT
-and CAPRESE is often similar, the procedures are different (in OT weights
-along edges can be directly interpreted as probabilities of transition
-\cite{Szabo2008}, whereas CAPRESE uses a probability raising notion of
-causation based on Suppes' probabilistic causation). CBN, MCCBN, and CAPRI
-code these restrictions using DAGs (i.e., each event can depend on
-multiple other events).  CBN and MCCBN are very closely related methods:
-the CBN version we use is H-CBN as described in \cite{Gerstung2009,
-  Gerstung2011}, and MCCBN is an implementation of the CBN model that uses
-a Monte Carlo Expectation-Maximization (EM) algorithm for fitting the CBN
-model \cite{montazeri_large-scale_2016} (instead of the simulated
-annealing nested within EM of H-CBN); in contrast, CAPRI tries to identify
-probability raising in the framework of Suppes' probabilistic
-causation. Common to most of these methods is a model of deterministic
-dependencies \cite{schill2020}, a model of accumulation of mutations where
-an event (a mutation) can only occur if all its dependencies are satisfied
-(though different methods allow for small error deviations from this
-requirement); CAPRI and CAPRESE try to recover ``probability raising''
-relations and, strictly, their trees/DAGs do not encode deterministic
-dependencies but will be treated as doing so in this paper, since
-obtaining probabilities of transition from them is not possible otherwise
----see details in section \nameref{S-supp-trans-prob} in the Supplementary
-Material.
-
-For all of these methods, since a DAG of restrictions determines a fitness
-graph (see Fig.~\ref{fig:fig1} A-B and
-\cite{diaz-uriarte2018,diaz-uriarte2019a}), we can obtain the set of
-possible descendants for a genotype directly from this fitness graph.  For
-CAPRESE and CAPRI all transitions were set as equiprobable (as
-probabilities of transition are not available) but for CBN, MCCBN, and OT
-we can compute the probabilities of transition between genotypes. CBN and
-MCCBN return the parameters of the waiting time to occurrence of each
-mutation given its restrictions are satisfied which allows us to compute
-transition probabilities between genotypes using competing exponentials
-(see \cite{montazeri_large-scale_2016, hosseini2019a, diaz-uriarte2019a})
-and a similar procedure can be used with OT by using the edge weights
-(though, strictly, the OT model, being an untimed oncogenetic tree does
-not return these probabilities ---see \cite{diaz-uriarte2019a}). Mutual
-Hazard Networks (MHN) \cite{schill2020} differ from the previous methods
-in that events are modeled by a spontaneous rate of fixation and a
-multiplicative effect each of these events can have on other events (i.e.,
-pairwise interactions), which allows it to model both enabling and
-inhibiting dependencies. The output of MHN is a transition rate matrix; as
-for CBN and MCCBN, we can obtain the probability of transition to each
-descendant genotype, given a transition, using competing exponentials from
-the transition rate matrix (see equation 2 and Fig.~2 in
-\cite{schill2020}). Details on software and computation of transition
-probabilities are provided in the Supplementary Material, sections
-\nameref{S-supp-cpm-details} and \nameref{S-supp-trans-prob}.
-
-% \subsubsection{Time-discretized CPMs}\label{sec:td}
-
-None of the CPM models considered here allow us to accommodate local
-maxima: under all of the models, the genotype with all genes mutated is
-always reached with probability 1 as time goes to infinity (this is also
-the case for MHN, even if it can include inhibiting effects). To overcome
-this limitation without making specific additional assumptions about the
-sampling process, and for MHN, CBN, and MCCBN, we have used the
-uniformization method \cite{grassmann1977,vandijk2018} (see also
-\cite[Fig.\ 6]{schill2020}) to approximate the continuous-time Markov
-chain ($\mathbf{Q}$) by a discrete time Markov chain 
-(with transition matrix $= \mathbf{I} + \frac{1}{\gamma}\mathbf{Q}$, where
-$\gamma = \max(|q_{ii}|)$). We will interpret the diagonal entries of the
-time-discretized transition matrix as a lower bound on the probability
-that the given genotypes behave as local maxima. The output from CBN, MHN,
-and MCCBN after time discretization will be referred to as CBN\_td,
-MHN\_td, and MCCBN\_td, respectively, and we will differentiate this
-\textbf{time-discretized} results from those obtained, as explained above,
-using competing exponentials. We will refer to these methods as the
-\textbf{TD} methods (for time-discretized) whereas their counterparts,
-CBN, MHN, MCCBN, obtained via competing exponentials on the rate
-matrix, woll be referred below as the CE methods. OT, CAPRESE, and CAPRI do not
-belong to any of those two groups, though none of them can accommodate
-local maxima either; thus, when we refer to the ``non-TD'' methods, that
-includes MHN, CBN, MCCBN (in their ``non-TD'' versions) as well as OT,
-CAPRESE, and CAPRI.
-%%CEr
-%(``CE'' in Figure \ref{fig_biol_comp}).
-
-% \idea{We should justify that much better. I don't see how. And first
-%   sentence is the same as saying that the limiting distribution is
-%   $\pi_{gentoypeWithAllGenesMutated} = 1$. Not sure that clarifies things
-% for most readers.}
-
-In summary, a total of 13 methods have been used: CBN, in three variants
-(CBN, with transition probabilities using competing exponentials
-conditional on transition; CBN\_uw, obtained from the former by using
-equiprobable transitions; CBN\_td, the time-discretized version of CBN);
-MCCBN, again in three variants (MCCBN; MCCBN\_uw; MCCBN\_td);
-MHN in two variants (MHN and MHN\_td); OT in variants OT and OT\_uw
-(equiprobable transitions); CAPRESE; CAPRI, in two variants, CAPRI\_AIC,
-CAPRI\_BIC (which differ on the penalty used, AIC or BIC). Each of these
-methods returns a matrix $\mathbf{\hat{P}}$ of predicted transition
-probabilities.
-
-% \idea{MHN: give a few details of how we used it. Basically, default with
-%   parameters for balbla}\todo{explain something about the All others too?
-%   CBN, OT, etc? It will
-% be the fourth time I do it! Or refer to previous papers and just add a
-% sentence for MHN?
--- I would definitely just add the references, maybe a short general explanation of "traditional" CPMs and then add a couple more sentences for MHN (and maybe time-discretized versions of CBN and MCCBN)}
-
-% \idea{misra: discuss how it is similar or different. Or in the intro. Or
-%   in the methods. Who cites Misra?. Misra looks similar, but it gives a
-%   0/1 decision, and has a parameter, threshold, `` (any number between 0
-%   and 1), smaller numbers give you more detail (give a number between 0
-%   and 1 that specifies the level of detail in output path file, this file
-%   will have a .dot extension. We used 0.3 in the manuscript)'' And Misra
-%   seems to only find ancestor of existing genotypes. So genotypes that are
-%   not present (at least with many mutations) are never predicted to
-%   exist. It is a somewhat different method, with difficult to use code,
-%   and poorly explained arguments. Its output are not really probabilities
-%   of paths.  }
-
-\subsection{Quantification of similarity between predicted and true
-  transition probabilities} \label{sec:js}
-
-For every observed genotype $i$ we have compared the vector of true
-probabilities $\mathbf{p}_i$, from the $i$th row of $\mathbf{P}$ with the
-analogous vector of predictions from a CPM $\mathbf{\hat{p}}_i$ returned
-from the $i$th row of $\mathbf{\hat{P}}$ for each CPM. We measured the
-similarity between these two vectors using the Jensen-Shannon distance
-($JS$) \cite{crooks2017}, the square root of the Jensen-Shannon divergence
-\cite{lin1991}. The Jensen-Shannon divergence is a symmetrized
-Kullback-Leibler divergence between two probability distributions that is
-0 when the two distributions are identical and reaches its maximum value
-of 1 if the two distributions do not overlap (we used log of base 2). See
-further details in Supplementary Material, section \ref{S-supp-js}.
-
-% Consider the vector of true probabilities
-% $\mathbf{p}_i=\left(p_{ia},p_{ib},\cdots,p_{iz}\right)$, where
-% $a,b,\cdots,z$ are all genotypes with one mutation more than $i$: this is
-% nothing but the $i$-th row of the matrix $\mathbf{P}$ (excluding the
-% elements that do not satisfy $n_{mut}(j) = n_{mut}(i) + 1$). Consider also
-% the analogous vector of predictions from a CPM $\mathbf{\hat{p}}_i$. We
-% measured the similarity between these two vectors using the Jensen-Shannon
-% distance ($JS$) \cite{crooks2017}, the square root of the Jensen-Shannon
-% divergence \cite{lin1991}. The Jensen-Shannon divergence is a symmetrized
-% Kullback-Leibler divergence between two probability distributions that is
-% 0 when the two distributions are identical and reaches its maximum value
-% of 1 if the two distributions do not overlap (we used log of base 2); it
-% is defined even if the two distributions do not have the same sample space
-% (i.e., even if $p_i \ne 0$ and $q_i = 0$ or viceversa). (And the square
-% root of the Jensen-Shannon divergence is a metric between probability
-% distributions \cite{crooks2017})
-
-% \idea{CHECK: ``excluding the undefined elements that do not satisfy the
-%   condition in equation ref{eq:pij}'' : En realidad, no need to exclude
-%   them; just set them to 0. This is equivalent. Pero dejo el ``excluding
-%   the elements that do not satisfy $n_{mut}(j) = n_{mut}(i) + 1$'' que
-%   juega el mismo papel}.
-
-% Using the Jensen-Shannon distance, the square root
-% of the Jensen-Shannon divergence is a metric between probability
-% distributions.
-
-% The Jensen-Shannon distance between them is defined as \cite{crooks2017, lin1991}.
-
-% \begin{equation}
-% JS\left(\mathbf{p}_i,\mathbf{\hat{p}}_i\right) = \sqrt{\frac{1}{2}d_{KL}\left(\mathbf{p}_i,\mathbf{q}\right) + \frac{1}{2}d_{KL}\left(\mathbf{\hat{p}}_i,\mathbf{q}\right)}
-% \label{eq:js}
-% \end{equation}
-% %
-% where we have used the Kullback–Leibler divergence, defined for two arbitrary vectors $\mathbf{x}=\left(x_1,x_2,\cdots\right)$ and $\mathbf{y}=\left(y_1,y_2,\cdots\right)$ as
-
-% \begin{equation}
-% d_{KL}\left(\mathbf{x},\mathbf{y}\right) = \sum_i x_i \log_2 \frac{x_i}{y_i}
-% \label{eq:kl}
-% \end{equation}
-% %
-% $JS$ is bounded between 0 and 1, is minimal when $\mathbf{p}_i$ and
-% $\mathbf{\hat{p}}_i$ are identical and largest when they are maximally
-% different. \idea{We opted for this metric because... and it is defined
-%   even when pi or qi are 0 (?)}
-
-% \idea{Move this entire chunk to the supplementary and just leave a sentence like ``to quantify the concordance between the true probabilities and the CPMs predictions, we used the Jensen-Shannon distance $JS$.
-This is a measure of disparity across probability distributions which is zero when the two are identical and has an upper bound of 1 for maximal dissimilarity (Supplementary Methods).
-We opted for this metric because...
-(?)"}
-
-%% No longer relevant
-% \subsubsection*{Averaging statistics with weights}
-% \idea{Maybe move this to the supplementary and just leave a sentence here like "for those analyses in which the row-wise (genotype-wise?) statistics were averaged across the entire matrix (across all genotypes?), we gave each row (each genotype?) a weight according to the frequency with which it appeared in the corresponding sample provided to the CPMs (Supplementary Methods)"}
-
-% \subsubsection{Mismatches in genotype observability and null model}
-% \label{sec:mismatch_null}
-
-% In the simulations, the most common genotype at any of the sampling times
-% is always observed. But not all genotypes are always accessible (and,
-% thus, observable) for all CPMs; if a genotype $j$ is not observable
-% according to a CPM, all $\hat{p}_{jk}$ are undefined for all $k$ (no
-% prediction can be made regarding what comes after $j$, since $j$ is not
-% even observable). How do we compare the true probabilities ($p_{ij}$) with
-% the estimations provided by the CPM ($\hat{p}_{ij}$) in these cases? If
-% the genotype under consideration ($i$) is not observable in the
-% simulations we simply exclude it from further analyses: since it is not
-% going to be observed in practice, it is irrelevant what the CPM predicts.
-% If it is observable in the simulations but not deemed so by the CPM, since
-% no prediction is provided we use that of a simple null. Under this simple
-% null model, all transitions from genotype $i$ to all genotypes with
-% $n + 1$ mutations plus the $end$ and $none$ states are considered
-% equiprobable.  If the genotype is observable according to both simulations
-% and CPM, we calculate the Jensen-Shannon distance normally as explained
-% earlier.  See Supplementary Material, section
-% \ref{S-sm-mismatches} for further details.
-
-% % model instead ($\mathbf{p}^0_i$). See Supplementary Material, section
-% % \ref{S-sm-mismatches} for further details.
-
-% % Else, if the genotype is observable according to both
-% % simulations and CPM, we calculate the Jensen-Shannon distance normally as
-% % explained earlier.
-
-\subsection{Factors that affect predictive performance: linear mixed-effects model trees}
-\label{sec:linear-mixed-model}
-
-To examine the relevance of the different factors on the performance of
-methods we used linear mixed-effects model trees. These are an extension
-of recursive partitioning (or tree-based) methods, where observations are
-first split repeatedly according to the predictor variables, which play
-the role of partitioning variables, so that the dependent variable becomes
-more homogeneous within each node; the linear mixed-effects model allows
-for the addition of random effects \cite{fokkema2018, fokkema2020}. In our
-models, the dependent variable is the minimal JS over all methods (for
-each genotype by replicate by id by detection regime by sample size).  The
-possible partitioning variables include characteristics that are fitness
-landscape- and evolutionary dynamics-specific, sampling characteristics,
-genotype (within fitness landscape)-specific, and genotype (within fitness
-landscape) by sampling (sample size and detection regime) by
-replicate-specific variables.  The fitness landscape- and evolutionary
-dynamics-specific variables are % total number of
-% loci or genes (numGenes in the figures),
-SSWM regime (\texttt{SSWM} in the figures, measured as the average frequency of the
-most frequent genotype), %freq\_most\_freq\_mean\_no450}
-$\gamma$ (\texttt{gamma} ---this is not the same gamma as used in the
-time-discretization of transition rate matrices) a measure of the amount
-of epistasis in the fitness landscape, defined as ``the correlation of
-fitness effects of the same mutation in single-mutant neighbors''
-\cite{ferretti_measuring_2016} or, equivalently, ``the correlation in
-fitness effects between genotypes that only differ by 1 locus, averaged
-across the [fitness] landscape'' \cite{brouillet_magellan:_2015}, fraction
-of pairs of loci with reciprocal sign epistasis (\texttt{epistRSign}), and number
-of observed peaks in the fitness landscape (\texttt{numObservedPeaks}). The
-sampling characteristics are detection regime (\texttt{detect} in the figures) and
-sample size (\texttt{sample\_size}) ---recall that each fitness landscape was
-sampled over all nine combinations of detection regime and sampling
-size. The genotype (within fitness landscape)-specific variables are
-number of mutations (\texttt{nMut} in the figures), fitness rank (\texttt{fitnessRank},
-where the genotype of maximum fitness in a landscape has fitnessRank 1),
-and the proportion of times the genotype is a (possibly local) fitness
-maximum (\texttt{propLocalMax}) . The genotype (within fitness landscape) by
-sampling by replicate specific variables are proportion of
-times that genotype was observed in the sample (\texttt{observedProp}), and
-difference between the observed proportion and the true frequency of that
-genotype in the samples (where the true frequency is the frequency over
-the 20000 samples) (\texttt{diff\_obs\_prop}). Random effects for the global model
-are ID and, nested within ID, the crossed random effects of genotype and
-replicate. As the meaning of the number of mutations, nMut, is different
-in models with 7 and 10 genes, we fitted separate models for simulations
-with 7 and 10 genes.
-Models were fitted using the R package ``glmertree''
-\cite{fokkema2018}. The Bonferroni-corrected significance level for
-node-splitting was set at 0.01. We specified a minimal (weighted) size in
-leaves (or terminal nodes) of 1\%; as the returned trees had over 50
-leaves in all cases, to allow for interpretability we pruned the resulting
-tree by recursively merging (starting from the leaves) all children node
-with a fitted minimal JS $>$ 0.0677 (which corresponds to the 20\% best JS
----see Table \ref{S-tab_q_all} in Supplementary Material). In addition, we
-merged all children with a fitted minimal JS that differed by less than
-0.0677/4 (so as to collapse good performing nodes with minor
-differences). Further details about the fitting procedure are provided in
-the Supplementary Material section \ref{S-sec:data-proc-lmertree}.
-
-% The possible partitioning variables include characteristics that are
-% fitness landscape- and evolutionary dynamics-specific, sampling
-% characteristics, genotype (within fitness landscape)-specific, and
-% genotype (within fitness landscape) by sampling (sample size and detection
-% regime) by replicate-specific variables. Random effects for the global
-% model are ID and, nested within ID, the crossed random effects of genotype
-% and replicate.  Models were fitted using the R package ``glmertree''
-% \cite{fokkema2018}. The Bonferroni-corrected significance level for
-% node-splitting was set at 0.01. We specified a minimal (weighted) size in
-% leaves (or terminal nodes) of 1\%; as the returned trees had over 50
-% leaves in all cases, to allow for interpretability we pruned the resulting
-% tree by recursively merging (starting from the leaves) all children node
-% with a fitted minimal JS $>$ 0.0677 (which corresponds to the 20\% best JS
-% ---see Table \ref{S-tab_q_all} in Supplementary Material). In addition, we
-% merged all children with a fitted minimal JS that differed by less than
-% 0.0677/4 (so as to collapse good performing nodes with minor
-% differences). Further details about the fitting procedure are provided in
-% the Supplementary Material section \ref{S-sec:data-proc-lmertree}.
-
-\subsection{Cancer data sets}
-\label{sec:datasets}
-
-We have used a total of 25 data sets. Twenty two of them were used in
-\cite{diaz-uriarte2019a}. These 22 data sets include six different cancer
-types (breast, glioblastoma, lung, ovarian, colorectal, and pancreatic
-cancer), use different types of features (nonsynonymous somatic mutations,
-copy number alterations, or both) analyzed in terms of pathways,
-functional modules, genes, gene events, or mutations (yielding from 3 to
-192 different features), and have samples sizes from 27 to 594; the
-original sources are \cite{bamford2004, TCGA:2008-g, Jones2008,
-  parsons2008b, Wood2007, Brennan:2013, ding2008, TCGA2011, Knutsen:2005,
-  Piazza:2013, cgan2012} and all have been used previously in CPM research
-\cite{Attolini2010a, Gerstung2011, Cheng2012, Misra2014, caprese_2014,
-  capri_bioinformatics, capri_pnas, diaz-uriarte2019a}; for further
-details about those 22 data sets, see the supplementary material (file
-S5\_Text) in \cite{diaz-uriarte2019a}. We have also used
-three CGH data sets previously used in \cite{Gerstung2009, schill2020}
-that were originally obtained from the Progenetix database
-\cite{baudis2001}. These are \texttt{Breast-CGH}, from 871 breast cancer
-patients with 10 CGH alterations, \texttt{Renal-CGH}, from 251 renal cell
-carcinoma with 12 CGH alterations, and \texttt{Colon-CGH}, from 570
-colorectal cancers with 11 CGH alterations. These three data sets were
-obtained from the Supplementary data of \cite{schill2020}. Thus, these 25
-cancer data sets constitute a wide, representative example of data to
-which researchers have applied or might want to apply CPMs. We have
-analyzed all the data sets with all the methods considered; because
-computing time increases nonlinearly with number of features for some
-methods, and in particular CBN (and, to a lesser extent, MHN), have
-difficulties with 16 or more features, for the data sets with more than 15
-features, we have used the common procedure \cite{Diaz-Uriarte2015} of
-selecting the 15 most frequent features.
-
-%% Example of issues of MHN with many features:
-%% example-MHN-issues-with-many-features.Rout
-%% It is not the rates, but they provide no code to compute the transition
-%% between genotypes, and that is where things fail for me.
-
-% Using JS we computed the similarity for the prediction of the next
-% genotype between pairs of methods for all genotypes that were observed at
-% least once in the samples. % (These are not necessarily all possible
-% % genotypes; moreover, of the methods considered, MHN gill generate
-% % predictions for all possible genotypes, but for all other methods the
-% % accessible genotypes is generally a much smaller set of genotypes which
-% % need not be identical to the set of observed genotypes).
-% Based upon the results of the simulations, we will focus, for conciseness,
-% on the similarities between MHN and CBN as these are the methods than can
-% model more complex dependency relationships between mutations and show
-% better performance.
-
-% \idea{Explain variables in figure legends; see the figure
-%   legend itself}
-
-% Finally, we have used the
-
-% Glioblastoma: used by \cite{schill2020}. 261 patients, with 486 point
-% mutations (M), amplifications (A), or deletions (D). But this data set was
-% also used in \cite{cristea_pathtimex:_2016} (``pathTiMEx: Joint Inference of Mutually Exclusive Cancer
-% Pathways and Their Progression Dynamics SIMONA CRISTEA, 1'') where they
-% say ``consisting of point mutations and copy number aberrations''. But all
-% come from the TCGA. Which is the source for GBM CNA.
-
-% So possibly overlap.
-
-% \begin{figure}[H]
-%   \centering
-%   \includegraphics[width=14.5cm,keepaspectratio]{figs/figexp-crop}
-%   \internallinenumbers
-%   \caption[Figure title]{\textbf{Figure title.} Figure caption. \idea{Creo    que esto ya no es relevante? FIXME}} \label{fig:figexp}
-% \end{figure}
+This is an example cite \cite{Vetrovsky2013,nloptr}
 
@@
 
-% \subsection{Predictive performance on simulated data}%Linear mixed-effects
-% % model trees}
-% \label{sec:linear-mixed-model-results}
+\section{Discussion}\label{discussion}
 
-Figure \ref{fig:7_spec} shows the liner mixed-effects model tree for
-fitness landscapes with seven genes under the uniform detection regime and
-with the largest sample size (4000). In the Supplementary Material
-(Fig.~\ref{S-fig:10_spec}, section
-\ref{S-suppl-linear-mixed-model-results}) we show the results for 10
-genes. Figures \ref{S-fig:7_ALL} and \ref{S-fig:10_ALL}, in the
-Supplementary Material, present the models fitted to all detection regimes
-and sample sizes. In these figures we can see the relevance of detection
-regime and sample size: the smallest JS were generally achieved under
-the uniform detection regime and with the largest sample size (4000) (see
-also Supplementary material, Tables \ref{S-tab_minjs} to
-\ref{S-tab_q_rmf}).  
-% \ref{S-tab_minjs_fl_mean}, \ref{S-tab_minjs_fl_median}, and Tables
-% \ref{S-tab_q_all} to \ref{S-tab_q_rmf}).
-Moreover, there were
-interactions between detection regime and sample size on the one hand and
-genotype-specific and fitness landscape-specific variables on the other;
-these interactions can make it harder to understand the relevance of
-fitness landscape-specific and genotype-specific variables on the quality
-of predictions. Figures \ref{fig:7_spec} and \ref{S-fig:10_spec} isolate
-the effects of fitness landscape and genotype-specific variables, which is
-one of the main objectives of this paper.
+\section{Methods}\label{methods}
 
-[Fig.~\ref{fig:7_spec} about here]
+\section{Figures}\label{figs}
 
-Best achievable performance is generally larger (i.e., JS is lower) in the
-representable than in the local maxima and RMF fitness landscapes, as can
-be seen in node 4 of Fig.~\ref{fig:7_spec} (see also Tables
-\ref{S-tab_minjs_fl_mean}, \ref{S-tab_minjs_fl_median}). Moreover, gamma,
-reciprocal sign epistasis, and number of observed peaks, three of the
-variables that play important roles in the fitted trees (e.g., nodes 5,
-11, 14, 15, and 23 in Fig.~\ref{S-fig:7_ALL}; node 5 in
-Fig.~\ref{S-fig:10_spec}; nodes 12 and 15 in Fig.~\ref{S-fig:10_ALL}),
-allow to perfectly separate the RMF from the rest of the fitness
-landscape, and to separate most local maxima fitness landscapes from the
-representable ones (Fig.~\ref{S-fig:gamma_rsign_obs_peaks} in the
-Supplementary Material). The results indicate, though, that highly
-specific fitness landscape characteristics (fitness landscape itself,
-gamma, reciprocal sign epistasis, number of observed peaks) show complex
-interactions with genotype-specific characteristics. For example, in
-Fig.~\ref{fig:7_spec}, fitness landscape plays a role in node 4 only for
-genotypes with small fitness rank (i.e., large relative fitness of a
-genotype compared to the rest of the genotypes in the fitness landscape)
-and that are rarely local maxima. Likewise, number of observed peaks in
-node 19, which would separate representable from local maxima and RMF
-landscapes, is relevant for the one-before-last genotype. Similar
-interactions between genotype-specific and fitness landscape-specific
-characteristics can be seen in the figures in the Supplementary Material
-(Fig.~\ref{S-fig:7_ALL}: node 23, child of genotype-specific nodes 1 and
-19; Fig.~\ref{S-fig:10_ALL}: nodes 15 and 12, mediated by
-genotype-specific nodes 1, 11, and 14; Fig.~\ref{S-fig:10_spec}: node 5,
-affected and which affects ancestor and descendant nodes of
-genotype-specific characteristics). A general observation is that deviations
-from SSWM lead to poorer predictive ability; these effects can depend both on
-general fitness landscape characteristics (e.g., node 5 in Fig.~\ref{fig:7_spec})
-and genotype-specific characteristics (e.g., node 21 in Fig.~\ref{S-fig:7_ALL}
-and node 19 in Fig.~\ref{S-fig:10_ALL} in Supplementary Material).
-
-Not all genotype-specific characteristics had a relevant effect. Observed
-proportion of a genotype and the difference between its observed proportion and
-true frequency in the sample are variables that do not appear in the final trees
-(although the former did appear in the models before prunning). Number of
-mutations had a non-linear effect: genotypes with either none or close to
-maximum number of mutations lead to better performance, whereas genotypes with
-intermediate number of mutations had worse performance (e.g., nodes 1 and 14 in
-Fig.~\ref{fig:7_spec}, nodes 1, 3, 6 in Fig.~\ref{S-fig:10_spec}, nodes 2 and 6
-in Fig.~\ref{S-fig:10_ALL}). This indicates that prediction becomes harder when
-the number of potential next genotypes increases, which happens at intermediate
-numbers of mutations.
-%(\idea{rm this? Note that this pattern
-  % is present for all methods, both the MHN methods, that always provide
-  % $\hat{p}_{ij}$, as well as the rest of the methods where, as explained
-  % in section \ref{sec:mismatch_null}, we used a simple equiprobable null
-  % model when the observed genotype was not among the accessible genotypes:
-  % see Supplementary Material \ref{S-JS-num-mut}}).
-Smaller fitnessRank
-(i.e., large relative fitness of a genotype) was generally associated to
-better predictions (e.g., nodes 9 and 15 in Fig.~\ref{fig:7_spec}; nodes
-12, 13, and 4 in Fig.~\ref{S-fig:10_spec}) but, as with most other
-characteristics, the effect was strongly affected by other variables
-(node 1 in Fig.~\ref{S-fig:7_ALL}, nodes 11 and 21 in
-Fig.~\ref{S-fig:10_ALL}). Note that genotypes of fitness rank 1 are
-necessarily fitness maxima.
-
-In fact, how often a genotype is a local fitness maximum (propLocalMax) not only
-affected performance but also which methods were better. This is apparent in
-node 3 in Fig.~\ref{fig:7_spec}, node 2 in Fig.~\ref{S-fig:10_spec}, node 19 in
-Fig.~\ref{S-fig:7_ALL}, and node 1 in Fig.~\ref{S-fig:10_ALL}: when a genotype
-was frequently (over 20 to 50\%, depending on the data set) a local maximum, the
-TD methods lead to much better performance that the other methods. Moreover,
-once the frequency of local maxima was large, making it even larger lead to
-better performance (node 11 in Fig.~\ref{fig:7_spec}; nodes 14, 17, and 19 in
-Fig.~\ref{S-fig:10_spec}; node 19 in Fig.~\ref{S-fig:7_ALL}; node 14 in
-Fig.~\ref{S-fig:10_ALL}). This phenomenon can also be looked at from the
-perspective of global fitness landscape characteristics: an increase in the
-number of local fitness peaks (numObservedPeaks) can lead to the TD methods
-being better than their CE counterparts (node 19 in Fig.~\ref{fig:7_spec}) or to
-the TD methods improving their performance (node 15 in
-Fig.~\ref{S-fig:10_ALL}).This is also observed with increasing reciprocal sign
-epistasis (episRSign ---nodes 5, 11, 14 in Fig.~\ref{S-fig:7_ALL}), a parameter
-that can almost perfectly separate the RMF and local maxima landscapes from the
-representable ones (see Fig.~\ref{S-fig:gamma_rsign_obs_peaks}). Note that
-epistRSign no longer plays a role in a switch between methods in node 23, since
-this node only splits genotypes in fitness landscapes that cannot be
-representable (genotypes that have a probability of being local maxima $> 0$
-even when their fitness rank is $> 1$).
-
-Scenarios with very good TD performance (JS less than approximately 0.1)
-comprise, then, between 1 and 5\% of the cases (4\% in nodes 10 and 14 in
-Fig.~\ref{fig:7_spec}; 5\% in nodes 16, 20, and 21 in Fig.~\ref{S-fig:10_spec};
-1\% in node 24 of Fig.~\ref{S-fig:7_ALL}; 2\% in nodes 18 and 22 of
-Fig.~\ref{S-fig:10_ALL}; see also Tables~\ref{S-tab_q_all} to
-\ref{S-tab:xtabs_ce1}). In these cases, MHN\_td was generally slightly better
-performing than CBN\_td (and MCCBN\_td was generally the worse of the TD
-methods). Within the non-TD methods, when they could do well, most methods had
-roughly similar performance (except for CAPRI, that was normally the worse
-performer), with a slight advantage of CBN and MHN. CBN was generally among the
-best performers, and scenarios where MHN was the single best method were rare
-(but see node 20 in Fig.~\ref{fig:7_spec}). For methods with both a weighted and
-an unweigthed version (CBN\_uw, MCCBN\_uw, OT\_uw) using weights never lead to
-worse performance and sometimes improved it. As with the TD methods, the CE
-methods were consistently very good only in limited scenarios (see also
-Tables~\ref{S-tab_q_all} to \ref{S-tab:xtabs_ce1}. For CE methods these were
-nodes 7 and 20 in Fig.~\ref{fig:7_spec} which comprised about 7\% of the
-(weighted) observations, and slightly worse in nodes 6 and 16 ---where MHN's
-performance was surprisingly poor--- that comprised another 6\% of the
-observations. Similar patterns are seen in Figures \ref{S-fig:10_spec} (node 23,
-with 5\% of observations), \ref{S-fig:7_ALL} (node 17, and with worse
-performance 6, 12, 16, comprising overall 6\% of observations) and
-\ref{S-fig:10_ALL} (node 10 and, with worse performance, 9, both with 1\% of the
-observations).
-
-% In summary, then, there were complex interactions between fitness
-% landscape-specific characteristics, sampling, and genotype-specific
-% characteristics that affected the best achievable performance; for
-% scenarios where the best achievable performance was good, these three
-% factors also affected which methods were the best. In general, the best
-% performance was seen with large sample sizes, a uniform detection regime,
-% with representable fitness landscapes and under a SSWM regime. Genotypes
-% with maximum or close to maximum fitness (i.e., small fitness rank) with
-% either 0 or close to the maximum number of mutations, are those for which
-% predictions tend to be better. In terms of method choice, and when the
-% minimal achievable JS can be small (i.e., where it matters) there were
-% sharply contrasting differences between the 13 methods examined: either
-% the TD methods (those that can model local maxima) MHN\_td and CBN\_td
-% both do a very good or both do a very bad job, with MHN and CBN presenting
-% the opposite pattern (generally both a very bad job or a very good job).
-% If the genotype can be a local maximum, then the more often it is a local
-% maximum, the better the prediction, and when a genotype is frequently a
-% local maximum a simple approach like the TD approach with CBN or MHN can
-% perform well.
-
-% \idea{some of the above repeat in first paragraph of discussion}
-
-\section{Discussion: implications for the analysis of biological data sets} \label{discussion}
-
-Can we use cancer progression models (CPMs) to predict the short-term
-evolution of a tumor? Informally, we could word this question as ``what
-genotype comes next in the path to tumor fixation?''. More precisely, we
-can ask whether, conditional on us observing a particular genotype with
-$\mathrm{n}$ mutations as the most abundant genotype in a tumor, we can
-predict what genotype with $\mathrm{n}+1$ mutations is in the Line of
-Descent (LOD) to the fitness maximum (where fixation occurs). We have
-examined the performance of 13 different CPM procedures for predicting
-short-term evolution, focusing mainly on the best achievable performance
-over all 13 methods (i.e., on how close to the truth are the predictions
-of short-term evolution, regardless of method) and, secondarily, on the
-choice of method. The analysis of simulated data indicated that it was
-possible to provide predictions of short-term evolution that closely
-matched the true probabilities of short-term evolution only under some
-specific combinations of genotype characteristics and fitness landscape
-characteristics. And when  good performance was possible, no method or
-set of methods was consistently superior, making choice of best method
-also dependent on genotype and fitness landscape properties.
-
-%\change{And when  good performance was possible, no method or
-%set of methods was consistently superior, making choice of best method
-%dependent on genotype and fitness landscape characteristics.}{These characteristics
-%also determined the choice of the best method or group of methods, since even in
-%the cases where good performance was possible no specific method was consistently
-%superior.}\textcolor{orange}{Hummm... no estoy seguro del rewording: it misses the
-%emphasis en la dependencia de la decisión. Y el empezar la frase con el ``And''
-%era para añadir ``dramatismo'' a la cosa} % Application
-%% of the best performing methods to 25 biological data sets showed wide
-%% variability in predictions and the difficulty of choosing the specific set
-%% of predictions to trust because of the lack of detailed information about
-%% genotype and fitness landscape characteristics.
-
-% With the simulated data (i.e., data for which we can measure the deviation
-% of predictions from the truth) b
-Best achievable performance was affected by
-complex interactions between fitness landscape-specific
-characteristics% (type of fitness landscape and associated variables gamma ---correlation
-% of fitness effects in single-mutant neighbors---, reciprocal sign
-% epistasis, and number of peaks in the landscape)
-, deviations from the
-strong selection and weak mutation (SSWM) regime, sampling%  (sample size
-% and whether tumors are sampled early, late, or uniformly during tumor
-% evolution)
-, and genotype-specific characteristics%  (fitness rank, number of
-% mutations, and probability of being a local fitness maximum)
-.  The
-best performance was seen with large sample sizes, a uniform detection
-regime, with representable fitness landscapes and under a SSWM
-regime.  % In general (see
-% Fig.~\ref{fig:7_spec} and Supplementary Material
-% Figs.~\ref{S-fig:10_spec}, ~\ref{S-fig:7_ALL}, ~\ref{S-fig:10_ALL})
-Genotypes with maximum or close to maximum fitness (i.e., small fitness
-rank) with either 0 or close to the maximum number of mutations, are those
-for which predictions tend to be better.  For scenarios where the best
-achievable performance was good (i.e., where it matters), these three
-factors also affected which methods were the best. There were sharply
-contrasting differences between methods, in particular, methods
-that can model local maxima ---the TD methods MHN\_td and CBN\_td--- tended to
-either both
-perform very well or both perform very bad. The corresponding two methods
-that do not model local maxima ---MHN and CBN--- generally present the opposite
-pattern, performing well when the TD methods do not and vice-versa.
-
-Since our focus are predictions conditional on the observed genotype, we
-can also summarize the results by dividing genotypes into three groups: A)
-Genotypes for which using the TD methods lead to very good predictions (JS
-$< 0.1$); B) Genotypes for which the CE methods can provide very good
-predictions (JS $< 0.1$); C) The remaining cases, which constitute the
-majority of the cases, where performance was generally poor (JS $> 0.25$).
-Group A consists of genotypes that are more often than not local maxima
-(probability of being a local maximum $>0.5$) and of large fitness (small
-fitnessRank) or that are often local maxima (probability of being a local
-maximum $>0.25$) while deviations from SSWM are minor. Fitness
-landscape-specific characteristics that make these genotypes 
-more likely to be observed
-are large number of local fitness peaks and large reciprocal sign epistasis.
-For these genotypes, the TD methods can
-correctly predict being at the end of the evolutionary process, something
-that the remaining methods cannot account for, which explains the very
-poor performance of the non-TD methods in these cases (JS $> 0.75$). In
-group B are genotypes of large relative fitness, no mutations or with all
-except one loci mutated, in evolutionary scenarios with SSWM and smooth
-fitness landscapes (little reciprocal sign epistasis, large gamma) and few
-local peaks (less than 3). For these genotypes, the TD methods make very
-poor predictions (JS $> 0.75$) while MHN and CBN do particularly good jobs, but
-CAPRESE and OT, two methods that can only model simpler patterns of
-dependencies, generally perform well too. In the remaining cases (group
-C), no method can provide good predictions. Assessing whether we can
-achieve good performance and, if so, deciding
-which method
-to use, requires detailed information about the genotype, the fitness
-landscape, and the evolutionary dynamics. For some small subset of
-scenarios performance could be very good but it could also be dismal if we
-use the incorrect method.
-
-% \idea{Somewhere, maybe in x1 or in x2, reintroduce the idea of two
-%   Flandscape traits that give advantage to TD:
-% ``an increase in the number of local fitness peaks (numObservedPeaks):
-% this can lead to the TD methods being better than their CE counterparts''
-% and ``increasing reciprocal sign epistasis
-% (epistRSign) can lead to the TD methods performing better than their CE
-% counterparts'': }
-
-% A) Genotypes for which using the TD methods lead to very good predictions
-% (JS $< 0.1$), which are genotypes that are more often than not local
-% maxima (probability of being a local maximum $>0.5$) and of large fitness
-% (small fitnessRank) or that are often local maxima (probability of being a
-% local maximum $>0.25$) while deviations from SSWM are minor; the TD
-% methods can correctly predict being at the end of the evolutionary
-% process, something that the remaining methods cannot account for, which
-% explains the very poor performance of the non-TD methods in these cases
-% (JS $> 0.75$). B) Genotypes for which the CE methods can provide very good
-% predictions (JD $< 0.1$), which are genotypes of large relative fitness,
-% no mutations or with all except one loci mutated, in evolutionary
-% scenarios with SSWM and smooth fitness landscapes (little reciprocal sign
-% epistasis, large gamma) and few local peaks (less than 3); for these
-% genotypes, the TD methods make very poor predictions (JS $> 0.75$); MHN
-% and CBN do particularly good jobs, but CAPRESE and OT, two methods that
-% can only model simpler patterns of dependencies, generally perform well
-% too. C) The remaining cases, which constitute the majority of the cases,
-% where performance was generally poor (JS $> 0.25$). Therefore, assessing
-% that we can achieve good performance and, if we can achieve it, deciding
-% which method to use, requires detailed information about the genotype, the
-% fitness landscape, and the evolutionary dynamics; for some small subset of
-% scenarios performance could be very good but it could also be dismal if we
-% use the incorrect method.
-
-Our results for short-term predictions highlight variables that have been
-found to be relevant for long-term prediction of the complete evolutionary
-paths.  Sampling regime and sample size have previously been found
-relevant by \cite{diaz-uriarte2019a,hosseini2019a} (see also
-\cite{Diaz-Uriarte2015, diaz-uriarte2018} for the effect of sampling regime
-in inferring the DAGs of restrictions \textit{per se}). Likewise,
-evolutionary scenarios with clonal interference in rugged fitness
-landscapes (small gamma, large reciprocal sign epistasis, large number of
-local fitness maxima) lead to poorer performance of CPMs compared to SSWM
-regimes in single-peaked fitness landscapes
-\cite{diaz-uriarte2019a,hosseini2019a}. % and the SSWM
-% assumption is explicit in how \cite{hosseini2019a} use CBN for predicting
-% tumor evolution.
-The role of smooth fitness landscapes for CPM's predictive ability is
-directly related to the CPM assumption that final fixation occurs in the
-genotype with all loci mutated and their inability to model reciprocal
-sign epistasis \cite{diaz-uriarte2018} (this problem also affects MHN,
-specially if there are higher-order epistatic
-interactions). Interestingly, these are the fitness landscapes where the
-TD methods (which allow us to capture local maxima) perform better than
-the CE methods. The role of SSWM for CPMs has been discussed before by
-\cite{Sprouffske2011, diaz-uriarte2019a} and \cite{hosseini2019a} used it
-explicitly for predicting tumor evolution. In fact, all of the methods
-considered here implicitly assume SSWM mainly for two reasons. First, CPMs
-are generally applied to cross sectional data from bulk sequencing (not
-single-cell sequencing) without undergoing any deconvolution step
-(i.e. without attempting to infer the clonal composition of the sequenced
-tumor), which reflects the implicit assumption that an actual existing
-genotype can be obtained from bulk sequencing (see
-\cite{alves_multiregional_2017} for discussion of how this assumption can
-fail with bulk samples as we collapse over existing genotypes); if this
-assumption does not hold, the question ``what genotype comes next'' is
-undefined because we are conditioning on a non-existing genotype.  Second,
-the observed genotypes are considered true steps in the path to tumor
-fixation, thus neglecting the possibility of tunneling effects
-\cite{devisser_empirical_2014,szendro_predictability_2013,sniegowski_beneficial_2010}. These
-two assumptions are true if SSWM holds (see Fig.~\ref{fig:fig1}, and note
-that the reverse is not necessarily true), in which case the most fit
-genotype at any given time is effectively the only one in the population,
-thus the only one observed, and an ancestor of the ultimately fixated
-one. Even when CPMs specifically include error estimations
-(e.g. \cite{Gerstung2009}), these typically refer to sequencing error but
-not necessarily to deviations from SSWM.  And our reformulation of the
-question ``what genotype comes next'' as we did in section
-\nameref{sec:intro} % and \nameref{sec:special}
-allows us to deal with the second violation, but of course not with the
-first. In our analysis of the simulated data the first assumption holds
-(we sample true, existing genotypes); yet, our results show that using
-CPMs to predict short-term evolution when the second assumption does not
-hold resulted in poorer performance. It must be emphasized, though, that
-the above comments refer to the effects of deviations from SSWM on the
-performance of the methods considered, not on predictability itself (the
-role of SSWM on predictability itself is more nuanced: see, e.g.,
-\cite{ferretti2018,devisser_empirical_2014,bailey2017,
-  szendro_predictability_2013,zhao2016, bank_predictability_2016}).
-
-% in the limit (if
-% only one mutation can improve fitness), evolutionary trajectories can even
-% be deterministic \cite{ferretti2018}, but deviations from SSWM can
-% actually improve predictability, such as those brought about by moderately
-% large (but not too large) population sizes \cite{zhao2016,
-%   devisser_empirical_2014, szendro_predictability_2013} which can increase
-% parallel evolution and the role that selection plays, as shown in
-% experimental evolution studies in yeast and bacteria \cite{bailey2017}.
-
-In addition to the above factors, which affect both short- and long-term
-predictions, our results highlight that when we want to make conditional,
-short-term predictions, features of the genotypes for which we make the
-predictions can be as relevant as, or more relevant than, fitness
-landscape-specific features (see also \cite{bank_predictability_2016,
-  visser2018,ferretti2018} for a comparison of short-term vs.\ long-term
-predictability). We have already mentioned the combined importance of
-fitness rank and probability of being a local maximum, and the latter,
-together with reciprocal sign epistasis, illustrate factors that can
-increase short-term predictability while decreasing global-path
-predictability. The role of number of mutations, where predictions are
-worse from genotypes with intermediate numbers of mutations, can be
-understood as a consequence of the inverted U-shape relationship between
-the number of entries that can be non-zero in $\mathbf{P}$ and the number
-of mutations in a genotype (${N}\choose{n + 1}$: see section
-\nameref{sec:methods:sim}): the problem of prediction becomes harder at
-intermediate numbers of mutations.  The role of genotype-specific factors
-is a noteworthy feature because it suggests that CPMs might be leveraged
-to provide short-term predictions even when global, long-term predictions
-are difficult because fitness landscape- and evolutionary model-specific
-assumptions are violated.
-
-% So, if the genotype can be a local maximum, a simple approach like the
-% TD approach can perform well, and it will perform better the more often
-% the genotype is a local maximum.
-
-%% Simplifying the results, we can divide genotypes in three groups.
-
-%% \idea{EH! Some of that could be the result of the methods we use!!! Rewrite}
-
-What are the implications of the above results for the analysis of cancer
-data? We have compared the predictions from the best performing methods on
-25 biological data sets: even if we do not know the true transition
-probabilities, we can examine the consequences of assuming that we choose
-one method when another method is the one that gives the best (or even
-perfect) predictions. This is what we do in
-Fig.~\ref{fig:biol_boxplots-3-comps}. The left panel
-of Fig.~\ref{fig:biol_boxplots-3-comps}A,
-for example, shows
-the difference in predictions (measured using JS) if the method that gave
-perfect predictions were CBN but we had used MHN (or viceversa).  If we
-found that, for a given genotype, two or more methods gave similar
-predictions (i.e. comparing their predicted probabilities yielded a small
-JS) we could argue that they could all be potentially providing good
-results. On the other hand,
-very different predictions would indicate that at least one of them
-(if not all) is failing, and in that case
-there are no solid grounds to choose one or the
-other since we lack access to the underlying factors that affect
-performance.  As can be seen, for most data sets there is wide
-variability in the consequences of using the wrong method, and often the
-difference in predictions is very large, specially if we use a TD vs.\ a
-non-TD method (MHN vs.\ MHN\_td, right-most panel). 
-The high variability between method
-predictions shown in Fig.~\ref{fig:biol_boxplots-3-comps}A ---even across genotypes 
-within a same dataset--- highlights the importance
-of conditioning predictions to the observed genotype, as different methods can make
-very similar or substantially different predictions depending on it.
-Particularly in the latter case, if any of the methods was indeed providing accurate
-predictions, determining which one is the correct one would depend on the properties
-of the observed genotype.
-
-[Figs.~\ref{fig:biol_boxplots-3-comps} and \ref{fig:biol_freqs} about
-here]
-
-However, in general we do not have access to the underlying parameters
-  that would allow us to make an informed decision on method (or method family)
-  choice. So how do the above results contribute to the applied analysis of actual
-  cancer data? Fig.~\ref{fig:biol_boxplots-3-comps}B shows the dissimilarity
-between the predictions of two specific CE methods (CBN and MHN), two specific
-TD methods (CBN\_td and MHN\_td), and two methods from different groups (MHN and
-MHN\_td).
-Many genotypes accumulate in the area of the plot where the predictions of the
-two CE methods, as well as those of the two TD methods, are very different, with variable
-similarity between CE and TD predictions. But a few genotypes escape this trend.
-Looking back at the simulations, we have seen that there
-is a sharp difference in behavior of TD and CE methods when the best
-achievable performance is good; and necessarily CBN and MHN (or CBN\_td
-and MHN\_td) can only both make correct predictions if both are making
-similar predictions. This same pattern ---predictions
-of methods within the same
-  group are similar and predictions of methods from different groups are not--- is
-  observed in the biological data for a small subset of genotypes.
-  What about limiting our interpretation to those genotypes?
-This approach by itself will most likely yield only modest improvements
-in performance (see Tables
-\ref{S-tab:xtabs_td0} to \ref{S-tab:xtabs_ce1} showing that, for TD, the
-proportion of cases with very good TD performance could at most increase
-from an unconditional 0.04 to a conditional 0.06, and the proportion of
-cases with very good performance of CE from an unconditional 0.04 to a
-conditional 0.16).
-%But there is a more fundamental problem illustrated in Fig.~\ref{fig:biol_freqs}.
-Fig.~\ref{fig:biol_freqs} shows the fraction of genotypes,
-for each data set, that meet fairly stringent conditions of similarity
-either within the CE methods, i.e. between CBN and MHN, or within the TD
-methods, i.e. between CBN\_td and MHN\_td (see Supplementary Material,
-figures in Section \ref{S-sm-indiv-biol-data-plots}, for similarity for
-each individual genotype for each data
-set% ; note that these patterns are not related to
-% sample size ---see Fig.~\ref{S-fig:biol_freqs_sampsize}
-).  For 11 of the
-data sets, there are more than 10\% of the genotypes that fall in either
-of these categories: which among the predictions that fulfill those
-similarity conditions should we trust? The difficulty of this choice is
-even more dramatic when the same genotype simultaneosuly shows large similarity
-within TD methods, large similarity within CE methods, and large
-dissimilarity between the CE and TD methods, as indicated by the gray
-points in Fig.~\ref{fig:biol_freqs}. In these cases, predictions between families of
-methods are very different and yet method (or method family) choice would
-require fitness landscape-specific, evolutionary model-specific and
-genotype-specific information. % Finally, the question ``what genotype comes
-% next'' could actually be undefined if the genotypes for which we ask the
-% question do not really exist but, rather, are an artifact from using bulk
-% sequencing and collapsing over genotypes.
-
-  Even if this information is generally unknown in practice,
-  our results provide a foundation to narrow down the
-  search for genotypes for which good performance is possible.
-  Furthermore, we have demonstrated that we do not need complete
-  information on the general properties of the fitness landscape; instead, even partial
-  information on those specific genotypes could serve to make informed decisions
-  on method choice. Gathering information on the properties of individual
-  genotypes is, in principle, a more reasonable objective than obtaining a full
-  description of the complete fitness landscape for every cancer type.
-
-%\idea{This previous paragraph could maybe be put in a more positive perspective?
-%Yes, it is true that even when CE predictions are similar + TD predictions are
-%similar + TD vs CE predictions are dissimilar, we do not have access to the 
-%underlying parameters that would allow us to make an informed decision on method
-%(or method family) choice. But at least we are seeing the same pattern as in the
-%simulations for a subset of genotypes: predictions of
-%methods within the same group are similar and predictions of methods from different
-%groups are not (maybe there is a more clear way to show this in the figure?).
-%We know which specific genotypes follow this pattern, so we
-%know that those are the genotypes for which potentially at least one method (or family)
-%could be providing good results. So we actually know which genotypes we should be
-%looking at.}
-%
-%\idea{And sure, we do not have access to the parameters that would allow us to know which
-%specific method we should be using.
-%But that "is not our problem": we know what conditions need to be met to
-%achieve good performance, and we know that when good performance is possible, one
-%family of methods does well and the other does not. We know which genotypes specifically
-%satisfy this behavior, so we can narrow down the search for genotypes for which
-%good performance is possible.
-%Furthermore, we have demonstrated that we do not need
-%complete information on the general properties of the fitness landscape; instead, even
-%partial information on those specific genotypes could serve to make informed
-%decisions on method choice. Gathering information on the properties of specific genotypes 
-%could, in principle, be a more realistic goal than obtaining a full description of
-%the complete fitness landscape for each cancer type. Even if "how to do this" is beyond
-%the scope of our work, our findings provide a foundation for "what we should be looking
-%for".}
-
-%\textcolor{orange}{Given that we do not have access to the underlying parameters
-%  that would allow us to make an informed decision on method (or method family)
-%  choice, how do the above results contribute to the applied analysis of actual
-%  cancer data? In the biological data sets, we see the same pattern as in the
-%  simulations for a subset of genotypes: predictions of methods within the same
-%  group are similar and predictions of methods from different groups are not.
-%  We can identify which specific genotypes follow this pattern, so we know that those
-%  are the genotypes for which at least one method (or method family) could potentially
-%  be providing good results. This provides a foundation to narrow down the
-%  search for genotypes for which good performance is possible.}
-%
-%  \textcolor{orange}{Furthermore, we have demonstrated that we do not need complete
-%  information on the general properties of the fitness landscape; instead, even partial
-%  information on those specific genotypes could serve to make informed decisions
-%  on method choice. Gathering information on the properties of specific
-%  genotypes could, in principle, be a more realistic goal than obtaining a full
-%  description of the complete fitness landscape for each cancer type. Even if
-%  "how to do this" is beyond the scope of our work, our findings provide a
-%  foundation for "what we should be looking for".}
-
-\subsection{Conclusion}
-\label{sec:conclusion}
-
-CPMs could, under very specific combinations of genotype characteristics
-and fitness landscape characteristics, be used to obtain good predictions
-of the short-term evolution of a tumor, even when long-term predictions
-are not possible because of violations of assumptions of CPM
-procedures. But method choice and assessing if the predictions obtained
-are to be trusted, for example to guide therapeutic decisions, requires
-detailed knowledge which is not available with empirical data. More
-generally, our work shows the promises and practical difficulties of using
-cross-sectional data in an evolutionary context to guide individualized
-therapies, even when cross-sectional data recapitulate the equivalent of
-many repeated evolutionary processes.  % , lack of detailed knowledge about key
-% parameters and variables do not allow direct application of the insights
-% gained from repeated evolution experiments in other systems.
-Exploiting the potential of these methods will require further work to, first,
-examine the interpretation of their output and the consequences of their
-usage when key assumptions do not hold; second,
-identify if the
-methods' output, by itself or in combination with data set features, can
-provide indicators of performance; and third, develop
-strategies to determine the characteristics of individual genotypes in real tumors and adapt
-method use consequently.
-For all three tasks, it will be
-important to make explicit the relationship between assumptions with
-respect to the evolutionary model (e.g., SSWM), fitness landscape
-characteristics (e.g., local maxima and reciprocal sign epistasis),
-sampling regimes (late vs.\ early tumors), and predictions, all in the
-context of well defined concepts such as lines of descent and for
-specified tasks such as interventions.
-
-% \begin{figure}[!h]
-%   \centering
-%   \includegraphics[width=15.5cm,keepaspectratio]{figs/biol-data_js-vs-nMut_multi-glmm.pdf}
-%   \caption{\textbf{Similarity of predictions between methods and relative
-%       number of mutations of genotypes.} Y-axis: similarity, measured with
-%     JS, of predictions of next genotype between different methods. X-axis:
-%     the number of mutations of the source genotype relative to the total
-%     number of genes (or loci that could be mutated) for the give data
-%     set. Different colors identify different data sets.  Upper panel, TD
-%     vs.\ TD: average similarity between CBN\_td and MHN\_td and MCCBN\_td
-%     and MHN\_td (``TD'' stands for ``time-discretized''). Middle panel:
-%     average similarity between CBN and MHN and MCCBN and MHN (``CE'':
-%     competing exponentials). Lower panel: average similarity of CBN and
-%     CBN\_td, MCCBN and MCCBN\_td, and MHN and MHN\_td (i.e., between the
-%     CE and TD versions of each method).  Lines shown represent the output
-%     from a generalized linear mixed effects model for a beta-distributed
-%     response (JS); we used natural cubic splines to model the relationship
-%     with relative number of mutations of genotypes in both the fixed and
-%     random effects (data set id) components. The black thick line shows
-%     the fixed effect fit (and the colored, thin, lines the data
-%     set-specific fits ---i.e., including the random effects). Models
-%     fitted using glmmTMB \cite{brooks2017}. See further details in the
-%     Supplementary Material (``Model fits to similarity of predictions
-%     between'').}
-
-%   \label{fig_biol_comp} \end{figure}
-
-\section{Acknowledgments}
-% \idea{Ask Achim Zeileis and Rudolf Schill}
-
-We thank A.~Zeileis for answers about the use of the glmertree package and
-R.~Schill for answers about the use of MHN, and R.~Guantes and C.~Lazaro-Perea
-for comments on the ms.
-%% assistance me parecía too much. Fueron dos respuestas.
-
-\section{Funding}
-Supported by BFU2015-67302-R (MINECO/FEDER, EU) and
-PID2019-111256RB-I00/AEI/10.13039/501100011033 
-to RDU. JDC supported by
-PEJD-2018-POST/BMD-8960 from Comunidad de Madrid.
-
-\section{Data \& code availability}
-
-Code for the analyses in this article is available at
-\url{https://github.com/rdiaz02/what_genotype_next}.
-%% \url{github.com/jdiazc9/what-genot-next}.
-
-% Details and code for the simulations available from the Supplementary
-% Material of Díaz-Uriarte and
-% Vasallo, 2019 \cite{diaz-uriarte2019a}. Code for the analyses in this
-% article is available at \idea{\url{github.com/jdiazc9/oncolab}: change
-%   that to the definitive one}.
-
-% \end{document}
-
-% REFERENCES
-{\small
-  %%\bibliography{refs}
-  \bibliography{extracted-draft}
-\bibliographystyle{mybibwithurl} %% so that Crooks URL always shows
-%%\bibliographystyle{mystyle} % choose this for: "This is some text in the paper blablabla [1-3], this is how it is cited [4]."
-% \bibliographystyle{apalike} % choose this for: "This is some text in the paper blablabla [Author1, 2009, Author2 and Author3, 2016], this is how it is cited [Author4 et al., 2012]."
-}
-
+% references
 \clearpage
-
-\section{Figures}
-\clearpage
-% \todo{Hay que quitar el trim de las figuras. Deberían ser figuras que
-%   están ya cropped ---y que serán las que enviemos}
-
-%%\idea{We need to add an MHN and change the figure accordingly}
-% \idea{el panel B deberia tener flechas que indiquen que se mueve hacia la
-%   derecha (i.e., ganando mutaciones)}
-
-\begin{figure}[H]
-	\centering
-	% \fcolorbox{white}{white}{\includegraphics[width=0.8\linewidth,trim={4cm
-        %     6.5cm 4cm 5.5cm}]{figs/final-layout/fig1}}
-        \includegraphics[width=0.99\linewidth,keepaspectratio]{figs/fig1-crop-witharrows-compact}
-	\internallinenumbers
-	\caption[Inferring evolutionary trajectories using Cancer
-        Progression Models.]{\textbf{Inferring evolutionary trajectories
-            using Cancer Progression Models.} \textbf{A.} Example of a
-          directed acyclic graph (DAG) of restrictions in the order of
-          accumulation of mutations, analogous to the ones produced by
-          most CPMs, such as CBN, MCCBN, OT, CAPRESE, and CAPRI. Given
-          four driver genes (numbered 1 to 4), the DAG indicates the
-          constraints in the order of accumulation of mutations
-          (arrows). Here, a mutation in the 4th loci is conditioned to the
-          2nd and 3rd ones having been previously mutated.  MHN uses a
-          network to encode both promoting and inhibiting effects between
-          pairs of events; the network could be identical to the one in A
-          (meaning that mutations 2 and 3 have a promoting effect on
-          mutation 4), but it could also include inhibiting effects; in
-          contrast to DAGs of restrictions, MHNs do not denote
-          deterministic dependencies.  \textbf{B.} Restrictions in the
-          order of accumulation of mutations limit the amount of available
-          evolutionary
-          trajectories and stochastic dependencies change the
-          probabilities of trajectories.  In this network, nodes represent
-          different genotypes of four driver genes and edges connect those
-          that are a single mutation away. Genotype names are assigned
-          according to which genes are mutated (1) or not mutated
-          (0). Without restrictions many mutational pathways are possible
-          (any route through both solid and/or dashed edges, from left to
-          right).  However, only a subset of genotypes satisfy the
-          constraints in panel A (solid nodes); thus, only trajectories
-          connecting these are predicted to be accessible pathways (solid
-          edges). For example, given a specific genotype (1000, red node),
-          restrictions in the mutational order limit the possible clones
-          that can evolve from it consecutively after (red edges). With
-          CBN, MCCBN, and OT, in addition to the set of accessible
-          pathways, it is possible to obtain transition probabilities
-          along the black lines.  Under the MHN model, as there are no
-          deterministic dependencies, all transitions (to genotypes with
-          one additional mutation) are possible, but transitions affected
-          by inhibitory stochastic dependencies will have very low
-          probability (in this simplified depiction they would be the gray
-          lines) and those affected by promoting stochastic dependencies
-          will have higher probability (black lines); MHN provides
-          transition probabilities for all the lines in the figure.
-          \textbf{C.}  Example of cancer progression with four driver
-          genes under SSWM conditions. Progression towards the fitness
-          maximum occurs through a defined sequence of parent-child
-          genotypes, each one of them establishing as the most abundant
-          rapidly after evolving. \textbf{D.} Example of cancer
-          progression with four driver genes under conditions of no
-          SSWM. Here, the sequence of parent-child genotypes from the
-          wild-type with no driver mutations to the final one (i.e. the
-          Line Of Descent or LOD) is: 0000 $\rightarrow$ 1000
-          $\rightarrow$ 1010 $\rightarrow$ 1110. Notice that in the most
-          general case, the ultimately fixated genotype need not be the
-          one with all driver mutations (1111).  Some genotypes can be
-          most abundant in the tumor during prolonged periods of time
-          despite not being part of the LOD (0010, in green), and
-          genotypes that do belong to the LOD may never be most abundant
-          (1010, in blue). In all models and figures we do assume that
-          there are no back mutations and that crossing valleys in the
-          fitness landscape using a single multi-mutation step is not
-          possible. }\label{fig:fig1}
-\end{figure}
-
-\begin{landscape}
-  \begin{figure}[!h]
-    \vspace*{-45pt}
-%  \centering
-\includegraphics[width=22.5cm,keepaspectratio]{figs/weighted-7-specific-no-split-01-sqrt-all_methods-js.pdf}
-  \internallinenumbers
-  \caption{\textbf{Linear mixed-effects model tree fitted to data from
-      fitness landscapes with 7 loci, sample size of 4000 and uniform
-      sampling.}  Internal nodes show the partitioning variables and the
-    values that lead to each partition; light blue denotes the values or
-    categories that lead down the left split and salmon the values or
-    categories that lead down the right split (partitioning criteria are
-    also shown in the edges leading down a node). Leaves or terminal nodes
-    show (weighted) boxplots of the dependent variable for the
-    observations that fall in that leave and, as label, the
-    name of the node and the sample size (sum of weighted observations and
-    \% of the total data set for the fit). The dot plots underneath the
-    boxplots show the (weighted) mean JS for each method for that leave.}
-  \label{fig:7_spec} \end{figure}
-% \restoregeometry
-\end{landscape}
-
-\begin{figure}[!h]
-  \centering
-  %\includegraphics[width=15.5cm,keepaspectratio]{figs/boxplots-three-comps_all_datasets_combined.pdf}
-  \includegraphics[width=15.5cm,keepaspectratio]{figs/biol_data_box_scatter.pdf}
-
-  \caption{\textbf{Similarity between predictions of three pairs of
-      methods for the 25 biological data sets.} For each data set, the JS
-    divergence between the predictions of a pair of methods is represented in
-    \textbf{A.}
-    each of the three panels or \textbf{B.} each of the three axes. 
-    Dots correspond to all genotypes
-    for which we computed predictions, which are the genotypes observed in the
-    samples, except for the genotype with all genes mutated if it was observed.
-    In parentheses,
-    after the data set, we indicate the total sample size (n) and number of genes (g).}
-
-  \label{fig:biol_boxplots-3-comps} \end{figure}
-
-\begin{figure}[!h]
-  \centering
-  \includegraphics[width=15.5cm,keepaspectratio]{figs/fraction_CE_TD_similar_all_datasets.pdf}
-
-  \caption{\textbf{Percentage of genotypes with similar TD and similar CE
-      predictions for the 25 biological data sets.} For each data set, the
-    orange points show the percentage of genotypes in which predictions
-    had both similar CE ($JS_{CBN, MHN} \le 0.0677$) and different CE-TD
-    ($(1/2)\ (JS_{CBN, CBN\_td} + JS_{MHN, MHN\_td}) \ge 0.7$),  the
-    blue dots those with similar TD ($JS_{CBN\_td, MHN\_td} \le 0.0677$)
-    and different CE-TD, and the gray points the percentage of genotypes
-    that had similar CE, similar TD, and different CE-TD. Values shown are the percentage relative to the
-    total number of genotypes for which we computed predictions, which are
-    the genotypes observed in the samples, except for the genotype with
-    all genes mutated,
-    if it was observed. In parentheses, after the data set, we indicate
-    the total number of genotypes observed (GT).  % For a given data set, genotypes that have boot similar CE and
-    % similar TD contribute to both counts.  Panels B) and C): scatter plots
-    % of Similar TD and Similar CE, shown in panel A, against sample
-    % size.
-  }
-
-  \label{fig:biol_freqs} \end{figure}
+\bibliographystyle{mystyle}
+%\bibliographystyle{unsrt}
+\bibliography{refs}
 
 \end{document}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%% Old results stuff
-
-% So: when we can achieve good performance depends crucially on
-% balblaba. (Give one or two). And when we should used one type of method
-% vs. other too (blablba)
-
-% How is this relevant for real data sets, for which we know nothing of
-% those characts? We have examined if this disyuntive could actually be
-% relevant. And it is. Blabla.
-
-% % \subsection{Results trees, II} \label{res2}
-
-% % \begin{enumerate}
-% % \item If prop local max is large, then a simple approach like TD works
-% %   (CBN and MHN)
-% %   % \begin{enumerate}
-% %   % \item Fig 2 node 10 (2\% ), to lesser degree nodes 12, 13 (2 and 3
-% %   %   perc.)
-% %   % \item Suppl Fig 3: nodes 19 (and 20 and 21: 1 and 2 percent), node 16 (2
-% %   %   perc) and 15 (1) and 22 (5 \% ) to lessser degree
-% %   % \item Even in less than ideal situations that is the case. Suppl Fig 1:
-% %   %   nodes 24 and 25 (with 1 and 5 \% )
-% %   % \item Suppl Fig 2, nodes 18 and 22 (1 and 1 \% )
-% %   % \item Lesser degree 16, 17, 23 (with 2, 1, 1,\% ): ehh??
-% %   % \end{enumerate}
-% % \item What about CE?
-% %   \begin{enumerate}
-% %   \item Cases where MHN is really the single best method are rare
-% %   \item When CE can do well (JS < 0.2) most methods about equially good
-% %     (except CAPRI), though CBN generally best (but recall simul scenarios
-% %     might favor CBN). So when good CE performance, then pretty much any of
-% %     the reasonable methods (all except CAPRI) will work. OT not the best,
-% %     and MHN sometimes not best.
-% %   \item CE is consistently gopod only in limited cases
-% %     \begin{enumerate}
-% %     \item Fig2: nodes 6, 7, 16, 20 (with 4, 1, 2, 6\% ); in node 16 MHN is
-% %       poor
-% %     \item Suppl Fig 1: Maybe node 6, 12, 16, 17 (with 1, 1, 2, 2\% data )
-% %     \item Suppl Fig 2: nodes 10 and maybe 9 (sum of 2\% )
-% %     \item Suppl Fig 3: node 23 (5\% ). OJO AQUI: probablemente escamotea
-% %       que tenemos mezcla de buenos TD y buenos CE, como el nodo 19 en la
-% %       figura del paper.
-% %     \end{enumerate}
-% %   \end{enumerate}
-% % \end{enumerate}
-
-% \subsection{Results: connect simuls-biol} \label{biol1}
-
-% \begin{enumerate}
-% \item If TD is good, the td methods are similar
-% \item If CE is good, the CE methods are similar
-% \item The converse is, of course, not true
-% \item Particularly the case if TD and CE are very different.
-% \item Refer to section ``Categorization of similarities between methods and
-%   performance'' in suppl mat (i.e., use that info judiciously in the
-%   paper, citing key numbers)
-% \item Thus, we analyze biol data and focus on similar TD, similar CE,
-%   dissimilar CE-TD: how often is this seen in the data?
-% \end{enumerate}
-
-% The fitness-landscape specific variables are
-
-% - SSWM regime
-% -  $\gamma$
-% - (epistRSign)
-% - numObservedPeaks
-
-% - detection regime
-% - sample\_size)
-
-% - nMut
-% - fitnessRank
-% - propLocalMax
-
-% - observedProp
-% - diff\_obs\_prop
-
-% epistRSign can flip which method is best: Node 11 in
-% Fig.~\ref{S-fig:7_ALL}; smaller is better (also node 23 in the figure)
-
-% numObservedPeaks can flip which method is best: Node 19 in
-% Fig.~\ref{fig:7_spec}
-
-% , ; notice that the effect of fitness landscape type
-% is not always apparent in these trees, but three of the variables included
-% in the fits, and that appear in the fitted trees, gamma, reciprocal sign
-% epistasis, and number of observed peaks,
-
-% The fitted trees show (see supporting information, Figures \todo{blabla})
-% show that \todo{blabla: sample size, detection regime}. The effect of
-% fitness landscape type is not \todo{always, often} apparent in these
-% trees, but three of the variables included in the fits, and that appear in
-% the fitted trees, gamma, reciprocal sign epistasis, and number of observed
-% peaks, allow to perfectly separate the RMF from the rest of the fitness
-% landscape, and to separate most local maxima fitness landscapes from the
-% representable ones (see figure \todo{gamma-rsign-obs-peaks-FL.pdf} in the
-% supplementary material).
-
-% The results for the tree fitted to seven genes, uniform, sample size =
-% 4000, show \todo{blablabla}.
-
-% Use the 7 genes, specific.
-
-% number of peaks, ruggeddness of landscape matters: see node 28 vs. 29. Not
-% unlike the pattern of 14 (here max fitness, so can only jump to end). node
-% 28 also kinda matches node 21 in the 10 genes.  ode 28 is smooth, so it
-% jumps to 7 mutations, it is often the single option. Note also node
-% 15. And note that these cases add up to close to 8\% of observations
-
-% Very obvious two different groups of methods, that, where it matters,
-% either do a very good or a very bad job (with the other group doing a very
-% bad or very good job)
-
-% To prevent detection regime and sample size from hindering
-% interpretation, we have fitted the models to the best performing
-% conditions, sample size of 4000 and uniform detection regime.
-
-% to prevent the effects of sample size and detection regime from
-% obscuring key patterns.
-
-% Our main objective is understanding the relevance of fitness
-% landscape-specific and genotype-specific variables for on the quality of
-% predictions. Thus, we have fitted the models to the best performing
-% conditions, sample size of 4000 and uniform detection regime, to prevent
-% the effects of sample size and detection regime from obscuring key
-% patterns.
-
-% The linear mixed-effects model trees fitted to the complete data sets for
-% 7 and 10 genes showed that sample size and detection regime had a strong
-% effect on performance (see Supplementary Material, Figures fig\_10\_ALL
-% and fig\_7\_ALL), with better performance achieved under the uniform
-% detection regime and with the largest sample size (4000) (see also Table
-% tab\_minjs in Supplementary material); there are interactions,
-% blablaba. We focus on something else.
-
-% For both 7 and 10 genes, JS was smallest in the uniform detection regime
-% with sample size of 4000 \todo{give numbers; do we even need this?}.
-
-% \begin{figure}[!h]
-%   \centering
-%   \includegraphics[width=15.5cm,keepaspectratio]{figs/fraction_CE_TD_similar_all_datasets_combined.pdf}
-
-%   \caption{\textbf{Percentage of genotypes with Similar TD or Similar CE
-%       predictions for the 25 biological data sets.} A) For each data set,
-%     the orange dots show the percentage of genotypes in which predictions
-%     had both similar CE
-%     ($(1/3) \ (JS_{CBN, MHN} + JS_{CBN, MCCBN} + JS_{MCCBN, MHN}) \le
-%     0.0677$) and Different CE-TD
-%     ($(1/2)\ (JS_{CBN, CBN\_td} + JS_{MHN, MHN\_td}) \ge 0.7$), and the
-%     blue dots those with similar TD ($JS_{CBN\_td, MHN\_td} \le 0.0677$)
-%     and Different CE-TD. Values shown are the percentage relative to the
-%     total number of genotypes for which we computed predictions, which are
-%     the genotypes observed in the samples, except for the last genotype,
-%     if it was observed. In parentheses, after the data set, we indicate
-%     the total number of genotypes observed (GT).  For a given data set, genotypes that have boot similar CE and
-%     similar TD contribute to both counts.  Panels B) and C): scatter plots
-%     of Similar TD and Similar CE, shown in panel A, against sample
-%     size. }
-
-%   \label{fig_biol_freqs} \end{figure}
-
-%% FIXME Place in Suppl Mat
-% \begin{figure}[!h]
-%   \centering
-  % \includegraphics[width=15.5cm,keepaspectratio]{figs/fraction_CE_TD_similar_all_datasets_sample_size.pdf}
-  % \caption{\textbf{Percentage of genotypes with Similar TD or Similar CE
-  %     predictions for the 24 biological data sets: relationship to sample size.} For each data set,
-  %   the orange dots show the percentage of genotypes in which predictions had both similar CE
-  %   ($(1/3) \ (JS_{CBN, MHN} + JS_{CBN, MCCBN} + JS_{MCCBN, MHN}) \le
-  %   0.2$) and Different CE-TD ($(1/2)\  (JS_{CBN, CBN\_td} + JS_{MHN,
-  %     MHN\_td}) \ge 0.6$), and the blue dots those with similar TD
-  %   ($JS_{CBN\_td, MHN\_td} \le 0.2$) and Different CE-TD. Values shown
-  %   are the percentage relative to the total number of genotypes for which we
-  %   computed predictions, which are  the genotypes observed in the
-  %   samples; the total number of genotypes observed is indicated to the
-  %   right of the data set name as ``(n = x)''. For a given data set, the
-  %   sum of the two numbers can be greater that 100, since genotypes that
-  %   have boot similar CE and similar TD contribute to both counts.
-  %    }
-
-  % \label{fig_biol_freqs_sampsize} \end{figure}
-
-%% From analyze-biol-data-3.R
-
-% Figures \ref{S-fig:7_ALL} and \ref{S-fig:10_ALL} show the models
-% fitted to all detection regimes and sample sizes; these figures showed the
-% relevance of detection regime and sample size, with the smallest JS were
-% generally achieved under the uniform detection regime and with the largest
-% sample size (4000) (see also Supplementary material, Table
-% \ref{S-tab_minjs} in section \nameref{S-suppl_minimal_js_ssize_det}).
-% Moreover, there were interactions between detection regime and sample size
-% on the one hand and genotype-specific and fitness landscape-specific
-% variables. These interactions can make it harder to understand the
-% relevance of fitness landscape-specific and genotype-specific variables on
-% the quality of predictions,
-
-% . In the Supplementary Material
-% (Fig.~\ref{S-fig:10_ALL}, section
-% \ref{S-suppl-linear-mixed-model-results}) we show the results for 10
-% genes.  These figures showed the relevance of detection regime and sample
-% size, with the smallest JS were generally achieved under the uniform
-% detection regime and with the largest sample size (4000) (see also
-% Supplementary material, Table \ref{S-tab_minjs} in section
-% \nameref{S-suppl_minimal_js_ssize_det}). Moreover, there are interactions
-% between detection regime and sample size on the one hand and
-% genotype-specific and fitness landscape-specific variables. These
-% interactions can make it harder to understand the relevance of fitness
-% landscape-specific and genotype-specific variables on the quality of
-% predictions, which is one of the main objectives of this paper. Thus, we
-% have fitted the linear mixed-effects model trees under the best performing
-% conditions, sample size of 4000 and uniform detection regime, to isolate
-% the effects of fitness landscape and genotype-specific variables; results
-% are shown in Figure \ref{S-fig:7_spec} for fitness landscapes with seven
-% genes, and in the Supplementary Material (Figure fig\_10\_spec) for
-% fitness landscapes with 10 genes.
-
-% The smallest JS were generally achieved under the uniform detection regime
-% and with the largest sample size (4000) (see Supplementary material,
-% ``Minimal JS for combinations of number of genes, sample size, and
-% detection region'') %%Table tab\_minjs in
-% a pattern that can also be observed in the linear
-% mixed-effects model trees fitted to the complete data sets for 7 and 10
-% genes (Supplementary Material, ``Linear mixed-effects model trees for
-% complete data''). These figures also showed the presence of interactions
-% between these two variables and other genotype-specific and fitness
-% landscape-specific variables. These interactions can make it harder to
-% understand the relevance of fitness landscape-specific and
-% genotype-specific variables on the quality of predictions, which is one of
-% the main objectives of this paper. Thus, we will focus below on models
-% fitted under the best performing conditions, sample size of 4000 and
-% uniform detection regime, to isolate the effects of fitness landscape and
-% genotype-specific variables; results are shown in Figure \ref{S-fig:7_spec}
-% for fitness landscapes with seven genes, and in the Supplementary Material
-% (Figure fig\_10\_spec) for fitness landscapes with 10 genes.
-
-%  of this same model nor
-% in node 12 in Fig.~\ref{S-fig:10_ALL} since those nodes split genotypes landscapes
-% that are not representable ().
-
-% Ojo: nodo 19 does not split con los de 1 peak, pero es 1 o 2, on the left!
-% Some of these results are easy to understand. For example, in
-% Fig.~\ref{fig:7_spec} the relevance of numObservedPeaks in node 19 for
-% determining whether TD methods are better than their CE counterparts. Node
-% 19 splits genotypes with 6 mutations (in fitness landscapes of 7
-% genes). Now, if there is only peak, we are in a representable fitness
-% landscape, and the next transition will necessarily be to the genotype
-% with 7 mutations; the TD methods, that can predict that the genotype with
-% 6 mutations is a local maximum will therefore fail. NOPE not that simple:
-% a) this is 1 or 2 on the left, so not just representable; b) notice
-% CAPRI's behavior.
-
-% \idea{Ya no tengo nada claro que mostrar la figura que mostramos tenga
-%   sentido. Podríamos mostrar solo la Suppl 1 y Suppl 2, en el main text.
-%   No tengo claro que esté bien justificado (por que centrarse en los
-%   mejores?) ni que ayude a clarificar la discusion. O lo ayudaría al
-%   revés: mostar el Suppl Fig 1 en resultados, lo demças en suppl mat}
-
-% \idea{La discusion podria ser muy distinta: complex interactions; SSWM,
-%   rank, and other genotype specific matter. Done}
-
-% They can be very
-% different, but we still would need to choose. How? Using info we do not
-% have about SSWM, fitness landscape, etc.
-
-% Fig.~\ref{fig:biol_freqs} shows blablba \idea{add to that
-%   plots those genotype where all three conditions are true; } So the CE
-% can be similar, the TD can be similar: should we use them? When we observe
-% similar CE and good performance depends on features we do not see; ditto
-% for TD. And, even more dramatially, what if the same genotype shows both
-% patterns?  They can be very different, but we still would need to
-% choose. How? Using info we do not have about SSWM, fitness landscape, etc.
-
-% since Fig.~\ref{fig:biol_freqs} based on the necessarily true conclusion
-% that CBN and MHN can only both make correct predictions if both are making
-% similar predictions, and similarly for CBN\_td and MHN\_td. Reversing the
-% conditioning, however, does not guarantee a large improvement in
-% performance.
-
-% Using this simpleminded approach, we can in fact improve the
-% performance, relative to baseline levels, if we condition on similar CE
-% (or similar TD) but different CE-TD, but that improvement is rather modest
-% \idea{blablaba ``Categorization of similarities between methods and
-%   performance'' in suppl mat, and this is overestimate of performance,
-%   etc, etc}. We can try to restrict the analysis to such
-% conditions. Fig.~\ref{fig:biol_freqs} shows blablba \idea{add to that
-%   plots those genotype where all three conditions are true; } So the CE
-% can be similar, the TD can be similar: should we use them? When we observe
-% similar CE and good performance depends on features we do not see; ditto
-% for TD. And, even more dramatially, what if the same genotype shows both
-% patterns?  They can be very different, but we still would need to
-% choose. How? Using info we do not have about SSWM, fitness landscape, etc.
-
-% \subsection{Results biol} \label{biol01}
-
-% \begin{enumerate}
-% \item In figure  \ref{fig:biol_freqs}
-% \item Details for each data set (similarity CE, similarity TD, difference
-%   CE-TD) in suppl mat (``Similarity of predictions for all genotypes for the biological
-%   data sets'').
-% \item Many cases where a large percentage genotypes very simlar TD or very
-%   similar CE and very different CE to TD
-% \item Wide variation among data sets
-% \item Not related to sample size (panels B and C)
-% \item Some of those preds could correspond to good predictions but telling
-%   would require knowing more.
-% \end{enumerate}
-
-\section{Como hacer el conectar con datos biol: miscel, other ideas, what
-  not to do, etc}
-\label{sec:como-hacer-el}
-
-Some options
-
-\begin{enumerate}
-\item Classify all output from simuls in four groups
-  \begin{itemize}
-  \item CE w.r.t. real approx 0, TD approx 1
-  \item TD w.r.t. real approx 0, CE approx 1
-  \item All $> 0.2$
-  \item All $< 0.2$
-  \end{itemize}
-  \begin{itemize}
-  \item But that sucks because we have a multi-class classification
-    problem and that is a PITA with mixed effects
-    models, etc.
-\item Moreover, this
-  would be circular thing (as it is based on previously looking at the
-  data with the other trees). 
-\item And of course we don't have anything similar with real data.
-
-  \end{itemize}
-\item Look at similarities between methods. What we didd with the other
-  trees. That nothing came out, is it a result? Nope, since this is
-  actually dealing with another issue. And it could be explored in much
-  more detail. This ain't the point of the paper.
-\item Focuse on some transitions/cases: like the genotype before
-  last. E.g., node 19, 20, 21 in Fig 2. Pero eso es coger la cosas por los
-  pelos. Compatre with node 23 in Suppl Fig 3. 
-\item Try to do a principled search, over biological data, of regions with
-  CE vs TD huge, within-CE small and within-TD small.
-
-  Tghis requires defining things in a principled way. So the whole mess of
-  a PCA if descriptive, PLS-like for predictive, etc. And we are back,
-  like above, in another issue. Not what we are dealing with.
-\item Focus on the key idea: choice of method/quality of predictions ought
-  to depend on details of evol model etc that we ignore. Show how this
-  matters. Show that it matters (Fig 2 of the paper) and show that there
-  are many cases in biol data that likely fall there. \textbf{This is the
-    connecting logic}.
-
-  Emphasize: \textbf{we are not focusing on choice of single best method,
-    just in seeing what is possible. Choice of best is a different
-    question}.
-
-  In results go over what is in \ref{res2}. Then, say:
-  \begin{enumerate}
-  \item If TD is good, the td methods are similar
-  \item If CE is good, the CE methods are similar
-  \item The converse is, of course, not true
-  \item Refer to section ``Categorization of similarities between methods and
-  performance'' in suppl mat (i.e., use that info judiciously in the
-  paper, citing key numbers)
-\item Now say: we will focus on cases in the biological data where the CEs
-  are similar and the TDs are similar, as those can be good.
-\item In the biological data, show statistics, for the data sets, of how
-  often we have similar TD + dissimular CE-TD and similar CE + dissimilar
-  CE-TD. Comment that ``choosing the best method and, thus, choosing
-  predictions depends on key featuyres fo the evol model''. This is figure \ref{fig:biol_boxplots-3-comps}
-
-  Show figure where it can be seen that we have used a whole bunch of data
-  sets.
-
-  We are done.
-
-  \end{enumerate}
-
-  \item Es tentador hacer más modelos de árboles con solo los datos
-    simulados con
-    buena predicción (JS $le 0.2$), y hacer modelo logístico para separar:
-    ``los mejores son los TD'' vs.\ ``los mejores son los CE''. Pero esto:
-    \begin{itemize}
-    \item Es una pregunta distinta y el paper already too long
-    \item Añade más tiempo de análisis.
-    \item No está claro cómo se conecta con los datos biológicos.
-    \item ¿Merece la pena?
-    \end{itemize}
-
-\item Es tentador hacer, con los datos simulados, una figura del tipo de
-  \ref{fig:biol_freqs}. Dividiendo por type of landscape y por sampling
-  type y quizas sample size. Pero:
-  \begin{enumerate}
-  \item Más tiempo análisis
-  \item Alarga paper
-  \item Delicate issue de los pesos
-  \item Mejora la conexion pero no sé si merece la pena
-  \end{enumerate}
-
-\end{enumerate}
-
-\idea{Do not forget to mention inherent variability of CBN (and MCCBN)
-  which is very noticeable when relatively large number of variables, as
-  has been noted before (Diaz-Uriarte and Vasallo, 2019; Hosseini et al.,
-  2019). Ellaborate on this, etc, etc. Hosseini develops his strategy for
-  this. With real data might one several runs for predicting genotye? etc,
-etc. To be explored in further papers.}
-
-\idea{Beware: we do not want anyone asking for cross-validatino of
-  performance estimates! If they ask, we will say that outside of the
-  scope, and a moot point anyway. Suppl mat tables present upper limits on
-what is already modest performance}
-
-Arriba en resultados explicar porqué uno se queda con cualquier genotipo
-para el que hay rpedcciones de al menos 3. Uno de esos 3 siempre es MHN.
-
-%%%%%%%%%%%%%%%%%%%%% Former biological data stuff
-
-%  (note that being local maxima is often associated with having
-% large relative fitness ---i.e., small fitness rank--- and local maxima
-% genotypes are more common in fitness landscapes with a larger number of
-% local fitness maxima)
-
-% What those results emphasize, then, is that predicting if performance will
-% be acceptable, and what method to use to achieve that performance, depends
-% on features of the fitness landscape, the evolutionary process, and the
-% genotype. This highlights the difficulties of interpreting the
-
-The results from the simulated data have direct implications for the
-interpretation of the application of these methods to biological data. As
-Fig.~\ref{fig:biol_boxplots-3-comps} shows, there are often very large
-differences in the predictions of a TD and a non-TD method (MHN vs.\
-MHN\_td, right-most panel), something we could observe if we were dealing
-with genotypes of types A) or B) above. But we have no solid grounds to
-choose one method over the other, as that choice depends on information
-about the evolutionary model and the genotype itself that we
-ignore. Fig.~\ref{fig:biol_freqs} emphasizes this point: a tempting
-approach when analyzing data would be to focus interpretation on genotypes
-for which we see very dissimilar CE-TD predictions and at the same time
-either similar TD or similar CE (or both).  This conditioning, by itself,
-is unlikely to yield relevant improvements in performance (see Tables
-\ref{S-tab:xtabs_td0} to \ref{S-tab:xtabs_ce1}). And it leaves unanswered
-the question of which among the predictions that fulfill those similarity
-conditions we should trust. The difficulty of this choice is even more
-dramatic when the same genotype shows both large similarity within TD
-methods, large similarity within CE methods, and large dissimilarity
-between the CE and TD methods, as indicated by the gray points in
-Fig.~\ref{fig:biol_freqs}: predictions between families of methods are
-very different and yet choice of family of method would require fitness
-landscape-specific and genotype-specific information.
-
-\subsection{Analysis of biological data sets} \label{sec:biol-data}
-
-\idea{algunas cosas de debajo repetir o enfatiza más en discusión y menos aquí}
-
-% Do the above results matter for analysis of biological data? Yes they
-% do.
-When analyzing biological data, even if we do not know the true transition
-probabilities, we can examine the consequences of assuming that we choose
-one method when another method is the one that gives the best (or even
-perfect) predictions. This is what we do in
-Fig.~\ref{fig:biol_boxplots-3-comps}, where we use the four best
-performing methods to illustrate the consequences of choosing the wrong
-method. The left panel, for example, shows the difference in predictions
-(measured using JS) if the method that gave perfect predictions were CBN
-but we had used MHN (or viceversa).  As can be seen, for most data sets,
-there is wide variability in the consequences of using the wrong method,
-and often the difference in predictions is very large, specially if we use
-a TD vs.\ a non-TD method (MHN vs.\ MHN\_td, right-most
-panel).% And most of the time we have no
-% solid grounds to choose one method over the other, as that choice depends
-% on information about the evolutionary model and the genotype itself that
-% we ignore.
-
-% \idea{}The 25 biol data sets show ranges of variation
-% comparable to those from the simulated data \idea{DO THIS PLOT for suppl
-%   mat}.
-
-\idea{algunas cosas de debajo mover a discusion: ``more dramatic'', blablablba}
-
-% A naive approach might try to leverage on the similarity between methods, 
-% as we have seen % :
-% Fig.~\ref{fig:7_spec}, as well as Figures~\ref{S-fig:10_spec},
-% \ref{S-fig:7_ALL} and \ref{S-fig:10_ALL}, indicate
-We have seen, with the simulated data, that there is a sharp difference in
-behavior of TD and CE methods when the best achievable performance is
-good.  And necessarily CBN and MHN can only both make correct predictions
-if both are making similar predictions; likewise, CBN\_td and MHN\_td can
-only both make correct predictions if both are making similar prediction.
-A simple approach could try to limit analyses to genotypes that show very
-distinct CE-TD patterns and either  similar CE patterns, or similar
-TD patterns. Note that conditioning in these similarity patterns could
-bring at most very modest improvements in performance (see Tables
-\ref{S-tab:xtabs_td0} to \ref{S-tab:xtabs_ce1}).
-Fig.~\ref{fig:biol_freqs} shows the fraction of genotypes, for each data
-set, that meet fairly stringent conditions of similarity either within the
-CE methods (i.e., between CBN and MHN) or within the TD methods (i.e.,
-between CBN\_td and MHN\_td). (See Supplementary Material, figures in
-Section \ref{S-sm-indiv-biol-data-plots}, for similarity for each
-individual genotype for each data set; note that these patterns are not
-related to sample size ---see Fig.~\ref{S-fig:biol_freqs_sampsize}). There
-is wide variation among biological data sets for the fraction of genotypes
-that meet these criteria, but for several of the data sets, there are more
-than 20\% of the genotypes that fall in either of these categories,
-including genotypes for which all three criteria (dissimilar CE-TD,
-similar TD, similar CE) are met.
-
-% A naive approach might try to leverage on the similarity between methods, 
-% as we have seen % :
-% % Fig.~\ref{fig:7_spec}, as well as Figures~\ref{S-fig:10_spec},
-% % \ref{S-fig:7_ALL} and \ref{S-fig:10_ALL}, indicate
-% that there is a sharp
-% difference in behavior of methods when the best achievable performance is
-% good. And necessarily CBN and MHN can only both make correct predictions
-% if both are making similar predictions; likewise, CBN\_td and MHN\_td can
-% only both make correct predictions if both are making similar prediction.
-% But, by itself, conditioning in these similarity patterns within CE
-% methods and within TD methods could bring at most very modest improvements
-% in performance (see Tables \ref{S-tab:xtabs_td0} to
-% \ref{S-tab:xtabs_ce1}). Fig.~\ref{fig:biol_freqs} shows the fraction of
-% genotypes, for each data set, that meet fairly stringent conditions of
-% similarity either within the CE methods (i.e., between CBN and MHN) or
-% within the TD methods (i.e., between CBN\_td and MHN\_td). (See
-% Supplementary Material, figures in Section
-% \ref{S-sm-indiv-biol-data-plots}, for similarity for each individual
-% genotype for each data set; note that these patterns are not related to
-% sample size ---see Fig.~\ref{S-fig:biol_freqs_sampsize}). There is wide
-% variation among biological data sets, and for several of the data sets,
-% there is a large fraction of genotypes with very simlar TD or very similar
-% CE and very different CE to TD. Which among these predictions could we
-% trust?  The difficulty of this choice is even more dramatic when the same
-% genotype shows both large similarity within TD methods, large similarity
-% within CE methods, and large dissimilarity between the CE and TD methods,
-% as indicated by the gray points in Fig.~\ref{fig:biol_freqs}: predictions
-% between families of methods are very different and yet choice of family of
-% method would require fitness landscape-specific and genotype-specific
-% information, as we have seen in Fig.~\ref{fig:7_spec}, as well as
-% Figures~\ref{S-fig:10_spec}, \ref{S-fig:7_ALL} and \ref{S-fig:10_ALL}.
-
-\subsubsection*{SSWM}
-
-Deviations from SSWM are known to undermine the predictive power of CPMs \cite{diaz-uriarte2018,diaz-uriarte2019a}.
-The model underlying most CPMs seems to implicitly assume SSWM mainly for two reasons: \textit{a)} CPMs are powered with cross sectional data that do not undergo any deconvolution step (i.e.
-inference of clonal composition of the sequenced tumor), implying that a unique genotype\todo{``clone genotype"?} can be obtained from bulk sequencing, and \textit{b)} those observed clones are considered true steps in the path to tumor fixation, thus neglecting the possibility of tunneling effects.
-The two are true only if there is SSWM, in which case the most fit genotype at any given time is effectively the only one in the population and a direct ancestor of the ultimately fixated one.
-Even when CPMs specifically include error estimations (e.g.
-in Gerstrung et al., 2009\cite{Gerstung2009}), these typically refer to sequencing error but not necessarily to deviations from SSWM.
-
-\iffalse
-Most CPMs assume that cancer progresses under conditions of strong
-selection and weak mutation (SSWM).\cite{hosseini2019a}\todo{I do not
-find where Hosseini et al say that. Reword along:}
-\idea{The model underlying most CPMs seems to
-  implicitly assume SSWM: poorly written now, but a couple of key ideas:
-  observed data are used directly, without any deconvolution step, which
-  seems to indicate that even bulk sequencing can obtain a ``clone
-  genotype''; this would only work if SSWM; and then those observed clone
-  genotypes are considered true steps in the path to the maximum (i.e., no
-  tunneling); note that the error included in models, e.g., Gerstung et
-  al., 2009, is designed to capture sequencing error, but not necessarily
-  deviations from SSWM ---slightly more subtle in MCCBN, though. Reword
-  this properly, move stuff out to discussion, and leave here a simple
-  sentence. One part of that sentence could be ``This assumption is rarely
-  stated explicitly, but it is explicit in how \cite{hosseini2019a} use
-  CBN for predicting tumor evolution.''  } \idea{Recheck MHN paper for
-  SSWM}.
-\fi
-
-\idea{  One part of that sentence could be ``This assumption is rarely
-  stated explicitly, but it is explicit in how \cite{hosseini2019a} use
-  CBN for predicting tumor evolution.''  } \idea{Recheck MHN paper for
-  SSWM}.
-
-this number has an inverted U-shape with the
-number of mutated loci (for a genotype $i$, the number of non-zero
-$p_{i,j}$ is ${N}\choose{n + 1}$, where $n$ is the number of mutations of
-the genotype and $N$ the total number of loci).
-
-When we are interested in predicting the complete evolutionary path the
-only features that matter are global features.
- But when we want to make conditional, short-term
-predictions, features of the genotypes for which we make the conditional
-predictions can be as relevant as, or more relevant than, fitness
-landscape-specific features \idea{citar banks y quizas otros}. This is a
-noteworthy difference because it suggests that CPMs could be leveraged to
-provide short-term predictions even when global, long-term predictions are
-difficult because fitness landscape-specific assumptions are violated.
-
-Our results indicate that,
-were we to know a genotype can be a local maximum, specially if SSWM
-holds, then we could make
-
-Los resultados de que a veces algunos tipos de métodos dan valores muy
-similares enter sí, pero muy distintos a otros, no son solo cosa de las
-simulaciones sino tb. en datos empírciso; ojo entonces con la
-interpretación porque depende de factores que no conocemos.  \idea{plot el
-  MHN vs MHN\_td para simulados?  sea como
-  \ref{fig:biol_boxplots-3-comps}?}
-
-\idea{remember that the local
-maxima scenarios could slightly favor CBN, since the local maxima
-landscapes are built as deviations from the CBN, and in part from OT/CAPRESE ).Also representable? What is the model for MHN?}
-
-\idea{Null model no está. Either mention in results, when we mention
-  results for number of mutations. Or mention here. I think that to keep
-  ms short, null model figures should go in suppl mat, unless null model
-  becomes a more important part of the ms.}
-
-% \begin{figure}[H]
-%   \centering
-%   \fcolorbox{black}{white}{\includegraphics[width=0.8\linewidth,trim={4.5cm 10cm 4cm 9cm}]{figs/figexp}}
-%   \internallinenumbers
-%   \caption[Figure title]{\textbf{Figure title.} Figure caption. \idea{Esta
-%     figura tapa texto encima}}
-%   \label{figexp}
-% \end{figure}
-
-% The effects of the sampling and
-% fitness landscape characteristics in predicting short-term evolution (as
-% summarized in the linear mixed-effects trees (Fig.~\ref{fig:7_spec} and
-% Supplementary Material Figs.~\ref{S-fig:10_spec}, ~\ref{S-fig:7_ALL},
-% ~\ref{S-fig:10_ALL}) are, thus, not surprising. What is noteworthy is that
-% local, genotype-specific features, i.e., features of the genotypes for
-% which we make the conditional predictions, can be much more relevant than
-% fitness landscape-specific features \idea{citar banks y quizas
-%   otros}. These suggests that CPMs could be leveraged to provide
-% short-term predictions even when global, long-term predictions are
-% difficult: when we are interested in predicting the complete path
-% evolutionary path, of course, the only features that matter are global
-% features and these can often lead to poor performance
-% \cite{diaz-uriarte2019a}. Our results here, however, indicate that good
-% conditional predictions are possible even if fitness landscape-specific
-% assumptions are violated.
-
-% \idea{NO! Reword as}
-
-% The effects of the sampling and
-% fitness landscape characteristics in predicting short-term evolution (as
-% summarized in the linear mixed-effects trees (Fig.~\ref{fig:7_spec} and
-% Supplementary Material Figs.~\ref{S-fig:10_spec}, ~\ref{S-fig:7_ALL},
-% ~\ref{S-fig:10_ALL}) are, thus, not surprising.
-
-#####################   SSWM
-
-it is widely accepted that intra-tumor heterogeneity is
-common in cancer \cite{mcgranahan_clonal_2017, hinohara2019, zhao2016},
-which can be partly due to deviations from SSWM arising from clonal
-interference and stochastic tunneling
-\cite{devisser_empirical_2014,szendro_predictability_2013,sniegowski_beneficial_2010};
-for example, in multiple myeloma ``heteroeneoys clonal mixtures with
-shifting predominant clones'' have been described
-\cite[][p.~1067]{keats2012} %% could give example of Goto et al., 2018
-                            %% lung cancer
-suggesting the existence of non-SSWM regimes,
-and stochastic tunneling in
-cancer has, itself, been the focus of theoretical work
-\cite[e.g.,][]{Ashcroft2015,wodarz_dynamics_2014}.
-% deviations from SSWM and phenomena such as clonal
-% interference and stochastic tunneling (due to rugged fitness landscapes
-% where epistasis is common) are frequent
-% \cite{devisser_empirical_2014}\todo{more refs; in fact, this is not a
-%   reference about cancer}. \idea{search for refs of SSWM and tunneling in
-%   cancer; there is one by Michor and see also the book by Wodarz and
-%   komarova}.
-
-it is widely accepted that intra-tumor heterogeneity is
-common in cancer \cite{mcgranahan_clonal_2017, hinohara2019, zhao2016},
-which can be partly due to deviations from SSWM arising from clonal
-interference and stochastic tunneling
-\cite{devisser_empirical_2014,szendro_predictability_2013,sniegowski_beneficial_2010};
-for example, in multiple myeloma ``heteroeneoys clonal mixtures with
-shifting predominant clones'' have been described
-\cite[][p.~1067]{keats2012} %% could give example of Goto et al., 2018
-                            %% lung cancer
-suggesting the existence of non-SSWM regimes,
-and stochastic tunneling in
-cancer has, itself, been the focus of theoretical work
-\cite[e.g.,][]{Ashcroft2015,wodarz_dynamics_2014}.
-% deviations from SSWM and phenomena such as clonal
-% interference and stochastic tunneling (due to rugged fitness landscapes
-% where epistasis is common) are frequent
-% \cite{devisser_empirical_2014}\todo{more refs; in fact, this is not a
-%   reference about cancer}. \idea{search for refs of SSWM and tunneling in
-%   cancer; there is one by Michor and see also the book by Wodarz and
-%   komarova}.
-
-When we move away from SSWM, multiple clones can coexist at significant
-fractions for prolonged periods of time, and new higher-fitness clones can
-evolve from low fitness parents that represented a minority of the
-population. Genotypes with a large number of
-mutations %high mutational load
-can already be present at the time of sampling yet go undetected if their
-relative abundance is small (figure
-\ref{fig:fig1}D). % Deviations from SSWM
-% are known to undermine the predictive power of CPMs
-% \cite{diaz-uriarte2018,diaz-uriarte2019a}, as the model underlying most
-% CPMs seems to assume SSWM (see Discussion). This assumption, however, is
-% not always stated
-% explicitly.
-% , but it is explicit, for example, in Hosseini et al.'s
-% \cite{hosseini2019a} use of CPMs for predicting tumor evolution (see
-% \nameref{discussion}).
-
-Widely accepted that ITH common in cancer \cite{mcgranahan_clonal_2017}
-and there are examples that suggest presence of non-SSWM regimes
-\cite{keats2012} with ``heteroeneoys clonal mixtures with shifting
-predominant clones'' and stochastic tunneling has, itself, been the focus
-of theoretical work \cite{Ashcroft2015}
-
-\cite{bauer2015}: yeap, cite this for predict in
-non-SSWM
-
-###############################################################
-
-%%% Local Variables:
-%%% mode: latex
-%%% TeX-master: t
-%%% End:

### 2021-04-26 draft/draft.tex
adding methods to main text

@@
 }
-\renewcommand{\thefigure}{\textbf{S\arabic{figure}}} % preceed the Suppl Mat figures with S
-\renewcommand{\thetable}{\textbf{S\arabic{table}}} % preceed the Suppl Mat tables with S
-\renewcommand{\theequation}{S\arabic{equation}} % preceed the Suppl Mat equations with S
 \captionsetup[figure]{font=small, labelfont={bf}, labelformat={default},
@@
 
+\subsection{Simulations}\label{methods:sim}
+
+We used the Community Simulator package \cite{Marsland2020} and included new
+features for our simulations. In the package,
+species are characterized by their resource uptake rates ($c_{i\alpha}$ for
+species $i$ and resource $\alpha$), and they all
+share a common metabolic matrix $\mathbf{D}$.
+The element $D_{\alpha\beta}$
+of this matrix represents the fraction of energy in the form of resource $\alpha$
+secreted when resource $\beta$ is consumed.
+Here we implemented a new operation mode
+in which species can secrete different metabolites (and/or
+in different abundances) when consuming a same resource. Experimental observations
+support the idea of distinct species producing different sets of byproducts when
+growing in the same primary resource \mr. We let $D_{i\alpha\beta}$ denote the
+fraction of energy in the form of resource $\alpha$ secreted \textit{by species
+i} when consuming resource $\beta$ ---note that now $D_{i\alpha\beta}$ need not be
+equal to $D_{j\alpha\beta}$ if $i \neq j$, unlike in the original Community
+Simulator. In the package's underlying Microbial Consumer Resource Model
+\cite{Goldford2018,Marsland2019}, this just means that the energy flux
+$J^{out}_{i\beta}$ now takes the form
+
+\begin{equation}
+J^{out}_{i\beta} = \sum_\alpha D_{i\beta\alpha} l_\alpha J^{in}_{i_\alpha}
+\label{eq:jout}
+\end{equation}
+%
+The documentation for the Community Simulator package contains a detailed
+description of the model, parameters and package use. For the updated package with
+this new functionality, see \nameref{datacode}.
+
+For the simulations here,
+we first generate a library of 660 species (divided into three specialist
+families of 200 memebers each
+and a generalist family of 50 members)
+and 30 resources (divided into three classes of 10 members each).
+We randomly sample 50 species from each pool in equal ratios to seed
+100 resident and
+100 invasive communities respectively.
+We then grow and dilute the communities serially,
+replenishing the primary
+resource after each dilution.
+We repeat the process 20 times to ensure generational equilibrium is
+achieved.
+
+\section{Data \& code availability}\label{datacode}
+
+Experimental data and code for the analysis, as well as code for the simulations
+and the updated Community Simulator package with instructions to use the
+new features can be found in \url{github.com/jdiazc9/coalescence}.
+
 \section{Figures}\label{figs}

### 2021-04-27 draft/draft.tex
more methods in draft

@@
 
-\newcommand\legendtreefigs{Internal nodes show the partitioning variables   and the values that lead to each partition; light blue denotes the   values or categories that lead down the left split and salmon the values   or categories that lead down the right split (partitioning criteria are   also shown in the edges leading down a node).
-Leaves or terminal nodes   show (weighted) boxplots of the dependent variable for the observations   that belong that fall in that leave and, as label, the name of the node   ax_3,nd the sample size (sum of weighted observations and \% of the total   data set for the fit).
-The dot plots underneath the boxplots show the mean JS for all the methods for that leave.}
-
 \begin{document}
@@
 
+\subsection{Stabilization of environmental communities in simple synthetic environments}
+\label{methods:community-assembly}
+
+Communities were stabilized \textit{ex situ} as described in \cite{Goldford2018}.
+In short, environmental samples (soil, leaves...) within one meter radius in eight different
+geographical locations were collected with sterile
+tweezers or spatulas into 50mL sterile tubes (Fig. \mr).
+One gram of each sample was allowed to
+sit at room temperature in 10mL of phosphate buffered saline (1$\times$PBS) containing
+200$\upmu$g/mL cycloheximide to suppress eukaryotic growth.
+After 48h, samples were mixed 1:1 with 80\% glycerol and kept frozen at -80$^\circ$C.
+Starting microbial communities were prepared by scrapping the frozen stocks into
+200$\upmu$L of 1$\times$PBS and adding a volume of 4$\upmu$L to 500$\upmu$L
+of synthetic minimal media (1$\times$M9) supplemented with 200$\upmu$g/mL cycloheximide
+and 0.07 C-mol/L glutamine or sodium citrate as the carbon source in 96 deep-well plates
+(1.2mL; VWR).
+Cultures were then incubated still at 30$^\circ$C to allow for re-growth.
+After 48h, samples were fully homogenized and biomass increase was followed by measuring
+the optical density (620nm) of 100$\upmu$L of the cultures in a Multiskan FC plate reader
+(Thermo Scientific).
+Communities were stabilized \cite{Goldford2018} by passaging 4$\upmu$L of the cultures into
+500$\upmu$L of fresh media (1$\times$M9 with the carbon source)  every 48h for a total of
+12 transfers at a dilution factor of 1:100,
+roughly equivalent to 80 generations per culture (Fig. \mr).
+Cycloheximide was not added to the media after the first two transfers.
+
+\subsection{Isolation of dominant species}\label{methods:dominants}
+
+For each community, the most abundant colony morphotype at the end of the ninth transfer
+was selected, resuspended in 100$\upmu$L 1$\times$PBS and serially diluted (1:10).
+Next, 20$\upmu$L of the cells diluted to $10^{-6}$ were plated in the corresponding synthetic
+minimal media and allowed to regrow at 30$^\circ$C for 48h. Dominants were then inoculated
+into 500$\upmu$L of fresh media and incubated still at 30$^\circ$C for 48h.
+After this period, the communities stabilized for eleven transfers and the isolated dominants
+were ready for the competition experiments (Fig \mr) at the onset of the twelfth transfer.
+
 \subsection{Simulations}\label{methods:sim}
@@
 support the idea of distinct species producing different sets of byproducts when
-growing in the same primary resource \mr. We let $D_{i\alpha\beta}$ denote the
+growing in the same primary resource \mr. We call $D_{i\alpha\beta}$ to the
 fraction of energy in the form of resource $\alpha$ secreted \textit{by species
@@
 \cite{Goldford2018,Marsland2019}, this just means that the energy flux
-$J^{out}_{i\beta}$ now takes the form
+$J^{\mathrm{out}}_{i\beta}$ now takes the form
 
 \begin{equation}
-J^{out}_{i\beta} = \sum_\alpha D_{i\beta\alpha} l_\alpha J^{in}_{i_\alpha}
+J^{\mathrm{out}}_{i\beta} = \sum_\alpha D_{i\beta\alpha} l_\alpha J^{\mathrm{in}}_{i\alpha}
 \label{eq:jout}
@@
 %
-The documentation for the Community Simulator package contains a detailed
-description of the model, parameters and package use. For the updated package with
-this new functionality, see \nameref{datacode}.
+The documentation for the Community Simulator contains detailed
+descriptions of the model, parameters and package use. For the updated package with
+the new functionality, see \nameref{datacode}.
 
-For the simulations here,
+For our simulations,
 we first generate a library of 660 species (divided into three specialist
-families of 200 memebers each
-and a generalist family of 50 members)
+families of 200 members each
+and a generalist family of 60 members)
 and 30 resources (divided into three classes of 10 members each).
+We split this library into two non-overlapping pools of 330 species each.
 We randomly sample 50 species from each pool in equal ratios to seed
@@
 We repeat the process 20 times to ensure generational equilibrium is
-achieved.
+achieved \cite{Goldford2018}.
+We then perform the \textit{in silico} experiments by using the
+generationally stable communities to seed 100 coalesced communities
+that we again stabilize as described previously.
+Similarly, we identify the dominant (most
+abundant) species of every resident and invasive community to carry out pairwise
+competition and single invasion simulations.
+Most parameters are set to the defaults of the original Community Simulator
+package. Table \mr shows those that are given non-default values to ensure
+enough variation in the primary communities.
 

### 2021-04-28 draft/draft.tex
more methods in paper

@@
 \usepackage{mathtools} % nice matrices
+\usepackage{units} % nice fractions
 \usepackage{rotating} % rotated text
@@
 \captionsetup[figure]{font=small, labelfont={bf}, labelformat={default},
-   labelsep=period, name={Fig.}} % custom figure captions
+   labelsep=period, name={Figure}} % custom figure captions
 \captionsetup[table]{font=small, labelfont={bf}, labelformat={default},
@@
 
-\section{Introduction}\label{intro}
+\section*{Introduction}\label{intro}
 
@@
 
-\section{Results}\label{results}
+\section*{Results \& Discussion}\label{results-discussion}
 
-\section{Discussion}\label{discussion}
+This is how you refer to the \nameref{intro}.
 
-\section{Methods}\label{methods}
+\section*{Methods}\label{methods}
 
-\subsection{Stabilization of environmental communities in simple synthetic environments}
+\subsection*{Stabilization of environmental communities in simple synthetic environments}
 \label{methods:community-assembly}
@@
 
-\subsection{Isolation of dominant species}\label{methods:dominants}
+\subsection*{Isolation of dominant species}\label{methods:dominants}
 
@@
 
-\subsection{Simulations}\label{methods:sim}
+\subsection*{Dominant-dominant and community-community competitions}
+\label{methods:competitions}
+
+All possible pairwise dominant-dominant and community-community
+competition experiments
+were performed by mixing equal volumes (4$\upmu$L) of each of the eight
+communities or eight dominants at the onset of the twelfth transfer.
+Competitions were set up in their native media,
+i.e. in 500$\upmu$L of 1$\times$M9 supplemented with 0.07 C-mol/L of
+either glutamine or citrate in 96 deep-well plates.
+Plates were incubated at 30$^\circ$C for 48h.
+Pairwise competitions were further propagated for seven serial transfers
+(roughly 42 generations; Fig. \mr) by transferring 8$\upmu$L of
+each culture to fresh media (500$\upmu$L).
+
+\subsection*{Determination of community composition by 16S sequencing}
+\label{methods:sequencing}
+
+The sequencing protocol was identical to that described in \cite{Goldford2018}.
+Community samples were collected by spinning down at 3500rpm for 25min
+in a bench-top centrifuge at room temperature;
+cell pellets were stored at -80$^\circ$C before processing.
+To maximize Gram-positive bacteria cell wall lysis,
+the cell pellets were re-suspended and incubated at 37$^\circ$C for 30min
+in enzymatic lysis buffer (20mM Tris-HCl, 2mM sodium EDTA, 1.2\% Triton X-100)
+and 20 mg/mL of lysozyme from chicken egg white (Sigma-Aldrich).
+After cell lysis, the DNA extraction and purification was performed using the
+DNeasy 96 protocol for animal tissues (Qiagen).
+The clean DNA in 100$\upmu$L elution buffer of 10mM Tris-HCl, 0.5mM EDTA
+at pH 9.0 was quantified using Quan-iT PicoGreen dsDNA Assay Kit
+(Molecular Probes, Inc.)
+and normalized to 5ng/$\upmu$L in nuclease-free water (Qiagen)
+for subsequent 16S rRNA illumina sequencing.
+16S rRNA amplicon library preparation was performed following a dual-index
+paired-end approach \cite{Kozich2013}.
+Briefly, PCR amplicon libraries of V4 regions of the 16S rRNA were prepared 
+sing dual-index primers (F515/R805), then pooled and sequenced
+using the Illumina MiSeq chemistry and platform.
+Each sample went through a 30-cycle PCR in duplicate of 20$\upmu$L
+reaction volumes using 5ng of DNA each, dual index primers, and AccuPrime Pfx SuperMix (Invitrogen).
+The thermocycling procedure includes a 2min initial denaturation step at
+95$^\circ$C, and 30 cycles of the following PCR scheme:
+(a) 20 second denaturation at 95$^\circ$C,
+(b) 15 second annealing at 55$^\circ$C,
+and (c) 5 minute extension at 72$^\circ$C.
+The duplicate PCR products of each sample were pooled, purified, and normalized
+using SequalPrep PCR cleanup and normalization kit (Invitrogen).
+Barcoded amplicon libraries were then pooled and sequenced using
+Illumina Miseq v2 reagent kit, which generated 2$\times$250bp paired-end reads
+at the Yale Center for Genome Analysis (YCGA).
+The sequencing reads were demultiplexed on QIIME 1.9.0 \cite{Caporaso2010}.
+The barcodes, indexes, and primers were removed from raw reads,
+producing FASTQ files with both the forward and reverse reads for each sample,
+ready for DADA2 analysis \cite{Callahan2017}.
+DADA2 version 1.1.6 was used to infer unique biological exact sequence variants
+(ESVs) for each sample
+and na{\"i}ve Bayes was used to assign taxonomy using the SILVA version 123
+database \cite{Wang2007,Quast2013}.
+
+\subsection*{Metrics of community distance}
+\label{methods:metrics}
+
+Beta-diversity indexes between the invasive and coalesced communities
+or the resident and coalesced communities were performed using Bray–Curtis or
+Jensen-Shannon similarity metrics. For two abrbitrary communities with
+ESV abundances represented by the vectors
+$\mathbf{x} = \left( x_1, x_2, \cdots, x_N \right)$
+and
+$\mathbf{y} = \left( y_1, y_2, \cdots, y_N \right)$
+(where $x_i$ and $y_i$ represent the relative abundance of the $i$th ESV in
+each community respectively and $N$ is the total number of ESVs), the Bray-Curtis
+similarity $\mathrm{BC} \left( \mathbf{x}, \mathbf{y} \right)$ is calculated as
+\cite{Bray1957}
+
+\begin{equation}
+\mathrm{BC} \left( \mathbf{x}, \mathbf{y} \right) = 
+\sum_i \mathrm{min} \left( x_i , y_i \right)
+\label{eq:bray-curtis}
+\end{equation}
+%
+and the Jensen-Shannon similarity
+$\mathrm{JS} \left( \mathbf{x}, \mathbf{y} \right)$
+is defined as one minus the Jensen-Shannon distance (which is, in turn,
+the square root of the Jensen-Shannon divergence \cite{Lin1991})
+
+\begin{equation}
+\mathrm{JS} \left( \mathbf{x}, \mathbf{y} \right) = 
+1 - \sqrt{\frac{1}{2}\mathrm{KL} \left( \mathbf{x}, \mathbf{m} \right) +
+\frac{1}{2}\mathrm{KL} \left( \mathbf{y}, \mathbf{m} \right)}
+\label{eq:jensen-shannon}
+\end{equation}
+%
+where $\mathbf{m} = \left( \mathbf{x} + \mathbf{y} \right)/2$ and KL denotes the
+Kullback-Leibler divergence \cite{Kullback1951}
+
+\begin{equation}
+\mathrm{KL} \left( \mathbf{x}, \mathbf{y} \right) = 
+\sum_i x_i \; \mathrm{log_2} \left( \frac{x_i}{y_i} \right)
+\label{eq:kullback-leibler}
+\end{equation}
+
+Additionally, we used alternative metrics that do not account for the specific
+fractions of each ESV (that is, only discern whether a ESV is present or absent
+in a community). These are the Jaccard similarity
+$\mathrm{J} \left( \mathbf{x}, \mathbf{y} \right)$ \cite{Jaccard1912}
+and the fraction of the endemic cohort of the invasive community
+that persists in the coalesced one,
+denoted as
+$\mathrm{E} \left( \mathbf{x}_\mathrm{I},
+\mathbf{x}_\mathrm{R},
+\mathbf{x}_\mathrm{C} \right)$
+where the subindices I, R and C correspond to the invasive, resident and coalesced
+communities respectively.
+
+\begin{equation}
+\mathrm{J} \left( \mathbf{x}, \mathbf{y} \right) = 
+\frac{\left|\; \mathbf{x} \cap \mathbf{y} \;\right|}
+{\left|\; \mathbf{x} \cup \mathbf{y} \;\right|}
+\label{eq:jaccard}
+\end{equation}
+
+\begin{equation}
+\mathrm{E} \left( \mathbf{x}_\mathrm{I},
+\mathbf{x}_\mathrm{R},
+\mathbf{x}_\mathrm{C} \right) = 
+\left.
+\sum_i
+f \left( x_{\mathrm{I}i}, x_{\mathrm{R}i}, x_{\mathrm{C}i} \right)
+\;
+\middle/
+\;
+\sum_i
+g \left( x_{\mathrm{I}i}, x_{\mathrm{R}i} \right)
+\right.
+\label{eq:endemic}
+\end{equation}
+%
+where we have defined
+
+\begin{equation}
+\begin{split}
+f \left( x_{\mathrm{I}i}, x_{\mathrm{R}i}, x_{\mathrm{C}i} \right) & =
+\begin{cases}
+    1 & \text{if } x_{\mathrm{I}i} > 0 \text{ and }
+                   x_{\mathrm{R}i} = 0 \text{ and }
+                   x_{\mathrm{C}i} > 0 \\
+    0 & \text{otherwise}
+\end{cases} \\
+g \left( x_{\mathrm{I}i}, x_{\mathrm{R}i} \right) & =
+\begin{cases}
+    1 & \text{if } x_{\mathrm{I}i} > 0 \text{ and }
+                   x_{\mathrm{R}i} = 0 \\
+    0 & \text{otherwise}
+\end{cases}
+\end{split}
+\label{eq:endemic-f}
+\end{equation}
+
+\subsection*{Simulations}\label{methods:sim}
 
@@
 
-\section{Data \& code availability}\label{datacode}
+\section*{Data \& code availability}\label{datacode}
 
 Experimental data and code for the analysis, as well as code for the simulations
-and the updated Community Simulator package with instructions to use the
+and the updated Community Simulator package with instructions for the
 new features can be found in \url{github.com/jdiazc9/coalescence}.
 
-\section{Figures}\label{figs}
+\section*{Acknowledgements}
+
+The authors wish to thank Joshua Goldford, Pankaj Mehta, Wenping Cui,
+Robert Marsland and all members of the Sanchez laboratory
+for many helpful discussions.
+We also wish to express our gratitude to the Goodman laboratory at Yale
+for technical help during the early stages of this project.
+The funding for this work partly results from a Scialog Program
+sponsored jointly by the Research Corporation for Science Advancement and
+the Gordon and Betty Moore Foundation through grants to Yale University by the
+Research Corporation and the Simons Foundation.
+
+\clearpage
+
+\section*{Figures}\label{figs}
 

### 2021-04-29 draft/draft.tex
changes to fig1, adjustments in analyzeData.R for easier to streamline plots

@@
 \usepackage[margin=3cm]{geometry} % margins
-\usepackage[nottoc,numbib]{tocbibind} % add references to table of contents
 \usepackage{lineno} % line numbers
@@
 200$\upmu$g/mL cycloheximide to suppress eukaryotic growth.
-After 48h, samples were mixed 1:1 with 80\% glycerol and kept frozen at -80$^\circ$C.
+After 48h, samples were mixed 1:1 with 80\% glycerol and kept frozen at $-80^\circ$C.
 Starting microbial communities were prepared by scrapping the frozen stocks into
-200$\upmu$L of 1$\times$PBS and adding a volume of 4$\upmu$L to 500$\upmu$L
-of synthetic minimal media (1$\times$M9) supplemented with 200$\upmu$g/mL cycloheximide
+$200\upmu$L of 1$\times$PBS and adding a volume of $4\upmu$L to $500\upmu$L
+of synthetic minimal media (1$\times$M9) supplemented with $200\upmu$g/mL cycloheximide
 and 0.07 C-mol/L glutamine or sodium citrate as the carbon source in 96 deep-well plates
 (1.2mL; VWR).
-Cultures were then incubated still at 30$^\circ$C to allow for re-growth.
+Cultures were then incubated still at $30^\circ$C to allow for re-growth.
 After 48h, samples were fully homogenized and biomass increase was followed by measuring
-the optical density (620nm) of 100$\upmu$L of the cultures in a Multiskan FC plate reader
+the optical density (620nm) of $100\upmu$L of the cultures in a Multiskan FC plate reader
 (Thermo Scientific).
-Communities were stabilized \cite{Goldford2018} by passaging 4$\upmu$L of the cultures into
-500$\upmu$L of fresh media (1$\times$M9 with the carbon source)  every 48h for a total of
+Communities were stabilized \cite{Goldford2018} by passaging $4\upmu$L of the cultures into
+$500\upmu$L of fresh media (1$\times$M9 with the carbon source)  every 48h for a total of
 12 transfers at a dilution factor of 1:100,
@@
 For each community, the most abundant colony morphotype at the end of the ninth transfer
-was selected, resuspended in 100$\upmu$L 1$\times$PBS and serially diluted (1:10).
-Next, 20$\upmu$L of the cells diluted to $10^{-6}$ were plated in the corresponding synthetic
-minimal media and allowed to regrow at 30$^\circ$C for 48h. Dominants were then inoculated
-into 500$\upmu$L of fresh media and incubated still at 30$^\circ$C for 48h.
+was selected, resuspended in $100\upmu$L 1$\times$PBS and serially diluted (1:10).
+Next, $20\upmu$L of the cells diluted to $10^{-6}$ were plated in the corresponding synthetic
+minimal media and allowed to regrow at $30^\circ$C for 48h. Dominants were then inoculated
+into $500\upmu$L of fresh media and incubated still at $30^\circ$C for 48h.
 After this period, the communities stabilized for eleven transfers and the isolated dominants
@@
 competition experiments
-were performed by mixing equal volumes (4$\upmu$L) of each of the eight
+were performed by mixing equal volumes ($4\upmu$L) of each of the eight
 communities or eight dominants at the onset of the twelfth transfer.
 Competitions were set up in their native media,
-i.e. in 500$\upmu$L of 1$\times$M9 supplemented with 0.07 C-mol/L of
+i.e. in $500\upmu$L of 1$\times$M9 supplemented with 0.07 C-mol/L of
 either glutamine or citrate in 96 deep-well plates.
-Plates were incubated at 30$^\circ$C for 48h.
+Plates were incubated at $30^\circ$C for 48h.
 Pairwise competitions were further propagated for seven serial transfers
-(roughly 42 generations; Fig. \mr) by transferring 8$\upmu$L of
-each culture to fresh media (500$\upmu$L).
+(roughly 42 generations; Fig. \mr) by transferring $8\upmu$L of
+each culture to fresh media ($500\upmu$L).
 
@@
 in a bench-top centrifuge at room temperature;
-cell pellets were stored at -80$^\circ$C before processing.
+cell pellets were stored at $-80^\circ$C before processing.
 To maximize Gram-positive bacteria cell wall lysis,
-the cell pellets were re-suspended and incubated at 37$^\circ$C for 30min
+the cell pellets were re-suspended and incubated at $37^\circ$C for 30min
 in enzymatic lysis buffer (20mM Tris-HCl, 2mM sodium EDTA, 1.2\% Triton X-100)
-and 20 mg/mL of lysozyme from chicken egg white (Sigma-Aldrich).
+and 20mg/mL of lysozyme from chicken egg white (Sigma-Aldrich).
 After cell lysis, the DNA extraction and purification was performed using the
 DNeasy 96 protocol for animal tissues (Qiagen).
-The clean DNA in 100$\upmu$L elution buffer of 10mM Tris-HCl, 0.5mM EDTA
+The clean DNA in $100\upmu$L elution buffer of 10mM Tris-HCl, 0.5mM EDTA
 at pH 9.0 was quantified using Quan-iT PicoGreen dsDNA Assay Kit
@@
 using the Illumina MiSeq chemistry and platform.
-Each sample went through a 30-cycle PCR in duplicate of 20$\upmu$L
+Each sample went through a 30-cycle PCR in duplicate of $20\upmu$L
 reaction volumes using 5ng of DNA each, dual index primers, and AccuPrime Pfx SuperMix (Invitrogen).
 The thermocycling procedure includes a 2min initial denaturation step at
-95$^\circ$C, and 30 cycles of the following PCR scheme:
-(a) 20 second denaturation at 95$^\circ$C,
-(b) 15 second annealing at 55$^\circ$C,
-and (c) 5 minute extension at 72$^\circ$C.
+$95^\circ$C, and 30 cycles of the following PCR scheme:
+(a) 20 second denaturation at $95^\circ$C,
+(b) 15 second annealing at $55^\circ$C,
+and (c) 5 minute extension at $72^\circ$C.
 The duplicate PCR products of each sample were pooled, purified, and normalized
@@
 
-\section*{Figures}\label{figs}
-
 % references
@@
 
+\clearpage
+
+\section*{Figures}\label{figs}
+
 \end{document}

### 2021-04-29 data/analyzeData.R

@@ -266,7 +266,7 @@ myplots[['community-compostion_rankplots']] <-
          aes(x=rank,y=fraction,
              group=interaction(carbon_source,community,sample),
              color=carbon_source)) +
-    geom_line(size=0.5) +
+    geom_line(size=0.25) +
     scale_y_continuous(trans='log10',
                        name='Relative abundance',
                        limits=10^c(-4,0),
@@ -282,10 +282,13 @@ myplots[['community-compostion_rankplots']] <-
     theme_bw() +
     theme(panel.grid=element_blank(),
           legend.title=element_blank(),
-          legend.position=c(0.7,0.9),
+          legend.position=c(0.6,0.9),
           legend.background=element_rect(fill='transparent'),
           text=element_text(size=15),
-          axis.text=element_text(size=15))
+          axis.text=element_text(size=15),
+          axis.line=element_blank(), #axis.line=element_line(size=0.25),
+          axis.ticks=element_line(size=0.25),
+          panel.border=element_rect(size=0.25))
 
 # display and save plots
 if (display_plots) {
@@ -302,8 +305,8 @@ if (save_plots) {
   ggsave(file.path('.','plots','community-compostion_rankplots.pdf'),
          plot=myplots[['community-compostion_rankplots']],
          device='pdf',
-         height=90,
-         width=90,
+         height=70,
+         width=70,
          units='mm')
 }
 

### 2021-04-30 draft/draft.tex
more fig1 and draft

@@
 
+\begin{figure}[!h]
+\centering
+\includegraphics[width=12cm,keepaspectratio]{figs/fig1.pdf}
+\caption{\textbf{Overview of the experimental protocol.}
+\textbf{a.}~Environmental samples obtained from eight different locations were used
+to inoculate our communities.
+\textbf{b.}~Communities were stabilized in serial batch culture bioreactors
+\cite{Goldford2018} in minimal synthetic media with glutamine or citrate as the
+only supplied carbon source.
+\textbf{c.}~Communities were plated in minimal media agar plates and the most abundant
+species (the ``dominants") from each community were isolated. We refer to the set of
+sub-dominant species as the ``cohorts".
+\textbf{d.}~Rank-frequency distributions of all eight communities stabilized in either
+glutamine (red) or citrate (blue), sequenced at a depth of $10^{-4}$ reads.
+Three biological replicates per community are shown.
+Community compositions are skewed and long-tailed.
+\textbf{e.}~Our hypothesis is that ecological co-selection can take place from the top-down,
+i.e. the dominant co-selecting the cohort, or from the bottom-up, i.e. the cohort co-selecting
+the dominant. Both forms of co-selection can be positive (recruitment) or negative
+(antagonism).
+\textbf{f.}~Illustration of the protocol of our coalescence experiments.}
+\label{fig1}
+\end{figure}
+
+\clearpage
+
 \end{document}

### 2021-05-01 draft/draft.tex
final fig1, initializing fig2, samll changes to methods

@@
 This is how you refer to the \nameref{intro}.
+
+\clearpage
 
@@
 Beta-diversity indexes between the invasive and coalesced communities
-or the resident and coalesced communities were performed using Bray–Curtis or
-Jensen-Shannon similarity metrics. For two abrbitrary communities with
+or the resident and coalesced communities were computed using various
+similarity metrics. For two arbitrary communities with
 ESV abundances represented by the vectors
@@
 %
-and the Jensen-Shannon similarity
+The Jensen-Shannon similarity
 $\mathrm{JS} \left( \mathbf{x}, \mathbf{y} \right)$
@@
 \end{equation}
-
-Additionally, we used alternative metrics that do not account for the specific
-fractions of each ESV (that is, only discern whether a ESV is present or absent
-in a community). These are the Jaccard similarity
+%
+The Jaccard similarity is given by
 $\mathrm{J} \left( \mathbf{x}, \mathbf{y} \right)$ \cite{Jaccard1912}
-and the fraction of the endemic cohort of the invasive community
-that persists in the coalesced one,
-denoted as
-$\mathrm{E} \left( \mathbf{x}_\mathrm{I},
-\mathbf{x}_\mathrm{R},
-\mathbf{x}_\mathrm{C} \right)$
-where the subindices I, R and C correspond to the invasive, resident and coalesced
-communities respectively.
 
@@
 \end{equation}
+%
+For all the metrics above, we quantify the relative similarity between the invasive
+and the coalesced communities using relative metrics (Q):
+
+\begin{equation}
+\mathrm{Q} \left( \mathbf{x}_\mathrm{I},
+\mathbf{x}_\mathrm{R},
+\mathbf{x}_\mathrm{C} \right) = 
+\frac{\mathrm{F} \left( \mathbf{x}_\mathrm{I},\mathbf{x}_\mathrm{C}\right)}
+{\mathrm{F} \left( \mathbf{x}_\mathrm{I},\mathbf{x}_\mathrm{C}\right)
++
+\mathrm{F} \left( \mathbf{x}_\mathrm{R},\mathbf{x}_\mathrm{C}\right)}
+\label{eq:q}
+\end{equation}
+%
+where the subindices I, R and C correspond to the invasive, resident and coalesced
+communities respectively, and F represents one of BC (Bray-Curtis similarity), JS
+(Jensen-Shannon similarity) or J (Jaccard similarity) defined in equations
+\ref{eq:bray-curtis} to \ref{eq:jaccard}.
+
+Additionally, we quantify coalescence outcomes by examining the fraction of the endemic
+cohort of the invasive community that persists in the coalesced one. We denote this
+metric as
+$\mathrm{E} \left( \mathbf{x}_\mathrm{I},
+\mathbf{x}_\mathrm{R},
+\mathbf{x}_\mathrm{C} \right)$.
 
@@
 \caption{\textbf{Overview of the experimental protocol.}
-\textbf{a.}~Environmental samples obtained from eight different locations were used
+\textbf{a.}~Environmental samples collected from eight different locations were used
 to inoculate our communities.
@@
 (antagonism).
-\textbf{f.}~Illustration of the protocol of our coalescence experiments.}
+\textbf{f.}~Illustration of the protocol of our coalescence experiments. All pairs of
+communities were inoculated into fresh minimal media supplemented with the same carbon
+source where communities had been previously stabilized. The coalesced (C) and original
+resident (R) and invasive (I) communities were then serially diluted and allowed to grow
+for seven additional transfers.}
 \label{fig1}

### 2021-05-04 draft/draft.tex
figure 2 initialized, changes to analyzeData.R for smoother streamlining of plots

@@
 $95^\circ$C, and 30 cycles of the following PCR scheme:
-(a) 20 second denaturation at $95^\circ$C,
-(b) 15 second annealing at $55^\circ$C,
-and (c) 5 minute extension at $72^\circ$C.
+(a) 20-second denaturation at $95^\circ$C,
+(b) 15-second annealing at $55^\circ$C,
+and (c) 5-minute extension at $72^\circ$C.
 The duplicate PCR products of each sample were pooled, purified, and normalized

### 2021-05-04 data/analyzeData.R

@@ -286,7 +286,7 @@ myplots[['community-compostion_rankplots']] <-
           legend.background=element_rect(fill='transparent'),
           text=element_text(size=15),
           axis.text=element_text(size=15),
-          axis.line=element_blank(), #axis.line=element_line(size=0.25),
+          axis.line=element_blank(),
           axis.ticks=element_line(size=0.25),
           panel.border=element_rect(size=0.25))
 
@@ -1062,7 +1062,7 @@ plot_this$community_2 <- factor(plot_this$community_2,levels=community_names)
 plot_this$metric <- factor(plot_this$metric,levels=unique(plot_this$metric))
 
 # parameters for annotations of plots
-poly <- list(w=0.15,
+poly <- list(w=0.10,
              aperture=0.15,
              color=c(rgb(220,245,220,maxColorValue=255),
                      rgb(230,230,230,maxColorValue=255),
@@ -1073,15 +1073,18 @@ myplots[['q-vs-pairwise_bray-curtis']] <-
   ggplot(data=plot_this[plot_this$metric=='q_bray_curtis',],
          aes(x=f_pairwise,y=value,
              color=carbon_source)) +
-  geom_point(size=2) +
+  geom_point(size=3,
+             shape=1,
+             stroke=0.5) +
   geom_smooth(formula = y ~ x,
               method = 'lm',
-              se = FALSE) +
+              se = FALSE,
+              size = 0.5) +
   scale_y_continuous(name='Q\nCoalesced - Invasive',
                      limits=c(0,1),
                      breaks=c(0,0.5,1),
                      labels=c('0','0.5','1')) +
-  scale_x_continuous(name='Frequency of invasive dominant species\nin pairwise competition',
+  scale_x_continuous(name='Frequency of\ninvasive dominant species\nin pairwise competition',
                      limits=c(0,1),
                      breaks=c(0,0.5,1),
                      labels=c('0','0.5','1')) +
@@ -1092,22 +1095,28 @@ myplots[['q-vs-pairwise_bray-curtis']] <-
         legend.position=c(0.2,0.9),
         legend.background=element_rect(fill='transparent'),
         text=element_text(size=15),
-        axis.text=element_text(size=15),) +
-  coord_fixed() +
-  ggtitle('Bray-Curtis similarity')
+        axis.text=element_text(size=15),
+        axis.line=element_blank(),
+        axis.ticks=element_line(size=0.25),
+        panel.border=element_rect(size=0.25)) +
+  coord_fixed() # +
[…]

### 2021-05-05 draft/draft.tex
fig2 complete and added to draft, changes to coalescence.py and analyzeData.R for smoother streamlining of figures

@@
 \definecolor{lightgray}{gray}{0.9}
+\newcommand{\figref}[2][]{%
+  \hyperref[{#2}]{%
+    Figure~\ref*{#2}%
+    \ifx\\#1\\%
+    \else
+      #1%
+    \fi
+  }%
+}
+\newcommand{\tableref}[1]{%
+  \hyperref[{#1}]{%
+   Table~\ref*{#1}%
+  }%
+}
+\newcommand{\methodsref}[1]{%
+  \hyperref[{#1}]{%
+   Methods:~\nameref*{#1}%
+  }%
+}
 
@@
 
-This is an example cite \cite{Vetrovsky2013,nloptr}
+This is an example cite \cite{Vetrovsky2013,nloptr}.
+This is how you refer to \figref{fig1}.
+This is how you refer to \figref[a]{fig1}.
+This is how you refer to \tableref{fig1}.
+This is how you refer to the section \nameref{methods:sim} of the \nameref{methods}.
+This is a cleaner way to refer to the \methodsref{methods:sim}.
 
 \section*{Results \& Discussion}\label{results-discussion}
-
-This is how you refer to the \nameref{intro}.
 
@@
 each community respectively and $N$ is the total number of ESVs), the Bray-Curtis
-similarity $\mathrm{BC} \left( \mathbf{x}, \mathbf{y} \right)$ is calculated as
+similarity $BC \left( \mathbf{x}, \mathbf{y} \right)$ is calculated as
 \cite{Bray1957}
@@
 \begin{equation}
-\mathrm{BC} \left( \mathbf{x}, \mathbf{y} \right) = 
+BC \left( \mathbf{x}, \mathbf{y} \right) = 
 \sum_i \mathrm{min} \left( x_i , y_i \right)
@@
 The Jensen-Shannon similarity
-$\mathrm{JS} \left( \mathbf{x}, \mathbf{y} \right)$
+$JS \left( \mathbf{x}, \mathbf{y} \right)$
 is defined as one minus the Jensen-Shannon distance (which is, in turn,
@@
 \begin{equation}
-\mathrm{JS} \left( \mathbf{x}, \mathbf{y} \right) = 
-1 - \sqrt{\frac{1}{2}\mathrm{KL} \left( \mathbf{x}, \mathbf{m} \right) +
-\frac{1}{2}\mathrm{KL} \left( \mathbf{y}, \mathbf{m} \right)}
+JS \left( \mathbf{x}, \mathbf{y} \right) = 
+1 - \sqrt{\frac{1}{2}KL \left( \mathbf{x}, \mathbf{m} \right) +
+\frac{1}{2}KL \left( \mathbf{y}, \mathbf{m} \right)}
 \label{eq:jensen-shannon}
@@
 \begin{equation}
-\mathrm{KL} \left( \mathbf{x}, \mathbf{y} \right) = 
+KL \left( \mathbf{x}, \mathbf{y} \right) = 
 \sum_i x_i \; \mathrm{log_2} \left( \frac{x_i}{y_i} \right)
@@
 The Jaccard similarity is given by
-$\mathrm{J} \left( \mathbf{x}, \mathbf{y} \right)$ \cite{Jaccard1912}
+$J \left( \mathbf{x}, \mathbf{y} \right)$ \cite{Jaccard1912}
 
 \begin{equation}
-\mathrm{J} \left( \mathbf{x}, \mathbf{y} \right) = 
+J \left( \mathbf{x}, \mathbf{y} \right) = 
 \frac{\left|\; \mathbf{x} \cap \mathbf{y} \;\right|}
@@
 %
-For all the metrics above, we quantify the relative similarity between the invasive
-and the coalesced communities using relative metrics (Q):
+Additionally, we quantify coalescence outcomes by examining the fraction of the endemic
+cohort of the original communities that persists in the coalesced one. We call
+$E \left( \mathbf{x}, \mathbf{y} \right)$ to the fraction of endemic species of
+$\mathbf{x}$ that are also found in $\mathbf{y}$.
 
-\begin{equation}
-\mathrm{Q} \left( \mathbf{x}_\mathrm{I},
-\mathbf{x}_\mathrm{R},
-\mathbf{x}_\mathrm{C} \right) = 
-\frac{\mathrm{F} \left( \mathbf{x}_\mathrm{I},\mathbf{x}_\mathrm{C}\right)}
-{\mathrm{F} \left( \mathbf{x}_\mathrm{I},\mathbf{x}_\mathrm{C}\right)
-+
-\mathrm{F} \left( \mathbf{x}_\mathrm{R},\mathbf{x}_\mathrm{C}\right)}
-\label{eq:q}
-\end{equation}
-%
-where the subindices I, R and C correspond to the invasive, resident and coalesced
-communities respectively, and F represents one of BC (Bray-Curtis similarity), JS
-(Jensen-Shannon similarity) or J (Jaccard similarity) defined in equations
-\ref{eq:bray-curtis} to \ref{eq:jaccard}.
+\iffalse
 
-Additionally, we quantify coalescence outcomes by examining the fraction of the endemic
-cohort of the invasive community that persists in the coalesced one. We denote this
+We denote this
 metric as
-$\mathrm{E} \left( \mathbf{x}_\mathrm{I},
+$E \left( \mathbf{x}_\mathrm{I},
 \mathbf{x}_\mathrm{R},
@@
 \begin{equation}
-\mathrm{E} \left( \mathbf{x}_\mathrm{I},
+E \left( \mathbf{x}_\mathrm{I},
 \mathbf{x}_\mathrm{R},
@@
 \end{equation}
+
+\fi
+
+For all the metrics above, we quantify the relative similarity between the invasive
+and the coalesced communities using relative metrics ($Q$):
+
+\begin{equation}
+Q \left( \mathbf{x}_\mathrm{I},
+\mathbf{x}_\mathrm{R},
+\mathbf{x}_\mathrm{C} \right) = 
+\frac{F \left( \mathbf{x}_\mathrm{I},\mathbf{x}_\mathrm{C}\right)}
+{F \left( \mathbf{x}_\mathrm{I},\mathbf{x}_\mathrm{C}\right)
++
+F \left( \mathbf{x}_\mathrm{R},\mathbf{x}_\mathrm{C}\right)}
+\label{eq:q}
+\end{equation}
+%
+where the subindices I, R and C correspond to the invasive, resident and coalesced
+communities respectively, and $F$ represents one of $BC$ (Bray-Curtis similarity), $JS$
+(Jensen-Shannon similarity), $J$ (Jaccard similarity) or $E$ (endemic survival) defined above.
 
@@
 \centering
-\includegraphics[width=12cm,keepaspectratio]{figs/fig1.pdf}
+\includegraphics[width=10cm,keepaspectratio]{figs/fig1.pdf}
 \caption{\textbf{Overview of the experimental protocol.}
@@
 
+\begin{figure}[!h]
+\centering
+\includegraphics[width=15cm,keepaspectratio]{figs/fig2.pdf}
+\caption{\textbf{Co-selection in microbial community coalescence.}
+\textbf{a.}~Coalescence outcomes are quantified by the relative Bray-Curtis similarity
+($Q$) between the coalesced and invasive communities (\methodsref{methods:metrics}).
+These outcomes are predicted by the pairwise competition between the invasive and
+resident dominant
+species ($R^2=X$ for glutamine and $X$ for citrate). This is consistent with a scenario of
+top-down positive co-selection where dominants recruit their cohorts for the final
+coalesced community.
+\textbf{b.}~We represent the frequency reached by
+the invasive dominant species when they invade the resident communities on their own versus
+when they are in the company of their cohort.
+Three scenarios are possible: green and red
+shaded areas represent limit cases of positive (recruitment) or negative (antagonism)
+bottom-up co-selection, gray area corresponds to situations where invasive dominant
+species can invade with equal success regardless of the presence of their cohorts.
+Data shows that positive co-selection is common, whereas antagonistic
+co-selection is rare in our experiments.
+\textbf{c-d.} Simulations of community coalescence with a consumer-resource model
+are able to capture these trends.}
+\label{fig2}
+\end{figure}
+
+\clearpage
+
 \end{document}

### 2021-05-05 coalescence.py

@@ -233,12 +233,12 @@ def mysim(p,q,pq):
 # assumptions
 
 """
-# defaults for this simulations (these work)
+# quasi-default assumptions
 assumptions = community_simulator.usertools.a_default.copy()
-assumptions['n_wells'] = 20 # number of communities
-assumptions['S'] = 100 # number of species sampled at initialization
-assumptions['SA'] = [100, 100, 100] # [100, 100, 100] # number of species per specialist family
-assumptions['Sgen'] = 30 # number of generalists
+assumptions['n_wells'] = 100 # number of communities
+assumptions['S'] = 50 # number of species sampled at initialization
+assumptions['SA'] = [400, 400, 400] # [100, 100, 100] # number of species per specialist family
+assumptions['Sgen'] = 120 # 30 # number of generalists
 assumptions['l'] = 0.8 # leakage fraction
 assumptions['MA'] = [10, 10, 10] # [30, 30, 30] # number of resources per resource class
 
@@ -265,8 +265,8 @@ assumptions['rs'] = 0.0 # control parameter for resource secretion: 0 means rand
 assumptions = community_simulator.usertools.a_default.copy()
 assumptions['n_wells'] = 100 # number of communities
 assumptions['S'] = 50 # number of species sampled at initialization
-assumptions['SA'] = [200, 200, 200] # [100, 100, 100] # number of species per specialist family
-assumptions['Sgen'] = 60 # 30 # number of generalists
+assumptions['SA'] = [400, 400, 400] # [100, 100, 100] # number of species per specialist family
+assumptions['Sgen'] = 120 # 30 # number of generalists
 assumptions['l'] = 0.8 # leakage fraction
 assumptions['MA'] = [10, 10, 10] # [30, 30, 30] # number of resources per resource class
 
@@ -282,7 +282,7 @@ assumptions['c0'] = 0.0 # background consumption rate in binary model
 assumptions['c1'] = 1.0 # specific consumption rate in binary model
 assumptions['sigc'] = 3 #3 # standard deviation of sum of consumption rates for Gaussian and Gamma models
 
-assumptions['sparsity'] = 0.8 #0.05 # variability in secretion fluxes among resources (must be less than 1)  
+assumptions['sparsity'] = 0.6 #0.05 # variability in secretion fluxes among resources (must be less than 1)  
 assumptions['fs'] = 0.45 #0.45 # fraction of secretion flux to resources of the same type as the consumed one
[…]

### 2021-05-05 data/analyzeData.R

@@ -301,13 +301,13 @@ if (save_plots) {
          device='pdf',
          height=297,
          width=210,
-         units='mm')
+         units='mm',dpi=600)
   ggsave(file.path('.','plots','community-compostion_rankplots.pdf'),
          plot=myplots[['community-compostion_rankplots']],
          device='pdf',
          height=70,
          width=70,
-         units='mm')
+         units='mm',dpi=600)
 }
 
 ### Pie plots show that community compositions are very conserved across
@@ -712,7 +712,7 @@ if (save_plots) {
          device='pdf',
          height=270,
          width=300,
-         units='mm')
+         units='mm',dpi=600)
 }
 
 ### Our criterion will be so that we mantain maximum consistency with the experimental
@@ -1248,25 +1248,25 @@ if (save_plots) {
          device='pdf',
          height=90,
          width=90,
-         units='mm')
+         units='mm',dpi=600)
   ggsave(file.path('.','plots','q-vs-pairwise_other-metrics.pdf'),
          plot=myplots[['q-vs-pairwise_other-metrics']],
          device='pdf',
          height=90,
          width=180,
-         units='mm')
+         units='mm',dpi=600)
   ggsave(file.path('.','plots','q-vs-pairwise_cohorts.pdf'),
          plot=myplots[['q-vs-pairwise_cohorts']],
          device='pdf',
          height=180,
          width=180,
-         units='mm')
+         units='mm',dpi=600)
   ggsave(file.path('.','plots','alone-vs-together.pdf'),
          plot=myplots[['alone-vs-together']],
          device='pdf',
          height=90,
          width=90,
-         units='mm')
+         units='mm',dpi=600)
 }
 
 
@@ -1361,13 +1361,13 @@ if (save_plots) {
          device='pdf',
          height=90,
          width=90,
-         units='mm')
+         units='mm',dpi=600)
   ggsave(file.path('.','plots','alone-vs-together_simul.pdf'),
          plot=myplots[['alone-vs-together_simul']],
          device='pdf',
          height=90,
          width=90,
-         units='mm')
+         units='mm',dpi=1000)
 }
 
 

### 2021-05-05 data/analyzeData.R
cleaner, higher quality and lighter figure 2b

@@ -1190,7 +1190,11 @@ myplots[['q-vs-pairwise_cohorts']] <-
   coord_fixed()
 
 myplots[['alone-vs-together']] <-
-  ggplot(data=plot_this,
+  ggplot(data=unique(plot_this[,c('carbon_source',
+                                  'community_1',
+                                  'community_2',
+                                  'f_singleinv',
+                                  'f_multiinv')]),
          aes(x=f_singleinv,y=f_multiinv,
              color=carbon_source)) +
   annotate('polygon',
@@ -1367,7 +1371,7 @@ if (save_plots) {
          device='pdf',
          height=90,
          width=90,
-         units='mm',dpi=1000)
+         units='mm',dpi=600)
 }
 
 

### 2021-05-05 coalescence.py
quasi-standard parameters updated

@@ -254,7 +254,7 @@ assumptions['c0'] = 0.0 # background consumption rate in binary model
 assumptions['c1'] = 1.0 # specific consumption rate in binary model
 assumptions['sigc'] = 3 #3 # standard deviation of sum of consumption rates for Gaussian and Gamma models
 
-assumptions['sparsity'] = 0.6 #0.05 # variability in secretion fluxes among resources (must be less than 1)  
+assumptions['sparsity'] = 0.9 #0.05 # variability in secretion fluxes among resources (must be less than 1)  
 assumptions['fs'] = 0.45 #0.45 # fraction of secretion flux to resources of the same type as the consumed one
 assumptions['fw'] = 0.45 #0.45 # fraction of secretion flux to waste resources
 assumptions['metabolism'] = 'specific' # 'common' uses a common D matrix for all species, 'specific' uses a different matrix D for each species
@@ -282,7 +282,7 @@ assumptions['c0'] = 0.0 # background consumption rate in binary model
 assumptions['c1'] = 1.0 # specific consumption rate in binary model
 assumptions['sigc'] = 3 #3 # standard deviation of sum of consumption rates for Gaussian and Gamma models
 
-assumptions['sparsity'] = 0.6 #0.05 # variability in secretion fluxes among resources (must be less than 1)  
+assumptions['sparsity'] = 0.9 #0.05 # variability in secretion fluxes among resources (must be less than 1)  
 assumptions['fs'] = 0.45 #0.45 # fraction of secretion flux to resources of the same type as the consumed one
 assumptions['fw'] = 0.45 #0.45 # fraction of secretion flux to waste resources
 assumptions['metabolism'] = 'specific' # 'common' uses a common D matrix for all species, 'specific' uses a different matrix D for each species

### 2021-05-05 coalescence.py
minor canges in simulation parameters, new simulated data to include in fig2

@@ -263,10 +263,10 @@ assumptions['rs'] = 0.0 # control parameter for resource secretion: 0 means rand
 
 # try these assumptions
 assumptions = community_simulator.usertools.a_default.copy()
-assumptions['n_wells'] = 100 # number of communities
+assumptions['n_wells'] = 50 # number of communities
 assumptions['S'] = 50 # number of species sampled at initialization
-assumptions['SA'] = [400, 400, 400] # [100, 100, 100] # number of species per specialist family
-assumptions['Sgen'] = 120 # 30 # number of generalists
+assumptions['SA'] = [800, 800, 800] # [100, 100, 100] # number of species per specialist family
+assumptions['Sgen'] = 240 # 30 # number of generalists
 assumptions['l'] = 0.8 # leakage fraction
 assumptions['MA'] = [10, 10, 10] # [30, 30, 30] # number of resources per resource class
 

### 2021-05-06 draft/draft.tex
added missing Rsquared values to draft

@@
 resident dominant
-species ($R^2=X$ for glutamine and $X$ for citrate). This is consistent with a scenario of
+species ($R^2=0.15$ for glutamine and $R^2=0.57$ for citrate). This is consistent with a scenario of
 top-down positive co-selection where dominants recruit their cohorts for the final

### 2021-05-06 coalescence.py
coalescence.py: commented out (some) plotting commands to speed up

@@ -1,7 +1,6 @@
 # -*- coding: utf-8 -*-
 """
 Created on Wed Sep 23 18:28:37 2020
-
 @author: Juan
 """
 
@@ -149,7 +148,7 @@ def bray_curtis(p,q):
     c = sum([min(p[i],q[i]) for i in range(len(p)) if p[i]>0 and q[i]>0])
     
     # Bray-Curtis similarity
-    bc = 1 - 2*c/(sum(p) + sum(q))
+    bc = 2*c/(sum(p) + sum(q))
     return bc
 
 # Kullback-Leibler divergence - using log2 so result is within [0,1]
@@ -241,19 +240,16 @@ assumptions['SA'] = [400, 400, 400] # [100, 100, 100] # number of species per sp
 assumptions['Sgen'] = 120 # 30 # number of generalists
 assumptions['l'] = 0.8 # leakage fraction
 assumptions['MA'] = [10, 10, 10] # [30, 30, 30] # number of resources per resource class
-
 assumptions['response'] = 'type I'
 assumptions['regulation'] = 'independent' # 'independent' is standard, 'energy' or 'mass' enable preferential consumption of resources
 assumptions['sampling'] = 'Gamma' # 'Binary', 'Gaussian', 'Uniform' or 'Gamma' (sampling of the matrix c)
 assumptions['supply'] = 'off'
 assumptions['R0_food'] = 1000
 assumptions['m'] = 0 # turn off mortality (?)
-
 assumptions['q'] = 0.9 #0.9 # preference strength (0 for generalist and 1 for specialist)
 assumptions['c0'] = 0.0 # background consumption rate in binary model
 assumptions['c1'] = 1.0 # specific consumption rate in binary model
 assumptions['sigc'] = 3 #3 # standard deviation of sum of consumption rates for Gaussian and Gamma models
-
 assumptions['sparsity'] = 0.9 #0.05 # variability in secretion fluxes among resources (must be less than 1)  
 assumptions['fs'] = 0.45 #0.45 # fraction of secretion flux to resources of the same type as the consumed one
 assumptions['fw'] = 0.45 #0.45 # fraction of secretion flux to waste resources
@@ -302,6 +298,7 @@ def dRdt(N,R,params):
 dynamics = [dNdt,dRdt]
 
 # plot matrices c and D
+'''
 fig,ax=plt.subplots()
[…]

### 2021-05-07 draft/draft.tex
started introduction, added some supplementary figures, coalescence.py ready for simulations with simple, complex and intermediate primary resource

@@
 \usepackage{zref-xr,zref-user} % more cross-references formatting
+\zxrsetup{toltxlabel}
+\zexternaldocument*{supp}
 \usepackage[square,numbers,sort&compress]{natbib}
@@
 \usepackage{lineno} % line numbers
+\usepackage{textcomp} % special characters (some references break if this isn't loaded?)
 
@@
 
-\section*{Introduction}\label{intro}
+\section*{Test section}
 
-This is an example cite \cite{Vetrovsky2013,nloptr}.
+This is an example cite \cite{Luo2020,Vass2021}.
 This is how you refer to \figref{fig1}.
@@
 This is how you refer to \tableref{fig1}.
+This is how you refer to \figref{figS1}.
 This is how you refer to the section \nameref{methods:sim} of the \nameref{methods}.
 This is a cleaner way to refer to the \methodsref{methods:sim}.
+
+\section*{Introduction}\label{intro}
+
+Microbial communities often invade one another.
+This has been observed, for instance, 
+in river courses where terrestrial microbial communities mix with aquatic microorganisms
+\cite{Luo2020,Vass2021}
+or in soil communities being invaded by aerially dispersed bacteria and funghi \cite{Evans2019}.
+The human digestive system can even get invaded several times a day by the microbial consortia
+that reside on the ingested food.
+Ecologists have long contemplated the idea that interactions between multiple co-invading species
+can produce correlated invasional outcomes
+\cite{Gilpin1994,Grosholz2005,Green2011,Livingston2013,Rillig2015,OLoughlin2017}.
+One example is the hypothesis known as \textit{invasional meltdown}, which proposes that
+positive interactions between co-invading species can enhance their invasive success and
+facilitate future invasions \cite{Simberloff1999,Simberloff2006,Gurevitch2006}.
+
+The phenomenon by which entire microbiomes invade one another has been termed
+\textit{community coalescence} \cite{Rillig2015}.
+In spite of its clear potential importance, the role of coalescence in microbiome assembly is
+only beginning to be addressed and little is known about the mechanisms that govern it and its
+potential implications.
+Early mathematical models of community-community invasions \cite{Gilpin1994,Toquenaga1997},
+as well as more recent work including environmental construction (e.g., consumer-resource models)
+\cite{Tikhonov2016,Tikhonov2017,Lechon2021}, suggest that high-order invasion effects are common
+during community coalescence. Communities that have a previous history of coexistence may exhibit an
+emergent ``cohesiveness'' which produces correlated invasional outcomes among species from the
+same community \cite{Livingston2013,Sierocinski2017}.
+The situation where ecological partners in the invading community recruit one another into the final
+coalesced community has been called \textit{ecological co-selection} \cite{Rillig2017,Sierocinski2017}.
 
@@
 \bibliographystyle{mystyle}
-%\bibliographystyle{unsrt}
 \bibliography{refs}
@@
 \centering
+\internallinenumbers
 \includegraphics[width=10cm,keepaspectratio]{figs/fig1.pdf}
@@
 \centering
+\internallinenumbers
 \includegraphics[width=15cm,keepaspectratio]{figs/fig2.pdf}
@@
 top-down positive co-selection where dominants recruit their cohorts for the final
-coalesced community.
+coalesced community. Two biological replicates per experiment are plotted individually.
 \textbf{b.}~We represent the frequency reached by
@@
 co-selection is rare in our experiments.
+Two biological replicates per experiment are plotted individually.
 \textbf{c-d.} Simulations of community coalescence with a consumer-resource model
@@
 
+\section*{Supplementary Figures}
+
+\renewcommand{\thefigure}{S\arabic{figure}}
+\setcounter{figure}{0}
+
+\begin{figure}[!h]
+\centering
+\internallinenumbers
+\includegraphics[width=15cm,keepaspectratio]{figs/figS1.pdf}
+\caption{\textbf{Community compositions after seven additional transfers without coalescence.}
+Each color of the pie plots corresponds to a different exact sequence variant
+(\methodsref{methods:sequencing}).
+Replicate 2 of community 1 from glutamine,
+as well as replicate 2 of community 8 from citrate (highlighted) were removed based on their
+dissimilarity to the other two replicates (details in code for data analysis, see
+\nameref{datacode}).
+Communities clustered in dashed boxes shared the same dominant species as revealed by
+sequencing data.
+For communities enclosed in red boxes, sequencing data showed that the species isolated by plating
+was not detectable in the community after seven additional transfers (i.e. the dominant was
+incorrectly identified) and were therefore excluded from downstream analyses.}
+\label{figS1}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[!h]
+\centering
+\internallinenumbers
+\includegraphics[width=12.5cm,keepaspectratio]{figs/figS2.pdf}
+\caption{\textbf{Alternative metrics of community distance.}
+Quantifying coalescence outcomes using different metrics of community similarity
+(\methodsref{methods:metrics}) gives similar results to those shown in \figref[a]{fig2}.
+Metrics that account for the relative species abundances (\textbf{a.} Jensen-Shannon similarity or
+\figref[a]{fig2} Bray-Curtis similarity) yield higher correlations than less quantitative metrics that only
+account for species presence/absence (\textbf{b.} Jaccard similarity or \textbf{c.} Fraction of endemic
+invasive species persisting in the coalesced community).}
+\label{figS2}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[!h]
+\centering
+\internallinenumbers
+\includegraphics[width=10cm,keepaspectratio]{figs/figS3.pdf}
+\caption{\textbf{Dominant species have limited effects on coalescence outcomes.}
+We repeated the analyses shown in \figref[a]{fig2} and \figref{figS2}, but this time we removed the
+dominants from the compositional data prior to quantifying community distances.
+The trends observed before are maintained.}
+\label{figS3}
+\end{figure}
+
 \end{document}

### 2021-05-07 draft/supp.tex
started introduction, added some supplementary figures, coalescence.py ready for simulations with simple, complex and intermediate primary resource

--- 2021-04-15 draft/supp.tex
@@
 \usepackage{zref-xr,zref-user} % more cross-references formatting
+\zxrsetup{toltxlabel}
+\zexternaldocument*{draft}
 \usepackage[square,numbers,sort&compress]{natbib}
-\usepackage{xcolor} % text color
+\usepackage[table,xcdraw]{xcolor} % text and table background colors
 \usepackage[margin=3cm]{geometry} % margins
@@
 \hypersetup{citecolor={blue}}
+\definecolor{lightgray}{gray}{0.9}
+\newcommand{\figref}[2][]{%
+  \hyperref[{#2}]{%
+    Figure~\ref*{#2}%
+    \ifx\\#1\\%
+    \else
+      #1%
+    \fi
+  }%
+}
+\newcommand{\tableref}[1]{%
+  \hyperref[{#1}]{%
+   Table~\ref*{#1}%
+  }%
+}
+\newcommand{\methodsref}[1]{%
+  \hyperref[{#1}]{%
+   Methods:~\nameref*{#1}%
+  }%
+}
 
@@
 \maketitle
-\tableofcontents
 \newpage
 
-\section{Supplementary Methods}\label{supp-methods}
+\section*{Supplementay Figures}\label{supp-figs}
 
-\subsection{Data processing}\label{supp-methods:data-proc}
+This is how you reference something from the draft...
 
-\subsubsection{Examining community composition}
-\label{supp-methods:data-proc:community-composition}
-
-We first analyze the composition of the communities assembled in each carbon
-source by examining the relative abundances of the exact sequence variants
-(ESVs) obtained from 16S rRNA gene sequencing. We define a vector $\mathbf{s}
-= \left( s_1, s_2, \cdots, s_M \right)$ such that $s_i$ represents the relative
-abundance
-of the $i$-th ESV, being $M$ the total number of unique ESVs. We assembled
-a total of 16 communities, 8 in glutamine
-and 8 in citrate (see Methods \mr), with 3 biological replicates of each. We
-start
-by comparing the composition of replicate communities in terms of their ESV
-relative abundances. To do this, we measure the distance $d$ between two
-communities (with ESV abundances $\mathbf{s}$ and $\mathbf{s}'$ respectively) as
-
-\begin{equation}
-d \left( \mathbf{s}, \mathbf{s}' \right) =
-\sqrt{\sum_i{\left(s_i - s'_i\right)^2}}
-\label{supp-eqn:esvdist}
-\end{equation}
-%
-We compute all the distances across replicates of the same community, obtaining
-a set of 48 distances (16 communities $\times$ 3 pairwise distances between the
-three replicates) that we will denote as $\left\{d\right\}$
-
-\begin{equation} 
-\left\{d\right\} = 
-\left\{ d\left(\mathrm{community} \; i \; \mathrm{replicate} \; j,
-\mathrm{community} \; i \; \mathrm{replicate} \; j' \right) \;
-\text{for all $i$, $j$, $j' \neq j$} \right\}
-\end{equation}
-%
-We then define a threshold $d_T$ as
-
-\begin{equation}
-d_T = \mathrm{Q}4 \left( \left\{d\right\} \right) + 
-\mathrm{IQR} \left( \left\{d\right\} \right)
-\label{supp-eqn:threshold}
-\end{equation}
-%
-Where Q4 and IQR represent the fourth quantile and the interquartile range
-respectively. Note that this threshold is analogous to the one used in
-standard boxplots to identify outliers.
-
-Finally, we discard the samples that are at a distance larger than $d_T$ from
-all other replicates of the same community. These are replicate 2 of community
-1 in glutamine and replicate 2 of community 8 in citrate (see Figure S1 \mr).
-The remaining replicates were used in further analyses. We took the average
-ESV abundance across replicates when indicated.
-
-\subsubsection{Special cases}\label{supp-methods:data-proc:special}
-
-Blablabla.
-
-\subsection{Inferring species abundance from sequencing data}
-\label{supp-methods:species-from-seq}
-
-Consider a community of $N$ species with relative abundances represented
-by the components of a vector $\mathbf{x} = \left( x_1, x_2, \dotsc,
-x_N\right)$ such that $x_i$ is the relative abundance of the $i$-th species.
-Naturally, the conditions
-%
-\begin{equation}
-\sum_i x_i = 1
-\label{supp-eqn:normx_1}
-\end{equation}
-%
-and
-%
-\begin{equation}
-0 \leq x_i \leq 1 \; \textrm{for all} \; i
-\label{supp-eqn:normx_2}
-\end{equation}
-%
-are satisfied.
-
-Performing 16S rRNA gene sequencing on such a community yields a list of
-sequences (exact sequence variants or ESVs) and their respective abundance.
-Normalizing by the total number of sequences, we can obtain a vector $\mathbf{s}
-= \left( s_1, s_2, \dotsc \right)$ that also satisfies the normalization
-conditions of equations \ref{supp-eqn:normx_1} and \ref{supp-eqn:normx_2}.
-Only if there are as many ESVs as species does $\mathbf{s}$ have length $N$
-(the same as $\mathbf{x}$), but in the most general case it is possible that
-\textit{a)} multiple species share a same ESV and/or \textit{b)} species carrying
-multiple copies
-of the 16S rRNA gene have different sequences for each copy \cite{Vetrovsky2013}. We denote
-the length of $\mathbf{s}$ as $M$, which need not be equal to $N$.
-
-By sequencing each species individually, we can build a $M \times N$ 
-matrix $\mathbf{Q}$ such
-that the element in the $(i,j)$ position, $q_{ij}$, represents the frequency of
-the $i$-th ESV when the $j$-th species is sequenced. $\mathbf{Q}$ can be used
-to determine the fraction of sequences of a given ESV that are obtained from
-sequencing a community with any arbitrary composition. In other words,
-$\mathbf{Q}$ maps $\mathbf{x}$ to $\mathbf{s}$:
-
-\begin{equation}
-\mathbf{Q} \cdot \mathbf{x} = \mathbf{s}
-\label{supp-eqn:Qxs}
-\end{equation}
-
-Note that equation \ref{supp-eqn:Qxs} is only true if all species have the same
-number of copies of the 16S rRNA gene. Otherwise, species with high copy numbers
-will yield more sequences, thus leading to an over-representation and vice-versa.
-Information on the 16S copy number for every species in the communities would be
-required to overcome this limitation, but this is not feasible when dealing with
-highly diverse communities. For our purpose in this
-work, we will need to assume that the copy numbers of the 16S are relatively
-conserved across species and equation \ref{supp-eqn:Qxs} holds. Additionally,
-we have sequencing data from a subset of the species in our
-communities, but not from all of them. This means that we cannot build a complete
-$\mathbf{Q}$ matrix. To address this, we first identify every ESV obtained
-from community sequencing that we cannot map back to (at least) one of the
-species for which we have single-species sequencing information. We then
-assume that each
-of those ESVs maps uniquely to a single species. This gives us a $\mathbf{Q}$
-matrix of the form:
-
-\begin{equation}
-\mathbf{Q} = \; \; \; \; \; \begin{matrix}
- &
-
-\begin{matrix}
-\; \; \;
-\rotatebox{90}{\scriptsize species 1} \; \; \; & 
-\rotatebox{90}{\scriptsize species 2} \; \; \; &
-\rotatebox{90}{\scriptsize species 3} \; \; &
-\cdots \; \; &
-\rotatebox{90}{\scriptsize species $N-2$} \! \! & 
-\rotatebox{90}{\scriptsize species $N-1$} \! \! &
-\rotatebox{90}{\scriptsize species $N$}
-\end{matrix}
-
-\\
-
-\begin{matrix*}[r]
-\rotatebox{0}{\scriptsize sequence 1} \\
-\rotatebox{0}{\scriptsize sequence 2} \\
-\rotatebox{0}{\scriptsize sequence 3} \\
-\vdots \; \; \; \\
-\rotatebox{0}{\scriptsize sequence $M-2$} \\ 
-\rotatebox{0}{\scriptsize sequence $M-1$} \\
-\rotatebox{0}{\scriptsize sequence $M$}
-\end{matrix*}
-
-&
-
-\begin{pmatrix}
-q_{11} & q_{12} & q_{13} &  & 0 & 0 & 0 \\
-q_{21} & q_{22} & q_{23} & \cdots & 0 & 0 & 0 \\
-q_{31} & q_{32} & q_{33} &  & 0 & 0 & 0 \\
- & \vdots &  & \ddots  &  &  &  \\
-0 & 0 & 0 &  & 1 & 0 & 0 \\ 
-0 & 0 & 0 &  & 0 & 1 & 0 \\ 
-0 & 0 & 0 &  & 0 & 0 & 1 
-\end{pmatrix}
-
-\end{matrix}
-\label{supp-eqn:Q_modified}
-\end{equation}
-%
-While a one-to-one mapping between species and ESVs may not be true in all cases,
-in practice the ESVs for which we make this assumption have low relative
-abundances in the communities, minimizing the impact of any potential artifact.
-
-Having built our $\mathbf{Q}$ matrix according to equation
-\ref{supp-eqn:Q_modified}, obtaining species abundances from sequencing data
-is relatively straigthforward: we need to find the $\mathbf{x}$ that satisfies
-equation \ref{supp-eqn:Qxs} for a given $\mathbf{s}$. However, in some cases
-sequencing error can introduce deviations in $\mathbf{s}$ that make it so the
-vector $\mathbf{x}$ that solves equation \ref{supp-eqn:Qxs} does not meet the
-boundary conditions in equation \ref{supp-eqn:normx_2}, i.e., some $x_i$ may be
-negative or greater than 1. These cases are obviously problematic, but we can
-avoid them by accounting for potential deviations in $\mathbf{s}$. We do this
-by solving the following nonlinear optimization problem: we want to obtain an
-estimate of the true community composition
-(we will denote the estimate as $\mathbf{\hat{x}}$) so that the product
-$\mathbf{Q} \cdot \mathbf{\hat{x}}$ is as close as possible to the $\mathbf{s}$
-obtained from sequencing while
-having $\mathbf{\hat{x}}$ satisfy the normalization condition in equation
-\ref{supp-eqn:normx_1} and the boundary conditions in equation
-\ref{supp-eqn:normx_2}. For every possible $\mathbf{\hat{x}}$, we can define a vector
-$\boldsymbol{\upvarepsilon} = \boldsymbol{\upvarepsilon} \left( \mathbf{\hat{x}} \right) = 
-\left( \varepsilon_1, \varepsilon_2, \cdots, \varepsilon_M \right)$ as
-
-\begin{equation}
-\boldsymbol{\upvarepsilon} \left( \mathbf{\hat{x}} \right) =
-\mathbf{Q} \cdot \mathbf{\hat{x}} - \mathbf{s}
-\end{equation}
-%
-and a function $f \left( \mathbf{\hat{x}} \right)$ as
-
-\begin{equation}
-f \left( \mathbf{\hat{x}} \right) =
-\left| \boldsymbol{\upvarepsilon} \left( \mathbf{\hat{x}} \right) \right| ^2 =
-\sum_i \varepsilon_i^2 \left( \mathbf{\hat{x}} \right)
-\end{equation}
-%
-We can also define a function $h \left( \mathbf{\hat{x}} \right)$ as
-
-\begin{equation}
-h \left( \mathbf{\hat{x}} \right) = 1 - \sum_i\hat{x}_i
-\end{equation}
-%
-The best estimation for the species composition of a community will be the
-$\mathbf{\hat{x}}$ that minimizes $f \left( \mathbf{\hat{x}} \right)$ while
-satisfying the normalization condition (eq. \ref{supp-eqn:normx_1}) and the
-boundary conditions (eq. \ref{supp-eqn:normx_2}). We solve
-this problem using the \texttt{nloptr} package for R \cite{nloptr}, passing it
-$f \left( \mathbf{\hat{x}} \right)$ as the function to minimize,
-$h \left( \mathbf{\hat{x}} \right) = 0$ as an equality constraint,
-and 0 and 1
-as the lower and upper bounds for the entries of $\mathbf{\hat{x}}$, and using
-the augmented Lagrangian algorithm \cite{auglag1,auglag2}. Further
-details can be found in the code for the analysis (see section 
-\nameref{supp-datacode} of this Supplementary Material).
-
-\clearpage
-
-\section{Data \& code availability}
-\label{supp-datacode}
-
-Data and code for the analyses in this article, as well as code for the
-\texttt{community-simulator} package updated with the new functionalities,
-is available at \url{https://github.com/jdiazc9/coalescence}.
-
-\clearpage
-
-\section{Supplementary Results}\label{supp-results}
-
-Blabla
-
-\subsection{Example subsection}
-\label{supp-results:example}
-
-Blabla
-
-\clearpage
-
-\section{Supplementay Figures}\label{supp-figs}
-
-Blablabla.
+\begin{figure}[!h]
+\centering
+\includegraphics[width=15cm,keepaspectratio]{figs/figS1.pdf}
+\caption{\textbf{Community compositions after seven additional transfers without coalescence.}
+Each color of the pie plots corresponds to a different exact sequence variant
+(\zref{fig1}).}
+\label{figS1}
+\end{figure}
 

### 2021-05-07 coalescence.py

@@ -229,42 +229,12 @@ def mysim(p,q,pq):
 
 ### SEED AND STABILIZE COMMUNITIES
 
-# assumptions
-
-"""
-# quasi-default assumptions
+# general assumptions
 assumptions = community_simulator.usertools.a_default.copy()
 assumptions['n_wells'] = 100 # number of communities
 assumptions['S'] = 50 # number of species sampled at initialization
-assumptions['SA'] = [400, 400, 400] # [100, 100, 100] # number of species per specialist family
-assumptions['Sgen'] = 120 # 30 # number of generalists
-assumptions['l'] = 0.8 # leakage fraction
-assumptions['MA'] = [10, 10, 10] # [30, 30, 30] # number of resources per resource class
-assumptions['response'] = 'type I'
-assumptions['regulation'] = 'independent' # 'independent' is standard, 'energy' or 'mass' enable preferential consumption of resources
-assumptions['sampling'] = 'Gamma' # 'Binary', 'Gaussian', 'Uniform' or 'Gamma' (sampling of the matrix c)
-assumptions['supply'] = 'off'
-assumptions['R0_food'] = 1000
-assumptions['m'] = 0 # turn off mortality (?)
-assumptions['q'] = 0.9 #0.9 # preference strength (0 for generalist and 1 for specialist)
-assumptions['c0'] = 0.0 # background consumption rate in binary model
-assumptions['c1'] = 1.0 # specific consumption rate in binary model
-assumptions['sigc'] = 3 #3 # standard deviation of sum of consumption rates for Gaussian and Gamma models
-assumptions['sparsity'] = 0.9 #0.05 # variability in secretion fluxes among resources (must be less than 1)  
-assumptions['fs'] = 0.45 #0.45 # fraction of secretion flux to resources of the same type as the consumed one
-assumptions['fw'] = 0.45 #0.45 # fraction of secretion flux to waste resources
-assumptions['metabolism'] = 'specific' # 'common' uses a common D matrix for all species, 'specific' uses a different matrix D for each species
-assumptions['rs'] = 0.0 # control parameter for resource secretion: 0 means random secretions, 1 means species only secrete resources they can consume (relevant only if 'metabolism' is 'specific')
-"""
-
-# try these assumptions
-assumptions = community_simulator.usertools.a_default.copy()
-assumptions['n_wells'] = 50 # number of communities
-assumptions['S'] = 50 # number of species sampled at initialization
[…]

### 2021-05-08 draft/draft.tex
more intro, more refs

@@
 in river courses where terrestrial microbial communities mix with aquatic microorganisms
-\cite{Luo2020,Vass2021}
-or in soil communities being invaded by aerially dispersed bacteria and funghi \cite{Evans2019}.
-The human digestive system can even get invaded several times a day by the microbial consortia
-that reside on the ingested food.
+\cite{Mansour2018,Luo2020,Vass2021}
+or in soil communities being invaded as a result of
+tillage and outplanting \cite{Rillig2016} or by
+aerially dispersed bacteria and funghi \cite{Evans2019}.
+The human digestive system can get invaded several times a day by the microbial consortia
+that reside on the ingested food, and the skin microbiota is also subject to
+invasions when making contact with environmental sources of microbes \cite{Vandegrift2019}.
+
+The phenomenon by which entire microbiomes invade one another has been termed
+\textit{community coalescence} \cite{Rillig2015}.
 Ecologists have long contemplated the idea that interactions between multiple co-invading species
 can produce correlated invasional outcomes
-\cite{Gilpin1994,Grosholz2005,Green2011,Livingston2013,Rillig2015,OLoughlin2017}.
+\cite{Gilpin1994,Grosholz2005,Green2011,Livingston2013,Rillig2015,OLoughlin2017,Castledine2020}.
 One example is the hypothesis known as \textit{invasional meltdown}, which proposes that
@@
 facilitate future invasions \cite{Simberloff1999,Simberloff2006,Gurevitch2006}.
-
-The phenomenon by which entire microbiomes invade one another has been termed
-\textit{community coalescence} \cite{Rillig2015}.
-In spite of its clear potential importance, the role of coalescence in microbiome assembly is
+However, and
+in spite of its clear potential importance, the role of coalescence in microbiome assembly is
 only beginning to be addressed and little is known about the mechanisms that govern it and its
 potential implications.
-Early mathematical models of community-community invasions \cite{Gilpin1994,Toquenaga1997},
-as well as more recent work including environmental construction (e.g., consumer-resource models)
-\cite{Tikhonov2016,Tikhonov2017,Lechon2021}, suggest that high-order invasion effects are common
+Early mathematical models of community-community invasions \cite{Gilpin1994,Toquenaga1997}
+as well as more recent work
+\cite{Tikhonov2016,Tikhonov2017,Vila2019,Lechon2021}
+suggest that high-order invasion effects are common
 during community coalescence. Communities that have a previous history of coexistence may exhibit an
@@
 same community \cite{Livingston2013,Sierocinski2017}.
-The situation where ecological partners in the invading community recruit one another into the final
+The situation where ecological partners in the invading community recruit each other into the final
 coalesced community has been called \textit{ecological co-selection} \cite{Rillig2017,Sierocinski2017}.
 
+The mechanisms of ecological co-selection during community coalescence are still poorly understood.
+Do a few key species recruit everyone else, or are collective interactions among all species
+(including the rarer members of the community) relevant for coalescence outcomes?
+While it is reasonable to expect species with larger population sizes to have a proportionally oversized effect,
+natural communities tend to be highly diverse \cite{Louca2016}
+and
+the role played by the less abundant species has long been subject to debate \cite{Winfree2015}.
+Laboratory cultures have also been found to contain uneven distributions of multiple strains that feed
+off the metabolic secretions of the dominant species \cite{Rosenzweig1994,Goldford2018}.
+The fate of these sub-dominant taxa may be dependent on the invasional success of their dominant
+species, or, alternatively, the dominant itself may owe its dominance (at least in part) to cross-feeding
+or other forms of facilitation from the rarer members of the population.
+These scenarios would give rise to ``top-down'' or ``bottom-up'' community cohesiveness, respectively.
+Either of these forms of co-selection could, in principle, be positive (recruitment) or negative (antagonism),
+as illustrated in \figref[e]{fig1}.
+Which of these situations are typically found in nature?
+Addressing this question has been experimentally challenging in the past \cite{Rillig2017,Sierocinski2017}.
+
+\vspace{1.5cm}
+
+Check this reference \cite{Castledine2020}
+
+Uncovering the rules of microbial community invasions \cite{Vila2019}
+
 \section*{Results \& Discussion}\label{results-discussion}
+
+Resource availability modulates biodiversity–invasion relationships by altering competitive interactions \cite{Yang2017}
+
+Resource pulses can alleviate the biodiversity–invasion relationship in soil microbial communities \cite{Mallon2015}
 

### 2021-05-10 draft/draft.tex
starting results, cleaning repo

@@
 \newcommand{\methodsref}[1]{%
-  \hyperref[{#1}]{%
-   Methods:~\nameref*{#1}%
+  \hyperref[{methods:#1}]{%
+   Methods:~\nameref*{methods:#1}%
   }%
@@
 
-\linenumbers
+%\linenumbers
 
@@
 \end{abstract}
-
-\section*{Test section}
-
-This is an example cite \cite{Luo2020,Vass2021}.
-This is how you refer to \figref{fig1}.
-This is how you refer to \figref[a]{fig1}.
-This is how you refer to \tableref{fig1}.
-This is how you refer to \figref{figS1}.
-This is how you refer to the section \nameref{methods:sim} of the \nameref{methods}.
-This is a cleaner way to refer to the \methodsref{methods:sim}.
 
@@
 
-\vspace{1.5cm}
-
-Check this reference \cite{Castledine2020}
-
-Uncovering the rules of microbial community invasions \cite{Vila2019}
+In previous work, we have shown that a large amount of soil and plant microbiomes can be cultured
+\textit{ex situ} in synthetic minimal environments with a single supplied limiting resource under serial
+growth-dilution cycles \cite{Goldford2018} (\figref[a-b]{fig1}). Under these conditions, environmental
+microbiomes spontaneously re-assemble into complex multi-species communities sustained by dense
+cross-feeding facilitation networks \cite{Goldford2018}. In addition,
+and just like in natural assemblies,
+species abundance distributions
+in these communities are generally long-tailed and uneven (\figref[d]{fig1} and \figref{figS1}),
+with the dominant (most abundant) species typically comprising most of the biomass ($\text{median}=46\%$,
+\figref{figS1}). Because these communities are easy to manipulate and grow in high throughput, and
+are largely made up by culturable members, they represent good test cases to investigate ecological
+co-selection during community coalescence. Here we focus on the dominants and ask whether they can
+co-select or be co-selected by the sub-dominant species in their communities (henceforth referred to as
+their \textit{cohorts}, \figref[c]{fig1}).
+Our results confirm that positive top-down co-selection is common, but its effects are weak.
+In contrast, bottom-up co-selection can be very strong, and positive co-selection is far more common
+than negative co-selection.
+We then turn to a microbial consumer-resource model (microCRM)
+\cite{Goldford2018,Marsland2019,Marsland2020}
+that is able to capture the dynamics of microbial communities dominated by metabolic interactions,
+as is the case for the ones assembled in our experimental conditions \cite{Goldford2018}.
+We show that the empirically observed trends in ecological co-selection are reproduced
+with minimal model assumptions, and that
+tuning the complexity of the metabolic interactions in our \textit{in silico} communities can modulate
+the recurrence of top-down or bottom-up co-selection.
+Our findings indicate that collective interactions play an important role at dictating community structure
+during coalescence.
 
@@
 
-Resource availability modulates biodiversity–invasion relationships by altering competitive interactions \cite{Yang2017}
+We collected eight natural microbiomes from different soil and plant environmental samples
+(\figref[a]{fig1})
+and used them to inoculate our synthetic communities, which were stabilized 
+in serial batch-culture biorreactors for 84 generations in synthetic minimal media containing either
+glutamine or citrate as the only supplied carbon source
+(\figref[b]{fig1}, \methodsref{community-assembly}).
+We isolated the dominant species of every community
+(\methodsref{dominants})
+and identified them by Sanger-sequencing their 16S rRNA gene
+(\methodsref{sequencing}), which correctly matched the dominant
+Exact Sequence Variant (ESV) \cite{Callahan2016,Callahan2017} found through
+community-level 16S illumina sequencing
+(\figref{figS1}).
+These dominants remained at high frequency after seven additional transfers
+with the exception of two of the citrate communities and one of the glutamine
+communities (where the dominants were presumably a transiently dominating species)
+that were excluded from further analysis
+(\figref{figS1}).
+Similarly, pairs of communities where the dominants shared a same 16S sequence
+and had similar colony morphology were excluded
+(\figref{figS1}).
 
-Resource pulses can alleviate the biodiversity–invasion relationship in soil microbial communities \cite{Mallon2015}
+\subsubsection*{Top-dow co-selection is weak}
+
+If communities being coalesced were highly cohesive from the top-down,
+the dominant species would co-select the rarer members of its community during coalescence
+(\figref[e]{fig1}, left panels).
+In this scenario, we would expect the outcome of community coalescence to be predicted by
+which of the two dominants is most competitive in pairwise competition.
+Coalescence outcomes can be quantified by the similarity between the coalesced and the
+invasive communities
+(\methodsref{metrics}).
+To test this hypothesis, we performed all pairwise competitions between dominant species
+in glutamine and citrate environments by mixing them 1:1 on their native media and propagating
+the cultures for seven serial transfers, roughly 42 generations
+(\methodsref{competitions}).
+We then performed all possible pairwise community coalescence experiments by mixing
+equal volumes of the communities and propagating the resulting cultures for seven extra
+transfers
+(\figref[f]{fig1}).
+The frequencies of all species in both community-community and dominant-dominant competitions
+were determined by 16S illumina sequencing
+(\methodsref{sequencing}).
+We found that the pairwise competitive ability of an invasive dominant
+is only weakly predictive of the performance of the invasive community in coalescence,
+as quantified by the relative Bray-Curtis similarity between the coalesced and invasive communities
+(\figref[a]{fig2}, $R^2=0.15$, $p<$ for glutamine and $R^2=0.15$, $p<$ for citrate).
+Alternative quantifications of community distance also yield weak effects,
+and even more so when the metric used
+accounts only for the presence/absence of specific species and not for their relative abundance in
+the communities
+(\figref{figS2}).
 
@@
 \textbf{a.}~Coalescence outcomes are quantified by the relative Bray-Curtis similarity
-($Q$) between the coalesced and invasive communities (\methodsref{methods:metrics}).
+($Q$) between the coalesced and invasive communities (\methodsref{metrics}).
 These outcomes are predicted by the pairwise competition between the invasive and
@@
 Each color of the pie plots corresponds to a different exact sequence variant
-(\methodsref{methods:sequencing}).
+(\methodsref{sequencing}).
 Replicate 2 of community 1 from glutamine,
@@
 Quantifying coalescence outcomes using different metrics of community similarity
-(\methodsref{methods:metrics}) gives similar results to those shown in \figref[a]{fig2}.
+(\methodsref{metrics}) gives similar results to those shown in \figref[a]{fig2}.
 Metrics that account for the relative species abundances (\textbf{a.} Jensen-Shannon similarity or
@@
 
+\section*{Test section}
+
+This is an example cite \cite{Luo2020,Vass2021}.
+This is how you refer to \figref{fig1}.
+This is how you refer to \figref[a]{fig1}.
+This is how you refer to \tableref{fig1}.
+This is how you refer to \figref{figS1}.
+This is how you refer to the section \nameref{methods:sim} of the \nameref{methods}.
+This is a cleaner way to refer to the \methodsref{sim}.
+
+Check this reference \cite{Castledine2020}
+
+Resource availability modulates biodiversity–invasion relationships by altering competitive interactions \cite{Yang2017}
+
+Resource pulses can alleviate the biodiversity–invasion relationship in soil microbial communities \cite{Mallon2015}
+
 \end{document}

### 2021-05-10 multimatrix.py

@@ -1,204 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-Created on Tue Dec  1 17:53:14 2020
-
-@author: Juan
-"""
-
-### INITIALIZIATION
-
-# reset variables etc.
-from IPython import get_ipython
-get_ipython().magic('reset -sf')
-
-# libraries
-import community_simulator
-import community_simulator.usertools
-import community_simulator.visualization
-import numpy as np
-import pandas as pd
-# import matplotlib
-#import matplotlib.pyplot as plt
-#import seaborn as sns
-#colors = sns.color_palette()
-# %matplotlib inline
-
-import random
-#import math
-#import copy
-
-# set random and numpy's seed manually for reproducibility
-np.random.seed(0)
-random.seed(0)
-
-
-
-
-
-# assumptions
-assumptions = community_simulator.usertools.a_default.copy()
-assumptions['n_wells'] = 10
-assumptions['S'] = 100 # number of species sampled at initialization
-assumptions['SA'] = 200 # [100, 100, 100] # number of species per specialist family
-assumptions['Sgen'] = 0 # number of generalists
-assumptions['l'] = 0.8 # leakage fraction
-assumptions['MA'] = 100 # [30, 30, 30] # number of resources per resource class
-assumptions['response'] = 'type I'
-assumptions['sampling'] = 'Binary' # 'Gamma'
-assumptions['supply'] = 'off'
-assumptions['R0_food'] = 1000
-assumptions['m'] = 0 # turn off mortality (?)
-assumptions['c0'] = 0.01 # background consumption rate in binary model
-assumptions['c1'] = 1 # specific consumption rate in binary model
-
-params = community_simulator.usertools.MakeParams(assumptions)
-Dlist = [params['D'], params['D'], params['D']]
-
-N, R = community_simulator.usertools.MakeInitialState(assumptions)
-N = N.iloc[:,0]
-R = R.iloc[:,0]
-
-
-
-
-
-# taken from community_simulator.usertools as is
-def MakeConsumerDynamics(assumptions):
-    """
-    Construct resource dynamics. 'assumptions' must be a dictionary containing at least
-    three entries:
-    
-    response = {'type I', 'type II', 'type III'} specifies nonlinearity of growth law
-    
-    regulation = {'independent','energy','mass'} allows microbes to adjust uptake
-        rates to favor the most abundant accessible resources (measured either by
-        energy or mass)
-    
[…]

### 2021-05-10 myFunctions.py

@@ -1,44 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-Created on Wed Oct 28 15:55:43 2020
-
-@author: Juan
-"""
-
-# libraries
-import math
-
-# Kullback-Leibler divergence - using log2 so result is within [0,1]
-def kldiv(p,q):
-        
-    # remove elements that are zero in both p and q (unused dimensions)
-    n = [i for i in range(len(p)) if p[i]>0 or q[i]>0]
-    p = [p[i] for i in n]
-    q = [q[i] for i in n]
-    
-    # define this operation to handle the p[i]->0 limit
-    def mylog(a,b):
-        if a==0:
-            return(0)
-        else:
-            return(a*math.log2(a/b))
-    
-    d = sum([mylog(p[i],q[i]) for i in range(len(p))])
-    return d
-
-# Jensen-Shannon distance
-def js(p,q):
-    
-    # return an error if...
-    if (not len(p)==len(q) or
-    not sum(p)==1 or
-    not sum(q)==1 or
-    all(p_i==0 for p_i in p) or
-    all(q_i==0 for q_i in q)):
-        return "error"
-    
-    else:
-        m = [(p[i]+q[i])/2 for i in range(len(p))]
-        js_div = 1/2 * kldiv(p,m) + 1/2 * kldiv(q,m)
-        js_dist = math.sqrt(js_div)
-        return js_dist

### 2021-05-10 originals.py

@@ -1,83 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-Created on Wed Dec  9 15:09:01 2020
-
-@author: Juan
-"""
-# original versions of community-simulator methods
-def CompressParams(not_extinct_consumers,not_extinct_resources,params,dimensions,S,M):
-    params_comp = params.copy()
-    if 'SxM' in dimensions.keys():
-        for item in dimensions['SxM']:
-            if item in params_comp.keys():
-                assert np.shape(params_comp[item])==(S,M), 'Invalid shape for ' + item + '. Please update dimensions dictionary with correct dimensions.'
-                params_comp[item]=params_comp[item][not_extinct_consumers,:]
-                params_comp[item]=params_comp[item][:,not_extinct_resources]
-    if 'MxM' in dimensions.keys():
-        for item in dimensions['MxM']:
-            if item in params_comp.keys():
-                assert np.shape(params_comp[item])==(M,M), 'Invalid shape for ' + item + '. Please update dimensions dictionary with correct dimensions.'
-                params_comp[item]=params_comp[item][not_extinct_resources,:]
-                params_comp[item]=params_comp[item][:,not_extinct_resources]
-    if 'SxS' in dimensions.keys():
-        for item in dimensions['SxS']:
-            if item in params_comp.keys():
-                assert np.shape(params_comp[item])==(S,S), 'Invalid shape for ' + item + '. Please update dimensions dictionary with correct dimensions.'
-                params_comp[item]=params_comp[item][not_extinct_consumers,:]
-                params_comp[item]=params_comp[item][:, not_extinct_consumers]
-    if 'S' in dimensions.keys():
-        for item in dimensions['S']:
-            if item in params_comp.keys():
-                if type(params_comp[item]) == np.ndarray:
-                    assert len(params_comp[item])==S, 'Invalid length for ' + item + '. Please update dimensions dictionary with correct dimensions.'
-                    params_comp[item]=params_comp[item][not_extinct_consumers]
-    if 'M' in dimensions.keys():
-        for item in dimensions['M']:
-            if item in params_comp.keys():
-                if type(params_comp[item]) == np.ndarray:
-                    assert len(params_comp[item])==M, 'Invalid length for ' + item + '. Please update dimensions dictionary with correct dimensions.'
-                    params_comp[item]=params_comp[item][not_extinct_resources]  
-    return params_comp
-
-
-# original versions of community-simulator methods
[…]

### 2021-05-10 temp.py

@@ -1,11 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-Spyder Editor
-
-This is a temporary script file.
-"""
-if True:
-    print("hola mundo")
-else:
-    print("fallo")
-        

### 2021-05-10 toyModel_metabolism.py

@@ -1,293 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-Created on Tue Jan 12 15:24:54 2021
-
-@author: Juan
-"""
-
-
-
-### INITIALIZIATION
-
-# reset variables etc.
-from IPython import get_ipython
-get_ipython().magic('reset -sf')
-
-# libraries
-import community_simulator
-import community_simulator.usertools
-import community_simulator.visualization
-import numpy as np
-import pandas as pd
-# import matplotlib
-import matplotlib.pyplot as plt
-import seaborn as sns
-colors = sns.color_palette()
-# %matplotlib inline
-
-import random
-# import math
-# import copy
-
-# set random and numpy's seed manually for reproducibility
-np.random.seed(0)
-random.seed(0)
-
-import time
-start_time = time.time()
-
-
-
-### USER-DEFINED FUNCTIONS
-
-# stabilize communities through serial passaging
-def stabilizeCommunities(plate):
-    
-    # parameters
-    dilution_factor = 1/100
-    n_transfers = 20
-    time_per_cycle = 1
-    # extinction_threshold = 1e-4 # we don't use this because we don't use StadyState()
-    
-    # run stabilization protocol
-    Ntraj, Rtraj = plate.RunExperiment(f = np.diag(np.ones(plate.n_wells))*dilution_factor,
-                                       T = time_per_cycle,
-                                       npass = n_transfers,
-                                       compress_resources=False,
-                                       compress_species=True)
-    
-    return([plate.N,plate.R])
-
-# stack plot with bars
-def myStackPlot(N):
-
-    # check number of wells and species
-    n_species = N.shape[0]
-    n_wells = N.shape[1]
-    
-    # normalize each column to get fractions
-    N = N/N.sum()
-    
-    # numpy data frame to list of lists (list of as many lists as species in total)
-    N = N.values.tolist()
-    
-    # colormap (adjusted to number of species)
-    random.seed(0)
-    cmap = plt.get_cmap("gist_stern")
-    cmap = cmap(np.linspace(0,1,num=n_species)).tolist()
-    random.shuffle(cmap)
-    
-    alpha = 0.75
-    cmap = [[227/255,0,0,alpha],
-            [255/255,161/255,77/255,alpha],
-            [51/255,153/255,237/255,alpha],
[…]

### 2021-05-10 draft/draft.tex
more results and figS4 added --am I getting overexcited with the interpretation of the ressults?

@@
 
-\subsubsection*{Top-dow co-selection is weak}
+\subsubsection*{The strength of top-down co-selection is dependent on the primary supplied resource}
 
@@
 (\methodsref{sequencing}).
-We found that the pairwise competitive ability of an invasive dominant
-is only weakly predictive of the performance of the invasive community in coalescence,
-as quantified by the relative Bray-Curtis similarity between the coalesced and invasive communities
-(\figref[a]{fig2}, $R^2=0.15$, $p<$ for glutamine and $R^2=0.15$, $p<$ for citrate).
-Alternative quantifications of community distance also yield weak effects,
-and even more so when the metric used
+
+We found that, for communities assembled in the glutamine environment,
+the pairwise competitive ability of an invasive dominant
+is only weakly predictive of the performance of the invasive community in coalescence
+(as quantified by the relative Bray-Curtis similarity between the coalesced and invasive communities),
+but this correlation was stronger for the citrate communities
+(\figref[a]{fig2}, $R^2=0.15$, $p<0.05$ for glutamine and $R^2=0.57$, $p<10^{-4}$ for citrate).
+Alternative quantifications of community distance yield similar results,
+with weaker effects when the metric used
 accounts only for the presence/absence of specific species and not for their relative abundance in
@@
 (\figref{figS2}).
+All these metrics include the presence of the dominant species themselves. To better disentangle the effect
+that these dominants have on the other member of their communities, we repeated the analysis this time
+excluding the dominant species from the compositional data, finding that our results still hold
+(\figref{figS3}).
+Furthermore, we observed that, in the glutamine communities,
+the relative frequency of a dominant against another in head-to-head pairwise competition
+is barely predictive of its relative frequency against that same other dominant when the
+cohorts are present too, i.e. after coalescence
+($R^2=0.04$, $p>0.05$, \figref{figS4}).
+On the other hand, pairwise competition between dominants of the citrate communities
+is highly correlated with the relative abundance of said dominants in community coalescence
+ ($R^2=0.83$, $p<10^{-8}$, \figref{figS4}).
+
+Together, these results suggest that the strength of top-down co-selection depends on
+the primary resource supplied to the coalesced communities.
+Communities assembled with citrate as the primary supplied resource
+display a strong degree of top-down co-selection,
+with the fates of the sub-dominant species determined to a large extent by dominant-dominant
+pairwise competition. This competition is, in turn, only weakly affected by the presence of the cohorts.
+For glutamine communities, although some degree of top-down co-selection is consistent with our data,
+the cohorts do not appear to be passively responding to their dominants.
+Instead, evidence suggests that the cohorts might be playing an active role in community coalescence.
+This finding led us to investigate the potential role of bottom-up ecological co-selection
+(\figref[e]{fig1}, right panels), i.e. whether in some cases the dominants may be co-selected
+for or against by their cohorts.
 
@@
 
-\subsection*{Stabilization of environmental communities in simple synthetic environments}
+\subsubsection*{Stabilization of environmental communities in simple synthetic environments}
 \label{methods:community-assembly}
@@
 geographical locations were collected with sterile
-tweezers or spatulas into 50mL sterile tubes (Fig. \mr).
+tweezers or spatulas into 50mL sterile tubes (\figref[a]{fig1}).
 One gram of each sample was allowed to
@@
 12 transfers at a dilution factor of 1:100,
-roughly equivalent to 80 generations per culture (Fig. \mr).
+roughly equivalent to 80 generations per culture (\figref[b]{fig1}).
 Cycloheximide was not added to the media after the first two transfers.
 
-\subsection*{Isolation of dominant species}\label{methods:dominants}
+\subsubsection*{Isolation of dominant species}\label{methods:dominants}
 
@@
 Next, $20\upmu$L of the cells diluted to $10^{-6}$ were plated in the corresponding synthetic
-minimal media and allowed to regrow at $30^\circ$C for 48h. Dominants were then inoculated
-into $500\upmu$L of fresh media and incubated still at $30^\circ$C for 48h.
+minimal media and allowed to regrow at $30^\circ$C for 48h. Dominants were then identified
+(\figref[c]{fig1}),
+inoculated into $500\upmu$L of fresh media and incubated still at $30^\circ$C for 48h.
 After this period, the communities stabilized for eleven transfers and the isolated dominants
-were ready for the competition experiments (Fig \mr) at the onset of the twelfth transfer.
+were ready for the competition experiments at the onset of the twelfth transfer.
 
-\subsection*{Dominant-dominant and community-community competitions}
+\subsubsection*{Dominant-dominant and community-community competitions}
 \label{methods:competitions}
@@
 Pairwise competitions were further propagated for seven serial transfers
-(roughly 42 generations; Fig. \mr) by transferring $8\upmu$L of
+(roughly 42 generations, \figref[f]{fig1}) by transferring $8\upmu$L of
 each culture to fresh media ($500\upmu$L).
 
-\subsection*{Determination of community composition by 16S sequencing}
+\subsubsection*{Determination of community composition by 16S sequencing}
 \label{methods:sequencing}
@@
 
-\subsection*{Metrics of community distance}
+\subsubsection*{Metrics of community distance}
 \label{methods:metrics}
@@
 
-\subsection*{Simulations}\label{methods:sim}
+\subsubsection*{Simulations}\label{methods:sim}
 
@@
 (\methodsref{metrics}) gives similar results to those shown in \figref[a]{fig2}.
-Metrics that account for the relative species abundances (\textbf{a.} Jensen-Shannon similarity or
-\figref[a]{fig2} Bray-Curtis similarity) yield higher correlations than less quantitative metrics that only
-account for species presence/absence (\textbf{b.} Jaccard similarity or \textbf{c.} Fraction of endemic
-invasive species persisting in the coalesced community).}
+Metrics that account for the relative species abundances (Bray-Curtis or Jensen-Shannon similarities)
+yield higher correlations than less quantitative metrics that only
+account for species presence/absence (Jaccard similarity or the fraction of endemic
+invasive species persisting in the coalesced community).
+\textbf{a.} Relative Jensen-Shannon similarity 
+($R^2=0.15$, $p<0.05$ for glutamine and $R^2=0.53$, $p<5\times10^{-4}$ for citrate)
+\textbf{b.} Relative Jaccard similarity
+($R^2=0.08$, $p>0.05$ for glutamine and $R^2=0.13$, $p>0.05$ for citrate)
+\textbf{c.} Relative survival of invasive endemic species after coalescence
+($R^2=0.16$, $p<0.05$ for glutamine and $R^2=0.04$, $p>0.05$ for citrate).}
 \label{figS2}
@@
 \includegraphics[width=10cm,keepaspectratio]{figs/figS3.pdf}
-\caption{\textbf{Dominant species have limited effects on coalescence outcomes.}
+\caption{\textbf{Dominant species have limited effects on coalescence outcomes quantification.}
 We repeated the analyses shown in \figref[a]{fig2} and \figref{figS2}, but this time we removed the
 dominants from the compositional data prior to quantifying community distances.
-The trends observed before are maintained.}
+The trends observed before are maintained.
+\textbf{a.} Relative Bray-Curtis similarity
+($R^2=0.20$, $p<0.01$ for glutamine and $R^2=0.34$, $p<0.005$ for citrate)
+\textbf{b.} Relative Jensen-Shannon similarity 
+($R^2=0.24$, $p<0.005$ for glutamine and $R^2=0.36$, $p<0.005$ for citrate)
+\textbf{c.} Relative Jaccard similarity
+($R^2=0.09$, $p>0.05$ for glutamine and $R^2=0.11$, $p>0.05$ for citrate)
+\textbf{d.} Relative survival of invasive endemic species after coalescence
+($R^2=0.18$, $p<0.05$ for glutamine and $R^2=0.08$, $p>0.05$ for citrate).}
 \label{figS3}
 \end{figure}
+
+\clearpage
+
+\begin{figure}[!h]
+\centering
+\internallinenumbers
+\includegraphics[width=8cm,keepaspectratio]{figs/figS4.pdf}
+\caption{\textbf{Pairwise competition of dominants with or without their cohorts.}
+In the horizontal axis, we plot the relative frequency of the invasive dominant species in head-to-head
+pairwise competition. In the vertical axis, we plot the same relative frequency when competition takes
+place in the presence of the cohorts, i.e. during community coalescence.
+$R^2=0.04$, $p>0.05$ for glutamine and $R^2=0.83$, $p<10^{-8}$ for citrate.}
+\label{figS4}
+\end{figure}
+
+\clearpage
 

### 2021-05-10 data/analyzeData.R

@@ -853,7 +853,9 @@ plot_this <- data.frame(carbon_source = character(0),
                         q_jensen_shannon_cohort = numeric(0),
                         q_jaccard_cohort = numeric(0),
                         q_endemic_cohort = numeric(0),
-                        f_singleinv = numeric(0))
+                        f_singleinv = numeric(0),
+                        f_multiinv = numeric(0),
+                        f_residentdom_multiinv = numeric(0))
 
 # run through carbon sources and communities
 for (cs in carbon_sources) {
@@ -1018,6 +1020,9 @@ for (cs in carbon_sources) {
         # fraction of dominant 1 (invasive) invading resident community with cohort
         f_multiinv <- as.numeric(cc_coalesced[dom_1,])
         
+        # fraction of dominant 2 (resident) when invasive community invades with cohort
+        f_residentdom_multiinv <- as.numeric(cc_coalesced[dom_2,])
+        
         # add to plotting table
         plot_this <- rbind(plot_this,
                            data.frame(
@@ -1040,7 +1045,8 @@ for (cs in carbon_sources) {
                              q_jaccard_cohort = q_jaccard_cohort,
                              q_endemic_cohort = q_endemic_cohort,
                              f_singleinv = f_singleinv,
-                             f_multiinv = f_multiinv
+                             f_multiinv = f_multiinv,
+                             f_residentdom_multiinv = f_residentdom_multiinv
                            ))
         
       }
@@ -1048,6 +1054,9 @@ for (cs in carbon_sources) {
     }
   }
 }
+
+# pseudo-pairwise competition of dominants with cohorts
+plot_this$f_multiinv_rel <- plot_this$f_multiinv/(plot_this$f_multiinv + plot_this$f_residentdom_multiinv)
 
 # check that sample positioning in wells is consistent
 stopifnot(all(plot_this$well_pairwise == plot_this$well_coalescence))
@@ -1189,6 +1198,41 @@ myplots[['q-vs-pairwise_cohorts']] <-
                                       color='transparent')) +
   coord_fixed()
 
+myplots[['pairwise-vs-coalescence']] <-
+  ggplot(data=unique(plot_this[,c('carbon_source','community_1','community_2','f_pairwise','f_multiinv_rel')]),
+       aes(x=f_pairwise,y=f_multiinv_rel,
+           color=carbon_source)) +
+  geom_point(size=3,
+             shape=1,
[…]

### 2021-05-11 draft/draft.tex
finished draft, time to run through alvaro

@@
 % title, authors, affiliations
-\title{Provisional Title}
+\title{Cohesiveness in microbial community coalescence}
 \author[1,2,3]{Authors}
@@
 aerially dispersed bacteria and funghi \cite{Evans2019}.
-The human digestive system can get invaded several times a day by the microbial consortia
-that reside on the ingested food, and the skin microbiota is also subject to
+Gut microbiomes can invade external communities
+through the host animal secretions \cite{Dutton2021},
+and the skin microbiota is also subject to
 invasions when making contact with environmental sources of microbes \cite{Vandegrift2019}.
@@
 can produce correlated invasional outcomes
-\cite{Gilpin1994,Grosholz2005,Green2011,Livingston2013,Rillig2015,OLoughlin2017,Castledine2020}.
+\cite{Gilpin1994,Simberloff1999,Grosholz2005,Simberloff2006,Gurevitch2006,Green2011,
+Livingston2013,Prior2015,Rillig2015,OLoughlin2017,Castledine2020}.
+\iffalse
 One example is the hypothesis known as \textit{invasional meltdown}, which proposes that
@@
 facilitate future invasions \cite{Simberloff1999,Simberloff2006,Gurevitch2006}.
+\fi
 However, and
@@
 Which of these situations are typically found in nature?
-Addressing this question has been experimentally challenging in the past \cite{Rillig2017,Sierocinski2017}.
+Previous theoretical and computational studies suggest that the answer is determined by the type and
+strength of the interactions of the community members with one another and with the environment
+\cite{Tikhonov2016,Vila2019,Lechon2021},
+but addressing this question has been experimentally challenging in the past \cite{Rillig2017,Sierocinski2017}.
 
@@
 cross-feeding facilitation networks \cite{Goldford2018}. In addition,
-and just like in natural assemblies,
+and just like in natural consortia,
 species abundance distributions
@@
 their \textit{cohorts}, \figref[c]{fig1}).
+
+Our results indicate that co-selection varies in direction and strength depending on the supplied limiting
+resource.
+This primary resource, in turn, has been shown to
+shape the structure and interactions of the communities \cite{Estrela2020}.
+We observe that, when top-down co-selection is weak, bottom-up co-selection can be very strong,
+with positive co-selection being far more common than negative co-selection.
+\iffalse
 Our results confirm that positive top-down co-selection is common, but its effects are weak.
@@
 than negative co-selection.
+\fi
 We then turn to a microbial consumer-resource model (microCRM)
@@
 that is able to capture the dynamics of microbial communities dominated by metabolic interactions,
-as is the case for the ones assembled in our experimental conditions \cite{Goldford2018}.
+as is the case for the ones assembled in our experimental conditions \cite{Goldford2018,Estrela2020}.
 We show that the empirically observed trends in ecological co-selection are reproduced
@@
 (\figref[b]{fig1}, \methodsref{community-assembly}).
+We chose these two carbon sources because they are metabolized through different
+pathways in bacteria \cite{Dimroth2004,Forchhammer2007},
+and we hypothesize that communities assembled in either resource will be supported by
+cross-feeding networks of distinct sets of metabolites \cite{Goldford2018,Estrela2020}
+thus leading to potentially variable degrees of community cohesiveness
+and coalescence outcomes \cite{Tikhonov2016,Tikhonov2017,Castledine2020,Lechon2021}.
 We isolated the dominant species of every community
@@
 Exact Sequence Variant (ESV) \cite{Callahan2016,Callahan2017} found through
-community-level 16S illumina sequencing
+community-level 16S Illumina sequencing
 (\figref{figS1}).
@@
 
-\subsubsection*{The strength of top-down co-selection is dependent on the primary supplied resource}
+\subsubsection*{Top-down ecological co-selection}
 
@@
 which of the two dominants is most competitive in pairwise competition.
-Coalescence outcomes can be quantified by the similarity between the coalesced and the
-invasive communities
-(\methodsref{metrics}).
 To test this hypothesis, we performed all pairwise competitions between dominant species
@@
 transfers
-(\figref[f]{fig1}).
+(\figref[f]{fig1}),
+and we quantified coalescence outcomes using the similarity between the coalesced and the
+invasive communities
+(\methodsref{metrics}).
 The frequencies of all species in both community-community and dominant-dominant competitions
-were determined by 16S illumina sequencing
+were determined by 16S Illumina sequencing
 (\methodsref{sequencing}).
@@
 (\figref[a]{fig2}, $R^2=0.15$, $p<0.05$ for glutamine and $R^2=0.57$, $p<10^{-4}$ for citrate).
-Alternative quantifications of community distance yield similar results,
+Alternative quantifications of community distance yielded similar results,
 with weaker effects when the metric used
-accounts only for the presence/absence of specific species and not for their relative abundance in
+accounted only for the presence/absence of specific species and not for their relative abundance in
 the communities
@@
 All these metrics include the presence of the dominant species themselves. To better disentangle the effect
-that these dominants have on the other member of their communities, we repeated the analysis this time
+that these dominants have on the other members of their communities, we repeated the analysis this time
 excluding the dominant species from the compositional data, finding that our results still hold
@@
 cohorts are present too, i.e. after coalescence
-($R^2=0.04$, $p>0.05$, \figref{figS4}).
+(\figref{figS4}, $R^2=0.04$, $p>0.05$).
 On the other hand, pairwise competition between dominants of the citrate communities
 is highly correlated with the relative abundance of said dominants in community coalescence
- ($R^2=0.83$, $p<10^{-8}$, \figref{figS4}).
+ (\figref{figS4}, $R^2=0.83$, $p<10^{-8}$).
 
@@
 pairwise competition. This competition is, in turn, only weakly affected by the presence of the cohorts.
-For glutamine communities, although some degree of top-down co-selection is consistent with our data,
+For glutamine communities, although some level of top-down co-selection is consistent with our data,
 the cohorts do not appear to be passively responding to their dominants.
@@
 for or against by their cohorts.
+
+\subsubsection*{Bottom-up co-selection during community coalescence}
+
+To study the effect of bottom-up co-selection, we compared the invasion success of the dominant
+from one of the communities when invading the other community in isolation versus when doing so
+accompanied by its cohort.
+In situations where positive bottom-up ecological co-selection is strong, we expect to see dominants
+invading more effectively with their cohorts than by themselves
+(\figref[b]{fig2}, green shaded region).
+On the other hand, a high degree of negative bottom-up co-selection would result in dominants reaching
+higher invasion success alone than in the presence of their cohorts
+(\figref[b]{fig2}, red shaded region).
+Alternatively, if both forms of bottom-up co-selection were weak, we would see a similar invasion
+success of the dominant regardless of whether it is invading with its cohort
+(\figref[b]{fig2}, gray shaded region).
+
+We performed a new round of invasion experiments where each dominant invaded every
+other community in isolation (\methodsref{competitions}). The dominant-invaded
+communities were then propagated for seven serial transfers, and the final
+community compositions were determined by 16S Illumina sequencing.
+We compared the relative abundances reached by the dominants in these experiments
+with the relative abundances of the same dominants in community coalescence
+(i.e. when they invade with their cohorts).
+\figref[b]{fig2} shows that, for communities assembled in citrate environments,
+the cohorts had little effect and the dominants were able to invade with comparable
+success regardless of being accompanied by them.
+This was expected since we had seen that coalescence between citrate communities
+tended to be determined by top-down co-selection.
+However, glutamine communities displayed a different behavior.
+A large number of dominants (roughly 50\%) that were not able to invade by themselves
+(or could only do so at very low final relative abundances, below 0.1)
+were able to reach much higher frequencies when they invaded together with their cohorts,
+which indicates that positive bottom-up co-selection is common in our experiments.
+In contrast, negative co-selection appears to be rare and no instances of it were
+observed in our experiments.
+
+\subsubsection*{Simulations of community coalescence with a consumer-resource model}
+
+We ran a set of simulations of community coalescence to better understand the mechanisms
+of ecological co-selection in our experiments. We used a microbial consumer-resource
+model (microCRM) \cite{Goldford2018,Marsland2019} implemented in the Community Simulator package
+for Python \cite{Marsland2020}. We chose this modeling framework because
+communities assembled under our experimental conditions (natural microbiomes re-assembled into
+multispecies communities through serial growth-dilution cycles in synthetic minimal media
+with a single carbon source)
+have been shown to be
+sustained by dense metabolic cross-feeding networks \cite{Goldford2018}, for which the
+microCRM provides a good description.
+We implemented a new feature in the package that makes individual species
+able to secrete different sets of byproducts when consuming a same resource
+(\methodsref{sim}).
+We also used sparse metabolic matrices, equivalent to assuming that species secrete only a few
+types of byproducts when consuming a resource. These assumptions are supported by the experimental
+observation that individual species can indeed secrete different sets of few metabolites
+to the environment when consuming a primary resource
+\cite{Harcombe2014,Pinu2018}.
+We reason that these choices are necessary to potentially generate community cohesiveness:
+if the secretions of all species were roughly similar, the effect of interchanging members of
+a community would be minimal. But in practice, species with a history of coexistence make up
+cohesive communities with highly specific cross-feeding structures \cite{Goldford2018,Estrela2020}.
+In short,
+we replicated our experimental protocol \textit{in silico}
+by generating a library of species that we divided into two pools. Each pool was used to seed a
+collection of invasive and resident communities, that were then allowed to stabilize through 20
+growth-dilution cycles. Stable communities were then used to reproduce our coalescence
+and dominant-dominant competition experiments,
+as well as the invasion experiments from the dominants alone.
+A full description of our modeling framework can be found in the \methodsref{sim} section.
+
+We carried out two sets of simulations under two different regimes.
+In one case, we generated a library of 30 resources divided into 3 resource classes
+(\methodsref{sim}). This corresponds to a situation where the digestion of a resource
+can yield a wide variety of secreted byproducts.
+We refer to this scenario as the \textit{complex resource} regime.
+Our second set of simulations was performed under a \textit{simple resource} regime, where we
+generated a library of only 15 resources belonging to a same class.
+For the simple resource, we found that coalescence outcomes were highly correlated with
+the competitive ability of the community dominants in pairwise competition
+(\figref[c]{fig2}),
+indicating a large degree of top-down positive co-selection.
+These dominants were able to reach similar fractions in the final communities when invading alone
+with respect to when invading with their cohorts
+(\figref[d]{fig2}),
+suggesting that bottom-up co-selection is weak in this scenario.
+On the other hand,
+simulations under the complex resource regime evidenced weak top-down co-selection but 
+common and strong bottom-up recruitment (\figref[e-f]{fig2}), with negative co-selection remaining rare.
+Simulations suggest that community cohesiveness is dependent on the metabolic interactions among species,
+which is in line with previous findings \cite{Goldford2018, Estrela2020}.
+These interactions are determined by the metabolic profile of the primary resource in the conditions
+of our simulations and, in turn, modulate the strength and direction of ecological co-selection.
+
+\subsubsection*{Conclusion}
+
+Understanding the mechanisms underlying the responses of microbial communities to invasions is
+an essential but poorly understood question in microbial ecology \cite{Rillig2015}.
+Theory has suggested that communities may exhibit an emergent cohesiveness
+\cite{Gilpin1994,Livingston2013,Tikhonov2016,Tikhonov2017},
+leading to members of the same community recruiting one another during
+community-community invasions.
+Our results provide direct experimental evidence of ecological co-selection in a large number
+of community coalescence experiments,
+and highlight the critical role played by the rarer, sub-dominant species in the generation
+of community cohesiveness.
+Our data suggests that the strength and direction of ecological co-selection is modulated
+by the supplied primary carbon source,
+which in turn shapes the cross-feeding network that dictates
+the structure of our communities \cite{Goldford2018,Estrela2020}.
+These observations,
+together with previous experimental results in different systems \cite{Sierocinski2017} as well as
+theoretical predictions
+\cite{Gilpin1994,Toquenaga1997,Tikhonov2016,Tikhonov2017,Vila2019,Lechon2021},
+suggest that collective interactions arising from metabolic feedbacks between microbes
+and the environment should be generically expected to produce ecological co-selection
+during community coalescence.
+
+Further work is required to clarify the relationship between metabolic interactions,
+community structure, cohesiveness and co-selection.
+The experimental system that we introduced in this work can be easily expanded so that
+large numbers of community coalescence experiments can be carried out in parallel.
+It thus represents a promising tool to explore the properties of microbial community
+coalescence in high throughput
+and test quantitative theories about its role in microbiome assembly.
 
@@
 
-\subsubsection*{Dominant-dominant and community-community competitions}
+\subsubsection*{Coalescence, competition and invasion experiments}
 \label{methods:competitions}
@@
 and normalized to 5ng/$\upmu$L in nuclease-free water (Qiagen)
-for subsequent 16S rRNA illumina sequencing.
+for subsequent 16S rRNA Illumina sequencing.
 16S rRNA amplicon library preparation was performed following a dual-index
@@
 support the idea of distinct species producing different sets of byproducts when
-growing in the same primary resource \mr. We call $D_{i\alpha\beta}$ to the
+feeding off the same primary resource
+\cite{Harcombe2014,Pinu2018}.
+We call $D_{i\alpha\beta}$ to the
 fraction of energy in the form of resource $\alpha$ secreted \textit{by species
@@
 For our simulations,
-we first generate a library of 660 species (divided into three specialist
-families of 200 members each
-and a generalist family of 60 members)
-and 30 resources (divided into three classes of 10 members each).
-We split this library into two non-overlapping pools of 330 species each.
+we first generated a library of 2400 species divided into three specialist
+families of 800 members each
+and a generalist family of 240 members.
+We split this library into two non-overlapping pools of 1320 species each.
 We randomly sample 50 species from each pool in equal ratios to seed
@@
 100 invasive communities respectively.
-We then grow and dilute the communities serially,
+We then let grow and diluted the communities serially,
 replenishing the primary
 resource after each dilution.
-We repeat the process 20 times to ensure generational equilibrium is
+We repeated the process 20 times to ensure generational equilibrium is
 achieved \cite{Goldford2018}.
-We then perform the \textit{in silico} experiments by using the
+We then performed the \textit{in silico} experiments by using the
 generationally stable communities to seed 100 coalesced communities
 that we again stabilize as described previously.
-Similarly, we identify the dominant (most
+Similarly, we identified the dominant (most
 abundant) species of every resident and invasive community to carry out pairwise
 competition and single invasion simulations.
-Most parameters are set to the defaults of the original Community Simulator
-package. Table \mr shows those that are given non-default values to ensure
-enough variation in the primary communities.
+
+Two sets of simulations were carried out, corresponding to a
+\textit{simple} or \textit{complex} externally replenished resource respectively
+(see details in main text).
+For the \textit{simple} resource scenario, we generated a library of 15 resources
+belonging to a same class. For the \textit{complex} resource scenario,
+the library contained 30 resources subdivided into 3 classes
+(each being preferentially consumed by one of the three species families).
+Our aim with this approach was to examine the implications of allowing for
+different degrees of density in the metabolic networks sustaining our communities.
+Most other parameters were set to the defaults of the original Community Simulator
+package, with the only exception of the maintenance costs ($m$) which are set to
+zero for all species (as cell death is negligible through the duration of our growth cycles)
+and the sparsity of the metabolic matrices ($s$) which is set to 0.9
+to generate significant variability in the secretion fluxes across different species
+(see main text).
 
@@
 and the updated Community Simulator package with instructions for the
-new features can be found in \url{github.com/jdiazc9/coalescence}.
+new features are deposited in \url{github.com/jdiazc9/coalescence}.
 
@@
 \internallinenumbers
-\includegraphics[width=15cm,keepaspectratio]{figs/fig2.pdf}
+\includegraphics[width=11cm,keepaspectratio]{figs/fig2.pdf}
 \caption{\textbf{Co-selection in microbial community coalescence.}
 \textbf{a.}~Coalescence outcomes are quantified by the relative Bray-Curtis similarity
-($Q$) between the coalesced and invasive communities (\methodsref{metrics}).
+($Q$) between the coalesced and invasive communities.
 These outcomes are predicted by the pairwise competition between the invasive and
-resident dominant
-species ($R^2=0.15$ for glutamine and $R^2=0.57$ for citrate). This is consistent with a scenario of
+resident dominant species
+($R^2=0.15$, $p<0.05$ for glutamine and $R^2=0.57$, $p<10^{-4}$ for citrate).
+A higher correlation is consistent with a scenario of
 top-down positive co-selection where dominants recruit their cohorts for the final
@@
 \textbf{c-d.} Simulations of community coalescence with a consumer-resource model
-are able to capture these trends.}
+and a metabolically simple supplied resource yield a strong degree of top-down positive
+co-selection
+($R^2=0.81$, $p<10^{-15}$)
+and few instances of bottom-up co-selection.
+\textbf{e-f.} Simulations with a metabolically complex resource result in weaker
+top-down co-selection
+($R^2=0.22$, $p<10^{-5}$)
+but frequent bottom-up positive co-selection.}
 \label{fig2}
@@
 
-\section*{Test section}
-
-This is an example cite \cite{Luo2020,Vass2021}.
-This is how you refer to \figref{fig1}.
-This is how you refer to \figref[a]{fig1}.
-This is how you refer to \tableref{fig1}.
-This is how you refer to \figref{figS1}.
-This is how you refer to the section \nameref{methods:sim} of the \nameref{methods}.
-This is a cleaner way to refer to the \methodsref{sim}.
-
-Check this reference \cite{Castledine2020}
-
-Resource availability modulates biodiversity–invasion relationships by altering competitive interactions \cite{Yang2017}
-
-Resource pulses can alleviate the biodiversity–invasion relationship in soil microbial communities \cite{Mallon2015}
-
 \end{document}

### 2021-05-11 coalescence.py

@@ -233,6 +233,7 @@ def mysim(p,q,pq):
 assumptions = community_simulator.usertools.a_default.copy()
 assumptions['n_wells'] = 100 # number of communities
 assumptions['S'] = 50 # number of species sampled at initialization
+assumptions['SA'] = [800, 800, 800] # [100, 100, 100] # number of species per specialist family
 assumptions['Sgen'] = 240 # 30 # number of generalists
 assumptions['l'] = 0.8 # leakage fraction
 
@@ -256,14 +257,9 @@ assumptions['rs'] = 0.0 # control parameter for resource secretion: 0 means rand
 
 # assumptions that depend on the nature of the supplied primary resource
 # 'complex' resource
-#assumptions['SA'] = [800, 800, 800] # [100, 100, 100] # number of species per specialist family
 #assumptions['MA'] = [10, 10, 10] # [30, 30, 30] # number of resources per resource class
 # 'simple' resource
-#assumptions['SA'] = 2400 # [100, 100, 100] # number of species per specialist family
-#assumptions['MA'] = 30 # [30, 30, 30] # number of resources per resource class
-# 'intermediate' resource
-assumptions['SA'] = [1200, 1200] # [100, 100, 100] # number of species per specialist family
-assumptions['MA'] = [15, 15] # [30, 30, 30] # number of resources per resource class
+assumptions['MA'] = 15 # [30, 30, 30] # number of resources per resource class
 
 print(assumptions)
 
@@ -542,6 +538,7 @@ kratio_resident = [math.log10(x) for x in kratio_resident]
 kratio_invasive = [math.log10(x) for x in kratio_invasive]
 kratio_diff = [kratio_invasive[i] - kratio_resident[i] for i in range(assumptions['n_wells'])]
 
+'''
 # get cohort invasiveness (ci)
 ci = ['NA' for i in range(assumptions['n_wells'])]
 N_cohortinv = N_cohortinv.droplevel(0)
@@ -584,6 +581,7 @@ J_in = lambda R,params: (u[assumptions['regulation']](params['c']*R,params)
 J_out = {'common': lambda R,params: (params['l']*J_in(R,params)).dot(params['D'].T),
[…]

### 2021-05-11 data/analyzeData.R

@@ -1330,7 +1330,7 @@ if (save_plots) {
 ### ----------------------------------------------------------------------
 
 # load data from simulations
-plot_this <- read.table('simul_data.txt',header=TRUE)
+plot_this <- read.table('simul_data_complex.txt',header=TRUE)
 
 # make plots
 myplots[['q-vs-pairwise_bray-curtis_simul']] <-

### 2021-05-11 data/analyzeData.R
minor changes

@@ -1330,7 +1330,7 @@ if (save_plots) {
 ### ----------------------------------------------------------------------
 
 # load data from simulations
-plot_this <- read.table('simul_data_complex.txt',header=TRUE)
+plot_this <- read.table('simul_data_simple.txt',header=TRUE)
 
 # make plots
 myplots[['q-vs-pairwise_bray-curtis_simul']] <-

### 2021-05-12 draft/draft.tex
minor changes
updated draft

@@
 % title, authors, affiliations
-\title{Cohesiveness in microbial community coalescence}
-\author[1,2,3]{Authors}
+\title{Top-down versus bottom-up cohesiveness
+in microbial community coalescence}
+\author[1,2,*]{Juan Diaz-Colunga}
+\author[1,2,*]{Nanxi Lu}
+\author[1,2,*]{Alicia Sanchez-Gorostiaga}
+\author[3]{Joshua E. Goldford}
+\author[4,5]{Mikhail Tikhonov}
 \author[1,2,$\dagger$]{Álvaro Sánchez}
@@
 Yale University, New Haven, CT, USA}
-\affil[3]{Other affiliations...}
-\affil[$\dagger$]{To whom correspondence should be addressed: \normalfont alvaro.sanchez@yale.edu}
+\affil[3]{Cell Metabolism, Agios Pharmaceuticals, Inc.,
+Cambridge, MA, USA}
+\affil[4]{Department of Applied Physics,
+Stanford University, Stanford, CA, USA}
+\affil[5]{Department of Physics,
+Washington University, St. Louis, MO, USA}
+
+\affil[$\dagger$]{\normalfont To whom correspondence should be addressed: alvaro.sanchez@yale.edu}
+\affil[*]{\normalfont These authors contributed equally}
 \date{}
@@
 
-%\linenumbers
+\linenumbers
 
@@
 
-We performed a new round of invasion experiments where each dominant invaded every
+To test which of these situations was more common in our conditions,
+we performed a new round of invasion experiments where each dominant invaded every
 other community in isolation (\methodsref{competitions}). The dominant-invaded
@@
 
-We ran a set of simulations of community coalescence to better understand the mechanisms
-of ecological co-selection in our experiments. We used a microbial consumer-resource
+To better understand the mechanisms of ecological co-selection in our experiments,
+we ran a set of simulations of community coalescence. We used a microbial consumer-resource
 model (microCRM) \cite{Goldford2018,Marsland2019} implemented in the Community Simulator package
@@
 to the environment when consuming a primary resource
-\cite{Harcombe2014,Pinu2018}.
+\cite{Harcombe2014,Pinu2018}. \mr cite Sylvie here?
 We reason that these choices are necessary to potentially generate community cohesiveness:
@@
 Our data suggests that the strength and direction of ecological co-selection is modulated
-by the supplied primary carbon source,
-which in turn shapes the cross-feeding network that dictates
-the structure of our communities \cite{Goldford2018,Estrela2020}.
+by the underlying metabolic network that shapes the structure of
+communities assembled in synthetic minimal conditions
+\cite{Goldford2018,Estrela2020}.
+The architecture of this network is dependent on the supplied primary carbon source.
 These observations,
@@
 \cite{Gilpin1994,Toquenaga1997,Tikhonov2016,Tikhonov2017,Vila2019,Lechon2021},
-suggest that collective interactions arising from metabolic feedbacks between microbes
+suggest that collective interactions between microbes
 and the environment should be generically expected to produce ecological co-selection
@@
 
-Further work is required to clarify the relationship between metabolic interactions,
-community structure, cohesiveness and co-selection.
+Further work will be necessary to clarify the relationship between
+metabolic feedbacks, community cohesiveness, and
+the strength and direction of ecological co-selection.
 The experimental system that we introduced in this work can be easily expanded so that
@@
 
-The authors wish to thank Joshua Goldford, Pankaj Mehta, Wenping Cui,
+The authors wish to thank Pankaj Mehta, Wenping Cui,
 Robert Marsland and all members of the Sanchez laboratory

### 2021-05-13 draft/draft.tex
updated affiliations (but check with co-authors)
new figure 2 (incorporated former figure S4), added figure 3, notes in the draft for later editing

@@
 \author[3]{Joshua E. Goldford}
-\author[4,5]{Mikhail Tikhonov}
+\author[4]{Mikhail Tikhonov}
 \author[1,2,$\dagger$]{Álvaro Sánchez}
@@
 Cambridge, MA, USA}
-\affil[4]{Department of Applied Physics,
-Stanford University, Stanford, CA, USA}
-\affil[5]{Department of Physics,
-Washington University, St. Louis, MO, USA}
+\affil[4]{Department of Physics,
+Center for Science \& Engineering of Living Systems,
+Washington University in St. Louis, St. Louis, MO, USA}
 
@@
 Livingston2013,Prior2015,Rillig2015,OLoughlin2017,Castledine2020}.
-\iffalse
-One example is the hypothesis known as \textit{invasional meltdown}, which proposes that
-positive interactions between co-invading species can enhance their invasive success and
-facilitate future invasions \cite{Simberloff1999,Simberloff2006,Gurevitch2006}.
-\fi
 However, and
@@
 which of the two dominants is most competitive in pairwise competition.
+Analogously, competition between dominants should be affected only weakly by the presence or
+absence of the cohorts, that would play a passive role under these conditions.
 To test this hypothesis, we performed all pairwise competitions between dominant species
@@
 (\methodsref{sequencing}).
+
+\mr Rearrange this section to follow the flow of the new figure 2.
+Also introduce the model here and cook a nice experiment-theory-experiment sandwich later.
 
@@
 cohorts are present too, i.e. after coalescence
-(\figref{figS4}, $R^2=0.04$, $p>0.05$).
+(\figref[a]{fig2} red dots, $R^2=0.04$, $p>0.05$).
 On the other hand, pairwise competition between dominants of the citrate communities
 is highly correlated with the relative abundance of said dominants in community coalescence
- (\figref{figS4}, $R^2=0.83$, $p<10^{-8}$).
+ (\figref[a]{fig2} blue dots, $R^2=0.83$, $p<10^{-8}$).
 
@@
 have been shown to be
-sustained by dense metabolic cross-feeding networks \cite{Goldford2018}, for which the
-microCRM provides a good description.
+sustained by dense metabolic cross-feeding networks
+\cite{Goldford2018,Estrela2020},
+for which the microCRM provides a good description.
 We implemented a new feature in the package that makes individual species
@@
 In short,
-we replicated our experimental protocol \textit{in silico}
+we reproduced our experimental protocol \textit{in silico}
 by generating a library of species that we divided into two pools. Each pool was used to seed a
-collection of invasive and resident communities, that were then allowed to stabilize through 20
-growth-dilution cycles. Stable communities were then used to reproduce our coalescence
+collection of invasive and resident communities, that were allowed to stabilize through 20
+growth-dilution cycles. We then mixed pairs of these stable communities to carry out our coalescence
 and dominant-dominant competition experiments,
@@
 
+\iffalse
 We carried out two sets of simulations under two different regimes.
@@
 of our simulations and, in turn, modulate the strength and direction of ecological co-selection.
+\fi
+
+\mr Reframe the entire simulations section (include it in previous two sections?).
+Sandwich: the model predicts a) that positive bottom-up co-selection is much more common than
+negative bottom-up co-selection... and we see it in the experiments; and b) that in the instances
+where bottom-up co-selection is strong (green area in figure 3...), the relative frequency of the dominants
+in head-to-head pairwise competition is a very poor predictor of community coalescence outcomes...
+which we confirm experimentally.
+Conclusions: results and previous theoretical and experimental work on different systems (and on the
+general relationship between metabolic interactions and community structure) suggest that...
+the strength and direction of ecological co-selection is determined by the underlying metabolic networks that
+shape the structure and interactions of communities assembled in synthetic minimal environments
+during coalescence.
+These networks are modulated by the supplied primary resource in our conditions.
+When top-down co-selection is weak, bottom-up positive co-selection can be very strong, while negative
+co-selection remains rare in our experiments.
 
@@
 
+\mr Update with new simulation protocol.
+
 We used the Community Simulator package \cite{Marsland2020} and included new
@@
 \internallinenumbers
-\includegraphics[width=11cm,keepaspectratio]{figs/fig2.pdf}
-\caption{\textbf{Co-selection in microbial community coalescence.}
-\textbf{a.}~Coalescence outcomes are quantified by the relative Bray-Curtis similarity
+\includegraphics[width=15cm,keepaspectratio]{figs/fig2_v3.pdf}
+\caption{\textbf{Top-down co-selection in microbial community coalescence.}
+\textbf{a.} Pairwise competition of dominants with or without their cohorts.
+In the horizontal axis, we plot the frequency of the invasive dominant species
+in head-to-head pairwise competition with the resident dominant.
+In the vertical axis, we plot the same relative frequency when the two species compete
+in the presence of their cohorts, i.e. during community coalescence.
+$R^2=0.04$, $p>0.05$ for glutamine and $R^2=0.83$, $p<10^{-8}$ for citrate.
+\textbf{b.} Coalescence outcomes are quantified by the relative Bray-Curtis similarity
 ($Q$) between the coalesced and invasive communities.
 These outcomes are predicted by the pairwise competition between the invasive and
-resident dominant species
-($R^2=0.15$, $p<0.05$ for glutamine and $R^2=0.57$, $p<10^{-4}$ for citrate).
-A higher correlation is consistent with a scenario of
-top-down positive co-selection where dominants recruit their cohorts for the final
+resident dominant species.
+Left panel (red): glutamine communities, $R^2=0.15$, $p<0.05$.
+Middle panel (blue): citrate communities, $R^2=0.57$, $p<10^{-4}$.
+A high correlation is consistent with a scenario of
+strong top-down positive co-selection where dominants recruit their cohorts for the final
 coalesced community. Two biological replicates per experiment are plotted individually.
-\textbf{b.}~We represent the frequency reached by
-the invasive dominant species when they invade the resident communities on their own versus
-when they are in the company of their cohort.
-Three scenarios are possible: green and red
-shaded areas represent limit cases of positive (recruitment) or negative (antagonism)
-bottom-up co-selection, gray area corresponds to situations where invasive dominant
-species can invade with equal success regardless of the presence of their cohorts.
-Data shows that positive co-selection is common, whereas antagonistic
-co-selection is rare in our experiments.
-Two biological replicates per experiment are plotted individually.
-\textbf{c-d.} Simulations of community coalescence with a consumer-resource model
-and a metabolically simple supplied resource yield a strong degree of top-down positive
-co-selection
-($R^2=0.81$, $p<10^{-15}$)
-and few instances of bottom-up co-selection.
-\textbf{e-f.} Simulations with a metabolically complex resource result in weaker
-top-down co-selection
-($R^2=0.22$, $p<10^{-5}$)
-but frequent bottom-up positive co-selection.}
+Right panel (black): simulations with a microbial consumer-resource model are able
+to capture these trends ($R^2=0.22$, $p<10^{-5}$).}
 \label{fig2}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[!h]
+\centering
+\internallinenumbers
+\includegraphics[width=12cm,keepaspectratio]{figs/fig3.pdf}
+\caption{\textbf{Trade offs between bottom-up and top-down ecological co-selection.}
+\textbf{a.} Blablabla.
+\textbf{b.} Blablabla.
+\textbf{c.} Blablabla.
+\textbf{d.} Blablabla.
+\textbf{e.} Blablabla.
+\textbf{f.} Blablabla.}
+\label{fig3}
 \end{figure}
@@
 
-\begin{figure}[!h]
-\centering
-\internallinenumbers
-\includegraphics[width=8cm,keepaspectratio]{figs/figS4.pdf}
-\caption{\textbf{Pairwise competition of dominants with or without their cohorts.}
-In the horizontal axis, we plot the relative frequency of the invasive dominant species in head-to-head
-pairwise competition. In the vertical axis, we plot the same relative frequency when competition takes
-place in the presence of the cohorts, i.e. during community coalescence.
-$R^2=0.04$, $p>0.05$ for glutamine and $R^2=0.83$, $p<10^{-8}$ for citrate.}
-\label{figS4}
-\end{figure}
-
-\clearpage
-
 \end{document}

### 2021-05-13 data/analyzeData.R

@@ -1055,11 +1055,16 @@ for (cs in carbon_sources) {
   }
 }
 
+# check that sample positioning in wells is consistent
+stopifnot(all(plot_this$well_pairwise == plot_this$well_coalescence))
+
 # pseudo-pairwise competition of dominants with cohorts
 plot_this$f_multiinv_rel <- plot_this$f_multiinv/(plot_this$f_multiinv + plot_this$f_residentdom_multiinv)
 
-# check that sample positioning in wells is consistent
-stopifnot(all(plot_this$well_pairwise == plot_this$well_coalescence))
+# strength of bottom-up co-selection
+plot_this$bucs <- plot_this$f_singleinv < 0.15 & plot_this$f_multiinv > 0.05
+plot_this$bucs <- c('neutral','strong')[1 + plot_this$bucs]
+plot_this$bucs <- factor(plot_this$bucs,levels=c('neutral','strong'))
 
 # reshape plotting table (this makes it easier to create multipanel plots)
 plot_this <- gather(plot_this,metric,value,q_bray_curtis:q_endemic_cohort)
@@ -1073,9 +1078,9 @@ plot_this$metric <- factor(plot_this$metric,levels=unique(plot_this$metric))
 # parameters for annotations of plots
 poly <- list(w=0.10,
              aperture=0.15,
-             color=c(rgb(220,245,220,maxColorValue=255),
-                     rgb(230,230,230,maxColorValue=255),
-                     rgb(250,220,220,maxColorValue=255)))
+             color=c('#d6e4d4',
+                     '#e6e6e5',
+                     '#fdd7c8'))
 
 # make plots
 myplots[['q-vs-pairwise_bray-curtis']] <-
@@ -1110,6 +1115,72 @@ myplots[['q-vs-pairwise_bray-curtis']] <-
         panel.border=element_rect(size=0.25)) +
   coord_fixed() # +
   # ggtitle('Bray-Curtis similarity')
+
+myplots[['q-vs-pairwise_bray-curtis_glutamine']] <-
+  ggplot(data=plot_this[plot_this$metric=='q_bray_curtis'
+                        & plot_this$carbon_source=='Glutamine',],
+         aes(x=f_pairwise,y=value,
+             color=carbon_source)) +
+  geom_point(size=3,
+             shape=1,
+             stroke=0.5) +
+  geom_smooth(formula = y ~ x,
+              method = 'lm',
+              se = FALSE,
+              size = 0.5) +
[…]

### 2021-05-13 draft/draft_v2.tex
box 1 added

--- 2021-05-13 draft/draft.tex
@@
 \usepackage{textcomp} % special characters (some references break if this isn't loaded?)
-
-\usepackage{pdflscape} %% Landscape
-\usepackage{dcolumn}
-\usepackage{booktabs}
-\usepackage{savesym}
-\savesymbol{iint}
-\restoresymbol{TXF}{iint}
-%% \usepackage{pdfpages} %% to include a pdf file
+\usepackage{mdframed} % to easily handle boxes
 
@@
 in microbial community coalescence}
-\author[1,2,*]{Juan Diaz-Colunga}
-\author[1,2,*]{Nanxi Lu}
+\author[1,*]{Juan Diaz-Colunga}
+\author[1,*]{Nanxi Lu}
 \author[1,2,*]{Alicia Sanchez-Gorostiaga}
@@
 \author[4]{Mikhail Tikhonov}
-\author[1,2,$\dagger$]{Álvaro Sánchez}
-\affil[1]{Department of Ecology \& Evolutionary Biology,
+\author[1,$\dagger$]{Álvaro Sánchez}
+\affil[1]{Department of Ecology \& Evolutionary Biology
+and
+Microbial Sciences Institute,
 Yale University, New Haven, CT, USA}
-\affil[2]{Microbial Sciences Institute,
-Yale University, New Haven, CT, USA}
-\affil[3]{Cell Metabolism, Agios Pharmaceuticals, Inc.,
-Cambridge, MA, USA}
+\affil[2]{Department of Microbial Biotechnology,
+Centro Nacional de Biotecnología (CNB-CSIC), Cantoblanco, Madrid, Spain}
+\affil[3]{Physics of Living Systems, Department of Physics,
+Massachusetts Institute of Technology, Cambridge, MA, USA}
 \affil[4]{Department of Physics,
@@
 \fi
-We then turn to a microbial consumer-resource model (microCRM)
+We then turn to a Microbial Consumer-Resource Model (MicroCRM)
 \cite{Goldford2018,Marsland2019,Marsland2020}
@@
 (\methodsref{competitions}).
-We then performed all possible pairwise community coalescence experiments by mixing
+We then carried out all possible pairwise community coalescence experiments by mixing
 equal volumes of the communities and propagating the resulting cultures for seven extra
 transfers
-(\figref[f]{fig1}),
+(\figref[f]{fig1}).
+\iffalse
 and we quantified coalescence outcomes using the similarity between the coalesced and the
@@
 (\methodsref{metrics}).
+\fi
 The frequencies of all species in both community-community and dominant-dominant competitions
@@
 
-\mr Rearrange this section to follow the flow of the new figure 2.
-Also introduce the model here and cook a nice experiment-theory-experiment sandwich later.
+We found that, for communities assembled in the glutamine environment,
+the relative frequency of a dominant against another in head-to-head pairwise competition
+is barely predictive of its relative frequency against that same other dominant when the
+cohorts are present too, i.e. during community coalescence
+(\figref[a]{fig2} red dots, $R^2=0.04$, $p>0.05$).
+This correlation is significantly higher for the citrate communities
+ (\figref[a]{fig2} blue dots, $R^2=0.83$, $p<10^{-8}$).
+This suggests that, in the glutamine environments, head-to-head competition of dominants
+is heavily influenced by higher order effects introduced by the rare taxa of the communities.
+On the other hand, the cohorts seem to play a more passive role in the citrate environments.
+To test the effects of top-down co-selection at the community level,
+we quantified the distances between the invasive and coalesced communities
+using the relative Bray-Curtis similarity (\methodsref{metrics})
+and compared them to the outcomes of the pairwise competitions between dominants alone.
+We again noticed differences between glutamine and citrate communities:
+for the former,
+the pairwise competitive ability of an invasive dominant
+is only weakly predictive of the performance of the invasive community in coalescence
+(\figref[b]{fig2} left panel, $R^2=0.15$, $p<0.05$).
+For the latter,
+the structure of the coalesced communities tends to be more strongly dictated by the
+result of the dominant-dominant competition
+(\figref[b]{fig2} middle panel, $R^2=0.57$, $p<10^{-4}$).
+Alternative quantifications of community distance yield similar results,
+with weaker effects when the metric used
+accounts only for the presence/absence of specific species and not for their relative abundance in
+the communities
+(\figref{figS2}).
+All these metrics include the presence of the dominant species themselves. To better disentangle the effect
+that these dominants have on the other members of their communities, we repeated the analysis this time
+excluding the dominant species from the compositional data, finding that our results still hold
+(\figref{figS3}).
 
+Together, these results suggest that the strength of top-down co-selection depends on
+the environment where communities are assembled and coalescence takes place.
+Communities stabilized with citrate as the primary supplied resource
+display a strong degree of top-down cohesiveness,
+with the fates of the sub-dominant species determined to a large extent by dominant-dominant
+pairwise competition. This competition is, in turn, only weakly affected by the presence of the cohorts.
+For glutamine communities, although some level of top-down co-selection is consistent with our data,
+the cohorts do not appear to just be passively responding to their dominants.
+Instead, evidence suggests that the cohorts might be playing an active role in community coalescence.
+To investigate the determinants of top-down co-selection and the factors modulating its strength,
+we ran a set of simulations of community coalescence. We used a Microbial Consumer-Resource
+Model (MicroCRM) \cite{Goldford2018,Marsland2019} implemented in the Community Simulator package
+for Python \cite{Marsland2020}. We chose this modeling framework because
+communities assembled under our experimental conditions (natural microbiomes re-assembled into
+multispecies communities through serial growth-dilution cycles in synthetic minimal media
+with a single carbon source)
+have been shown to be sustained by dense metabolic cross-feeding networks
+\cite{Goldford2018,Estrela2020},
+for which the MicroCRM provides a good description.
+We carried out our simulations enabling a large variation across species (consumers) in terms of
+their metabolic characterization.
+This choice is discussed in \hyperref[box1]{Box 1}.
+
+\clearpage
+
+\begin{mdframed}
+\internallinenumbers
+\subsubsection*{Box 1: Community cohesiveness in a Microbial Consumer-Resource Model}
+\label{box1}
+
+The Microbial Consumer-Resource Model (MicroCRM) \cite{Goldford2018,Marsland2019,Marsland2020}
+is a modeling framework based on the classic MacArthur's consumer resource model
+\cite{MacArthur1970}.
+It encodes the dynamics of a system with $S$ species and $M$ resources
+in terms of a consumer preference matrix $\mathbf{c}$ and a metabolic matrix $\mathbf{D}$,
+with an additional set of parameters controlling
+the species maintenance costs ($m_i$ for species $i$),
+the resource energy densities ($w_\alpha$ for resource $\alpha$),
+the energy to growth rate conversion factor ($g_i$ for species $i$)
+or the leakage fraction, i.e. the amount of energy lost as byproducts when a resource is consumed
+($l_\alpha$ for resource $\alpha$).
+The element $c_{i\alpha}$ of the consumer preference matrix
+represents the uptake rate of resource $\alpha$ by
+species $i$ (although the relationship between $c_{i\alpha}$ and the uptake rate can be more
+complex in modeling scenarios that are not considered here,
+see \cite{Goldford2018,Marsland2019,Marsland2020}).
+The element $D_{\beta\alpha}$ of the metabolic matrix
+represents the amount of energy secreted in the form of resource $\beta$
+as a result of the metabolism of resource $\alpha$.
+
+This formulation assumes that all species metabolize resources through similar pathways,
+with any two species secreting the same byproducts when consuming a same resource.
+However, experimental evidence suggests that individual species can secrete different sets of metabolites
+to the environment when growing on a same primary resource
+\cite{Harcombe2014,Pinu2018,Estrela2020}. This observation motivated us to introduce a new
+feature in the model:
+in short, we now define $\mathbf{D}$ as a three-dimensional matrix
+where the element $D_{i\beta\alpha}$ represents the energy flux in the form of resource $\beta$
+that is secreted \textit{by species $i$} when it metabolizes resource $\alpha$.
+Note that now $D_{i\beta\alpha}$ need not be equal to $D_{j\beta\alpha}$ if $i \neq j$
+(see illustration below).
+
+\bigskip
+\begin{center}
+\includegraphics[width=8cm,keepaspectratio]{figs/figBox.png}
+\end{center}
+\bigskip
+
+We reason that the choice of species-specific metabolic fluxes is necessary to potentially generate
+cohesiveness at the community level during coalescence.
+If the secretions of all species were identical (or only slightly different),
+higher order cross-feeding effects would be very unspecific:
+the establishment of new invasive species
+--given that they could outcompete resident taxa within their metabolic niches,
+i.e. more effectively feed off the same resource or set of resources as them--
+would not alter (or only do so moderately)
+the metabolic flows through the rest of the cross-feeding network.
+On the other hand, said network could undergo a profound and further-reaching restructuring
+if the invasive species secreted very different sets of metabolites with respect to the resident ones,
+potentially disabling existing niches and/or enabling new ones
+where more invaders could be co-selected.
+For a similar reason, we argue that the sparsity of the metabolic matrix could also modulate
+the emergence of cohesiveness in the face of coalescence.
+A dense metabolic matrix corresponds to a situation where all species secrete a wide
+variety of byproducts.
+New-coming invasive species that secrete similar byproducts as resident ones
+--even if they do so in different amounts--
+might only induce moderate quantitative changes in the metabolic fluxes.
+But if the sets of secretions are qualitatively different, co-selection of species adapted to
+each of those sets becomes possible.
+These ideas are supported by experimental observations
+suggesting that species with a history of coexistence make up
+cohesive communities with highly specific cross-feeding configurations
+\cite{Rosenzweig1994,Goldford2018,Estrela2020}.
+
+\bigskip
+\end{mdframed}
+
+\clearpage
+
+\iffalse
 We found that, for communities assembled in the glutamine environment,
@@
 Instead, evidence suggests that the cohorts might be playing an active role in community coalescence.
+
 This finding led us to investigate the potential role of bottom-up ecological co-selection
@@
 for or against by their cohorts.
+\fi
 
@@
 
-\subsubsection*{Simulations of community coalescence with a consumer-resource model}
+\subsubsection*{Simulations of community coalescence with a Consumer-Resource Model}
 
 To better understand the mechanisms of ecological co-selection in our experiments,
-we ran a set of simulations of community coalescence. We used a microbial consumer-resource
-model (microCRM) \cite{Goldford2018,Marsland2019} implemented in the Community Simulator package
+we ran a set of simulations of community coalescence. We used a Microbial Consumer-Resource
+Model (MicroCRM) \cite{Goldford2018,Marsland2019} implemented in the Community Simulator package
 for Python \cite{Marsland2020}. We chose this modeling framework because
@@
 \cite{Goldford2018,Estrela2020},
-for which the microCRM provides a good description.
+for which the MicroCRM provides a good description.
 We implemented a new feature in the package that makes individual species
@@
 coalesced community. Two biological replicates per experiment are plotted individually.
-Right panel (black): simulations with a microbial consumer-resource model are able
+Right panel (black): simulations with a Microbial Consumer-Resource Model are able
 to capture these trends ($R^2=0.22$, $p<10^{-5}$).}
@@
 \caption{\textbf{Trade offs between bottom-up and top-down ecological co-selection.}
-\textbf{a.} Blablabla.
-\textbf{b.} Blablabla.
-\textbf{c.} Blablabla.
-\textbf{d.} Blablabla.
-\textbf{e.} Blablabla.
-\textbf{f.} Blablabla.}
+\textbf{a.} We hypothesize that three scenarios are possible regarding bottom-up co-selection:
+sub-dominant species could co-select for (green) or against (red) their dominant in coalescence,
+or they could have no effect in the invasion success of the dominant taxa (gray).
+\textbf{b.} Simulations with a Microbial Consumer-Resource Model suggest that positive bottom-up
+co-selection is common and can be very strong, whereas negative bottom-up co-selection is rare.
+We plot the frequency reached by the invasive dominants when invading the resident communities
+in isolation versus the same frequency when invading together with their cohorts,
+i.e. in community coalescence.
+Points in the green/red area represent instances where the invasive dominant is able to invade
+with higher/lower success when accompanied by its cohort, evidencing positive/negative
+bottom-up co-selection.
+Points around the diagonal (gray area) correspond to cases where the success of the invasive dominant
+is only weakly affected by the presence or absence of its cohort.
+\textbf{c.} We divided the data from our simulations into two sets according to whether
+positive or no bottom-up co-selection was observed (that is, whether points fell into the green or gray
+areas of panel b). Here we reproduce the plots in \figref[b]{fig2} for each set, representing the
+result of the dominant head-to-head pairwise competition versus the outcome of community coalescence.
+Left panel: when positive bottom-up co-selection is strong, head-to-head pairwise competition of dominants
+is poorly predictive of coalescence outcomes
+($R^2=0.00$, $p>0.05$).
+Right panel: on the other hand, when bottom-up co-selection is weak coalescence outcomes are more
+strongly dictated by the result of the dominant-dominant competition
+($R^2=0.34$, $p<10^{-6}$).
+\textbf{d.} Experiments show that in our conditions, positive bottom-up co-selection is indeed more
+frequent and strong than negative bottom-up co-selection.
+\textbf{e.} We reproduce the plots in panel c for our experimental data. This is equivalent to recreating
+\figref[b]{fig2}, this time splitting our data by the strength of bottom-up co-selection
+instead of by the carbon source provided to the communities.
+Like simulations predicted, instances of community coalescence displaying strong positive bottom-up co-selection
+yield a poor correspondence between dominant pairwise competition and coalescence outcomes
+(left panel, $R^2=0.07$, $p>0.05$)
+and vice-versa
+(right panel, $R^2=0.37$, $p<10^{-4}$).}
 \label{fig3}

### 2021-05-14 draft/draft_v2.tex
more results re-written, methods now in their final form (eliminated redundancies with box 1)

@@
 For glutamine communities, although some level of top-down co-selection is consistent with our data,
-the cohorts do not appear to just be passively responding to their dominants.
-Instead, evidence suggests that the cohorts might be playing an active role in community coalescence.
+the cohorts do not appear to just be passively responding to their dominants
+but rather playing an active role in community coalescence.
+
 To investigate the determinants of top-down co-selection and the factors modulating its strength,
@@
 have been shown to be sustained by dense metabolic cross-feeding networks
-\cite{Goldford2018,Estrela2020},
+\cite{Goldford2018,Estrela2020}
 for which the MicroCRM provides a good description.
-We carried out our simulations enabling a large variation across species (consumers) in terms of
-their metabolic characterization.
-This choice is discussed in \hyperref[box1]{Box 1}.
+We carried out our simulations enabling a large variation across species in terms of
+their metabolic architectures.
+This choice is discussed in \hyperref[box1]{Box~1}.
+To reproduce our experimental protocol \textit{in silico},
+we first generated a library of resources and two non-overlapping pools of species.
+Each pool was used to seed a collection of 100 invasive and resident communities respectively,
+that were allowed to stabilize through 20 growth-dilution cycles.
+We then mixed these stable communities in pairs to carry out our coalescence
+and dominant-dominant competition experiments.
+A full description of our modeling framework can be found in the \methodsref{sim} section.
+We found that the MicroCRM was able to capture a correspondence between
+the head-to-head pairwise competition of dominants
+and the outcome of community coalescence
+(\figref[b]{fig2}, right panel).
+
+[Maybe this ``outro'' sentence is not even necessary]
+Our experiments had shown that this correlation varied
+across communities assembled in the glutamine and citrate environments
+(\figref[b]{fig2}, left and middle panels),
+which led us to investigate the determinants of the strength and direction
+of ecological co-selection in our model.
 
@@
 
-We reason that the choice of species-specific metabolic fluxes is necessary to potentially generate
+We reason that the choice of species-specific metabolic architectures is necessary to potentially generate
 cohesiveness at the community level during coalescence.
@@
 would not alter (or only do so moderately)
-the metabolic flows through the rest of the cross-feeding network.
+the metabolic flows through the rest of the community's cross-feeding network.
 On the other hand, said network could undergo a profound and further-reaching restructuring
@@
 New-coming invasive species that secrete similar byproducts as resident ones
---even if they do so in different amounts--
+--even if they do so in different relative amounts--
 might only induce moderate quantitative changes in the metabolic fluxes.
@@
 Instead, evidence suggests that the cohorts might be playing an active role in community coalescence.
-
 This finding led us to investigate the potential role of bottom-up ecological co-selection
@@
 
-To study the effect of bottom-up co-selection, we compared the invasion success of the dominant
-from one of the communities when invading the other community in isolation versus when doing so
-accompanied by its cohort.
+Experimental evidence suggests that
+the cohorts might have a strong effect in dictating coalescence outcomes.
+At least in some instances,
+head-to-head pairwise competition of dominants can be heavily influenced by the presence of
+sub-dominant species
+(\figref[a]{fig2}).
+These findings led us to investigate the potential role of bottom-up ecological co-selection
+(\figref[e]{fig1}, right panels), i.e. whether the dominants may be co-selected
+for or against by their cohorts.
+We asked whether our model could provide insight regarding the role of sub-dominant taxa in coalescence.
+To test this, we ran a new set of simulations
+this time invading resident communities with the dominants alone,
+as opposed to whole communities invading together (\methodsref{sim}).
+We compared the invasion success of the dominants in this new scenario with respect to our previous
+simulations where they invaded accompanied by their cohorts.
+The invasion success of the dominant was quantified by its relative abundance
+in the final coalesced community.
 In situations where positive bottom-up ecological co-selection is strong, we expect to see dominants
-invading more effectively with their cohorts than by themselves
-(\figref[b]{fig2}, green shaded region).
-On the other hand, a high degree of negative bottom-up co-selection would result in dominants reaching
-higher invasion success alone than in the presence of their cohorts
-(\figref[b]{fig2}, red shaded region).
-Alternatively, if both forms of bottom-up co-selection were weak, we would see a similar invasion
-success of the dominant regardless of whether it is invading with its cohort
-(\figref[b]{fig2}, gray shaded region).
-
+reaching higher invasion success with their cohorts than by themselves
+(\figref[b]{fig3}, green shaded region).
+On the other hand, a high degree of antagonistic bottom-up co-selection would result in dominants
+invading more effectively alone than in the presence of their cohorts
+(\figref[b]{fig3}, red shaded region).
+Alternatively, if both forms of bottom-up co-selection are weak, we would see a similar invasion success
+regardless of the presence or absence of the cohort
+(\figref[b]{fig3}, gray shaded region).
+
+\figref[b]{fig3} shows that, in the simulations,
+many dominants could not invade on their own
+(or could only do so at very low final relative abundances, below 0.1)
+but were able to reach high frequencies when they were accompanied by their cohorts.
+This is a sign of positive bottom-up co-selection being common and potentially very strong,
+with negative bottom-up co-selection being rarely observed.
+We then asked...
+
+\iffalse
 To test which of these situations was more common in our conditions,
@@
 observed in our experiments.
-
+\fi
+
+\iffalse
 \subsubsection*{Simulations of community coalescence with a Consumer-Resource Model}
@@
 
-\iffalse
 We carried out two sets of simulations under two different regimes.
@@
 
-\mr Reframe the entire simulations section (include it in previous two sections?).
-Sandwich: the model predicts a) that positive bottom-up co-selection is much more common than
-negative bottom-up co-selection... and we see it in the experiments; and b) that in the instances
-where bottom-up co-selection is strong (green area in figure 3...), the relative frequency of the dominants
-in head-to-head pairwise competition is a very poor predictor of community coalescence outcomes...
-which we confirm experimentally.
-Conclusions: results and previous theoretical and experimental work on different systems (and on the
+\subsubsection*{Conclusion}
+
+Ideas for the conclusions: results and previous theoretical and experimental work on different systems (and on the
 general relationship between metabolic interactions and community structure) suggest that...
@@
 co-selection remains rare in our experiments.
-
-\subsubsection*{Conclusion}
+ALSO: we have only discussed metabolic interactions here. Other forms of facilitation are possible
+(and probably more relevant in specific natural settings)... but cohesiveness FOR SURE arises from metabolic interactions
+in our experiments, and we know the effect is ultimately metabolic because the MicroCRM captures it
+(actually both forms of it, top-down and bottom-up).
 
@@
 For each community, the most abundant colony morphotype at the end of the ninth transfer
-was selected, resuspended in $100\upmu$L 1$\times$PBS and serially diluted (1:10).
+was selected (\figref[c]{fig1}),
+resuspended in $100\upmu$L 1$\times$PBS and serially diluted (1:10).
 Next, $20\upmu$L of the cells diluted to $10^{-6}$ were plated in the corresponding synthetic
-minimal media and allowed to regrow at $30^\circ$C for 48h. Dominants were then identified
-(\figref[c]{fig1}),
+minimal media and allowed to regrow at $30^\circ$C for 48h. Dominants were then identified,
 inoculated into $500\upmu$L of fresh media and incubated still at $30^\circ$C for 48h.
@@
 %
+Using base-two logarithms ensures that the metric is bounded between 0 and 1.
 The Jaccard similarity is given by
@@
 %
-Additionally, we quantify coalescence outcomes by examining the fraction of the endemic
+Additionally, we quantified coalescence outcomes by examining the fraction of the endemic
 cohort of the original communities that persists in the coalesced one. We call
@@
 
-For all the metrics above, we quantify the relative similarity between the invasive
-and the coalesced communities using relative metrics ($Q$):
+For all the metrics above, we quantified the relative similarity between the invasive
+and the coalesced communities using relative metrics (denoted as $Q$):
 
@@
 
-\mr Update with new simulation protocol.
-
 We used the Community Simulator package \cite{Marsland2020} and included new
@@
 in which species can secrete different metabolites (and/or
-in different abundances) when consuming a same resource. Experimental observations
-support the idea of distinct species producing different sets of byproducts when
-feeding off the same primary resource
-\cite{Harcombe2014,Pinu2018}.
+in different abundances) when consuming a same resource.
 We call $D_{i\alpha\beta}$ to the
 fraction of energy in the form of resource $\alpha$ secreted \textit{by species
-i} when consuming resource $\beta$ ---note that now $D_{i\alpha\beta}$ need not be
-equal to $D_{j\alpha\beta}$ if $i \neq j$, unlike in the original Community
-Simulator. In the package's underlying Microbial Consumer Resource Model
-\cite{Goldford2018,Marsland2019}, this just means that the energy flux
-$J^{\mathrm{out}}_{i\beta}$ now takes the form
+i} when consuming resource $\beta$.
+In the Community Simulator underlying Microbial Consumer-Resource Model,
+this means that the energy flux 
+$J^{\mathrm{out}}_{i\beta}$ \cite{Goldford2018,Marsland2019}
+now takes the form
 
@@
 The documentation for the Community Simulator contains detailed
-descriptions of the model, parameters and package use. For the updated package with
+descriptions of the model formulation,
+parameters and package use.
+For the updated package with
 the new functionality, see \nameref{datacode}.
@@
 We split this library into two non-overlapping pools of 1320 species each.
-We randomly sample 50 species from each pool in equal ratios to seed
+We randomly sampled 50 species from each pool in equal ratios to seed
 100 resident and
@@
 resource after each dilution.
-We repeated the process 20 times to ensure generational equilibrium is
+We repeated the process 20 times to ensure generational equilibrium was
 achieved \cite{Goldford2018}.
@@
 generationally stable communities to seed 100 coalesced communities
-that we again stabilize as described previously.
+that were again stabilized as described previously.
 Similarly, we identified the dominant (most
@@
 
-Two sets of simulations were carried out, corresponding to a
-\textit{simple} or \textit{complex} externally replenished resource respectively
-(see details in main text).
-For the \textit{simple} resource scenario, we generated a library of 15 resources
-belonging to a same class. For the \textit{complex} resource scenario,
-the library contained 30 resources subdivided into 3 classes
-(each being preferentially consumed by one of the three species families).
-Our aim with this approach was to examine the implications of allowing for
-different degrees of density in the metabolic networks sustaining our communities.
 Most other parameters were set to the defaults of the original Community Simulator
 package, with the only exception of the maintenance costs ($m$) which are set to
-zero for all species (as cell death is negligible through the duration of our growth cycles)
+zero for all species
+(equivalent to assuming cell death is negligible through the duration of our growth cycles)
 and the sparsity of the metabolic matrices ($s$) which is set to 0.9
@@
 Experimental data and code for the analysis, as well as code for the simulations
-and the updated Community Simulator package with instructions for the
+and the updated Community Simulator package with instructions for enabling the
 new features are deposited in \url{github.com/jdiazc9/coalescence}.
@@
 in the presence of their cohorts, i.e. during community coalescence.
-$R^2=0.04$, $p>0.05$ for glutamine and $R^2=0.83$, $p<10^{-8}$ for citrate.
+$R^2=0.04$, $p>0.05$ for glutamine (red)
+and $R^2=0.83$, $p<10^{-8}$ for citrate (blue).
 \textbf{b.} Coalescence outcomes are quantified by the relative Bray-Curtis similarity

### 2021-05-17 draft/draft_v2.tex
version 2 of draft added

@@
 % title, authors, affiliations
-\title{Top-down versus bottom-up cohesiveness
-in microbial community coalescence}
+\title{Cohesiveness in microbial community coalescence}
 \author[1,*]{Juan Diaz-Colunga}
@@
 
-Together, these results suggest that the strength of top-down co-selection depends on
+Together, these observations suggest that the strength of top-down co-selection depends on
 the environment where communities are assembled and coalescence takes place.
@@
 we ran a set of simulations of community coalescence. We used a Microbial Consumer-Resource
-Model (MicroCRM) \cite{Goldford2018,Marsland2019} implemented in the Community Simulator package
+Model (MicroCRM) \cite{Goldford2018,Marsland2019} as implemented in the Community Simulator package
 for Python \cite{Marsland2020}. We chose this modeling framework because
@@
 We carried out our simulations enabling a large variation across species in terms of
-their metabolic architectures.
-This choice is discussed in \hyperref[box1]{Box~1}.
+their metabolic architectures;
+this choice is discussed in \hyperref[box1]{Box~1}.
 To reproduce our experimental protocol \textit{in silico},
@@
 that were allowed to stabilize through 20 growth-dilution cycles.
-We then mixed these stable communities in pairs to carry out our coalescence
-and dominant-dominant competition experiments.
-A full description of our modeling framework can be found in the \methodsref{sim} section.
-We found that the MicroCRM was able to capture a correspondence between
+We then mixed these stable communities in pairs to simulate our coalescence
+and dominant-dominant competition experiments
+(\methodsref{sim}).
+We found that the MicroCRM was able to capture a correlation between
 the head-to-head pairwise competition of dominants
 and the outcome of community coalescence
-(\figref[b]{fig2}, right panel).
-
+(\figref[b]{fig2}, right panel),
+further supporting the idea that top-down ecological co-selection can consistently
+emerge from metabolic interactions across species.
+
+\iffalse
 [Maybe this ``outro'' sentence is not even necessary]
@@
 
+\fi
 \clearpage
@@
 \begin{center}
-\includegraphics[width=8cm,keepaspectratio]{figs/figBox.png}
+\includegraphics[width=8cm,keepaspectratio]{figs/figBox_v2.png}
 \end{center}
@@
 
-Experimental evidence suggests that
-the cohorts might have a strong effect in dictating coalescence outcomes.
-At least in some instances,
-head-to-head pairwise competition of dominants can be heavily influenced by the presence of
-sub-dominant species
-(\figref[a]{fig2}).
-These findings led us to investigate the potential role of bottom-up ecological co-selection
-(\figref[e]{fig1}, right panels), i.e. whether the dominants may be co-selected
-for or against by their cohorts.
-We asked whether our model could provide insight regarding the role of sub-dominant taxa in coalescence.
-To test this, we ran a new set of simulations
-this time invading resident communities with the dominants alone,
-as opposed to whole communities invading together (\methodsref{sim}).
-We compared the invasion success of the dominants in this new scenario with respect to our previous
+Our data indicates that the primary resource supplied to the communities can modulate
+the effect of the cohorts in the dominants pairwise competition
+(\figref[a]{fig2})
+and the strength of top-down co-selection
+(\figref[b]{fig2}, left and middle panels).
+Our model suggests that this might be a result of the metabolic interactions
+between community members, including the rarer taxa
+(\hyperref[box1]{Box~1}).
+To investigate the potential role of the cohorts in coalescence,
+ i.e. whether the dominants may be co-selected for or against by them
+(\figref[e]{fig1}, right panels),
+we ran a new set of simulations
+this time invading resident communities with the dominants alone
+(\methodsref{sim}).
+We compared the invasion success of the dominants in isolation with respect to our previous
 simulations where they invaded accompanied by their cohorts.
-The invasion success of the dominant was quantified by its relative abundance
-in the final coalesced community.
-In situations where positive bottom-up ecological co-selection is strong, we expect to see dominants
+The invasion success of the dominants was quantified by their relative abundance
+in the final stabilized communities.
+Whenever positive bottom-up ecological co-selection is strong, we expect to see dominants
 reaching higher invasion success with their cohorts than by themselves
@@
 but were able to reach high frequencies when they were accompanied by their cohorts.
-This is a sign of positive bottom-up co-selection being common and potentially very strong,
-with negative bottom-up co-selection being rarely observed.
-We then asked...
+This indicates that positive bottom-up co-selection is frequent and potentially very strong,
+while negative bottom-up co-selection is far more uncommon.
+We then asked whether the ability of the pairwise competition of dominants
+to predict coalescence outcomes was dependent on the strength of bottom-up co-selection.
+We divided our simulations into two subsets:
+the first one was comprised of the instances where positive bottom-up co-selection was strong
+(i.e. dots in the green shaded region of \figref[b]{fig3}),
+the second set included all other cases
+(dots near the diagonal of \figref[b]{fig3}).
+We went back to our original simulations and plotted the frequency of the invasive dominant
+in pairwise competition with the resident dominant versus the relative similarity between
+the invasive and coalesced communities,
+i.e. the same plot as in \figref[b]{fig2}, for each subset.
+We found that when bottom-up positive co-selection is strong,
+the pairwise competition of dominants is not predictive of coalescence outcomes
+(\figref[c]{fig3}, left panel) and vice-versa (\figref[c]{fig3}, right panel).
+
+We then asked whether this trend was also observed \textit{in vitro}.
+We went back to our synthetic communities
+and carried out a new round of experiments where we invaded the resident communities
+with the invasive dominants alone
+(\methodsref{competitions}).
+After stabilization, we quantified species abundance through 16S Illumina sequencing
+(\methodsref{sequencing}).
+Again, we observed that bottom-up co-selection is far more common in its
+positive than in its negative form
+(\figref[d]{fig3}).
+Interestingly, bottom-up recruitment appears to be more frequent in the glutamine
+environments than in the citrate ones,
+consistent with our hypothesis that metabolic interactions among species
+are key in determining the strength and direction of ecological co-selection.
+We then repeated our analysis in \figref[c]{fig3},
+this time splitting our data according to the observed strength of bottom-up
+co-selection instead of the primary carbon source
+as we had done in \figref[b]{fig2}.
+Our findings were in line with the model prediction:
+pairwise competition between dominants is only predictive of coalescence outcomes
+if bottom-up co-selection is weak.
 
@@
 
-\subsubsection*{Conclusion}
-
+\subsubsection*{Conclusions}
+
+\iffalse
 Ideas for the conclusions: results and previous theoretical and experimental work on different systems (and on the
@@
 (actually both forms of it, top-down and bottom-up).
+\fi
 
@@
 of community cohesiveness.
+
 Our data suggests that the strength and direction of ecological co-selection is modulated
-by the underlying metabolic network that shapes the structure of
+by the underlying metabolic networks that shape the structure of
 communities assembled in synthetic minimal conditions
 \cite{Goldford2018,Estrela2020}.
-The architecture of this network is dependent on the supplied primary carbon source.
+This network is in turn regulated by the supplied primary carbon source.
+This idea is supported by the observation that a Microbial Consumer-Resource Model
+captures the trends observed experimentally when we enable a large variation
+in the metabolic fluxes across species.
+The model also predicts a trade-off between the strength of bottom-up co-selection
+and the ability of dominant-dominant pairwise competition to dictate coalescence outcomes,
+which we have confirmed experimentally.
 These observations,
-together with previous experimental results in different systems \cite{Sierocinski2017} as well as
+together with previous results in different systems \cite{Sierocinski2017} as well as
 theoretical predictions
@@
 
-Further work will be necessary to clarify the relationship between
-metabolic feedbacks, community cohesiveness, and
-the strength and direction of ecological co-selection.
+Additional work will be necessary to further clarify the relationship between
+metabolic feedbacks, community cohesiveness and ecological co-selection.
 The experimental system that we introduced in this work can be easily expanded so that
@@
 and test quantitative theories about its role in microbiome assembly.
+On the other hand, coalescence of communities under different settings
+(e.g. in spatially structured environments)
+might be governed by additional factors.
+Understanding them and quantifying their relative contributions in natural communities
+remains an open question in ecology.
 

### 2021-05-18 draft/draft_v2.tex
playing around with python code, checks to quantify bottom-up and top-down cohesiveness

@@
 \usepackage[table,xcdraw]{xcolor} % text and table background colors
-\usepackage[margin=3cm]{geometry} % margins
+\usepackage[margin=2.5cm]{geometry} % margins
 \usepackage{lineno} % line numbers
@@
 \usepackage{mdframed} % to easily handle boxes
+\usepackage[misc]{ifsym} % use \Letter symbol for corresponding author
 
@@
 \author[1,2,*]{Alicia Sanchez-Gorostiaga}
+\author[1]{Chang-Yu Chang}
+\author[1]{Helen S. Cai}
 \author[3]{Joshua E. Goldford}
 \author[4]{Mikhail Tikhonov}
-\author[1,$\dagger$]{Álvaro Sánchez}
+\author[1,\Letter]{Álvaro Sánchez}
 \affil[1]{Department of Ecology \& Evolutionary Biology
@@
 
-\affil[$\dagger$]{\normalfont To whom correspondence should be addressed: alvaro.sanchez@yale.edu}
+\affil[$\textrm{\Letter}$]{\normalfont To whom correspondence should be addressed: alvaro.sanchez@yale.edu}
 \affil[*]{\normalfont These authors contributed equally}
@@
 We show that the empirically observed trends in ecological co-selection are reproduced
-with minimal model assumptions, and that
-tuning the complexity of the metabolic interactions in our \textit{in silico} communities can modulate
-the recurrence of top-down or bottom-up co-selection.
+with minimal model assumptions, and that the recurrence of top-down and bottom-up co-selection
+is determined by the configuration of the cross-feeding networks in the MicroCRM.
 Our findings indicate that collective interactions play an important role at dictating community structure
@@
 \begin{center}
-\includegraphics[width=8cm,keepaspectratio]{figs/figBox_v2.png}
+\includegraphics[width=11cm,keepaspectratio]{figs/figBox_v3.png}
 \end{center}

### 2021-05-18 coalescence_v2.py

@@ -1,0 +1,927 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Wed Sep 23 18:28:37 2020
+@author: Juan
+"""
+
+
+
+### INITIALIZIATION
+
+# reset variables etc.
+from IPython import get_ipython
+get_ipython().magic('reset -sf')
+
+# libraries
+import community_simulator
+import community_simulator.usertools
+import community_simulator.visualization
+import numpy as np
+import pandas as pd
+# import matplotlib
+import matplotlib.pyplot as plt
+import seaborn as sns
+colors = sns.color_palette()
+# %matplotlib inline
+
+import random
+import math
+import copy
+import os
+
+# check if running on Windows
+if os.name == 'nt':
+    par = False
+else:
+    par = True
+
+# set random and numpy's seed manually for reproducibility
+#np.random.seed(0)
+#random.seed(0)
+
+import time
+start_time = time.time()
+
+
+
+### USER-DEFINED FUNCTIONS
+
+# stabilize communities through serial passaging
+def stabilizeCommunities(plate):
+    
+    # parameters
+    dilution_factor = 1/100
+    n_transfers = 20
+    time_per_cycle = 1
+    # extinction_threshold = 1e-4 # we don't use this because we don't use StadyState()
+    
+    # run stabilization protocol
+    Ntraj, Rtraj = plate.RunExperiment(f = np.diag(np.ones(plate.n_wells))*dilution_factor,
+                                       T = time_per_cycle,
+                                       npass = n_transfers,
+                                       compress_resources=False,
+                                       compress_species=True)
+    
+    return([plate.N,plate.R])
+
+# rank plot
+def rankPlot(N):
+    
+    # abundance threshold
+    u = 1e-4
+    
+    # check number of wells
+    n_wells = N.shape[1]
+    
+    # normalize to get fractions
+    N = N/N.sum()
+        
+    # initialize figure
+    fig, ax = plt.subplots(1)
+    
+    # plot
+    for i in range(n_wells):
+        
+        # get ith column of N, remove elements below threshold and sort it
+        Ni = N.iloc[:,i].to_numpy()
+        Ni = Ni[Ni>u]
+        Ni.sort()
+        Ni = Ni[::-1]
+        
+        # plot
+        ax.plot(Ni)
+    
+    ax.set_yscale("log")
[…]

### 2021-05-18 data/analyzeData_v2.R

@@ -1,0 +1,2065 @@
+### Juan Diaz-Colunga
+### Feb 16 2021
+
+### ----------------------------------------------------------------------
+### This script performs all the analyses and produces all the plots
+### in the paper.
+###    1. Community composition (pie plots & rank plots of ESV frequency)
+###    2. Characterization of isolated species from each community
+###    3. Matching isolated species to dominants based on sequencing data
+###    4. Fractions of isolated species in pairwise competition
+###    5. Similarity of coalesced, resident and invasive communities
+### ----------------------------------------------------------------------
+
+rm(list=ls())
+source('isolates_abundance_from_ESVs.R')
+source('myFunctions.R')
+library(matlib)
+library(testthat)
+library(ggplot2)
+library(ggh4x)
+library(tidyr)
+library(RColorBrewer)
+
+# initialize variables to store plots
+myplots <- vector(mode='list')
+if(!dir.exists('plots')) dir.create('plots')
+display_plots <- TRUE
+save_plots <- TRUE
+
+# general options
+deconvolute_ESVs <- FALSE # should species abundance be estimated from ESV abundance?
+
+
+### ----------------------------------------------------------------------
+### LOAD AND FORMAT DATA
+### ----------------------------------------------------------------------
+
+# load OTU table, taxonomy table and metadata
+otus <- read.csv('OTU_table.csv',stringsAsFactors=F)
+taxa <- read.csv('taxonomy_silva.csv',stringsAsFactors=F)
+metadata <- read.csv('metaData.csv',stringsAsFactors=F)
+
+### merge similar taxa?
+# merged <- mergeTaxa(otus,taxa)
+# otus <- merged[['otus']]
+# taxa <- merged[['taxa']]
+
+# check that OTU table and taxonomy table follow the same order
+stopifnot(all(otus$X==taxa$X))
+
+# keep only relevant metadata
+metadata <- metadata[,c("Sample","Plate","Row","Column","Experiment",
+                        "Comm1","StabilizingCarb1","Comm2","StabilizingCarb2",
+                        "Carbon","Rep","Transfer")]
+
+# assign IDs to each sequence
+otus <- cbind(seq_id = paste('seq_',1:nrow(otus),sep=''),
+              otus)
+
+# normalize OTU abundances (so entries in table represent fractions)
+otus[,3:ncol(otus)] <- otus[,3:ncol(otus)]/
[…]

### 2021-05-19 coalescence_v2.py
to do tomorrow: simulations break in the dominant monocultures, potentially because w is now a list instead of a scalar --check expression for J_out in the specirfic metabolism case

@@ -256,10 +256,10 @@ assumptions['fw'] = 0.45 #0.45 # fraction of secretion flux to waste resources
 assumptions['metabolism'] = 'specific' # 'common' uses a common D matrix for all species, 'specific' uses a different matrix D for each species
 assumptions['rs'] = 0.0 # control parameter for resource secretion: 0 means random secretions, 1 means species only secrete resources they can consume (relevant only if 'metabolism' is 'specific')
 
-assumptions['w'] = [0.5 + (1-0.5)*random.random() for i in range(assumptions['MA'].sum())]
+assumptions['w'] = [0.5 + (1-0.5)*random.random() for i in range(sum(assumptions['MA']))]
 assumptions['w'].sort(reverse=True)
 
-assumptions['g'] = [0.5 + (1-0.5)*random.random() for i in range(assumptions['SA'].sum() + assumptions['Sgen'])]
+assumptions['g'] = [0.5 + (1-0.5)*random.random() for i in range(sum(assumptions['SA']) + assumptions['Sgen'])]
 
 print(assumptions)
 

### 2021-05-19 coalescence.py
cleaned code a bit, initialized minimal model --to do: wrap into function for easy streamlining, make function return Q, f_pairwise, etc.

@@ -6,7 +6,8 @@ Created on Wed Sep 23 18:28:37 2020
 
 
 
-### INITIALIZIATION
+# %%
+### INITIALIZATION
 
 # reset variables etc.
 from IPython import get_ipython
@@ -44,6 +45,7 @@ start_time = time.time()
 
 
 
+# %%
 ### USER-DEFINED FUNCTIONS
 
 # stabilize communities through serial passaging
@@ -227,14 +229,16 @@ def mysim(p,q,pq):
 
 
 
-### SEED AND STABILIZE COMMUNITIES
+# %%
+### MODEL DEFINITION
 
 # general assumptions
 assumptions = community_simulator.usertools.a_default.copy()
-assumptions['n_wells'] = 100 # number of communities
+assumptions['n_wells'] = 50 # number of communities
 assumptions['S'] = 50 # number of species sampled at initialization
 assumptions['SA'] = [800, 800, 800] # [100, 100, 100] # number of species per specialist family
 assumptions['Sgen'] = 240 # 30 # number of generalists
+assumptions['MA'] = [10, 10, 10] # [30, 30, 30] # number of resources per resource class
 assumptions['l'] = 0.8 # leakage fraction
 
 assumptions['response'] = 'type I'
@@ -255,14 +259,6 @@ assumptions['fw'] = 0.45 #0.45 # fraction of secretion flux to waste resources
 assumptions['metabolism'] = 'specific' # 'common' uses a common D matrix for all species, 'specific' uses a different matrix D for each species
 assumptions['rs'] = 0.0 # control parameter for resource secretion: 0 means random secretions, 1 means species only secrete resources they can consume (relevant only if 'metabolism' is 'specific')
 
-# assumptions that depend on the nature of the supplied primary resource
-# 'complex' resource
-#assumptions['MA'] = [10, 10, 10] # [30, 30, 30] # number of resources per resource class
-# 'simple' resource
-assumptions['MA'] = 15 # [30, 30, 30] # number of resources per resource class
-
-print(assumptions)
-
 # parameters
 params = community_simulator.usertools.MakeParams(assumptions)
 
@@ -274,25 +270,10 @@ def dRdt(N,R,params):
 
 dynamics = [dNdt,dRdt]
 
-# plot matrices c and D
-'''
-fig,ax=plt.subplots()
[…]

### 2021-05-19 coalescence_minimalModel.py

@@ -1,0 +1,196 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Wed May 19 16:27:04 2021
+@author: Juan
+"""
+
+
+
+# %%
+### INITIALIZATION
+
+# reset variables etc.
+from IPython import get_ipython
+get_ipython().magic('reset -sf')
+
+# libraries
+import community_simulator
+import community_simulator.usertools
+import community_simulator.visualization
+import numpy as np
+import pandas as pd
+# import matplotlib
+import matplotlib.pyplot as plt
+import seaborn as sns
+colors = sns.color_palette()
+# %matplotlib inline
+
+import random
+import math
+import copy
+import os
+
+# check if running on Windows
+if os.name == 'nt':
+    par = False
+else:
+    par = True
+
+# set random and numpy's seed manually for reproducibility
+#np.random.seed(0)
+#random.seed(0)
+
+import time
+start_time = time.time()
+
+
+
+### USER-DEFINED FUNCTIONS
+
+# stabilize communities through serial passaging
+def stabilizeCommunities(plate):
+    
+    # parameters
+    dilution_factor = 1/100
+    n_transfers = 20
+    time_per_cycle = 1
+    # extinction_threshold = 1e-4 # we don't use this because we don't use StadyState()
+    
+    # run stabilization protocol
+    Ntraj, Rtraj = plate.RunExperiment(f = np.diag(np.ones(plate.n_wells))*dilution_factor,
+                                       T = time_per_cycle,
+                                       npass = n_transfers,
+                                       compress_resources=False,
+                                       compress_species=True)
+    
+    return([plate.N,plate.R])
+
+# Bray-Curtis similarity
+# p and q are lists (of absolute abundances, sum need not be 1)
+def bray_curtis(p,q):
+
+    # C_pq: sum of lesser values of common species
+    c = sum([min(p[i],q[i]) for i in range(len(p)) if p[i]>0 and q[i]>0])
+    
+    # Bray-Curtis similarity
+    bc = 2*c/(sum(p) + sum(q))
+    return bc
+
+# similarity index (relative Bray-Curtis similarity)
+# p: invasive
+# q: resident
+# pq: coalesced
+def mysim(p,q,pq):
+    
+    # normalization
+    p = [p_i/sum(p) for p_i in p]
+    q = [q_i/sum(q) for q_i in q]
+    pq = [pq_i/sum(pq) for pq_i in pq]
+    
[…]

### 2021-05-19 coalescence_v2.py

@@ -1,927 +1,0 @@
-# -*- coding: utf-8 -*-
-"""
-Created on Wed Sep 23 18:28:37 2020
-@author: Juan
-"""
-
-
-
-### INITIALIZIATION
-
-# reset variables etc.
-from IPython import get_ipython
-get_ipython().magic('reset -sf')
-
-# libraries
-import community_simulator
-import community_simulator.usertools
-import community_simulator.visualization
-import numpy as np
-import pandas as pd
-# import matplotlib
-import matplotlib.pyplot as plt
-import seaborn as sns
-colors = sns.color_palette()
-# %matplotlib inline
-
-import random
-import math
-import copy
-import os
-
-# check if running on Windows
-if os.name == 'nt':
-    par = False
-else:
-    par = True
-
-# set random and numpy's seed manually for reproducibility
-#np.random.seed(0)
-#random.seed(0)
-
-import time
-start_time = time.time()
-
-
-
-### USER-DEFINED FUNCTIONS
-
-# stabilize communities through serial passaging
-def stabilizeCommunities(plate):
-    
-    # parameters
-    dilution_factor = 1/100
-    n_transfers = 20
-    time_per_cycle = 1
-    # extinction_threshold = 1e-4 # we don't use this because we don't use StadyState()
-    
-    # run stabilization protocol
-    Ntraj, Rtraj = plate.RunExperiment(f = np.diag(np.ones(plate.n_wells))*dilution_factor,
-                                       T = time_per_cycle,
-                                       npass = n_transfers,
-                                       compress_resources=False,
-                                       compress_species=True)
-    
-    return([plate.N,plate.R])
-
-# rank plot
-def rankPlot(N):
-    
-    # abundance threshold
-    u = 1e-4
-    
-    # check number of wells
-    n_wells = N.shape[1]
-    
-    # normalize to get fractions
-    N = N/N.sum()
-        
-    # initialize figure
-    fig, ax = plt.subplots(1)
-    
-    # plot
-    for i in range(n_wells):
-        
-        # get ith column of N, remove elements below threshold and sort it
-        Ni = N.iloc[:,i].to_numpy()
-        Ni = Ni[Ni>u]
-        Ni.sort()
-        Ni = Ni[::-1]
-        
-        # plot
-        ax.plot(Ni)
-    
-    ax.set_yscale("log")
[…]

### 2021-05-19 community-simulator/community_simulator/essentialtools.py

@@ -42,12 +42,12 @@ def CompressParams(not_extinct_consumers,not_extinct_resources,params,dimensions
     if 'MxM' in dimensions.keys():
         for item in dimensions['MxM']:
             if item in params_comp.keys():
-                ### FIXME: added this condition: is D a np.ndarray or a list?
-                if isinstance(params_comp[item],np.ndarray): ### FIXME: if D is a np.ndarray, proceed normally (compress it)
+                ### added this condition: is D a np.ndarray or a list?
+                if isinstance(params_comp[item],np.ndarray): ### if D is a np.ndarray, proceed normally (compress it)
                     assert np.shape(params_comp[item])==(M,M), 'Invalid shape for ' + item + '. Please update dimensions dictionary with correct dimensions.'
                     params_comp[item]=params_comp[item][not_extinct_resources,:]
                     params_comp[item]=params_comp[item][:,not_extinct_resources]
-                elif isinstance(params_comp[item],list): ### FIXME: if D is a list, do: 1) compress the list itself (remove elements corresponding to extinct species) and 2) compress every element of the list (remove rows/columns corresponding to extinct resources) - keep the list structure to wrap the remaining compressed matrices
+                elif isinstance(params_comp[item],list): ### if D is a list, do: 1) compress the list itself (remove elements corresponding to extinct species) and 2) compress every element of the list (remove rows/columns corresponding to extinct resources) - keep the list structure to wrap the remaining compressed matrices
                     params_comp[item]=[params_comp[item][i] for i in range(len(params_comp[item])) if not_extinct_consumers[i]]
                     params_comp[item]=pd.Series(params_comp[item]).apply(lambda x: x[not_extinct_resources,:]).tolist()
                     params_comp[item]=pd.Series(params_comp[item]).apply(lambda x: x[:,not_extinct_resources]).tolist()

### 2021-05-20 coalescence_minimalModel.py
minimal model ready to start spitting data

@@ -51,17 +51,35 @@ start_time = time.time()
 def stabilizeCommunities(plate):
     
     # parameters
-    dilution_factor = 1/100
-    n_transfers = 20
+    dilution_factor = 0.01
     time_per_cycle = 1
-    # extinction_threshold = 1e-4 # we don't use this because we don't use StadyState()
     
-    # run stabilization protocol
+    # get starting abundance of species in plate
+    Nold = plate.N.copy()
+    
+    # run stabilization protocol for one transfer
     Ntraj, Rtraj = plate.RunExperiment(f = np.diag(np.ones(plate.n_wells))*dilution_factor,
                                        T = time_per_cycle,
-                                       npass = n_transfers,
+                                       npass = 1,
                                        compress_resources=False,
                                        compress_species=True)
+    
+    # get new abundances
+    Nnew = plate.N.copy()
+    
+    # maximum relative change in abundance
+    d = np.max(np.abs(Nnew-Nold)/Nold)
+    
+    # run stabilization cycles until maximum relative change in abundance is below 1% (generational equlibrium achieved)
+    while d[0] > 0.01:
+        Nold = plate.N.copy()
+        Ntraj, Rtraj = plate.RunExperiment(f = np.diag(np.ones(plate.n_wells))*dilution_factor,
+                                       T = time_per_cycle,
+                                       npass = 1,
+                                       compress_resources=False,
+                                       compress_species=True)
+        Nnew = plate.N.copy()
+        d = np.max(np.abs(Nnew-Nold)/Nold)
     
     return([plate.N,plate.R])
 
@@ -101,11 +119,11 @@ assumptions['S'] = 1
 assumptions['SA'] = 4
 assumptions['Sgen'] = 0 # 30 # number of generalists
 assumptions['MA'] = 5 # [30, 30, 30] # number of resources per resource class
-assumptions['l'] = 0.8 # leakage fraction
+assumptions['l'] = 0.5 # leakage fraction
 
+assumptions['sampling'] = 'Gamma'
 assumptions['response'] = 'type I'
 assumptions['regulation'] = 'independent' # 'independent' is standard, 'energy' or 'mass' enable preferential consumption of resources
-assumptions['sampling'] = 'Gamma' # 'Binary', 'Gaussian', 'Uniform' or 'Gamma' (sampling of the matrix c)
[…]

### 2021-05-23 draft/draft_v3.tex
defined community hierarchy in coalescence.py; new plots (figures 4 and 5); need to run the final simulations for the final figures

--- 2021-05-18 draft/draft_v2.tex
@@
 % title, authors, affiliations
-\title{Cohesiveness in microbial community coalescence}
+\title{Top-down versus bottom-up cohesiveness in microbial community coalescence}
 \author[1,*]{Juan Diaz-Colunga}
@@
 
-\affil[$\textrm{\Letter}$]{\normalfont To whom correspondence should be addressed: alvaro.sanchez@yale.edu}
+\affil[$\textrm{\Letter}$]{\normalfont alvaro.sanchez@yale.edu}
 \affil[*]{\normalfont These authors contributed equally}
@@
 This has been observed, for instance, 
-in river courses where terrestrial microbial communities mix with aquatic microorganisms
+in river courses where terrestrial microbes mix with aquatic microorganisms
 \cite{Mansour2018,Luo2020,Vass2021}
@@
 
+\begin{figure}[!h]
+\centering
+\internallinenumbers
+\includegraphics[width=13cm,keepaspectratio]{figs/fig4_v2.pdf}
+\caption{\textbf{A minimal model of community coalescence.}
+\textbf{a.} Blablabla...
+\textbf{b.} Blablabla...}
+\label{fig4}
+\end{figure}
+
+\clearpage
+
+\begin{figure}[!h]
+\centering
+\internallinenumbers
+\includegraphics[width=13cm,keepaspectratio]{figs/fig5.pdf}
+\caption{\textbf{Community hierarchies determine coalescence outcomes.}
+\textbf{a.} Blablabla...
+\textbf{b.} Blablabla...
+\textbf{b.} Blablabla... ($p=0.083$).}
+\label{fig4}
+\end{figure}
+
+\clearpage
+
 \section*{Supplementary Figures}

### 2021-05-23 coalescence.py

@@ -36,12 +36,16 @@ if os.name == 'nt':
 else:
     par = True
 
-# set random and numpy's seed manually for reproducibility
-#np.random.seed(0)
-#random.seed(0)
+# set random and numpy's seed for reproducibility
+myseed = 999  # 500 works pretty nicely
+np.random.seed(myseed)
+random.seed(myseed)
 
 import time
 start_time = time.time()
+
+# get community hierarchies? (this generates a complete data set but takes time)
+getHierarchies = True
 
 
 
@@ -230,11 +234,11 @@ def mysim(p,q,pq):
 
 
 # %%
-### MODEL DEFINITION
+### MODEL DEFINITION 
 
 # general assumptions
 assumptions = community_simulator.usertools.a_default.copy()
-assumptions['n_wells'] = 50 # number of communities
+assumptions['n_wells'] = 100 # number of communities
 assumptions['S'] = 50 # number of species sampled at initialization
 assumptions['SA'] = [800, 800, 800] # [100, 100, 100] # number of species per specialist family
 assumptions['Sgen'] = 240 # 30 # number of generalists
@@ -418,6 +422,7 @@ N_pairwise, R_pairwise = stabilizeCommunities(pairwise_plate)
 
 
 
+# %%
 ### DOMINANT SPECIES INVADING ALONE
 
 # make plate (added dilution factor)
@@ -435,27 +440,58 @@ N_singleinv, R_singleinv = stabilizeCommunities(singleinv_plate)
 
 
 
-### COHORT INVADING WITHOUT DOMINANT SPECIES
+# %%
+### HIERARCHY INDEX: HOW MUCH OF THE DOMINANTS' GROWTH COMES FROM THE PRIMARY RESOURCE?
 
-# make cohort plate
-invasive_cohorts_plate = invasive_plate.copy()
-invasive_cohorts_plate.N = invasive_cohorts_plate.N.droplevel(0)
-for i in range(assumptions['n_wells']):
-    invasive_cohorts_plate.N.loc[dominants_invasive[i]][i] = 0
-invasive_cohorts_plate.N.index = invasive_plate.N.index
+# define dNdt on resource R0 only
+def dNdt_R0(N,R,params):
+    return community_simulator.usertools.MakeConsumerDynamics_R0(assumptions)(N,R,params)
+
+# define function to calculate hierarchy (requires integration of dynamic equations; passed plate should be stabilized)
+def communityHierarchy(plate,well):
+
[…]

### 2021-05-23 coalescence_minimalModel.py

@@ -18,16 +18,12 @@ import community_simulator
 import community_simulator.usertools
 import community_simulator.visualization
 import numpy as np
-import pandas as pd
 # import matplotlib
 import matplotlib.pyplot as plt
 import seaborn as sns
 colors = sns.color_palette()
 # %matplotlib inline
 
-import random
-import math
-import copy
 import os
 
 # check if running on Windows
@@ -119,7 +115,7 @@ assumptions['S'] = 1
 assumptions['SA'] = 4
 assumptions['Sgen'] = 0 # 30 # number of generalists
 assumptions['MA'] = 5 # [30, 30, 30] # number of resources per resource class
-assumptions['l'] = 0.5 # leakage fraction
+assumptions['l'] = 0.8 # leakage fraction
 
 assumptions['sampling'] = 'Gamma'
 assumptions['response'] = 'type I'
@@ -147,7 +143,7 @@ dynamics = [dNdt,dRdt]
 ### COALESCENCE EXPERIMENTS SETUP
 
 # define experiment function (so it is iterable)
-def competeMinimalCommunities(c_11 = 1.5, c_31 = 1, c_13 = 0, c_35 = 0, c_22 = 1, c_44 = 1):
+def competeMinimalCommunities(c_11 = 1.5, c_31 = 1, c_13 = 0, c_35 = 0, c_22 = 100, c_44 = 100):
 
     # consumer matrix
     params['c'].iloc[:,:] = np.array([[c_11,    0, c_13,    0,    0],
@@ -201,12 +197,33 @@ def competeMinimalCommunities(c_11 = 1.5, c_31 = 1, c_13 = 0, c_35 = 0, c_22 = 1
     # stabilize coalesced community
     N_coa, R_coa = stabilizeCommunities(plate_coa)
     
-    # monocultures
-    N_mono = params['g']*params['w']*params['R0'][0]*(1 - params['l'])
+    # monoculture: resident dominant
+    N0_res_mono = init_state[0].copy()
+    N0_res_mono.iloc[:,0] = 0
+    N0_res_mono.iloc[0,0] = 1
+    plate_res_mono = community_simulator.Community((N0_res_mono,init_state[1]),
+                                                   dynamics,
+                                                   params,
+                                                   parallel=par,
[…]

### 2021-05-23 community-simulator/community_simulator/usertools.py

@@ -461,33 +461,13 @@ def MakeResourceDynamics(assumptions):
     """
     ### original in community-simulator package (single D matrix)
     J_out = lambda R,params: (params['l']*J_in(R,params)).dot(params['D'].T)
-    ###
-    """
-    
-    """
-    ### my first attempt at multiple matrices - very unefficient and slow bc it requires a lot of .dot() operations between large matrices
-    def J_out(R,params):
-        # get J_in so J_in(R,params) does not have to be re-evaluated for every element of the list
-        Jin = J_in(R,params)
-        # J_out is now a list of J_out_i matrices
-        # J_out_i is the result of the operation (l*J_in)*D_i'
-        # where D_i is the metabolic matrix for species i and ' indicates transposition
-        #J_out = pd.Series(params['D']).apply(lambda x: (params['l']*J_in(R,params)).dot(x.T)).tolist()
-        J_out = pd.Series(params['D']).apply(lambda x: (params['l']*Jin).dot(x.T)).tolist()
-        # the final J_out comes from combining all the J_out_i row by row
-        J_out = [J_out[i][i,:] for i in range(J_out[0].shape[0])]
-        J_out = np.array(J_out)
-        return J_out
-    ###
-    """
-    
-    """
-    ### my second attempt: using list comprehension and trimming J_in before .dot() to speed up
+
+    ### new version --accepts D as a list of matrices (using list comprehension and trimming J_in before .dot() to speed up)
     J_out = lambda R,params: params['l']*np.array([ J_in(R,params)[i,:].dot(params['D'][i].T) for i in range(len(params['D'])) ])
     ###
     """
     
-    ### incorporating both scenarios (D is a matrix or D is a list) in the same function
+    ### new version incorporating both scenarios (D can be a matrix or a list, depending on the choice of 'metabolism')
     J_out = {'common': lambda R,params: (params['l']*J_in(R,params)).dot(params['D'].T),
              'specific': lambda R,params: params['l']*np.array([ J_in(R,params)[i,:].dot(params['D'][i].T) for i in range(len(params['D'])) ])
             }
@@ -495,7 +475,6 @@ def MakeResourceDynamics(assumptions):
     return lambda N,R,params: (h[assumptions['supply']](R,params)
                                -(J_in(R,params)/params['w']).T.dot(N)
                                +(J_out[assumptions['metabolism']](R,params)/params['w']).T.dot(N))
-    ###
     
[…]

### 2021-05-23 data/analyzeData.R

@@ -1586,9 +1586,16 @@ if (save_plots) {
 plot_this <- read.table('simul_data.txt',header=TRUE)
 
 # strength of bottom-up co-selection
-plot_this$bucs <- plot_this$f_singleinv < 0.15 & plot_this$f_coalescence > 0.05
+plot_this$bucs <- plot_this$f_singleinv < 0.1 & plot_this$f_coalescence > 0.05
 plot_this$bucs <- c('neutral','strong')[1 + plot_this$bucs]
-plot_this$bucs <- factor(plot_this$bucs,levels=c('neutral','strong'))
+plot_this$bucs <- factor(plot_this$bucs,levels=c('strong','neutral'))
+
+# difference in hierarchies
+plot_this$h_diff = log10(plot_this$h_res/plot_this$h_inv)
+
+# hierarchy distribution
+h_distrib = data.frame(h=hist(c(plot_this$h_res,plot_this$h_inv),breaks=20)$mids,
+                       freq=hist(c(plot_this$h_res,plot_this$h_inv),breaks=20)$counts/(2*nrow(plot_this)))
 
 # make plots
 myplots[['q-vs-pairwise_bray-curtis_simul']] <-
@@ -1731,10 +1738,135 @@ myplots[['q-vs-pairwise_bray-curtis_simul_top-down-vs-bottom-up_neutral-bucs']] 
         panel.border=element_rect(size=0.25)) +
   coord_fixed()
 
+myplots[['simul_hierarchy-distribution']] <-
+  ggplot(data=h_distrib,
+         aes(x=h,y=freq)) +
+  geom_bar(stat='identity') +
+  scale_y_continuous(name='Frequency',
+                     limits=c(0,0.5),
+                     breaks=c(0,0.25,0.5),
+                     labels=c('0','0.25','0.5')) +
+  scale_x_continuous(name='Community hierarchy\n \n ',
+                     limits=c(0,1),
+                     breaks=c(0,0.5,1),
+                     labels=c('0','0.5','1')) +
+  theme_bw() +
+  theme(panel.grid=element_blank(),
+        text=element_text(size=15),
+        axis.text=element_text(size=15),
+        axis.line=element_blank(),
+        axis.ticks=element_line(size=0.25),
+        panel.border=element_rect(size=0.25),
+        legend.position='none') +
+  coord_fixed(2)
+
+myplots[['alone-vs-together_simul_hierarchy']] <-
+  ggplot(data=plot_this,
[…]

### 2021-05-24 draft/draft_v3.tex
more draft

@@
 \title{Top-down versus bottom-up cohesiveness in microbial community coalescence}
-\author[1,*]{Juan Diaz-Colunga}
-\author[1,*]{Nanxi Lu}
-\author[1,2,*]{Alicia Sanchez-Gorostiaga}
+\author[1 *]{Juan Diaz-Colunga}
+\author[1 *]{Nanxi Lu}
+\author[1,2 *]{Alicia Sanchez-Gorostiaga}
 \author[1]{Chang-Yu Chang}
@@
 \author[4]{Mikhail Tikhonov}
-\author[1,\Letter]{Álvaro Sánchez}
+\author[1 \Letter]{Álvaro Sánchez}
 \affil[1]{Department of Ecology \& Evolutionary Biology
@@
 
-\affil[$\textrm{\Letter}$]{\normalfont alvaro.sanchez@yale.edu}
-\affil[*]{\normalfont These authors contributed equally}
+\affil[$\textrm{\Letter}$]{\normalfont~alvaro.sanchez@yale.edu}
+\affil[*]{\normalfont~These authors contributed equally}
 \date{}
@@
 Model (MicroCRM) \cite{Goldford2018,Marsland2019} as implemented in the Community Simulator package
-for Python \cite{Marsland2020}. We chose this modeling framework because
+for Python \cite{Marsland2020}
+(\hyperref[box1]{Box~1}).
+We chose this modeling framework because
 communities assembled under our experimental conditions (natural microbiomes re-assembled into
@@
 for which the MicroCRM provides a good description.
-We carried out our simulations enabling a large variation across species in terms of
-their metabolic architectures;
+Based on previous experimental work
+\cite{Harcombe2014,Pinu2018,Estrela2020},
+we allowed different species in our model
+to secrete different sets of byproducts when metabolizing a same resource
+(\hyperref[box1]{Box~1}).
+\iffalse
 this choice is discussed in \hyperref[box1]{Box~1}.
+\fi
 To reproduce our experimental protocol \textit{in silico},
 we first generated a library of resources and two non-overlapping pools of species.
-Each pool was used to seed a collection of 100 invasive and resident communities respectively,
-that were allowed to stabilize through 20 growth-dilution cycles.
+Each pool was used to seed a collection of 100 invasive
+and 100 resident communities respectively
+by randomly choosing 50 species and
+allowing them to stabilize through 20 growth-dilution cycles.
 We then mixed these stable communities in pairs to simulate our coalescence
@@
 \internallinenumbers
-\subsubsection*{Box 1: Community cohesiveness in a Microbial Consumer-Resource Model}
+\subsubsection*{Box 1: A Microbial Consumer-Resource Model for community coalescence}
 \label{box1}
@@
 the energy to growth rate conversion factor ($g_i$ for species $i$)
-or the leakage fraction, i.e. the amount of energy lost as byproducts when a resource is consumed
+and the leakage fraction, i.e. the amount of energy lost as byproducts when a resource is consumed
 ($l_\alpha$ for resource $\alpha$).
@@
 see \cite{Goldford2018,Marsland2019,Marsland2020}).
-The element $D_{\beta\alpha}$ of the metabolic matrix
-represents the amount of energy secreted in the form of resource $\beta$
-as a result of the metabolism of resource $\alpha$.
-
-This formulation assumes that all species metabolize resources through similar pathways,
-with any two species secreting the same byproducts when consuming a same resource.
-However, experimental evidence suggests that individual species can secrete different sets of metabolites
+Experimental evidence suggests that individual species can secrete different sets of metabolites
 to the environment when growing on a same primary resource
-\cite{Harcombe2014,Pinu2018,Estrela2020}. This observation motivated us to introduce a new
-feature in the model:
-in short, we now define $\mathbf{D}$ as a three-dimensional matrix
+\cite{Harcombe2014,Pinu2018,Estrela2020}.
+Thus, we define $\mathbf{D}$ as a three-dimensional matrix
 where the element $D_{i\beta\alpha}$ represents the energy flux in the form of resource $\beta$
-that is secreted \textit{by species $i$} when it metabolizes resource $\alpha$.
-Note that now $D_{i\beta\alpha}$ need not be equal to $D_{j\beta\alpha}$ if $i \neq j$
+that is secreted by species $i$ when it metabolizes resource $\alpha$.
+Note that $D_{i\beta\alpha}$ need not be equal to $D_{j\beta\alpha}$ if $i \neq j$
 (see illustration below).
+\iffalse
+For our simulations,
+the elements of $\mathbf{c}$ are sampled from a gamma distibution,
+while the elements of $\mathbf{D}$ are sampled from a Dirichlet distribution
+to ensure that the normalization condition $\sum_\beta D_{i\beta\alpha} = 1$
+is satisfied for every $i$ and $\alpha$.
+\fi
+\iffalse
+The sampling is performed so that species belonging to a same family
+tend to secrete more similar byproducts when consuming a same resource
+(see \methodsref{sim} and \cite{Marsland2019,Marsland2020}),
+as supported by experimental observations
+\cite{Harcombe2014,Pinu2018,Estrela2020}.
+\fi
 
@@
 
-We reason that the choice of species-specific metabolic architectures is necessary to potentially generate
-cohesiveness at the community level during coalescence.
-If the secretions of all species were identical (or only slightly different),
-higher order cross-feeding effects would be very unspecific:
-the establishment of new invasive species
---given that they could outcompete resident taxa within their metabolic niches,
-i.e. more effectively feed off the same resource or set of resources as them--
-would not alter (or only do so moderately)
-the metabolic flows through the rest of the community's cross-feeding network.
-On the other hand, said network could undergo a profound and further-reaching restructuring
-if the invasive species secreted very different sets of metabolites with respect to the resident ones,
-potentially disabling existing niches and/or enabling new ones
-where more invaders could be co-selected.
-For a similar reason, we argue that the sparsity of the metabolic matrix could also modulate
-the emergence of cohesiveness in the face of coalescence.
-A dense metabolic matrix corresponds to a situation where all species secrete a wide
-variety of byproducts.
-New-coming invasive species that secrete similar byproducts as resident ones
---even if they do so in different relative amounts--
-might only induce moderate quantitative changes in the metabolic fluxes.
-But if the sets of secretions are qualitatively different, co-selection of species adapted to
-each of those sets becomes possible.
-These ideas are supported by experimental observations
-suggesting that species with a history of coexistence make up
-cohesive communities with highly specific cross-feeding configurations
-\cite{Rosenzweig1994,Goldford2018,Estrela2020}.
+The following equations describe the kinetics of the abundances
+of the $i$-th species (denoted as $N_i$)
+and the $\alpha$-th resource (denoted as $R_\alpha$):
+
+\begin{equation}
+\frac{dN_i}{dt} = 
+g_i N_i
+\left[
+\sum_\alpha \left( 1-l_\alpha \right)
+w_\alpha c_{i\alpha} R_\alpha
+- m_i
+\right]
+\label{eq:crm-n}
+\end{equation}
+
+\begin{equation}
+\frac{dR_\alpha}{dt} = 
+- \sum_j N_j c_{j\alpha} R_\alpha
++ \sum_j \sum_\beta N_j c_{j\beta} R_\beta
+\left[
+l_\beta D_{j\alpha\beta} \frac{w_\beta}{w_\alpha}
+\right]
+\label{eq:crm-r}
+\end{equation}
+%
+These equations can take slightly different forms in certain cases,
+e.g. if the primary resource is supplied continuously instead of at the beggining
+of each growth cycle \cite{Marsland2019,Marsland2020},
+but they are a good approximation for the community dynamics between consecutive
+serial dilutions in our setup.
+Here, we assembled \textit{in silico} communities by randomly
+sampling a set of species from a pool,
+then integrating equations \ref{eq:crm-n} and \ref{eq:crm-r},
+diluting the final abundances,
+replenishing the primary resource (by default denoted as $R_0$),
+and repeating the process until generational equilibrium was achieved
+(\methodsref{sim}).
+Coalescence simulations were carried out following the same logic,
+this time seeding the coalesced communities by mixing the invasive and resident
+communities instead of sampling from a species pool.
 
@@
 Our data indicates that the primary resource supplied to the communities can modulate
-the effect of the cohorts in the dominants pairwise competition
+the effect that the cohorts have in the dominants pairwise competition
 (\figref[a]{fig2})
@@
 (\figref[b]{fig2}, left and middle panels).
-Our model suggests that this might be a result of the metabolic interactions
-between community members, including the rarer taxa
-(\hyperref[box1]{Box~1}).
+The fact that
+our model captures these trends
+suggests that this might be a result of the metabolic interactions
+between community members, including the rarer taxa.
 To investigate the potential role of the cohorts in coalescence,
@@
 (\figref[b]{fig3}, green shaded region).
-On the other hand, a high degree of antagonistic bottom-up co-selection would result in dominants
+On the other hand, a high degree of bottom-up antagonism would result in dominants
 invading more effectively alone than in the presence of their cohorts
@@
 (dots near the diagonal of \figref[b]{fig3}).
-We went back to our original simulations and plotted the frequency of the invasive dominant
+We reexamined our original simulations and plotted the frequency of the invasive dominant
 in pairwise competition with the resident dominant versus the relative similarity between
@@
 We then asked whether this trend was also observed \textit{in vitro}.
-We went back to our synthetic communities
+We went back to our laboratory cultures
 and carried out a new round of experiments where we invaded the resident communities
@@
 pairwise competition between dominants is only predictive of coalescence outcomes
-if bottom-up co-selection is weak.
+if bottom-up co-selection is weak
+(\figref[e]{fig3}, $R^2=0.07$, $p>0.05$ when bottom-up co-selection is strong;
+$R^2=0.37$, $p<10^{-4}$ when bottom-up co-selection is weak).
 
@@
 
+\subsubsection*{Understanding the mechanisms of ecological co-selection: a minimal model of community coalescence}
+
+To better understand the underlying mechanisms that govern
+the emergence of ecological co-selection,
+we developed a minimal model of community coalescence.
+This model is comprised of two communities with two species each
+as illustrated in \figref[a]{fig4}.
+Within each community, the dominant species
+($\mathrm{s}_1$ and $\mathrm{s}_3$ in the resident and invasive communities
+respectively)
+is able to utilize the primary resource that is supplied externally ($\mathrm{R}_1$).
+As a result of $\mathrm{R}_1$ metabolism,
+the dominants secrete a single byproduct off which
+the sub-dominant species ($\mathrm{s}_2$ and $\mathrm{s}_4$ respectively) can feed.
+Finally, these sub-dominants secrete an additional resource that can again
+be metabolized by the respective dominants.
+This is the simplest structure where both top-dwn and bottom-up
+cohesiveness can emerge, with their relative strengths being determined
+by how effectively the species can utilize each other's metabolic byproducts.
+Several parameters can modulate the direction of community cohesiveness,
+and even in this simple model there can be complex interactions between them.
+For instance, a sub-dominant that is very efficient at metabolizing the dominant
+secretions will therefore be a strong producer of secondary secretions,
+in turn increasing the metabolic flow towards the dominant itself.
+
+\subsubsection*{Community hierarchy regulates the strength of bottom-up co-selction}
+
+How do the ideas above scale to more complex and diverse communities?
+
+We reason that the choice of species-specific metabolic architectures is necessary to potentially generate
+cohesiveness at the community level during coalescence.
+If the secretions of all species were identical (or only slightly different),
+higher order cross-feeding effects would be very unspecific:
+the establishment of new invasive species
+--given that they could outcompete resident taxa within their metabolic niches,
+i.e. more effectively feed off the same resource or set of resources as them--
+would not alter (or only do so moderately)
+the metabolic flows through the rest of the community's cross-feeding network.
+On the other hand, said network could undergo a profound and further-reaching restructuring
+if the invasive species secreted very different sets of metabolites with respect to the resident ones,
+potentially disabling existing niches and/or enabling new ones
+where more invaders could be co-selected.
+For a similar reason, we argue that the sparsity of the metabolic matrix could also modulate
+the emergence of cohesiveness in the face of coalescence.
+A dense metabolic matrix corresponds to a situation where all species secrete a wide
+variety of byproducts.
+New-coming invasive species that secrete similar byproducts as resident ones
+--even if they do so in different relative amounts--
+might only induce moderate quantitative changes in the metabolic fluxes.
+But if the sets of secretions are qualitatively different, co-selection of species adapted to
+each of those sets becomes possible.
+These ideas are supported by experimental observations
+suggesting that species with a history of coexistence make up
+cohesive communities with highly specific cross-feeding configurations
+\cite{Rosenzweig1994,Goldford2018,Estrela2020}.
+
 \subsubsection*{Conclusions}
@@
 \cite{Goldford2018,Estrela2020}.
-This network is in turn regulated by the supplied primary carbon source.
+This network is in turn regulated by the supplied primary carbon source
+in our minimal laboratory conditions.
 This idea is supported by the observation that a Microbial Consumer-Resource Model
@@
 ESV abundances represented by the vectors
-$\mathbf{x} = \left( x_1, x_2, \cdots, x_N \right)$
+$\mathbf{x} = \left( x_1, x_2, \cdots, x_S \right)$
 and
-$\mathbf{y} = \left( y_1, y_2, \cdots, y_N \right)$
+$\mathbf{y} = \left( y_1, y_2, \cdots, y_S \right)$
 (where $x_i$ and $y_i$ represent the relative abundance of the $i$th ESV in
-each community respectively and $N$ is the total number of ESVs), the Bray-Curtis
+each community respectively and $S$ is the total number of ESVs), the Bray-Curtis
 similarity $BC \left( \mathbf{x}, \mathbf{y} \right)$ is calculated as
@@
 and the updated Community Simulator package with instructions for enabling the
-new features are deposited in \url{github.com/jdiazc9/coalescence}.
+new features are in \url{github.com/jdiazc9/coalescence}.
 
@@
 \internallinenumbers
-\includegraphics[width=10cm,keepaspectratio]{figs/fig1.pdf}
+\includegraphics[scale=0.7,keepaspectratio]{figs/fig1.pdf}
 \caption{\textbf{Overview of the experimental protocol.}
@@
 \internallinenumbers
-\includegraphics[width=15cm,keepaspectratio]{figs/fig2_v3.pdf}
+\includegraphics[scale=0.7,keepaspectratio]{figs/fig2_v3.pdf}
 \caption{\textbf{Top-down co-selection in microbial community coalescence.}
@@
 \internallinenumbers
-\includegraphics[width=12cm,keepaspectratio]{figs/fig3.pdf}
+\includegraphics[scale=0.7,keepaspectratio]{figs/fig3.pdf}
 \caption{\textbf{Trade offs between bottom-up and top-down ecological co-selection.}
@@
 \internallinenumbers
-\includegraphics[width=13cm,keepaspectratio]{figs/fig4_v2.pdf}
+\includegraphics[scale=0.7,keepaspectratio]{figs/fig4_v2.pdf}
 \caption{\textbf{A minimal model of community coalescence.}
@@
 \internallinenumbers
-\includegraphics[width=13cm,keepaspectratio]{figs/fig5.pdf}
+\includegraphics[scale=0.7,keepaspectratio]{figs/fig5.pdf}
 \caption{\textbf{Community hierarchies determine coalescence outcomes.}
@@
 \textbf{b.} Blablabla... ($p=0.083$).}
-\label{fig4}
+\label{fig5}
 \end{figure}
@@
 \internallinenumbers
-\includegraphics[width=15cm,keepaspectratio]{figs/figS1.pdf}
+\includegraphics[scale=0.9,keepaspectratio]{figs/figS1.pdf}
 \caption{\textbf{Community compositions after seven additional transfers without coalescence.}
@@
 \internallinenumbers
-\includegraphics[width=12.5cm,keepaspectratio]{figs/figS2.pdf}
+\includegraphics[scale=0.9,keepaspectratio]{figs/figS2.pdf}
 \caption{\textbf{Alternative metrics of community distance.}
@@
 \internallinenumbers
-\includegraphics[width=10cm,keepaspectratio]{figs/figS3.pdf}
+\includegraphics[scale=0.9,keepaspectratio]{figs/figS3.pdf}
 \caption{\textbf{Dominant species have limited effects on coalescence outcomes quantification.}

### 2021-05-24 coalescence.py

@@ -37,7 +37,7 @@ else:
     par = True
 
 # set random and numpy's seed for reproducibility
-myseed = 999  # 500 works pretty nicely
+myseed = 9999  # 500 works pretty nicely
 np.random.seed(myseed)
 random.seed(myseed)
 
@@ -45,7 +45,7 @@ import time
 start_time = time.time()
 
 # get community hierarchies? (this generates a complete data set but takes time)
-getHierarchies = True
+getHierarchies = False
 
 
 

### 2021-05-25 draft/draft_v3.tex
minor changes to draft

@@
 diluting the final abundances,
-replenishing the primary resource (by default denoted as $R_0$),
+replenishing the primary resource,
 and repeating the process until generational equilibrium was achieved
@@
 this time seeding the coalesced communities by mixing the invasive and resident
-communities instead of sampling from a species pool.
+ones instead of sampling from a species pool.
 
@@
 We reexamined our original simulations and plotted the frequency of the invasive dominant
-in pairwise competition with the resident dominant versus the relative similarity between
+in pairwise competition
+\iffalse
+with the resident dominant
+\fi
+versus the relative similarity between
 the invasive and coalesced communities,

### 2021-05-25 draft/draft_v3.tex
coalescence.py now shows more plots; more draft written

@@
 As a result of $\mathrm{R}_1$ metabolism,
-the dominants secrete a single byproduct off which
-the sub-dominant species ($\mathrm{s}_2$ and $\mathrm{s}_4$ respectively) can feed.
-Finally, these sub-dominants secrete an additional resource that can again
+the dominants secrete a single byproduct
+($\mathrm{R}_2$ and $\mathrm{R}_4$ respectively)
+off which
+the sub-dominants ($\mathrm{s}_2$ and $\mathrm{s}_4$ respectively) can feed.
+Finally, these sub-dominants secrete an additional resource
+($\mathrm{R}_3$ and $\mathrm{R}_5$ respectively)
+that can again
 be metabolized by the respective dominants.
-This is the simplest structure where both top-dwn and bottom-up
+This is the simplest structure where both top-down and bottom-up
 cohesiveness can emerge, with their relative strengths being determined
 by how effectively the species can utilize each other's metabolic byproducts.
-Several parameters can modulate the direction of community cohesiveness,
+Several parameters can modulate the strength and direction of community cohesiveness,
 and even in this simple model there can be complex interactions between them.
-For instance, a sub-dominant that is very efficient at metabolizing the dominant
-secretions will therefore be a strong producer of secondary secretions,
+For instance, a sub-dominant that is very efficient at metabolizing the dominant's
+secretions will therefore be a strong producer of secondary byproducts,
 in turn increasing the metabolic flow towards the dominant itself.
+To keep both communities symmetrical and
+mantain tractability of the model's behavior, we chose to make
+the two sub-dominant species equally able consumers of the
+byproduct of their respective dominants
+(\methodsref{sim-min}).
+We also made it so species secrete only one type of byproduct each,
+i.e. the metabolic matrix is binary in this case
+(see \hyperref[box1]{Box~1}).
+The secretions of the resident and invasive communities are
+non-overlapping:
+\iffalse
+(i.e. $\mathrm{s}_1$ and $\mathrm{s}_2$ secretions
+differ from $\mathrm{s}_3$ and $\mathrm{s}_4$ secretions)
+\fi
+this ensures that coalescence always results in one of the two
+communities taking over and completely excluding the other one,
+which facilitates the interpretation of the model's outcomes.
+
+With these constraints, the model is specified by four rates:
+the resident dominant's uptake rate of the primary resource ($c_{11}$)
+and of the secondary byproduct secreted by $\mathrm{s}_2$ ($c_{13}$),
+as well as
+the invasive dominant's uptake rate of the primary resource ($c_{31}$)
+and of the secondary byproduct secreted by $\mathrm{s}_4$ ($c_{35}$).
+Analogous to what we did for our initial
+experiments and simulations of community coalescence,
+we first stabilized each community separately
+through serial growth-dilution cycles,
+and then mixed them 1:1 and stabilized again until generational
+equilibrium was achieved.
+Equations \ref{eq:crm-n} and \ref{eq:crm-r} in \hyperref[box1]{Box~1}
+describe the dynamics of this system between consecutive dilutions
+with the corresponding choices for the parameter values
+(\methodsref{sim-min}).
+We ran simulations for both community-community competitions and
+invasions from the invasive dominant alone.
+We considered four different scenarios illustrated in \figref[b-e]{fig4}:
+\textit{1)} both communities are cohesive
+from the top-down but not from the bottom-up,
+\textit{2)} the resident community is cohesive
+both from the top-down and from the bottom-up
+while
+the resident community is cohesive
+from the top-down only,
+\textit{3)} the resident community is cohesive
+from the top-down only
+while
+the invasive community is cohesive
+both from the top-down and from the bottom-up,
+and \textit{4)} both communities are cohesive
+from the top-down and from the bottom-up.
+The outcomes in each case are discussed below.
+
+\paragraph*{Top-down resident versus top-down invasive.}
+This is the simplest scenario.
+When both communities are cohesive strictly from the top-down
+($c_{13}=0$ and $c_{35}=0$),
+coalescence outcomes are determined by which dominant is a better consumer
+of the primary resource.
+The most competitive dominant excludes the least competitive one
+($\mathrm{s}_1$ excludes $\mathrm{s}_3$ if $c_{11}>c_{31}$ and vice-versa,
+\figref[b]{fig4})
+and,
+because the cross-feeding networks of the invasive and resident
+communities are non-overlapping,
+it co-selects its sub-dominant exclusively.
+The relative similarity between the invasive and the coalesced communities
+($Q$, see \methodsref{metrics})
+can only be 0 (if the resident dominant ``wins'' competition for
+$\mathrm{R}_1$) or 1 (if the opposite happens).
+The invasion success of the invasive dominant ($\mathrm{s}_3$) is
+unaffected by the presence or absence of its sub-dominant ($\mathrm{s}_4$)
+since it is not being cross-fed by it.
+Thus, again $\mathrm{s}_3$ is only able to invade alone when it
+outcompetes the resident dominant ($\mathrm{s}_1$) for the primary resource ($c_{11}>c_{31}$).
+
+\paragraph*{Bidirectional resident versus top-down invasive.}
+Blablabla...
+
+\paragraph*{Top-down resident versus bidirectional invasive.}
+Blablabla...
+
+\paragraph*{Bidirectional resident versus bidirectional invasive.}
+Blablabla...
 
@@
 
+\subsubsection*{Minimal model}\label{methods:sim-min}
+
 \section*{Data \& code availability}\label{datacode}
@@
 \textbf{b.}~Communities were stabilized in serial batch culture bioreactors
-\cite{Goldford2018} in minimal synthetic media with glutamine or citrate as the
+in minimal synthetic media with glutamine or citrate as the
 only supplied carbon source.
@@
 sub-dominant species as the ``cohorts".
-\textbf{d.}~Rank-frequency distributions of all eight communities stabilized in either
+\textbf{d.}~Rank-frequency distributions of the eight communities stabilized in either
 glutamine (red) or citrate (blue), sequenced at a depth of $10^{-4}$ reads.
@@
 or they could have no effect in the invasion success of the dominant taxa (gray).
-\textbf{b.} Simulations with a Microbial Consumer-Resource Model suggest that positive bottom-up
+\textbf{b.}
+\iffalse
+Simulations with a Microbial Consumer-Resource Model suggest that positive bottom-up
 co-selection is common and can be very strong, whereas negative bottom-up co-selection is rare.
-We plot the frequency reached by the invasive dominants when invading the resident communities
+\fi
+Simulations with a Microbial Consumer-Resource Model:
+we plot the frequency reached by the invasive dominants when invading the resident communities
 in isolation versus the same frequency when invading together with their cohorts,
@@
 result of the dominant head-to-head pairwise competition versus the outcome of community coalescence.
+Left panel: strong positive bottom-up co-selection
+($R^2=0.00$, $p>0.05$).
+Right panel: no bottom-up co-selection
+($R^2=0.34$, $p<10^{-6}$).
+\iffalse
 Left panel: when positive bottom-up co-selection is strong, head-to-head pairwise competition of dominants
@@
 ($R^2=0.34$, $p<10^{-6}$).
+\fi
 \textbf{d.} Experiments show that in our conditions, positive bottom-up co-selection is indeed more
 frequent and strong than negative bottom-up co-selection.
-\textbf{e.} We reproduce the plots in panel c for our experimental data. This is equivalent to recreating
-\figref[b]{fig2}, this time splitting our data by the strength of bottom-up co-selection
-instead of by the carbon source provided to the communities.
-Like simulations predicted, instances of community coalescence displaying strong positive bottom-up co-selection
-yield a poor correspondence between dominant pairwise competition and coalescence outcomes
-(left panel, $R^2=0.07$, $p>0.05$)
-and vice-versa
-(right panel, $R^2=0.37$, $p<10^{-4}$).}
+\textbf{e.} We reproduce the plots in panel c for our experimental data,
+i.e. we recreate
+\figref[b]{fig2} but this time splitting our data by the strength of bottom-up co-selection
+instead of by the carbon source supplied to the communities.
+Left panel: strong positive bottom-up co-selection
+($R^2=0.07$, $p>0.05$).
+Right panel: no bottom-up co-selection
+($R^2=0.37$, $p<10^{-4}$).}
 \label{fig3}
@@
 \caption{\textbf{A minimal model of community coalescence.}
-\textbf{a.} Blablabla...
-\textbf{b.} Blablabla...}
+\textbf{a.} Illustration of the model structure and parameters.
+The primary resource ($\mathrm{R}_1$) is replenished after
+each growth-dilution cycle (red arrows).
+Solid arrows indicate resource consumption, dashed arrows represent
+resource secretion.
+\textbf{b-e.} Coalescence outcomes in the minimal model
+under different relations of cohesiveness between the
+resident and the invasive communities.}
 \label{fig4}

### 2021-05-25 coalescence.py

@@ -36,16 +36,18 @@ if os.name == 'nt':
 else:
     par = True
 
+'''
 # set random and numpy's seed for reproducibility
-myseed = 9999  # 500 works pretty nicely
+myseed = 1000000 # 99
 np.random.seed(myseed)
 random.seed(myseed)
+'''
 
 import time
 start_time = time.time()
 
 # get community hierarchies? (this generates a complete data set but takes time)
-getHierarchies = False
+getHierarchies = True
 
 
 
@@ -629,10 +631,70 @@ def cohort_vs_alone():
     ax.set_yticks([0,0.5,1])
     
     fig
+    
+# Q vs fraction in pairwise competition, split by strength of bottom-up co-selection
+def q_vs_fraction_split():
+    
+    x = f_pairwise
+    y = stats['bray_curtis']  
+    
+    ### remove nan elements (this happens if pairwise competition ends up in extinction of both, need to investigate how this happens)
+    ### SOLVED: this happens when none of the two competing species are able to grow over the dilution factor during the propagation time.
+    ### It is an unusual case but it is possible and it does not imply code malfunction.
+    n = [ni for ni in range(len(x)) if ~np.isnan(x[ni]) and ~np.isnan(y[ni])]
+    x = [x[i] for i in n]
+    y = [y[i] for i in n]
+    
+    # split data
+    x_buc = [x[i] for i in range(len(x)) if f_singleinv[i]<0.1 and f_coalescence[i]>0.05]
+    y_buc = [y[i] for i in range(len(x)) if f_singleinv[i]<0.1 and f_coalescence[i]>0.05]
+    
+    x_nobuc = [x[i] for i in range(len(x)) if f_singleinv[i]>0.1 or f_coalescence[i]<0.05]
+    y_nobuc = [y[i] for i in range(len(x)) if f_singleinv[i]>0.1 or f_coalescence[i]<0.05]
+    
+    # start plot
+    fig, (ax1,ax2) = plt.subplots(1, 2)
+    
+    # positive bottom-up co-selection
+    ax1.scatter(x_buc,y_buc,edgecolors="black",facecolors="none")
+    z = np.polyfit(x_buc, y_buc, 1)
+    p = np.poly1d(z)
[…]

### 2021-05-26 draft/draft_v3.tex
draft almost done

@@
 
+\makeatother
+
 % title, authors, affiliations
@@
 that can again
-be metabolized by the respective dominants.
+be metabolized by the corresponding dominants.
 This is the simplest structure where both top-down and bottom-up
@@
 by how effectively the species can utilize each other's metabolic byproducts.
-Several parameters can modulate the strength and direction of community cohesiveness,
+Several parameters can modulate the intensity and direction of community cohesiveness,
 and even in this simple model there can be complex interactions between them.
@@
 which facilitates the interpretation of the model's outcomes.
+This means that
+the relative similarity between the invasive and the coalesced communities
+($Q$, see \methodsref{metrics})
+can only be 0 (if the resident community dominates in coalescence)
+or 1 (if the opposite happens).
 
@@
 experiments and simulations of community coalescence,
-we first stabilized each community separately
+we first stabilized each minimal community separately
 through serial growth-dilution cycles,
@@
 We ran simulations for both community-community competitions and
-invasions from the invasive dominant alone.
+invasions from the dominant alone.
 We considered four different scenarios illustrated in \figref[b-e]{fig4}:
@@
 
-\paragraph*{Top-down resident versus top-down invasive.}
+\textbf{Top-down resident versus top-down invasive.}
 This is the simplest scenario.
@@
 it co-selects its sub-dominant exclusively.
-The relative similarity between the invasive and the coalesced communities
-($Q$, see \methodsref{metrics})
-can only be 0 (if the resident dominant ``wins'' competition for
-$\mathrm{R}_1$) or 1 (if the opposite happens).
 The invasion success of the invasive dominant ($\mathrm{s}_3$) is
@@
 Thus, again $\mathrm{s}_3$ is only able to invade alone when it
-outcompetes the resident dominant ($\mathrm{s}_1$) for the primary resource ($c_{11}>c_{31}$).
-
-\paragraph*{Bidirectional resident versus top-down invasive.}
-Blablabla...
-
-\paragraph*{Top-down resident versus bidirectional invasive.}
-Blablabla...
-
-\paragraph*{Bidirectional resident versus bidirectional invasive.}
-Blablabla...
-
-\subsubsection*{Community hierarchy regulates the strength of bottom-up co-selction}
+outcompetes $\mathrm{s}_1$ for the primary resource ($c_{11}>c_{31}$).
+
+\textbf{Bidirectional resident versus top-down invasive.}
+When
+the resident community is cohesive both top-down and bottom-up ($c_{13}>0$)
+and the invasive community is only cohesive from the top-down ($c_{35}=0$),
+coalescence outcomes
+are determined by the strength of bottom-up cohesiveness in the resident community,
+i.e. the magnitude of $c_{13}$ in the minimal model
+(\figref[c]{fig4}).
+If $c_{13}$ is small (weak bottom-up cohesiveness), the system behaves as in the previous scenario,
+virtually as if both communities were only cohesive from the top-down.
+However, as $c_{13}$ increases (strong bottom-up cohesiveness) the resident community becomes more
+resistant to invasion.
+For large values of $c_{13}$, the invasive dominant being more competitive than the resident
+dominant for the primary resource does not ensure invasion success.
+Since the latter can also be cross-fed by its sub-dominant (through $\mathrm{R}_3$),
+it does not rely on the primary resource only ($\mathrm{R}_1$) to outcompete the invader.
+The same is true when the invasive dominant invades alone.
+
+\textbf{Top-down resident versus bidirectional invasive.}
+More outcomes become possible if the resident community is cohesive from the top-down only ($c_{13}=0$)
+and the invasive one is cohesive in both directions ($c_{35}>0$).
+Like in the first scenario we considered, the invasive dominant will only be successful by itself if it is a better
+competitor for $\mathrm{R}_1$ than the resident dominant ($c_{31}>c_{11}$).
+However, if the invasive sub-dominant is also present, the invasion can be successful even if $c_{31}<c_{11}$
+as long as the cross-feeding from $\mathrm{s}_4$ (invasive sub-dominant) to $\mathrm{s}_3$ (invasive dominant)
+is strong enough, i.e. if $c_{35}$ sufficiently large
+(\figref[d]{fig4}).
+In this scenario, we can see cases where the invasive dominant cannot invade in isolation but
+the invasive community still dominates during coalescence,
+similar to what we observed in our experiments and initial simulations.
+
+\textbf{Bidirectional resident versus bidirectional invasive.}
+If both communities are cohesive from the top-down and from the bottom-up ($c_{13}>0$ and $c_{35}>0$),
+coalescence outcomes are generically determined by the relations of strength between
+all forms of cohesiveness.
+In limit cases we can recover scenarios considered previously, e.g. if $c_{13} >> c_{35}$ the system
+behaves as if the resident community was cohesive from the bottom-up but the invasive community was not.
+Situations where the invasive dominant can invade when accompanied by its sub-dominant despite
+being unable to do so by itself are again possible
+if the invasive community is more strongly cohesive from the bottom-up than the resident community
+(\figref[e]{fig4}).
+
+This minimal coalescence model shows that
+bottom-up co-selection of a dominant that is unable to invade by itself
+is only possible if said dominant is strongly cross-fed by its cohort.
+The model also demonstrates that this behavior can only happen
+if the byproducts of the cohorts are to some extent non-overlapping
+and preferentially utilized by the respective dominants.
+Otherwise, if both species can utilize all secretions, bottom-up
+co-selection becomes unspecific with respect to the dominant identity.
+
+\subsubsection*{Community hierarchy regulates the strength of bottom-up co-selection}
 
@@
 
+Our minimal model is set within the same MicroCRM framework that we used
+for the previous simulations.
+As described in the main text, the model contains two communities
+of two species each
+($\mathrm{s}_1$ to $\mathrm{s}_4$),
+with five resources in total, out of which the first one ($\mathrm{R}_1$)
+is replenished externally at the beginning of each growth cycle
+and the rest correspond to the species' metabolic byproducts.
+Each species secretes a unique byproduct, meaning that the metabolic
+matrix $\mathbf{D}$ is binary in this case.
+The specific structure of $\mathbf{D}$ is displayed below
+--because it is a 3-dimensional matrix in our framework,
+we have ``sliced'' it into the five 2-dimensional matrices corresponding to
+our five species.
+
+\begin{equation*}
+\mathbf{D}_1 = \; \; \; \; \; \begin{matrix}
+ &
+
+\begin{matrix}
+\text{\scriptsize $\mathrm{R}_1$} \!\!\! & 
+\text{\scriptsize $\mathrm{R}_2$} \!\!\! & 
+\text{\scriptsize $\mathrm{R}_3$} \!\!\! & 
+\text{\scriptsize $\mathrm{R}_4$} \!\!\! & 
+\text{\scriptsize $\mathrm{R}_5$}
+\end{matrix}
+
+\\
+
+\begin{matrix*}[l]
+\text{\scriptsize $\mathrm{R}_1$} \!\!\! \\
+\text{\scriptsize $\mathrm{R}_2$} \!\!\! \\
+\text{\scriptsize $\mathrm{R}_3$} \!\!\! \\
+\text{\scriptsize $\mathrm{R}_4$} \!\!\! \\
+\text{\scriptsize $\mathrm{R}_5$} \!\!\!
+\end{matrix*}
+
+&
+
+\begin{pmatrix}
+0 & * & 0 & * & * \\
+1 & * & 0 & * & * \\
+0 & * & 1 & * & * \\
+0 & * & 0 & * & * \\
+0 & * & 0 & * & *
+\end{pmatrix}
+
+\end{matrix}
+\end{equation*}
+
+\bigskip
+
+\begin{equation*}
+\mathbf{D}_2 = \; \; \; \; \; \begin{matrix}
+ &
+
+\begin{matrix}
+\text{\scriptsize $\mathrm{R}_1$} \!\!\! & 
+\text{\scriptsize $\mathrm{R}_2$} \!\!\! & 
+\text{\scriptsize $\mathrm{R}_3$} \!\!\! & 
+\text{\scriptsize $\mathrm{R}_4$} \!\!\! & 
+\text{\scriptsize $\mathrm{R}_5$}
+\end{matrix}
+
+\\
+
+\begin{matrix*}[l]
+\text{\scriptsize $\mathrm{R}_1$} \!\!\! \\
+\text{\scriptsize $\mathrm{R}_2$} \!\!\! \\
+\text{\scriptsize $\mathrm{R}_3$} \!\!\! \\
+\text{\scriptsize $\mathrm{R}_4$} \!\!\! \\
+\text{\scriptsize $\mathrm{R}_5$} \!\!\!
+\end{matrix*}
+
+&
+
+\begin{pmatrix}
+* & 0 & * & * & * \\
+* & 0 & * & * & * \\
+* & 1 & * & * & * \\
+* & 0 & * & * & * \\
+* & 0 & * & * & *
+\end{pmatrix}
+
+\end{matrix}
+\end{equation*}
+
+\bigskip
+
+\begin{equation*}
+\mathbf{D}_3 = \; \; \; \; \; \begin{matrix}
+ &
+
+\begin{matrix}
+\text{\scriptsize $\mathrm{R}_1$} \!\!\! & 
+\text{\scriptsize $\mathrm{R}_2$} \!\!\! & 
+\text{\scriptsize $\mathrm{R}_3$} \!\!\! & 
+\text{\scriptsize $\mathrm{R}_4$} \!\!\! & 
+\text{\scriptsize $\mathrm{R}_5$}
+\end{matrix}
+
+\\
+
+\begin{matrix*}[l]
+\text{\scriptsize $\mathrm{R}_1$} \!\!\! \\
+\text{\scriptsize $\mathrm{R}_2$} \!\!\! \\
+\text{\scriptsize $\mathrm{R}_3$} \!\!\! \\
+\text{\scriptsize $\mathrm{R}_4$} \!\!\! \\
+\text{\scriptsize $\mathrm{R}_5$} \!\!\!
+\end{matrix*}
+
+&
+
+\begin{pmatrix}
+0 & * & * & * & 0 \\
+0 & * & * & * & 0 \\
+0 & * & * & * & 0 \\
+1 & * & * & * & 0 \\
+0 & * & * & * & 1
+\end{pmatrix}
+
+\end{matrix}
+\end{equation*}
+
+\bigskip
+
+\begin{equation*}
+\mathbf{D}_4 = \; \; \; \; \; \begin{matrix}
+ &
+
+\begin{matrix}
+\text{\scriptsize $\mathrm{R}_1$} \!\!\! & 
+\text{\scriptsize $\mathrm{R}_2$} \!\!\! & 
+\text{\scriptsize $\mathrm{R}_3$} \!\!\! & 
+\text{\scriptsize $\mathrm{R}_4$} \!\!\! & 
+\text{\scriptsize $\mathrm{R}_5$}
+\end{matrix}
+
+\\
+
+\begin{matrix*}[l]
+\text{\scriptsize $\mathrm{R}_1$} \!\!\! \\
+\text{\scriptsize $\mathrm{R}_2$} \!\!\! \\
+\text{\scriptsize $\mathrm{R}_3$} \!\!\! \\
+\text{\scriptsize $\mathrm{R}_4$} \!\!\! \\
+\text{\scriptsize $\mathrm{R}_5$} \!\!\!
+\end{matrix*}
+
+&
+
+\begin{pmatrix}
+* & * & * & 0 & * \\
+* & * & * & 0 & * \\
+* & * & * & 0 & * \\
+* & * & * & 0 & * \\
+* & * & * & 1 & *
+\end{pmatrix}
+
+\end{matrix}
+\end{equation*}
+%
+Asterisks indicate values of $D_{i\beta\alpha}$ that are irrelevant because
+species $i$ cannot utilize resource $\alpha$ (and so the metabolic flux from
+$\alpha$ to $\beta$ corresponding to that species will always be zero
+regardless of the value of $D_{i\alpha\beta}$).
+The consumer preference matrix $\mathbf{c}$ takes the following form:
+
+\begin{equation*}
+\mathbf{c} = \; \; \; \; \; \begin{matrix}
+ &
+
+\begin{matrix}
+\text{\scriptsize $\mathrm{R}_1$} \;\;\; & 
+\text{\scriptsize $\mathrm{R}_2$} \;\;\; & 
+\text{\scriptsize $\mathrm{R}_3$} \;\;\; & 
+\text{\scriptsize $\mathrm{R}_4$} \;\;\; & 
+\text{\scriptsize $\mathrm{R}_5$}
+\end{matrix}
+
+\\
+
+\begin{matrix*}[l]
+\text{\scriptsize $\mathrm{s}_1$} \!\!\! \\
+\text{\scriptsize $\mathrm{s}_2$} \!\!\! \\
+\text{\scriptsize $\mathrm{s}_3$} \!\!\! \\
+\text{\scriptsize $\mathrm{s}_4$} \!\!\!
+\end{matrix*}
+
+&
+
+\begin{pmatrix}
+c_{11} & 0 & c_{13} & 0 & 0 \\
+0 & 100 & 0 & 0 & 0 \\
+c_{31} & 0 & 0 & 0 & c_{35} \\
+0 & 0 & 0 & 100 & 0
+\end{pmatrix}
+
+\end{matrix}
+\end{equation*}
+%
+Where we made the sub-dominants equally strong consumers of their dominants' secretions
+($c_{22} = c_{44}$ = 100),
+and we varied all other uptake rates depending on the scenario we were considering (see main text).
+Whenever we were interested in the ratio between two rates (e.g. $c_{35}/c_{13}$ in \figref[e]{fig4})
+we gave the one in the denominator a fixed value of 1 and let one in the numerator range
+within the specified limits.
+
 \section*{Data \& code availability}\label{datacode}
@@
 
-\clearpage
-
 % references
@@
 \bibliography{refs}
-
 \clearpage
@@
 \internallinenumbers
-\includegraphics[scale=0.7,keepaspectratio]{figs/fig4_v2.pdf}
+\includegraphics[scale=0.7,keepaspectratio]{figs/fig4_v3.pdf}
 \caption{\textbf{A minimal model of community coalescence.}

### 2021-05-27 draft/draft_v3.tex
draft v3 done, run through Alvaro

@@
 % title, authors, affiliations
-\title{Top-down versus bottom-up cohesiveness in microbial community coalescence}
+\title{Top-down and bottom-up cohesiveness\\in microbial community coalescence}
 \author[1 *]{Juan Diaz-Colunga}
@@
 e.g. if the primary resource is supplied continuously instead of at the beggining
-of each growth cycle \cite{Marsland2019,Marsland2020},
-but they are a good approximation for the community dynamics between consecutive
+of each growth cycle \cite{Marsland2019,Marsland2020}.
+They represent a good approximation for the community dynamics between consecutive
 serial dilutions in our setup.
@@
 How do the ideas above scale to more complex and diverse communities?
-
-We reason that the choice of species-specific metabolic architectures is necessary to potentially generate
-cohesiveness at the community level during coalescence.
+In natural microbiomes and in our laboratory cultures,
+a large number of species can coexist and cross-feed each other,
+giving rise to facilitation networks that are far more dense
+than the ones in our minimal model.
+However, we argue that some conditions
+for the emergence of bottom-up ecological co-selection still apply:
+\textit{a)} the byproducts of the invasive cohort should be utilized preferentially
+by the invasive dominant (and thus said byproducts need to be species-specific
+to minimize overlap with the resident cohort secretions),
+and \textit{b)} the invasive dominant should be strongly cross-fed by its cohort.
+
+We reason that the choice of species-specific metabolic architectures is necessary
+to potentially generate bottom-up cohesiveness at the community level during coalescence.
 If the secretions of all species were identical (or only slightly different),
@@
 the metabolic flows through the rest of the community's cross-feeding network.
-On the other hand, said network could undergo a profound and further-reaching restructuring
+On the other hand, said network could undergo a deeper and further-reaching restructuring
 if the invasive species secreted very different sets of metabolites with respect to the resident ones,
@@
 For a similar reason, we argue that the sparsity of the metabolic matrix could also modulate
-the emergence of cohesiveness in the face of coalescence.
+the emergence of bottom-up cohesiveness in the face of coalescence.
 A dense metabolic matrix corresponds to a situation where all species secrete a wide
@@
 New-coming invasive species that secrete similar byproducts as resident ones
---even if they do so in different relative amounts--
+(even if they do so in different relative amounts)
 might only induce moderate quantitative changes in the metabolic fluxes.
@@
 \cite{Rosenzweig1994,Goldford2018,Estrela2020}.
+In fact, when we use the MicroCRM to run simulations of community coalescence
+but we use a dense metabolic matrix or one that is not species-specific
+($D_{i\beta\alpha}=D_{j\beta\alpha}$ for all $i$, $j$, $\alpha$ and $\beta$,
+see \hyperref[box1]{Box~1})
+we observe virtually no instances of bottom-up ecological co-selection
+(\figref{figS4}).
+
+To quantify the degree of cross-feeding from the cohorts towards the dominant,
+we defined a community hierarchy metric $h$ as
+
+\begin{equation}
+h = \frac{\Delta N_{\mathrm{dom}}^{\mathrm{R1}}}{\Delta N_{\mathrm{dom}}}
+\label{eq:h}
+\end{equation}
+%
+where $\Delta N_{\mathrm{dom}}$ represents the overall increase in dominant biomass
+between consecutive dilutions in a generationally stable community,
+and $\Delta N_{\mathrm{dom}}^{\mathrm{R1}}$
+represents the increase in said biomass resulting from the metabolism of the primary resource
+($\mathrm{R}_1$) only.
+If the dominant was just utilizing the primary resource, the community would be very
+hierarchical ($h \sim 1$),
+whereas if it was growing mostly on the secretions of other taxa,
+it would be more distributed ($h \sim 0$).
+We quantified the hierarchies of the resident and invasive communities in our MicroCRM,
+finding that $h$ follows a bimodal distribution (\figref[a]{fig5}).
+We therefore divided our simulations into four groups according to the relation
+of hierarchy between the resident and invasive communities (\figref[b]{fig5}).
+For each group, we evaluated the recurrence of bottom-up co-selection,
+i.e. the fraction of cases where a dominant that could not invade in isolation
+was successful when accompanied by its cohort
+(green area of \figref[b]{fig3}).
+We found that bottom-up ecological co-selection
+was significantly more frequent when the invasive community was
+non-hierarchical (\figref[c]{fig5}),
+in line with what the minimal model anticipated
+(\figref[d-e]{fig4}).
 
@@
 in the metabolic fluxes across species.
-The model also predicts a trade-off between the strength of bottom-up co-selection
+The model predicts a trade-off between the strength of bottom-up co-selection
 and the ability of dominant-dominant pairwise competition to dictate coalescence outcomes,
 which we have confirmed experimentally.
+It also suggests that rarer taxa may play a more prominent role in co-selecting
+dominant species when the metabolic fluxes across the community are
+distributed rather than hierarchical.
 These observations,
@@
 \cite{Gilpin1994,Toquenaga1997,Tikhonov2016,Tikhonov2017,Vila2019,Lechon2021},
-suggest that collective interactions between microbes
-and the environment should be generically expected to produce ecological co-selection
+suggest that collective interactions of microbes with one another
+and with the environment should be generically expected to produce ecological co-selection
 during community coalescence.
@@
 and test quantitative theories about its role in microbiome assembly.
-On the other hand, coalescence of communities under different settings
+Accurately characterizing the metabolic architectures of the species in
+this type of laboratory cultures through metabolomics techniques
+will also be necessary to verify the relationship between community hierarchy
+and the direction of ecological co-selection.
+Furthermore, community coalescence under different settings
 (e.g. in spatially structured environments)
-might be governed by additional factors.
+might be governed by additional factors that we have not considered here.
 Understanding them and quantifying their relative contributions in natural communities
@@
 \internallinenumbers
-\includegraphics[scale=0.7,keepaspectratio]{figs/fig5.pdf}
-\caption{\textbf{Community hierarchies determine coalescence outcomes.}
-\textbf{a.} Blablabla...
-\textbf{b.} Blablabla...
-\textbf{b.} Blablabla... ($p=0.083$).}
+\includegraphics[scale=0.7,keepaspectratio]{figs/fig5_v2.pdf}
+\caption{\textbf{Community hierarchy modulates the recurrence of bottom-up co-selection.}
+\textbf{a.} Distribution of community hierarchies for our \textit{in silico} communities.
+\textbf{b.} We divided our coalescence simulations into four groups according to the
+hierarchies of the resident ($h_\mathrm{R}$) and invasive ($h_\mathrm{I}$) communities
+as indicated by the dashed boxes.
+For every group, we calculated the fraction of cases where bottom-up co-selection was observed,
+i.e. the invasive dominant was unsuccessful when invading in isolation
+but successful when invading with its cohort.
+\textbf{c.} Bottom-up co-selection of the invasive dominant during coalescence
+is significantly more frequent when the invasive community is non-hierarchical.
+Error bars representing 95\% confidence intervals
+and p-values were computed by bootstrapping
+($p<10^{-4}$ where indicated).}
 \label{fig5}
@@
 
+\begin{figure}[!h]
+\centering
+\internallinenumbers
+\includegraphics[scale=0.9,keepaspectratio]{figs/figS4.pdf}
+\caption{\textbf{Bottom-up ecological co-selection is not observed
+when species have similar metabolic architectures.}
+We ran simulations of community coalescence following the same
+procedure described in the main text, but this time we
+used a dense metabolic matrix
+(left panel,
+sparsity = 0.05 in the Community Simulator package \cite{Marsland2020})
+or a species-unspecific metabolic matrix
+(right panel,
+$D_{i\beta\alpha}=D_{j\beta\alpha}$ for all $i$, $j$, $\alpha$ and $\beta$,
+see \hyperref[box1]{Box~1}).
+Virtually no bottom-up co-selection is observed in either case.}
+\label{figS4}
+\end{figure}
+
+\clearpage
+
 \end{document}

### 2021-05-27 coalescence.py

@@ -47,8 +47,7 @@ import time
 start_time = time.time()
 
 # get community hierarchies? (this generates a complete data set but takes time)
-getHierarchies = True
-
+getHierarchies = False
 
 
 # %%

### 2021-05-27 data/analyzeData.R

@@ -25,7 +25,7 @@ library(RColorBrewer)
 myplots <- vector(mode='list')
 if(!dir.exists('plots')) dir.create('plots')
 display_plots <- TRUE
-save_plots <- TRUE
+save_plots <- FALSE
 
 # general options
 deconvolute_ESVs <- FALSE # should species abundance be estimated from ESV abundance?
@@ -1590,13 +1590,6 @@ plot_this$bucs <- plot_this$f_singleinv < 0.1 & plot_this$f_coalescence > 0.05
 plot_this$bucs <- c('neutral','strong')[1 + plot_this$bucs]
 plot_this$bucs <- factor(plot_this$bucs,levels=c('strong','neutral'))
 
-# difference in hierarchies
-plot_this$h_diff = log10(plot_this$h_res/plot_this$h_inv)
-
-# hierarchy distribution
-h_distrib = data.frame(h=hist(c(plot_this$h_res,plot_this$h_inv),breaks=20)$mids,
-                       freq=hist(c(plot_this$h_res,plot_this$h_inv),breaks=20)$counts/(2*nrow(plot_this)))
-
 # make plots
 myplots[['q-vs-pairwise_bray-curtis_simul']] <-
   ggplot(data=plot_this[!is.na(plot_this$f_pairwise),],
@@ -1738,6 +1731,52 @@ myplots[['q-vs-pairwise_bray-curtis_simul_top-down-vs-bottom-up_neutral-bucs']] 
         panel.border=element_rect(size=0.25)) +
   coord_fixed()
 
+
+### ----------------------------------------------------------------------
+### SIMULATIONS DATA: HIERARCHY
+### ----------------------------------------------------------------------
+
+# difference in hierarchies
+plot_this$h_diff = log10(plot_this$h_res/plot_this$h_inv)
+
+# hierarchy distribution
+h_distrib = data.frame(h=hist(c(plot_this$h_res,plot_this$h_inv),breaks=20)$mids,
+                       freq=hist(c(plot_this$h_res,plot_this$h_inv),breaks=20)$counts/(2*nrow(plot_this)))
+
+# hierarchy groups
+n <- vector(mode='list',length=4)
+names(n) <- c('low_low','low_high','high_low','high_high')
+n[[1]] <- which(plot_this$h_res<0.25 & plot_this$h_inv<0.25 & !is.na(plot_this$f_singleinv) & !is.na(plot_this$f_coalescence))
[…]

### 2021-05-27 draft/draft_v3.tex
typos

@@
 --because it is a 3-dimensional matrix in our framework,
-we have ``sliced'' it into the five 2-dimensional matrices corresponding to
-our five species.
+we have ``sliced'' it into the four 2-dimensional matrices corresponding to
+our four species.
 

### 2021-05-28 data/analyzeData.R
final touches and README

@@ -25,7 +25,7 @@ library(RColorBrewer)
 myplots <- vector(mode='list')
 if(!dir.exists('plots')) dir.create('plots')
 display_plots <- TRUE
-save_plots <- FALSE
+save_plots <- TRUE
 
 # general options
 deconvolute_ESVs <- FALSE # should species abundance be estimated from ESV abundance?

### 2021-05-28 data/analyzeData_v2.R

@@ -1,2065 +1,0 @@
-### Juan Diaz-Colunga
-### Feb 16 2021
-
-### ----------------------------------------------------------------------
-### This script performs all the analyses and produces all the plots
-### in the paper.
-###    1. Community composition (pie plots & rank plots of ESV frequency)
-###    2. Characterization of isolated species from each community
-###    3. Matching isolated species to dominants based on sequencing data
-###    4. Fractions of isolated species in pairwise competition
-###    5. Similarity of coalesced, resident and invasive communities
-### ----------------------------------------------------------------------
-
-rm(list=ls())
-source('isolates_abundance_from_ESVs.R')
-source('myFunctions.R')
-library(matlib)
-library(testthat)
-library(ggplot2)
-library(ggh4x)
-library(tidyr)
-library(RColorBrewer)
-
-# initialize variables to store plots
-myplots <- vector(mode='list')
-if(!dir.exists('plots')) dir.create('plots')
-display_plots <- TRUE
-save_plots <- TRUE
-
-# general options
-deconvolute_ESVs <- FALSE # should species abundance be estimated from ESV abundance?
-
-
-### ----------------------------------------------------------------------
-### LOAD AND FORMAT DATA
-### ----------------------------------------------------------------------
-
-# load OTU table, taxonomy table and metadata
-otus <- read.csv('OTU_table.csv',stringsAsFactors=F)
-taxa <- read.csv('taxonomy_silva.csv',stringsAsFactors=F)
-metadata <- read.csv('metaData.csv',stringsAsFactors=F)
-
-### merge similar taxa?
-# merged <- mergeTaxa(otus,taxa)
-# otus <- merged[['otus']]
-# taxa <- merged[['taxa']]
-
-# check that OTU table and taxonomy table follow the same order
-stopifnot(all(otus$X==taxa$X))
-
-# keep only relevant metadata
-metadata <- metadata[,c("Sample","Plate","Row","Column","Experiment",
-                        "Comm1","StabilizingCarb1","Comm2","StabilizingCarb2",
-                        "Carbon","Rep","Transfer")]
-
-# assign IDs to each sequence
-otus <- cbind(seq_id = paste('seq_',1:nrow(otus),sep=''),
-              otus)
-
-# normalize OTU abundances (so entries in table represent fractions)
-otus[,3:ncol(otus)] <- otus[,3:ncol(otus)]/
[…]

### 2021-06-02 draft/draft_v4.tex
draft v4 added, need to polish a few details with Alvaro

--- 2021-05-27 draft/draft_v3.tex
@@
 \usepackage{nameref} % cross-references between draft and supplementary material
-\usepackage{zref-xr,zref-user} % more cross-references formatting
-\zxrsetup{toltxlabel}
-\zexternaldocument*{supp}
 \usepackage[square,numbers,sort&compress]{natbib}
@@
 \renewcommand\Affilfont{\itshape\small} % style of affiliations text
-\makeatletter\renewcommand{\@biblabel}[1]{#1.}\makeatother % change [number] for number. in reference list
+\makeatletter\renewcommand{\@biblabel}[1]{#1.}\makeatother % change [number] for number. in references
 \addto{\captionsenglish}{\renewcommand{\bibname}{References}}
@@
 \newcommand{\Burl}[1]{\url{#1}}
-\zxrsetup{toltxlabel} % cross references: draft & suppl mat
-\zexternaldocument*[M-]{draft}
 \hypersetup{
@@
    labelsep=period, name={Figure}} % custom figure captions
-\captionsetup[table]{font=small, labelfont={bf}, labelformat={default},
-   labelsep=period, name={Table}} % custom table captions
 \hypersetup{citecolor={blue}}
@@
 
-\affil[$\textrm{\Letter}$]{\normalfont~alvaro.sanchez@yale.edu}
-\affil[*]{\normalfont~These authors contributed equally}
+\affil[$\textrm{\Letter}$]{\normalfont alvaro.sanchez@yale.edu}
+\affil[*]{\normalfont These authors contributed equally}
 \date{}
@@
 This has been observed, for instance, 
-in river courses where terrestrial microbes mix with aquatic microorganisms
+in river courses where terrestrial microbes mix with aquatic microorganisms,
 \cite{Mansour2018,Luo2020,Vass2021}
@@
 Gut microbiomes can invade external communities
-through the host animal secretions \cite{Dutton2021},
+through the host's secretions \cite{Dutton2021},
 and the skin microbiota is also subject to
-invasions when making contact with environmental sources of microbes \cite{Vandegrift2019}.
+invasions when it makes contact with environmental sources of microbes \cite{Vandegrift2019}.
 
@@
 and
-the role played by the less abundant species has long been subject to debate \cite{Winfree2015}.
+the role played by the less abundant community members has long been subject to debate \cite{Winfree2015}.
 Laboratory cultures have also been found to contain uneven distributions of multiple strains that feed
 off the metabolic secretions of the dominant species \cite{Rosenzweig1994,Goldford2018}.
-The fate of these sub-dominant taxa may be dependent on the invasional success of their dominant
+The fate of these sub-dominant taxa may be dependent on the invasion success of their dominant
 species, or, alternatively, the dominant itself may owe its dominance (at least in part) to cross-feeding
 or other forms of facilitation from the rarer members of the population.
-These scenarios would give rise to ``top-down'' or ``bottom-up'' community cohesiveness, respectively.
+We refer to these two opposite scenarios as
+``top-down'' or ``bottom-up'' forms of community cohesiveness, respectively.
+Top-down cohesiveness emerges when the dominant invader co-selects other sub-dominant taxa
+into the final community during coalescence.
+Alternatively, bottom-up cohesiveness refers to the case when the dominant is co-selected by
+the more rare members of its community.
 Either of these forms of co-selection could, in principle, be positive (recruitment) or negative (antagonism),
@@
 with the dominant (most abundant) species typically comprising most of the biomass ($\text{median}=46\%$,
-\figref{figS1}). Because these communities are easy to manipulate and grow in high throughput, and
-are largely made up by culturable members, they represent good test cases to investigate ecological
+\figref{figS1}). Because these communities are easy to manipulate and grow in high throughput,
+they represent good test cases to investigate ecological
 co-selection during community coalescence. Here we focus on the dominants and ask whether they can
@@
 
-Our results indicate that co-selection varies in direction and strength depending on the supplied limiting
-resource.
-This primary resource, in turn, has been shown to
-shape the structure and interactions of the communities \cite{Estrela2020}.
-We observe that, when top-down co-selection is weak, bottom-up co-selection can be very strong,
+Our results indicate that
+when top-down co-selection is weak, bottom-up co-selection can be very strong,
 with positive co-selection being far more common than negative co-selection.
-\iffalse
-Our results confirm that positive top-down co-selection is common, but its effects are weak.
-In contrast, bottom-up co-selection can be very strong, and positive co-selection is far more common
-than negative co-selection.
-\fi
 We then turn to a Microbial Consumer-Resource Model (MicroCRM)
@@
 (\figref[a]{fig1})
-and used them to inoculate our synthetic communities, which were stabilized 
-in serial batch-culture biorreactors for 84 generations in synthetic minimal media containing either
-glutamine or citrate as the only supplied carbon source
-(\figref[b]{fig1}, \methodsref{community-assembly}).
+and used them to inoculate
+eight identical habitats containing minimal media with either glutamine or citrate as
+the only supplied carbon source.
 We chose these two carbon sources because they are metabolized through different
@@
 and we hypothesize that communities assembled in either resource will be supported by
-cross-feeding networks of distinct sets of metabolites \cite{Goldford2018,Estrela2020}
+cross-feeding networks of distinct sets of metabolites \cite{Goldford2018,Estrela2020},
 thus leading to potentially variable degrees of community cohesiveness
 and coalescence outcomes \cite{Tikhonov2016,Tikhonov2017,Castledine2020,Lechon2021}.
+After inoculation, all communities were serially passaged for 12 transfers (84 generations),
+with an incubation time of 48 hours and a dilution factor of 1:100.
+(\figref[b]{fig1}, \methodsref{community-assembly}).
+In previous work we have shown that under these conditions,
+12 transfers allow communities to approach a state of ``generational equilibrium'',
+where the community composition at the end of one batch incubation will be the same as in consecutive incubations.
 We isolated the dominant species of every community
@@
 
-If communities being coalesced were highly cohesive from the top-down,
-the dominant species would co-select the rarer members of its community during coalescence
+One form of cohesiveness may arise when the sub-dominant members of the community
+depend on the dominant species.
+This can occur, for instance, when the dominant provides resources (or stressors)
+that select for the sub-dominant taxa
 (\figref[e]{fig1}, left panels).
+If communities being coalesced are highly cohesive from the top-down,
+the fate of the sub-dominant community members will be tied to their dominant:
+if it gets excluded, they will be likely to fall with it, and if it is able to resist coalescence,
+they will be likely to follow suit.
 In this scenario, we would expect the outcome of community coalescence to be predicted by
 which of the two dominants is most competitive in pairwise competition.
-Analogously, competition between dominants should be affected only weakly by the presence or
-absence of the cohorts, that would play a passive role under these conditions.
+Likewise, competition between dominants should be affected only weakly by the presence or
+absence of sub-dominant species, which would play a passive role under these conditions.
 To test this hypothesis, we performed all pairwise competitions between dominant species
@@
 (\figref[f]{fig1}).
-\iffalse
-and we quantified coalescence outcomes using the similarity between the coalesced and the
-invasive communities
-(\methodsref{metrics}).
-\fi
 The frequencies of all species in both community-community and dominant-dominant competitions
@@
 
-We found that, for communities assembled in the glutamine environment,
-the relative frequency of a dominant against another in head-to-head pairwise competition
-is barely predictive of its relative frequency against that same other dominant when the
-cohorts are present too, i.e. during community coalescence
-(\figref[a]{fig2} red dots, $R^2=0.04$, $p>0.05$).
-This correlation is significantly higher for the citrate communities
- (\figref[a]{fig2} blue dots, $R^2=0.83$, $p<10^{-8}$).
-This suggests that, in the glutamine environments, head-to-head competition of dominants
-is heavily influenced by higher order effects introduced by the rare taxa of the communities.
-On the other hand, the cohorts seem to play a more passive role in the citrate environments.
 To test the effects of top-down co-selection at the community level,
@@
 and compared them to the outcomes of the pairwise competitions between dominants alone.
-We again noticed differences between glutamine and citrate communities:
-for the former,
+We noticed a difference between communities assembled in the glutamine and citrate environments:
+for the latter,
+the structure of the coalesced communities tends to be strongly dictated by the
+result of the dominant-dominant competition
+(\figref[a]{fig2} middle panel, $R^2=0.57$, $p<10^{-4}$).
+For the former,
 the pairwise competitive ability of an invasive dominant
 is only weakly predictive of the performance of the invasive community in coalescence
-(\figref[b]{fig2} left panel, $R^2=0.15$, $p<0.05$).
-For the latter,
-the structure of the coalesced communities tends to be more strongly dictated by the
-result of the dominant-dominant competition
-(\figref[b]{fig2} middle panel, $R^2=0.57$, $p<10^{-4}$).
+(\figref[a]{fig2} left panel, $R^2=0.15$, $p<0.05$).
 Alternative quantifications of community distance yield similar results,
@@
 (\figref{figS3}).
-
-Together, these observations suggest that the strength of top-down co-selection depends on
-the environment where communities are assembled and coalescence takes place.
-Communities stabilized with citrate as the primary supplied resource
+We then examined whether, as predicted by the top-down cohesiveness hypothesis,
+the cohorts would play a passive role on the competition between dominant species.
+We found that, for communities assembled in the citrate environments,
+the relative frequency of a dominant against another in head-to-head pairwise competition
+is highly predictive of its relative frequency against that same other dominant when the
+cohorts are present too, i.e. during community coalescence
+(\figref[b]{fig2} blue dots, $R^2=0.83$, $p<10^{-8}$).
+This is not the case for the glutamine communities
+(\figref[b]{fig2} red dots, $R^2=0.04$, $p>0.05$).
+This suggests that, in the glutamine environments, head-to-head competition of dominants
+is heavily influenced by interactions between those dominants and the rarer taxa of the communities.
+On the other hand, the cohorts seem to play a more passive role in the citrate environments.
+Together, these observations indicate that
+communities stabilized with citrate as the primary supplied resource
 display a strong degree of top-down cohesiveness,
@@
 communities assembled under our experimental conditions (natural microbiomes re-assembled into
-multispecies communities through serial growth-dilution cycles in synthetic minimal media
+multi-species communities through serial growth-dilution cycles in synthetic minimal media
 with a single carbon source)
@@
 for which the MicroCRM provides a good description.
-Based on previous experimental work
-\cite{Harcombe2014,Pinu2018,Estrela2020},
-we allowed different species in our model
-to secrete different sets of byproducts when metabolizing a same resource
-(\hyperref[box1]{Box~1}).
-\iffalse
-this choice is discussed in \hyperref[box1]{Box~1}.
-\fi
+We and others have previously found a strong concordance between the behavior of
+laboratory and natural microbial communities and the generic behavior of the MicroCRM
+\cite{Goldford2018,Marsland2019,Marsland2020,Marsland2020b,Estrela2021}.
 To reproduce our experimental protocol \textit{in silico},
@@
 (\methodsref{sim}).
-We found that the MicroCRM was able to capture a correlation between
+We found that the MicroCRM simulations naturally exhibit the observed correlation between
 the head-to-head pairwise competition of dominants
 and the outcome of community coalescence
-(\figref[b]{fig2}, right panel),
-further supporting the idea that top-down ecological co-selection can consistently
-emerge from metabolic interactions across species.
-
-\iffalse
-[Maybe this ``outro'' sentence is not even necessary]
-Our experiments had shown that this correlation varied
-across communities assembled in the glutamine and citrate environments
-(\figref[b]{fig2}, left and middle panels),
-which led us to investigate the determinants of the strength and direction
-of ecological co-selection in our model.
-
-\fi
+(\figref[a]{fig2}, right panel),
+further supporting the idea that top-down ecological co-selection consistently
+emerges from metabolic interactions across species.
+Moreover, we found that top-down co-selection is observed under a wide range of different simulation conditions
+and cross-feeding networks (\figref{figSX}), indicating that it is a robust phenomenon.
+
 \clearpage
@@
 (see illustration below).
-\iffalse
-For our simulations,
-the elements of $\mathbf{c}$ are sampled from a gamma distibution,
-while the elements of $\mathbf{D}$ are sampled from a Dirichlet distribution
-to ensure that the normalization condition $\sum_\beta D_{i\beta\alpha} = 1$
-is satisfied for every $i$ and $\alpha$.
-\fi
-\iffalse
-The sampling is performed so that species belonging to a same family
-tend to secrete more similar byproducts when consuming a same resource
-(see \methodsref{sim} and \cite{Marsland2019,Marsland2020}),
-as supported by experimental observations
-\cite{Harcombe2014,Pinu2018,Estrela2020}.
-\fi
 
@@
 
-\iffalse
-We found that, for communities assembled in the glutamine environment,
-the pairwise competitive ability of an invasive dominant
-is only weakly predictive of the performance of the invasive community in coalescence
-(as quantified by the relative Bray-Curtis similarity between the coalesced and invasive communities),
-but this correlation was stronger for the citrate communities
-(\figref[a]{fig2}, $R^2=0.15$, $p<0.05$ for glutamine and $R^2=0.57$, $p<10^{-4}$ for citrate).
-Alternative quantifications of community distance yielded similar results,
-with weaker effects when the metric used
-accounted only for the presence/absence of specific species and not for their relative abundance in
-the communities
-(\figref{figS2}).
-All these metrics include the presence of the dominant species themselves. To better disentangle the effect
-that these dominants have on the other members of their communities, we repeated the analysis this time
-excluding the dominant species from the compositional data, finding that our results still hold
-(\figref{figS3}).
-Furthermore, we observed that, in the glutamine communities,
-the relative frequency of a dominant against another in head-to-head pairwise competition
-is barely predictive of its relative frequency against that same other dominant when the
-cohorts are present too, i.e. after coalescence
-(\figref[a]{fig2} red dots, $R^2=0.04$, $p>0.05$).
-On the other hand, pairwise competition between dominants of the citrate communities
-is highly correlated with the relative abundance of said dominants in community coalescence
- (\figref[a]{fig2} blue dots, $R^2=0.83$, $p<10^{-8}$).
-
-Together, these results suggest that the strength of top-down co-selection depends on
-the primary resource supplied to the coalesced communities.
-Communities assembled with citrate as the primary supplied resource
-display a strong degree of top-down co-selection,
-with the fates of the sub-dominant species determined to a large extent by dominant-dominant
-pairwise competition. This competition is, in turn, only weakly affected by the presence of the cohorts.
-For glutamine communities, although some level of top-down co-selection is consistent with our data,
-the cohorts do not appear to be passively responding to their dominants.
-Instead, evidence suggests that the cohorts might be playing an active role in community coalescence.
-This finding led us to investigate the potential role of bottom-up ecological co-selection
-(\figref[e]{fig1}, right panels), i.e. whether in some cases the dominants may be co-selected
-for or against by their cohorts.
-\fi
-
 \subsubsection*{Bottom-up co-selection during community coalescence}
@@
 the effect that the cohorts have in the dominants pairwise competition
-(\figref[a]{fig2})
+(\figref[b]{fig2})
 and the strength of top-down co-selection
-(\figref[b]{fig2}, left and middle panels).
+(\figref[a]{fig2}, left and middle panels).
 The fact that
@@
 Whenever positive bottom-up ecological co-selection is strong, we expect to see dominants
-reaching higher invasion success with their cohorts than by themselves
+reaching higher invasion success with their cohorts than by themselves,
+with the strongest instances occurring when dominants are unable to invade on their own
+but reach high densities when invading together with their cohorts
 (\figref[b]{fig3}, green shaded region).
-On the other hand, a high degree of bottom-up antagonism would result in dominants
+Alternatively, a high degree of bottom-up antagonism would result in dominants
 invading more effectively alone than in the presence of their cohorts
 (\figref[b]{fig3}, red shaded region).
-Alternatively, if both forms of bottom-up co-selection are weak, we would see a similar invasion success
+Finally, if bottom-up co-selection is weak, we would see a similar invasion success
 regardless of the presence or absence of the cohort
@@
 
-\figref[b]{fig3} shows that, in the simulations,
-many dominants could not invade on their own
+In simulations of the MicroCRM, we find no instances of bottom-up antagonism
+but multiple such instances of positive bottom-up co-selection
+(\figref[b]{fig3}).
+Many dominant members of our \textit{in silico} communities
+could not invade another community on their own
 (or could only do so at very low final relative abundances, below 0.1)
-but were able to reach high frequencies when they were accompanied by their cohorts.
-This indicates that positive bottom-up co-selection is frequent and potentially very strong,
+but were able to reach high frequencies when they were accompanied by their cohorts
+in community coalescence.
+Thus, theory indicates that positive bottom-up co-selection is frequent and potentially very strong,
 while negative bottom-up co-selection is far more uncommon.
-We then asked whether the ability of the pairwise competition of dominants
-to predict coalescence outcomes was dependent on the strength of bottom-up co-selection.
-We divided our simulations into two subsets:
+Interestingly, our simulations suggest that strong bottom-up co-selection
+should only observed in communities where top-down co-selection is weak,
+while top down co-selection is only seen when bottom-up co-selection is weak.
+To better illustrate this prediction,
+we divided our simulations into two subsets:
 the first one was comprised of the instances where positive bottom-up co-selection was strong
@@
 (dots near the diagonal of \figref[b]{fig3}).
-We reexamined our original simulations and plotted the frequency of the invasive dominant
-in pairwise competition
-\iffalse
-with the resident dominant
-\fi
-versus the relative similarity between
-the invasive and coalesced communities,
-i.e. the same plot as in \figref[b]{fig2}, for each subset.
-We found that when bottom-up positive co-selection is strong,
+We reexamined our original simulations and
+found that when bottom-up positive co-selection is strong,
 the pairwise competition of dominants is not predictive of coalescence outcomes
-(\figref[c]{fig3}, left panel) and vice-versa (\figref[c]{fig3}, right panel).
-
-We then asked whether this trend was also observed \textit{in vitro}.
-We went back to our laboratory cultures
-and carried out a new round of experiments where we invaded the resident communities
+(\figref[c]{fig3}, left panel)
+indicating that top-down co-selection is weak.
+At the same time, when considering only those coalesced communities in the diagonal of \figref[b]{fig3}
+(where bottom-up co-selection is weak),
+our model predicts that the fates of the sub-dominant community members
+after coalescence are much more strongly determined by the head-to-head competition between dominants in isolation
+($R^2=0.34$ for instances where bottom-up co-selection is weak, \figref[c]{fig3} right panel;
+$R^2=0.22$ when all instances are considered, \figref[a]{fig2} right panel).
+
+We then asked whether this trend was also observed in our experimental communities.
+To address this question,
+we carried out a new round of experiments where we invaded the resident communities
 with the invasive dominants alone
 (\methodsref{competitions}).
-After stabilization, we quantified species abundance through 16S Illumina sequencing
+After stabilization
+(\methodsref{community-assembly}),
+we quantified species abundance through 16S Illumina sequencing
 (\methodsref{sequencing}).
-Again, we observed that bottom-up co-selection is far more common in its
+Consistent with the behavior of our model,
+we observed that bottom-up co-selection is far more common in its
 positive than in its negative form
@@
 Interestingly, bottom-up recruitment appears to be more frequent in the glutamine
-environments than in the citrate ones,
-consistent with our hypothesis that metabolic interactions among species
-are key in determining the strength and direction of ecological co-selection.
+environments, where top-down co-selection was weak,
+than in the citrate ones,
+where top-down co-selection was strong
+(\figref{fig2}).
 We then repeated our analysis in \figref[c]{fig3},
@@
 co-selection instead of the primary carbon source
-as we had done in \figref[b]{fig2}.
+as we had done in \figref[a]{fig2}.
 Our findings were in line with the model prediction:
@@
 $R^2=0.37$, $p<10^{-4}$ when bottom-up co-selection is weak).
-
-\iffalse
-To test which of these situations was more common in our conditions,
-we performed a new round of invasion experiments where each dominant invaded every
-other community in isolation (\methodsref{competitions}). The dominant-invaded
-communities were then propagated for seven serial transfers, and the final
-community compositions were determined by 16S Illumina sequencing.
-We compared the relative abundances reached by the dominants in these experiments
-with the relative abundances of the same dominants in community coalescence
-(i.e. when they invade with their cohorts).
-\figref[b]{fig2} shows that, for communities assembled in citrate environments,
-the cohorts had little effect and the dominants were able to invade with comparable
-success regardless of being accompanied by them.
-This was expected since we had seen that coalescence between citrate communities
-tended to be determined by top-down co-selection.
-However, glutamine communities displayed a different behavior.
-A large number of dominants (roughly 50\%) that were not able to invade by themselves
-(or could only do so at very low final relative abundances, below 0.1)
-were able to reach much higher frequencies when they invaded together with their cohorts,
-which indicates that positive bottom-up co-selection is common in our experiments.
-In contrast, negative co-selection appears to be rare and no instances of it were
-observed in our experiments.
-\fi
-
-\iffalse
-\subsubsection*{Simulations of community coalescence with a Consumer-Resource Model}
-
-To better understand the mechanisms of ecological co-selection in our experiments,
-we ran a set of simulations of community coalescence. We used a Microbial Consumer-Resource
-Model (MicroCRM) \cite{Goldford2018,Marsland2019} implemented in the Community Simulator package
-for Python \cite{Marsland2020}. We chose this modeling framework because
-communities assembled under our experimental conditions (natural microbiomes re-assembled into
-multispecies communities through serial growth-dilution cycles in synthetic minimal media
-with a single carbon source)
-have been shown to be
-sustained by dense metabolic cross-feeding networks
-\cite{Goldford2018,Estrela2020},
-for which the MicroCRM provides a good description.
-We implemented a new feature in the package that makes individual species
-able to secrete different sets of byproducts when consuming a same resource
-(\methodsref{sim}).
-We also used sparse metabolic matrices, equivalent to assuming that species secrete only a few
-types of byproducts when consuming a resource. These assumptions are supported by the experimental
-observation that individual species can indeed secrete different sets of few metabolites
-to the environment when consuming a primary resource
-\cite{Harcombe2014,Pinu2018}. \mr cite Sylvie here?
-We reason that these choices are necessary to potentially generate community cohesiveness:
-if the secretions of all species were roughly similar, the effect of interchanging members of
-a community would be minimal. But in practice, species with a history of coexistence make up
-cohesive communities with highly specific cross-feeding structures \cite{Goldford2018,Estrela2020}.
-In short,
-we reproduced our experimental protocol \textit{in silico}
-by generating a library of species that we divided into two pools. Each pool was used to seed a
-collection of invasive and resident communities, that were allowed to stabilize through 20
-growth-dilution cycles. We then mixed pairs of these stable communities to carry out our coalescence
-and dominant-dominant competition experiments,
-as well as the invasion experiments from the dominants alone.
-A full description of our modeling framework can be found in the \methodsref{sim} section.
-
-We carried out two sets of simulations under two different regimes.
-In one case, we generated a library of 30 resources divided into 3 resource classes
-(\methodsref{sim}). This corresponds to a situation where the digestion of a resource
-can yield a wide variety of secreted byproducts.
-We refer to this scenario as the \textit{complex resource} regime.
-Our second set of simulations was performed under a \textit{simple resource} regime, where we
-generated a library of only 15 resources belonging to a same class.
-For the simple resource, we found that coalescence outcomes were highly correlated with
-the competitive ability of the community dominants in pairwise competition
-(\figref[c]{fig2}),
-indicating a large degree of top-down positive co-selection.
-These dominants were able to reach similar fractions in the final communities when invading alone
-with respect to when invading with their cohorts
-(\figref[d]{fig2}),
-suggesting that bottom-up co-selection is weak in this scenario.
-On the other hand,
-simulations under the complex resource regime evidenced weak top-down co-selection but 
-common and strong bottom-up recruitment (\figref[e-f]{fig2}), with negative co-selection remaining rare.
-Simulations suggest that community cohesiveness is dependent on the metabolic interactions among species,
-which is in line with previous findings \cite{Goldford2018, Estrela2020}.
-These interactions are determined by the metabolic profile of the primary resource in the conditions
-of our simulations and, in turn, modulate the strength and direction of ecological co-selection.
-\fi
+Once the bottom-up communities are removed,
+the glutamine and citrate communities behave very similarly and both have similar degrees of top-down cohesiveness
+(\figref[e]{fig3} right panel,
+$R^2=0.27$ for the glutamine environments, in red; $R^2=0.54$ for the citrate environments, in blue;
+and $R^2=0.37$ when both environments are considered, black line).
+This suggests that the main difference between citrate and glutamine
+is that the latter is richer in communities exhibiting bottom-up cohesiveness than the former.
 
@@
 To better understand the underlying mechanisms that govern
-the emergence of ecological co-selection,
-we developed a minimal model of community coalescence.
+the emergence of ecological co-selection in our experiments,
+we found it useful to study a minimal model of community coalescence
+(\methodsref{sim-min}).
 This model is comprised of two communities with two species each
@@
 respectively)
-is able to utilize the primary resource that is supplied externally ($\mathrm{R}_1$).
-As a result of $\mathrm{R}_1$ metabolism,
-the dominants secrete a single byproduct
+are able to utilize the single externally supplied resource ($\mathrm{R}_1$).
+They secrete a single byproduct
 ($\mathrm{R}_2$ and $\mathrm{R}_4$ respectively)
@@
 ($\mathrm{R}_3$ and $\mathrm{R}_5$ respectively)
-that can again
+that can in turn
 be metabolized by the corresponding dominants.
-This is the simplest structure where both top-down and bottom-up
-cohesiveness can emerge, with their relative strengths being determined
-by how effectively the species can utilize each other's metabolic byproducts.
-Several parameters can modulate the intensity and direction of community cohesiveness,
-and even in this simple model there can be complex interactions between them.
-For instance, a sub-dominant that is very efficient at metabolizing the dominant's
-secretions will therefore be a strong producer of secondary byproducts,
-in turn increasing the metabolic flow towards the dominant itself.
-To keep both communities symmetrical and
-mantain tractability of the model's behavior, we chose to make
-the two sub-dominant species equally able consumers of the
-byproduct of their respective dominants
+With this structure, community members interact via simple cross-feeding networks that can be
+very vertical if the dominant cross-feeds the sub-dominant but the sub-dominant does not cross-feed
+the dominant (i.e. if the dominant cannot utilize the sub-dominant's secretions),
+or more horizontal (bidirectional) if the cohort also cross-feeds the dominant.
+We hypothesize that when two such communities coalesce,
+bottom-up ecological co-selection can emerge if the invasive community is
+sustained by a bidirectional cross-feeding network.
+
+In the limit case when the cross-feeding networks of both the invasive and resident communities
+are strictly vertical (that is, the sub-dominants are passively sustained by the dominants
+but do not cross-feed them),
+it is straightforward that the outcome of community coalescence will depend on the competitive
+ability of the dominants to grow on the single externally supplied resource.
+The most competitive dominant will co-select its sub-dominant (from the top-down)
+through the secretion of specific metabolic byproducts
+(\figref[b]{fig4}).
+If the resident community is maintained by a more horizontal cross-feeding network,
+it can display further resistance to invasion.
+In this scenario, even if the resident dominant is less competitive for the externally supplied resource
+than the invasive dominant,
+cross-feeding from the resident cohort can favor its success in coalescence.
+The stronger the metabolic flux from the resident cohort towards the dominant,
+the more prominent this effect can be
+(\figref[c]{fig4}).
+On the other hand, if the cross-feeding network of the invasive community is horizontal
+(i.e. the sub-dominant is cross-fed by and also cross-feeds the dominant),
+more complex behaviors can emerge.
+The invasive dominant may not be able to invade the resident community by itself
+if it is less competitive for the externally supplied resource than the resident dominant
+(\figref[d]{fig4}),
+or if despite being more competitive, cross-feeding from the resident cohort towards
+the resident dominant favors the success of the latter
+(\figref[e]{fig4}).
+But even then, the invasive community could dominate in coalescence (i.e. when the invasive
+sub-dominant is also present).
+In order for this to happen, cross-feeding from the invasive cohort towards the invasive dominant
+should be strong enough to overcome the competitive disadvantage that said dominant
+may have in isolation.
+
+In summary, coalescence outcomes are contingent on the direction of the
+cross-feeding networks sustaining the communities in this simple setting.
+We ran simulations of all scenarios described above with the minimal model of community coalescence
+implemented in the MicroCRM framework
 (\methodsref{sim-min}).
-We also made it so species secrete only one type of byproduct each,
-i.e. the metabolic matrix is binary in this case
-(see \hyperref[box1]{Box~1}).
-The secretions of the resident and invasive communities are
-non-overlapping:
-\iffalse
-(i.e. $\mathrm{s}_1$ and $\mathrm{s}_2$ secretions
-differ from $\mathrm{s}_3$ and $\mathrm{s}_4$ secretions)
-\fi
-this ensures that coalescence always results in one of the two
-communities taking over and completely excluding the other one,
-which facilitates the interpretation of the model's outcomes.
-This means that
-the relative similarity between the invasive and the coalesced communities
-($Q$, see \methodsref{metrics})
-can only be 0 (if the resident community dominates in coalescence)
-or 1 (if the opposite happens).
-
-With these constraints, the model is specified by four rates:
-the resident dominant's uptake rate of the primary resource ($c_{11}$)
-and of the secondary byproduct secreted by $\mathrm{s}_2$ ($c_{13}$),
-as well as
-the invasive dominant's uptake rate of the primary resource ($c_{31}$)
-and of the secondary byproduct secreted by $\mathrm{s}_4$ ($c_{35}$).
-Analogous to what we did for our initial
-experiments and simulations of community coalescence,
-we first stabilized each minimal community separately
-through serial growth-dilution cycles,
-and then mixed them 1:1 and stabilized again until generational
-equilibrium was achieved.
-Equations \ref{eq:crm-n} and \ref{eq:crm-r} in \hyperref[box1]{Box~1}
-describe the dynamics of this system between consecutive dilutions
-with the corresponding choices for the parameter values
-(\methodsref{sim-min}).
-We ran simulations for both community-community competitions and
-invasions from the dominant alone.
-We considered four different scenarios illustrated in \figref[b-e]{fig4}:
-\textit{1)} both communities are cohesive
-from the top-down but not from the bottom-up,
-\textit{2)} the resident community is cohesive
-both from the top-down and from the bottom-up
-while
-the resident community is cohesive
-from the top-down only,
-\textit{3)} the resident community is cohesive
-from the top-down only
-while
-the invasive community is cohesive
-both from the top-down and from the bottom-up,
-and \textit{4)} both communities are cohesive
-from the top-down and from the bottom-up.
-The outcomes in each case are discussed below.
-
-\textbf{Top-down resident versus top-down invasive.}
-This is the simplest scenario.
-When both communities are cohesive strictly from the top-down
-($c_{13}=0$ and $c_{35}=0$),
-coalescence outcomes are determined by which dominant is a better consumer
-of the primary resource.
-The most competitive dominant excludes the least competitive one
-($\mathrm{s}_1$ excludes $\mathrm{s}_3$ if $c_{11}>c_{31}$ and vice-versa,
-\figref[b]{fig4})
-and,
-because the cross-feeding networks of the invasive and resident
-communities are non-overlapping,
-it co-selects its sub-dominant exclusively.
-The invasion success of the invasive dominant ($\mathrm{s}_3$) is
-unaffected by the presence or absence of its sub-dominant ($\mathrm{s}_4$)
-since it is not being cross-fed by it.
-Thus, again $\mathrm{s}_3$ is only able to invade alone when it
-outcompetes $\mathrm{s}_1$ for the primary resource ($c_{11}>c_{31}$).
-
-\textbf{Bidirectional resident versus top-down invasive.}
-When
-the resident community is cohesive both top-down and bottom-up ($c_{13}>0$)
-and the invasive community is only cohesive from the top-down ($c_{35}=0$),
-coalescence outcomes
-are determined by the strength of bottom-up cohesiveness in the resident community,
-i.e. the magnitude of $c_{13}$ in the minimal model
-(\figref[c]{fig4}).
-If $c_{13}$ is small (weak bottom-up cohesiveness), the system behaves as in the previous scenario,
-virtually as if both communities were only cohesive from the top-down.
-However, as $c_{13}$ increases (strong bottom-up cohesiveness) the resident community becomes more
-resistant to invasion.
-For large values of $c_{13}$, the invasive dominant being more competitive than the resident
-dominant for the primary resource does not ensure invasion success.
-Since the latter can also be cross-fed by its sub-dominant (through $\mathrm{R}_3$),
-it does not rely on the primary resource only ($\mathrm{R}_1$) to outcompete the invader.
-The same is true when the invasive dominant invades alone.
-
-\textbf{Top-down resident versus bidirectional invasive.}
-More outcomes become possible if the resident community is cohesive from the top-down only ($c_{13}=0$)
-and the invasive one is cohesive in both directions ($c_{35}>0$).
-Like in the first scenario we considered, the invasive dominant will only be successful by itself if it is a better
-competitor for $\mathrm{R}_1$ than the resident dominant ($c_{31}>c_{11}$).
-However, if the invasive sub-dominant is also present, the invasion can be successful even if $c_{31}<c_{11}$
-as long as the cross-feeding from $\mathrm{s}_4$ (invasive sub-dominant) to $\mathrm{s}_3$ (invasive dominant)
-is strong enough, i.e. if $c_{35}$ sufficiently large
-(\figref[d]{fig4}).
-In this scenario, we can see cases where the invasive dominant cannot invade in isolation but
-the invasive community still dominates during coalescence,
-similar to what we observed in our experiments and initial simulations.
-
-\textbf{Bidirectional resident versus bidirectional invasive.}
-If both communities are cohesive from the top-down and from the bottom-up ($c_{13}>0$ and $c_{35}>0$),
-coalescence outcomes are generically determined by the relations of strength between
-all forms of cohesiveness.
-In limit cases we can recover scenarios considered previously, e.g. if $c_{13} >> c_{35}$ the system
-behaves as if the resident community was cohesive from the bottom-up but the invasive community was not.
-Situations where the invasive dominant can invade when accompanied by its sub-dominant despite
-being unable to do so by itself are again possible
-if the invasive community is more strongly cohesive from the bottom-up than the resident community
-(\figref[e]{fig4}).
-
-This minimal coalescence model shows that
+In line with our hypothesis, simulations indicate that
 bottom-up co-selection of a dominant that is unable to invade by itself
-is only possible if said dominant is strongly cross-fed by its cohort.
-The model also demonstrates that this behavior can only happen
-if the byproducts of the cohorts are to some extent non-overlapping
-and preferentially utilized by the respective dominants.
-Otherwise, if both species can utilize all secretions, bottom-up
-co-selection becomes unspecific with respect to the dominant identity.
+is only possible if said dominant is strongly cross-fed by its cohort
+(\figref{fig4}).
 
@@
 than the ones in our minimal model.
+To generalize the intuition gained in \figref{fig4} to communities with more than two species,
+we introduce a hierarchy index $h$ that quantifies how vertical a cross-feeding network is:
+
+\iffalse
 However, we argue that some conditions
@@
 we defined a community hierarchy metric $h$ as
+\fi
 
@@
 where $\Delta N_{\mathrm{dom}}$ represents the overall increase in dominant biomass
-between consecutive dilutions in a generationally stable community,
+within a single batch incubation for a generationally stable community,
 and $\Delta N_{\mathrm{dom}}^{\mathrm{R1}}$
@@
 ($\mathrm{R}_1$) only.
-If the dominant was just utilizing the primary resource, the community would be very
+If the dominant was just utilizing the primary resource, the cross-feeding network would be very
 hierarchical ($h \sim 1$),
 whereas if it was growing mostly on the secretions of other taxa,
-it would be more distributed ($h \sim 0$).
+it would be more distributed ($h << 1$).
 We quantified the hierarchies of the resident and invasive communities in our MicroCRM,
 finding that $h$ follows a bimodal distribution (\figref[a]{fig5}).
-We therefore divided our simulations into four groups according to the relation
-of hierarchy between the resident and invasive communities (\figref[b]{fig5}).
-For each group, we evaluated the recurrence of bottom-up co-selection,
+We therefore divided our simulations into four groups according to
+whether the cross-feeding networks of both resident and invasive communities were
+vertical (high $h$) or horizontal (low $h$) as shown in
+\figref[b]{fig5}.
+For each group, we evaluated the frequency of instances of bottom-up co-selection,
 i.e. the fraction of cases where a dominant that could not invade in isolation
@@
 We found that bottom-up ecological co-selection
-was significantly more frequent when the invasive community was
+is significantly more frequent when the invasive community is
 non-hierarchical (\figref[c]{fig5}),
@@
 \subsubsection*{Conclusions}
-
-\iffalse
-Ideas for the conclusions: results and previous theoretical and experimental work on different systems (and on the
-general relationship between metabolic interactions and community structure) suggest that...
-the strength and direction of ecological co-selection is determined by the underlying metabolic networks that
-shape the structure and interactions of communities assembled in synthetic minimal environments
-during coalescence.
-These networks are modulated by the supplied primary resource in our conditions.
-When top-down co-selection is weak, bottom-up positive co-selection can be very strong, while negative
-co-selection remains rare in our experiments.
-ALSO: we have only discussed metabolic interactions here. Other forms of facilitation are possible
-(and probably more relevant in specific natural settings)... but cohesiveness FOR SURE arises from metabolic interactions
-in our experiments, and we know the effect is ultimately metabolic because the MicroCRM captures it
-(actually both forms of it, top-down and bottom-up).
-\fi
 
@@
 of community coalescence experiments,
-and highlight the critical role played by the rarer, sub-dominant species in the generation
+and highlight the critical role that may be played by the rarer, sub-dominant species in the generation
 of community cohesiveness.
 
-Our data suggests that the strength and direction of ecological co-selection is modulated
-by the underlying metabolic networks that shape the structure of
-communities assembled in synthetic minimal conditions
+Our simulations suggest that the strength and direction of ecological co-selection is modulated
+by the underlying cross-feeding networks that shape the structure of
+communities in synthetic minimal environments
 \cite{Goldford2018,Estrela2020}.
-This network is in turn regulated by the supplied primary carbon source
-in our minimal laboratory conditions.
-This idea is supported by the observation that a Microbial Consumer-Resource Model
+This idea is supported by the observation that our Microbial Consumer-Resource Model
 captures the trends observed experimentally when we enable a large variation
@@
 during community coalescence.
-
+Testing this theoretical prediction would require to map the cross-feeding networks of all of our communities.
+Keeping track of every molecule secreted by every species in co-culture and by which species they are uptaken
+is still a low throughput process that is both labor intensive and expensive.
+Recent progress in metabolomic tools promise to help us test this hypothesis in future work.
+
+\iffalse
 Additional work will be necessary to further clarify the relationship between
@@
 remains an open question in ecology.
+\fi
 
@@
 
-\iffalse
-
-We denote this
-metric as
-$E \left( \mathbf{x}_\mathrm{I},
-\mathbf{x}_\mathrm{R},
-\mathbf{x}_\mathrm{C} \right)$.
-
-\begin{equation}
-E \left( \mathbf{x}_\mathrm{I},
-\mathbf{x}_\mathrm{R},
-\mathbf{x}_\mathrm{C} \right) = 
-\left.
-\sum_i
-f \left( x_{\mathrm{I}i}, x_{\mathrm{R}i}, x_{\mathrm{C}i} \right)
-\;
-\middle/
-\;
-\sum_i
-g \left( x_{\mathrm{I}i}, x_{\mathrm{R}i} \right)
-\right.
-\label{eq:endemic}
-\end{equation}
-%
-where we have defined
-
-\begin{equation}
-\begin{split}
-f \left( x_{\mathrm{I}i}, x_{\mathrm{R}i}, x_{\mathrm{C}i} \right) & =
-\begin{cases}
-    1 & \text{if } x_{\mathrm{I}i} > 0 \text{ and }
-                   x_{\mathrm{R}i} = 0 \text{ and }
-                   x_{\mathrm{C}i} > 0 \\
-    0 & \text{otherwise}
-\end{cases} \\
-g \left( x_{\mathrm{I}i}, x_{\mathrm{R}i} \right) & =
-\begin{cases}
-    1 & \text{if } x_{\mathrm{I}i} > 0 \text{ and }
-                   x_{\mathrm{R}i} = 0 \\
-    0 & \text{otherwise}
-\end{cases}
-\end{split}
-\label{eq:endemic-f}
-\end{equation}
-
-\fi
-
 For all the metrics above, we quantified the relative similarity between the invasive
@@
 \internallinenumbers
-\includegraphics[scale=0.7,keepaspectratio]{figs/fig2_v3.pdf}
+\includegraphics[scale=0.7,keepaspectratio]{figs/fig2_v4.pdf}
 \caption{\textbf{Top-down co-selection in microbial community coalescence.}
-\textbf{a.} Pairwise competition of dominants with or without their cohorts.
-In the horizontal axis, we plot the frequency of the invasive dominant species
-in head-to-head pairwise competition with the resident dominant.
-In the vertical axis, we plot the same relative frequency when the two species compete
-in the presence of their cohorts, i.e. during community coalescence.
-$R^2=0.04$, $p>0.05$ for glutamine (red)
-and $R^2=0.83$, $p<10^{-8}$ for citrate (blue).
-\textbf{b.} Coalescence outcomes are quantified by the relative Bray-Curtis similarity
+\textbf{a.} Coalescence outcomes are quantified by the relative Bray-Curtis similarity
 ($Q$) between the coalesced and invasive communities.
@@
 Right panel (black): simulations with a Microbial Consumer-Resource Model are able
-to capture these trends ($R^2=0.22$, $p<10^{-5}$).}
+to capture these trends ($R^2=0.22$, $p<10^{-5}$).
+\textbf{b.} Pairwise competition of dominants with or without their cohorts.
+In the horizontal axis, we plot the frequency of the invasive dominant species
+in head-to-head pairwise competition with the resident dominant.
+In the vertical axis, we plot the same relative frequency when the two species compete
+in the presence of their cohorts, i.e. during community coalescence.
+$R^2=0.04$, $p>0.05$ for glutamine (red)
+and $R^2=0.83$, $p<10^{-8}$ for citrate (blue).
+}
 \label{fig2}
@@
 \textbf{b.}
-\iffalse
-Simulations with a Microbial Consumer-Resource Model suggest that positive bottom-up
-co-selection is common and can be very strong, whereas negative bottom-up co-selection is rare.
-\fi
 Simulations with a Microbial Consumer-Resource Model:
@@
 positive or no bottom-up co-selection was observed (that is, whether points fell into the green or gray
-areas of panel b). Here we reproduce the plots in \figref[b]{fig2} for each set, representing the
+areas of panel b). Here we reproduce the plots in \figref[a]{fig2} for each set, representing the
 result of the dominant head-to-head pairwise competition versus the outcome of community coalescence.
@@
 ($R^2=0.34$, $p<10^{-6}$).
-\iffalse
-Left panel: when positive bottom-up co-selection is strong, head-to-head pairwise competition of dominants
-is poorly predictive of coalescence outcomes
-($R^2=0.00$, $p>0.05$).
-Right panel: on the other hand, when bottom-up co-selection is weak coalescence outcomes are more
-strongly dictated by the result of the dominant-dominant competition
-($R^2=0.34$, $p<10^{-6}$).
-\fi
 \textbf{d.} Experiments show that in our conditions, positive bottom-up co-selection is indeed more
@@
 i.e. we recreate
-\figref[b]{fig2} but this time splitting our data by the strength of bottom-up co-selection
+\figref[a]{fig2} but this time splitting our data by the strength of bottom-up co-selection
 instead of by the carbon source supplied to the communities.
@@
 under different relations of cohesiveness between the
-resident and the invasive communities.}
+resident and the invasive communities.
+We represent the relative Bray-Curtis similarity
+between the invasive and the coalesced communities ($Q$) as a function of
+the relevant model parameters.
+For the specific representative cases indicated by the hollow circles,
+we also show $Q$ as a function of the frequency of the invasive dominant
+in pairwise competition with the resident dominant,
+as well as the frequency of the invasive dominant invading alone versus invading
+accompanied by its cohort.}
 \label{fig4}

### 2021-06-03 draft/draft_v4.tex
draft v4 and fig4 v4 updated (pending update in overleaf)

@@
 
+\textcolor{red}{To do: figSX, re-write minimal model, check literature, check p-values.}
+
 \end{abstract}
@@
 This has been observed, for instance, 
-in river courses where terrestrial microbes mix with aquatic microorganisms,
-\cite{Mansour2018,Luo2020,Vass2021}
+in river courses where terrestrial microbes mix with aquatic microorganisms
+\cite{Mansour2018,Luo2020,Vass2021},
 or in soil communities being invaded as a result of
@@
 or other forms of facilitation from the rarer members of the population.
-We refer to these two opposite scenarios as
+We refer to these two opposite scenarios as the
 ``top-down'' or ``bottom-up'' forms of community cohesiveness, respectively.
@@
 Within each community, the dominant species
-($\mathrm{s}_1$ and $\mathrm{s}_3$ in the resident and invasive communities
+($\mathrm{S}_1$ and $\mathrm{s}_1$ in the resident and invasive communities
 respectively)
@@
 They secrete a single byproduct
-($\mathrm{R}_2$ and $\mathrm{R}_4$ respectively)
+($\mathrm{R}_2$ and $\mathrm{r}_2$ respectively)
 off which
-the sub-dominants ($\mathrm{s}_2$ and $\mathrm{s}_4$ respectively) can feed.
+the sub-dominants ($\mathrm{S}_2$ and $\mathrm{s}_2$ respectively) can feed.
 Finally, these sub-dominants secrete an additional resource
-($\mathrm{R}_3$ and $\mathrm{R}_5$ respectively)
+($\mathrm{R}_3$ and $\mathrm{r}_3$ respectively)
 that can in turn
@@
 It also suggests that rarer taxa may play a more prominent role in co-selecting
-dominant species when the metabolic fluxes across the community are
-distributed rather than hierarchical.
-These observations,
+dominant species when the cross-feeding interactions across community members are
+horizontal rather than hierarchical.
+Testing this theoretical prediction would require to map the cross-feeding networks of all of our communities.
+Keeping track of every molecule secreted by every species in co-culture and by which species they are uptaken
+is still a low throughput process that is both labor intensive and expensive.
+Recent progress in metabolomic tools promise to help us test this hypothesis in future work.
+Our findings,
 together with previous results in different systems \cite{Sierocinski2017} as well as
@@
 during community coalescence.
-Testing this theoretical prediction would require to map the cross-feeding networks of all of our communities.
-Keeping track of every molecule secreted by every species in co-culture and by which species they are uptaken
-is still a low throughput process that is both labor intensive and expensive.
-Recent progress in metabolomic tools promise to help us test this hypothesis in future work.
 
@@
 of two species each
-($\mathrm{s}_1$ to $\mathrm{s}_4$),
+($\mathrm{S}_1$ and $\mathrm{S}_2$ in the resident community,
+$\mathrm{s}_1$ and $\mathrm{s}_2$ in the invasive community),
 with five resources in total, out of which the first one ($\mathrm{R}_1$)
@@
 \text{\scriptsize $\mathrm{R}_3$} \!\!\! & 
-\text{\scriptsize $\mathrm{R}_4$} \!\!\! & 
-\text{\scriptsize $\mathrm{R}_5$}
+\text{\scriptsize $\mathrm{r}_2$} \!\!\! & 
+\text{\scriptsize $\mathrm{r}_3$}
 \end{matrix}
@@
 \text{\scriptsize $\mathrm{R}_3$} \!\!\! \\
-\text{\scriptsize $\mathrm{R}_4$} \!\!\! \\
-\text{\scriptsize $\mathrm{R}_5$} \!\!\!
+\text{\scriptsize $\mathrm{r}_2$} \!\!\! \\
+\text{\scriptsize $\mathrm{r}_3$} \!\!\!
 \end{matrix*}
@@
 \text{\scriptsize $\mathrm{R}_3$} \;\;\; & 
-\text{\scriptsize $\mathrm{R}_4$} \;\;\; & 
-\text{\scriptsize $\mathrm{R}_5$}
+\text{\scriptsize $\mathrm{r}_2$} \;\;\; & 
+\text{\scriptsize $\mathrm{r}_3$}
 \end{matrix}
@@
 \begin{matrix*}[l]
+\text{\scriptsize $\mathrm{S}_1$} \!\!\! \\
+\text{\scriptsize $\mathrm{S}_2$} \!\!\! \\
 \text{\scriptsize $\mathrm{s}_1$} \!\!\! \\
-\text{\scriptsize $\mathrm{s}_2$} \!\!\! \\
-\text{\scriptsize $\mathrm{s}_3$} \!\!\! \\
-\text{\scriptsize $\mathrm{s}_4$} \!\!\!
+\text{\scriptsize $\mathrm{s}_2$} \!\!\!
 \end{matrix*}
@@
 \begin{pmatrix}
-c_{11} & 0 & c_{13} & 0 & 0 \\
+C_{11} & 0 & C_{13} & 0 & 0 \\
 0 & 100 & 0 & 0 & 0 \\
-c_{31} & 0 & 0 & 0 & c_{35} \\
+c_{11} & 0 & 0 & 0 & c_{13} \\
 0 & 0 & 0 & 100 & 0
@@
 Where we made the sub-dominants equally strong consumers of their dominants' secretions
-($c_{22} = c_{44}$ = 100),
+($C_{22} = c_{22}$ = 100),
 and we varied all other uptake rates depending on the scenario we were considering (see main text).
-Whenever we were interested in the ratio between two rates (e.g. $c_{35}/c_{13}$ in \figref[e]{fig4})
+Whenever we were interested in the ratio between two rates (e.g. $c_{13}/C_{13}$ in \figref[e]{fig4})
 we gave the one in the denominator a fixed value of 1 and let one in the numerator range
@@
 \internallinenumbers
-\includegraphics[scale=0.7,keepaspectratio]{figs/fig4_v3.pdf}
+\includegraphics[scale=0.7,keepaspectratio]{figs/fig4_v4.pdf}
 \caption{\textbf{A minimal model of community coalescence.}

### 2021-06-03 draft/draft_v4.tex
changes to draft v4, adding code for supplementary figure 5 (4?)

@@
 
-\textcolor{red}{To do: figSX, re-write minimal model, check literature, check p-values.}
-
 \end{abstract}
@@
 or in soil communities being invaded as a result of
-tillage and outplanting \cite{Rillig2016} or by
-aerially dispersed bacteria and funghi \cite{Evans2019}.
+tillage and outplanting
+\cite{Rillig2016,Ramoneda2020,Rochefort2020}
+or by aerially dispersed bacteria and funghi \cite{Evans2019}.
 Gut microbiomes can invade external communities
@@
 only beginning to be addressed and little is known about the mechanisms that govern it and its
-potential implications.
+potential implications or applications
+\cite{Rillig2016b}.
 Early mathematical models of community-community invasions \cite{Gilpin1994,Toquenaga1997}
@@
 \begin{center}
-\includegraphics[width=11cm,keepaspectratio]{figs/figBox_v3.png}
+\includegraphics[width=11cm,keepaspectratio]{figs/figBox_v4.png}
 \end{center}
@@
 These equations can take slightly different forms in certain cases,
-e.g. if the primary resource is supplied continuously instead of at the beggining
+e.g. if the primary resource is supplied continuously instead of at the beginning
 of each growth cycle \cite{Marsland2019,Marsland2020}.
@@
 Once the bottom-up communities are removed,
-the glutamine and citrate communities behave very similarly and both have similar degrees of top-down cohesiveness
-(\figref[e]{fig3} right panel,
-$R^2=0.27$ for the glutamine environments, in red; $R^2=0.54$ for the citrate environments, in blue;
-and $R^2=0.37$ when both environments are considered, black line).
+both the glutamine and citrate communities
+display similar degrees of top-down cohesiveness
+(\figref[e]{fig3}, right panel).
 This suggests that the main difference between citrate and glutamine
@@
 
-To better understand the underlying mechanisms that govern
-the emergence of ecological co-selection in our experiments,
-we found it useful to study a minimal model of community coalescence
+In view of the success of our model in reproducing the
+experimentally observed trends in ecological co-selection,
+we set out to better understand the mechanisms for its emergence.
+In our experimental conditions and in the MicroCRM,
+comunities are sustained by dense cross-feeding facilitation networks.
+These networks can have a very vertical direction if
+a single species (the dominant) cross-feeds
+the rarer members of the community
+but these do not cross feed the dominant in return.
+Alternatively, if the dominant is strongly cross-fed by its cohort
+the network structure would be more horizontal.
+In the latter scenario, positive bottom-up co-selection of a dominant
+can take place if
+cross-feeding from its cohort allow it
+to persist in the final community after coalescence
+--even if it cannot invade succesfully in isolation.
+
+We found it useful to study a minimal model of community coalescence
+to test these ideas
 (\methodsref{sim-min}).
-This model is comprised of two communities with two species each
+This model is comprised of two communities
+(resident and invasive)
+with only two species each
 as illustrated in \figref[a]{fig4}.
 Within each community, the dominant species
-($\mathrm{S}_1$ and $\mathrm{s}_1$ in the resident and invasive communities
-respectively)
+($\mathrm{S}_1$ and $\mathrm{s}_1$ respectively)
 are able to utilize the single externally supplied resource ($\mathrm{R}_1$).
@@
 be metabolized by the corresponding dominants.
-With this structure, community members interact via simple cross-feeding networks that can be
-very vertical if the dominant cross-feeds the sub-dominant but the sub-dominant does not cross-feed
-the dominant (i.e. if the dominant cannot utilize the sub-dominant's secretions),
-or more horizontal (bidirectional) if the cohort also cross-feeds the dominant.
-We hypothesize that when two such communities coalesce,
-bottom-up ecological co-selection can emerge if the invasive community is
-sustained by a bidirectional cross-feeding network.
+The dominants' ability to utilize their sub-dominants' metabolic
+byproducts determines whether the structure of the cross-feeding
+networks of these minimal communities is vertical
+(if the dominants cannot utilize the cohort secretions
+and thus are not cross-fed by them)
+or horizontal
+(in the opposite scenario).
+The model parameters controlling how effectively the dominants
+can metabolize said byproducts modulate the direction of the
+cross-feeding networks
+(\figref[a]{fig4}).
 
@@
 than the invasive dominant,
-cross-feeding from the resident cohort can favor its success in coalescence.
+cross-feeding from the resident cohort can allow it to persist in coalescence.
 The stronger the metabolic flux from the resident cohort towards the dominant,
@@
 cross-feeding networks sustaining the communities in this simple setting.
-We ran simulations of all scenarios described above with the minimal model of community coalescence
+We ran simulations of all scenarios described above with our minimal model of community coalescence
 implemented in the MicroCRM framework
 (\methodsref{sim-min}).
-In line with our hypothesis, simulations indicate that
+In line with our initial proposition, simulations indicate that
 bottom-up co-selection of a dominant that is unable to invade by itself
-is only possible if said dominant is strongly cross-fed by its cohort
+is possible if said dominant is strongly cross-fed by its cohort
 (\figref{fig4}).
@@
 If the dominant was just utilizing the primary resource, the cross-feeding network would be very
-hierarchical ($h \sim 1$),
+vertical ($h \sim 1$),
 whereas if it was growing mostly on the secretions of other taxa,
-it would be more distributed ($h << 1$).
+it would be more horizontal ($h << 1$).
 We quantified the hierarchies of the resident and invasive communities in our MicroCRM,
@@
 
+\begin{figure}[!h]
+\centering
+\internallinenumbers
+\includegraphics[scale=0.9,keepaspectratio]{figs/figS4.pdf}
+\caption{\textbf{Top-down co-selection is robust in the MicroCRM.}
+We ran simulations of community coalescence following the same
+procedure described in the main text, but randomly sampling the
+following prameters of the MicroCRM
+uniformly within the indicated ranges:
+number of species per community sampled at initialization ($S$) between 10 and 90;
+total number of species families between 1 and 5;
+total number of species per family ($S_A$) between 400 and 1200;
+total number of generalist species ($S_{gen}$) between 100 and 380;
+specialist species preference strength ($q$) between 0.5 and 1;
+number of resource classes between 1 and 5;
+number of resources per class ($M_A$) between 3 and 17;
+leakage fraction ($l_{\alpha}$) between 0.45 and 0.95;
+maintenance cost ($m_i$) between 0 and 0.2;
+standard deviation of the sum of consumption rates ($\sigma_c$) between 2 and 4;
+sparsity of the metabolic matrix between 0.05 and 0.95;
+fraction of secretion flux to resources of the same type ($f_s$) or to waste resources ($f_w$) between 0.05 and 0.45.
+The metabolic matrix was randomly chosen to be such that $D_{i\beta\alpha}=D_{j\beta\alpha}$ or such that
+$D_{i\beta\alpha} \neq D_{j\beta\alpha}$ for all species ($i$, $j$) and resources ($\alpha$, $\beta$).
+Both options were given equal probabilities.
+For each randomly sampled set of parameters, we ran 100 simulations
+and quantified the ability of the pairwise competition
+of dominants to predict coalescence outcomes (see \figref[a]{fig2}), i.e. the strength of top-down co-selection
+in each regime.
+The histogram shows that top-down co-selection is robust throughout the parameter space in the MicroCRM.}
+\label{figSX}
+\end{figure}
+
+\clearpage
+
 \end{document}

### 2021-06-03 coalescence.py

@@ -36,13 +36,6 @@ if os.name == 'nt':
 else:
     par = True
 
-'''
-# set random and numpy's seed for reproducibility
-myseed = 1000000 # 99
-np.random.seed(myseed)
-random.seed(myseed)
-'''
-
 import time
 start_time = time.time()
 

### 2021-06-03 coalescence_randomParams.py

@@ -1,0 +1,335 @@
+# -*- coding: utf-8 -*-
+"""
+Created on Thu Jun  3 09:57:19 2021
+
+@author: juandc
+"""
+
+
+
+# %%
+### INITIALIZATION
+
+# reset variables etc.
+from IPython import get_ipython
+get_ipython().magic('reset -sf')
+
+# libraries
+import community_simulator
+import community_simulator.usertools
+import community_simulator.visualization
+import numpy as np
+#import pandas as pd
+# import matplotlib
+#import matplotlib.pyplot as plt
+# import seaborn as sns
+# colors = sns.color_palette()
+# %matplotlib inline
+
+import random
+# import math
+import copy
+import os
+
+# check if running on Windows
+if os.name == 'nt':
+    par = False
+else:
+    par = True
+
+import time
+start_time = time.time()
+
+
+
+# %%
+### USER-DEFINED FUNCTIONS
+
+# stabilize communities through serial passaging
+def stabilizeCommunities(plate):
+    
+    # parameters
+    dilution_factor = 1/100
+    n_transfers = 20
+    time_per_cycle = 1
+    # extinction_threshold = 1e-4 # we don't use this because we don't use StadyState()
+    
+    # run stabilization protocol
+    Ntraj, Rtraj = plate.RunExperiment(f = np.diag(np.ones(plate.n_wells))*dilution_factor,
+                                       T = time_per_cycle,
+                                       npass = n_transfers,
+                                       compress_resources=False,
+                                       compress_species=True)
+    
+    return([plate.N,plate.R])
+
+# Bray-Curtis similarity
+# p and q are lists (of absolute abundances, sum need not be 1)
+def bray_curtis(p,q):
+
+    # C_pq: sum of lesser values of common species
+    c = sum([min(p[i],q[i]) for i in range(len(p)) if p[i]>0 and q[i]>0])
+    
+    # Bray-Curtis similarity
+    bc = 2*c/(sum(p) + sum(q))
+    return bc
+
+# similarity index
+# p: invasive
+# q: resident
+# pq: coalesced
+def mysim(p,q,pq):
+    
+    # normalization
+    p = [p_i/sum(p) for p_i in p]
+    q = [q_i/sum(q) for q_i in q]
+    pq = [pq_i/sum(pq) for pq_i in pq]
+    
+    sim = bray_curtis(p,pq)/(bray_curtis(p,pq) + bray_curtis(q, pq))
[…]

### 2021-06-03 coalescence_randomParams.py
typo in code for supplementary fig 5

@@ -98,7 +98,7 @@ def mysim(p,q,pq):
 r_squared = ['NA' for i in range(100)]
 
 # repeat simulation 100 times
-for n_sim in range(1):
+for n_sim in range(100):
     
     # print progress
     print('Running simulation %s out of 100' % (n_sim+1))
@@ -106,7 +106,7 @@ for n_sim in range(1):
     
     # general assumptions (common to all simulations)
     assumptions = community_simulator.usertools.a_default.copy()
-    assumptions['n_wells'] = 100 # number of communities
+    assumptions['n_wells'] = 50 # number of communities
     
     assumptions['response'] = 'type I'
     assumptions['regulation'] = 'independent' # 'independent' is standard, 'energy' or 'mass' enable preferential consumption of resources
@@ -142,7 +142,8 @@ for n_sim in range(1):
     assumptions['sparsity'] = random.uniform(0.05,0.95) # variability in secretion fluxes among resources (must be less than 1) -- uniformly distributed between 0.05 and 0.95
     assumptions['fs'] = random.uniform(0.05,0.45) # fraction of secretion flux to resources of the same type as the consumed one -- uniformly distributed between 0.05 and 0.45
     assumptions['fw'] = random.uniform(0.05,0.45) # fraction of secretion flux to waste resources -- uniformly distributed between 0.05 and 0.45
-    assumptions['metabolism'] = ['common','specific'][random.randint(0,1)] # common or species-specific D matrix -- chosen arbitrarily with equal probabilities
+    #assumptions['metabolism'] = ['common','specific'][random.randint(0,1)] # common or species-specific D matrix -- chosen arbitrarily with equal probabilities
+    assumptions['metabolism'] = 'common'
     
     # parameters (make new matrices, etc)
     params = community_simulator.usertools.MakeParams(assumptions)

### 2021-06-03 draft/draft_v4.tex
typo

@@
 specialist species preference strength ($q$) between 0.5 and 1;
-number of resource classes between 1 and 5;
+number of resource classes between 2 and 5;
 number of resources per class ($M_A$) between 3 and 17;

### 2021-06-03 coalescence_randomParams.py

@@ -127,7 +127,7 @@ for n_sim in range(100):
     assumptions['SA'] = [n_species_per_family] * n_families
     assumptions['Sgen'] = random.randint(100,380) # number of generalists -- uniformly distributed between 100 and 380
     
-    n_resource_classes = random.randint(1,5) # number of resource classes -- uniformly distributed between 1 and 5
+    n_resource_classes = random.randint(2,5) # number of resource classes -- uniformly distributed between 1 and 5
     n_resources_per_class = random.randint(3,17) # number of resources per class -- uniformly distributed between 3 and 17
     assumptions['MA'] = [n_resources_per_class] * n_resource_classes
     
@@ -142,8 +142,7 @@ for n_sim in range(100):
     assumptions['sparsity'] = random.uniform(0.05,0.95) # variability in secretion fluxes among resources (must be less than 1) -- uniformly distributed between 0.05 and 0.95
     assumptions['fs'] = random.uniform(0.05,0.45) # fraction of secretion flux to resources of the same type as the consumed one -- uniformly distributed between 0.05 and 0.45
     assumptions['fw'] = random.uniform(0.05,0.45) # fraction of secretion flux to waste resources -- uniformly distributed between 0.05 and 0.45
-    #assumptions['metabolism'] = ['common','specific'][random.randint(0,1)] # common or species-specific D matrix -- chosen arbitrarily with equal probabilities
-    assumptions['metabolism'] = 'common'
+    assumptions['metabolism'] = ['common','specific'][random.randint(0,1)] # common or species-specific D matrix -- chosen arbitrarily with equal probabilities
     
     # parameters (make new matrices, etc)
     params = community_simulator.usertools.MakeParams(assumptions)

### 2021-06-03 draft/draft_v4.tex
typo

@@
 specialist species preference strength ($q$) between 0.5 and 1;
-number of resource classes between 2 and 5;
+number of resource classes between 3 and 8;
 number of resources per class ($M_A$) between 3 and 17;

### 2021-06-03 coalescence_randomParams.py

@@ -40,6 +40,9 @@ else:
 import time
 start_time = time.time()
 
+np.random.seed(1)
+random.seed(1)
+
 
 
 # %%
@@ -127,7 +130,7 @@ for n_sim in range(100):
     assumptions['SA'] = [n_species_per_family] * n_families
     assumptions['Sgen'] = random.randint(100,380) # number of generalists -- uniformly distributed between 100 and 380
     
-    n_resource_classes = random.randint(2,5) # number of resource classes -- uniformly distributed between 1 and 5
+    n_resource_classes = random.randint(3,8) # number of resource classes -- uniformly distributed between 1 and 5
     n_resources_per_class = random.randint(3,17) # number of resources per class -- uniformly distributed between 3 and 17
     assumptions['MA'] = [n_resources_per_class] * n_resource_classes
     
